### 2020-09-25 reports/v1_0/report.tex
building report in tex

\documentclass[a4paper]{article}
\usepackage{style}

\begin{document}

\title {HTM + RL: Project details}
\author {Petr Kuderov}
\maketitle

\section {Постановка задачи и среда}

Агент играет в среде GridWorld - лабиринт на квадратной сетке. Агент начинает игру в фиксированной точке старта, и ему требуется найти единственную награду, которая находится в фиксированном месте.

Рассматриваются два близких варианта GridWorld - с учетом направления взгляда агента и без.

В первом случае в каждый момент времени агент находится в некоторой клетке лабиринта, указывая лицом одно из 4х направлений: восток, север, запад, юг. Из текущего состояния он может совершить следующие действия: шагнуть прямо (в соседнюю клетку), повернуться на 90 градусов по или против часовой стрелки. По умолчанию, в наших экспериментах одно из направлений поворота, по часовой стрелке, было запрещено.

Во втором случае агент не имеет направления взгляда и может шагать непосредственно в любом из четырех направлений (восток, север, запад, юг), т.е. в одну из четырех соседних клеток.

Переходы между состояниями однозначны, т.е. среда описывается детерминированным MDP. Когда агент пытается шагнуть "в стенку", он остается на месте.

На каждом невыигрышном шаге агент получает небольшую отрицательную награду $-0.01$, на выигрышном - большую положительную $+1.0$.

\section{Agent}

Текущая реализация агента содержит, грубо говоря, две политики - случайную и спланированную. Спланированную политику предоставляет планировщик. В случае, если он не может предоставить политику, применяется случайная.

Таким образом основная логика агента содержится в реализации планировщика.

Задача планировщика - найти путь до награды, если она достижима за не более чем \verb|planning_horizon| шагов. Делается это в два этапа:

\begin{enumerate}
    \item Планирование до награды из текущего состояния
    \item Нахождение обратного пути из состояния с наградой в текущее
\end{enumerate}

Чтобы агент мог планировать, он наделен памятью для запоминания увиденных переходов между состояниями, которая эмулирует функцию динамики среды $f: (s, a) \rightarrow s'$. При этом память агента имеет некоторые особенности.

Главная особенность - она способна работать с суперпозициями (=множествами). Это означает, что память может отвечать на запросы вида \textit{имея набор состояний $\{s_i\}$ и набор действий $\{a_j\}$, в какие состояния $\{s'_m\}$ могут привести все возможные пары $(s_i, a_j)$}? В ответ на такой запрос память возвращает суперпозицию следующих состояний.

Вторая особенность - для каждого конкретного состояния $s'$ из суперпозиции в ответе на запрос к памяти есть возможность узнать $(s, a) = f^{-1}(s')$

Третья особенность - память агента дополнительно снабжена подмодулем памяти о выигрышных состояниях. На текущий момент этот подмодуль реализован с помощью наивного списка состояний, в которых агент получил положительную награду - мы называем этот подмодуль списком целей. Список целей имеет ограниченный размер S, являющийся гиперпараметром.

На основе этих особенностей и построен алгоритм планировщика.

\subsection{Планировщик. Шаг 1: планирование до цели}

Агент опирается на свой опыт и исходит из предположения, что награда может находиться в одном из выигрышных состояний предыдущих эпизодов. Поэтому состояния из списка целей являются приоритетными для исследования.

Пусть в некоторый момент времени $t_0$ агент находится в состоянии $s_0$. Чтобы ответить на вопрос, есть ли достижимое состояние из списка целей в радиусе $N$ действий, начнем планировать из суперпозиции $(s_0, A)$, где $A$ - все множество доступных действий агента.

После первого шага предсказания мы будем иметь множество состояний $\{s_i\}^1$, в которые можно попасть из текущего за одно действие агента. Продолжим планировать из суперпозиции этих состояний так же по всем возможным действиям. И так далее до тех пора пока либо не обнаружим хотя бы одно состояние из списка целей, либо не исчерпаем лимит горизонта планирования $N =$ \verb|planning_horizon|.

Если цель найдена, то планировщик переходит к следующему этапу.

NB: первый этап может обнаружить сразу несколько достижимых целевых состояний в радиусе $T$ шагов. Второй этап проводится последовательно для каждого из этих состояний до первого успеха. Выбор очередности перебора целей в текущей реализации явно не определен.

\subsection{Планировщик. Шаг 2: нахождение обратного пути}

Итак, на текущий момент, планировщик уверен, что некоторое целевое состояние $s_T$может быть достигнуто за $T$ шагов, но не знает в точности как - на первом этапе мы работали с суперпозициями и использовали все возможные действия, поэтому какие конкретно действия привели к целевому состоянию, неизвестно.

На момент окончания успешного планирования за $T$ шагов, имеем всю историю суперпозиций состояний на каждом из этих шагов $[s_0, \{s_i\}^1, .. \{s_i\}^t]$.

По целевому состоянию планировщик выбирает пару состояние-действие $(s_{T-1}, a_{T-1}) = f^{-1}(s_T)$, которые привели к нему. Таким образом для предпоследнего шага известно требуемое действие $a_{T-1}$ и состояние $s_{T-1}$, в которое нужно прийти.

Далее рекурсивно повторяем процедуру, на каждом шаге $t$ находя пару состояние-действие $(s_{t-1}, a_{t-1}) = f^{-1}(s_t)$, пока не вернемся в начальную позицию $s_0$.

Здесь, правда, стоит уточнить, что $f^{-1}$ в общем случае возвращает не единственную пару $(s, a)$, а множество пар, и не всякая пара рекурсивно приведет нас обратно в начальную позицию (т.е. реально достижима из начальной позиции). Поэтому рекурсивно проверяется каждая из возможных пар до первого успешного возвращения в начало.

Заметим, что в процессе возвращения назад алгоритм формирует однозначную последовательность действий $[a_0, .. , a_{T-1}]$, которая приведет агента из начального состояния $s_0$ в искомое целевое состояние $s_T$.

Если процесс нахождения обратного пути завершился успешно, планировщик имеет политику действий на $T$ шагов вперед, которую агент безусловно выполняет. В случае успеха агент находит награду, и эпизод завершается, иначе агент исключает рассмотренное целевое состояние из списка целей до окончания текущего эпизода и пытается заново спланировать действия.

\section{Тестирование}

Цель тестирования:

\begin{itemize}
    \item Исследовать влияние на производительность агента:
    \begin{itemize}
        \item длины горизонта планирования
        \item размера списка целей
    \end{itemize}

    \item Сравнить производительность агента с бейзлайнами:
    \begin{itemize}
        \item Random Walk агент
        \item DQN агент
    \end{itemize}
\end{itemize}

Тестирование агента содержало два различных набора экспериментов в соответствии с исследуемыми задачами transfer learning, отличающихся количеством сред в каждом из экспериментов.

В качестве реализации рандомной стратегии использовался частный случай htm агента с нулевым горизонтом планирования. У DQN агента в сравнении использовались результаты обеих стратегий - жадной и $\epsilon$-жадной.

\subsection{Тестирование на задаче transfer learning в одной среде}

В рамках одной среды проводилась серия испытаний по N эпизодов. В течение одного испытания из N эпизодов агент учился находить награду в некоторой ячейке. Между испытаниями местоположение награды изменялось. Исследовалась способность агентов адаптироваться к новому местоположению награды.

Для этого набора экспериментов были выбраны 3 среды типа Gridworld без учета направления взгляда агента.

Ниже к каждой из этих сред приведены графики результатов рассматриваемых агентов: а) награда за эпизод, б) количество шагов за эпизод и в) длительность выполнения одного эпизода (в секундах). На всех графиках изображена скользящая средняя с окном, указанным в заголовке графика (например, \verb|MA = 20|)

Легенда к названиям агентов на графиках:

\begin{itemize}
    \item \verb|htm| в названии обозначает htm агента
    \begin{itemize}
        \item суффикс \verb|_X| означает горизонт планирования \verb|X|, например:
        \begin{itemize}
            \item \verb|htm_0| - нулевой горизонт планирования, т.е. случайный агент
            \item \verb|htm_4| - горизонт планирования 4
        \end{itemize}

        \item суффикс \verb|_Xg| означает размер списка целей \verb|X|, например:
        \begin{itemize}
            \item \verb|htm_2_1g| - горизонт планирования 2, длина списка целей 1
            \item \verb|htm_1_16g| - горизонт планирования 1, длина списка целей 16
            \item если размер списка целей не указан явно, предполагается неограниченная длина списка
        \end{itemize}
    \end{itemize}

    \item dqn в названии обозначает DQN агента
    \begin{itemize}
        \item \verb|_greedy| - суффикс жадной стратегии
        \item \verb|_eps| - суффикс $\epsilon$-жадной стратегии
    \end{itemize}
\end{itemize}

Легенда к схемам тестовых сред:

\begin{itemize}
    \item \verb|-| пустая клетка,
    \item \verb|#| стена,
    \item \verb|@| агент,
    \item \verb|X| награда
\end{itemize}

\subsubsection {\texttt{multi\_way\_v0}}

Последовательность из четырех испытаний по 100 эпизодов. Положения награды в испытаниях было следующим:

\begin{verbatim}
  #####   #####   #####   #####
  #@--#   #@-X#   #@--#   #@--#
  #-#-#   #-#-#   #-#-#   #-#X#
  #--X#   #---#   #X--#   #---#
  #####   #####   #####   #####
\end{verbatim}

Сравнение агента с бейзлайнами:

% ![episode steps absolute](../../reports/v1_0/multi_way_transfer/multi_way_v0__steps.svg)
% ![episode steps log-relative](../../reports/v1_0/multi_way_transfer/multi_way_v0__steps_rel_htm_0.svg)

Сравнение DQN жадного и $\epsilon$-жадного агента:

% ![episode steps absolute](../../reports/v1_0/multi_way_transfer/multi_way_v0__dqn__steps.svg)
% ![episode steps log-relative](../../reports/v1_0/multi_way_transfer/multi_way_v0__dqn__steps_rel_dqn_greedy.svg)

Сравнение агента с разными горизонтами и размером списка целей:

% - горизонт планирования 1

%   ![episode steps absolute](../../reports/v1_0/multi_way_transfer/multi_way_v0__1__steps.svg)
%   ![episode steps log-relative](../../reports/v1_0/multi_way_transfer/multi_way_v0__1__steps_rel_htm_1_1g.svg)

% - горизонт планирования 2

%   ![episode steps absolute](../../reports/v1_0/multi_way_transfer/multi_way_v0__2__steps.svg)
%   ![episode steps log-relative](../../reports/v1_0/multi_way_transfer/multi_way_v0__2__steps_rel_htm_2_1g.svg)

% - горизонт планирования 4

%   ![episode steps absolute](../../reports/v1_0/multi_way_transfer/multi_way_v0__4__steps.svg)

\subsubsection{\texttt{multi\_way\_v1}}

Последовательность из пяти испытаний по 200 эпизодов. Положения награды:

\begin{verbatim}
  #######   #######   #######   #######   #######
  #@--###   #@--###   #@--###   #@--###   #@--###
  #-#-###   #-#-###   #-#X###   #-#-###   #-#-###
  #X----#   #-----#   #-----#   #-----#   #----X#
  ###-#-#   ###X#-#   ###-#-#   ###-#-#   ###-#-#
  ###---#   ###---#   ###---#   ###--X#   ###---#
  #######   #######   #######   #######   #######
\end{verbatim}

Сравнение агента с бейзлайнами:

% ![episode steps absolute](../../reports/v1_0/multi_way_transfer/multi_way_v1__steps.svg)
% ![episode steps log-relative](../../reports/v1_0/multi_way_transfer/multi_way_v1__steps_rel_htm_0.svg)

Сравнение DQN жадного и $\epsilon$-жадного агента:

% ![episode steps absolute](../../reports/v1_0/multi_way_transfer/multi_way_v1__dqn__steps.svg)
% ![episode steps log-relative](../../reports/v1_0/multi_way_transfer/multi_way_v1__dqn__steps_rel_dqn_greedy.svg)

Сравнение агента с разными горизонтами и размером списка целей:

% - горизонт планирования 1

%   ![episode steps absolute](../../reports/v1_0/multi_way_transfer/multi_way_v1__1__steps.svg)
%   ![episode steps log-relative](../../reports/v1_0/multi_way_transfer/multi_way_v1__1__steps_rel_htm_1_1g.svg)

% - горизонт планирования 2

%   ![episode steps absolute](../../reports/v1_0/multi_way_transfer/multi_way_v1__2__steps.svg)
%   ![episode steps log-relative](../../reports/v1_0/multi_way_transfer/multi_way_v1__2__steps_rel_htm_2_1g.svg)

% - горизонт планирования 4

%   ![episode steps absolute](../../reports/v1_0/multi_way_transfer/multi_way_v1__4__steps.svg)

\subsubsection{\texttt{multi\_way\_v1}}

Последовательность из шести испытаний по 200 эпизодов. Положения награды:

\begin{verbatim}
  #########    #########    #########    #########    #########    #########
  #---#####    #-X-#####    #---#####    #---#####    #---#####    #---#####
  #-#-#####    #-#-#####    #-#-#####    #-#-#####    #-#-#####    #-#-#####
  #--@---##    #--@---##    #--@--X##    #--@---##    #X-@---##    #--@---##
  ###--####    ###--####    ###--####    ###--####    ###--####    ###--####
  ###-#--X#    ###-#---#    ###-#---#    ###-#---#    ###-#---#    ###-#--X#
  ###---###    ###---###    ###---###    ###---###    ###---###    ###---###
  ###-#####    ###-#####    ###-#####    ###X#####    ###-#####    ###-#####
  #########    #########    #########    #########    #########    #########
\end{verbatim}

Сравнение агента с бейзлайнами:

% ![episode steps absolute](../../reports/v1_0/multi_way_transfer/multi_way_v2__steps.svg)
% ![episode steps log-relative](../../reports/v1_0/multi_way_transfer/multi_way_v2__steps_rel_htm_0.svg)

Сравнение DQN жадного и $\epsilon$-жадного агента:

% ![episode steps absolute](../../reports/v1_0/multi_way_transfer/multi_way_v2__dqn__steps.svg)
% ![episode steps log-relative](../../reports/v1_0/multi_way_transfer/multi_way_v2__dqn__steps_rel_dqn_greedy.svg)

Сравнение агента с разными горизонтами и размером списка целей:

% - горизонт планирования 2

%   ![episode steps absolute](../../reports/v1_0/multi_way_transfer/multi_way_v2__2__steps.svg)
%   ![episode steps log-relative](../../reports/v1_0/multi_way_transfer/multi_way_v2__2__steps_rel_htm_2_1g.svg)

% - горизонт планирования 4

%   ![episode steps absolute](../../reports/v1_0/multi_way_transfer/multi_way_v2__4__steps.svg)

% - горизонт планирования 8 и 12

%   ![episode steps absolute](../../reports/v1_0/multi_way_transfer/multi_way_v2__8-12__steps.svg)

\subsection{Тестирование на задаче transfer learning в нескольких средах}

Эксперименты данного режима проводились на наборе различных сред-лабиринтов. Схема каждого эксперимента следующая:

\begin{itemize}
  \item эксперимент проводится последовательно на $N_{env}$ средах
  \item при фиксированной среде последовательно рассматриваются $N_{rew}$ положений награды
  \item при фиксированном положении награды рассматривается $N_{s_0}$ начальных положений агента
  \item в фиксированной конфигурации агент играет $N_{ep}$ эпизодов
\end{itemize}

Исследовалась способность агента адаптироваться не только к новому местоположению награды в рамках одной среды, но и к новой среде.

Эксперименты тоже проводились на средах типа Gridworld без учета направления взгляда агента. В рамках каждого эксперимента все среды генерировались случайно на некотором фиксированном квадрате $n \times n$ (в основном $5 \times 5$ и $6 \times 6$).

Примеры сгенерированных сред

Легенда:

\begin{itemize}
  \item темно фиолетовый - стены
  \item желтый - начальная позиция агента
  \item салатовый - награда
\end{itemize}

Примеры сред:

% - 5x5
%   ![gridworld 5x5 #1](../../reports/v1_0/gridworld_transfer/gridworld_5x5_1_1_200_1337_map_0_1173222464.svg)
%   ![gridworld 5x5 #2](../../reports/v1_0/gridworld_transfer/gridworld_5x5_1_1_200_1337_map_2_1561234712.svg)

% - 8x8
%   ![gridworld 8x8 #1](../../reports/v1_0/gridworld_transfer/gridworld_8x8_10_1_10_1337_map_0_1173222464.svg)
%   ![gridworld 8x8 #2](../../reports/v1_0/gridworld_transfer/gridworld_8x8_10_1_10_1337_map_7_2032734714.svg)

Ключевым различием между экспериментами были суммарное число эпизодов с фиксированной наградой и суммарное число эпизодов с фиксированной средой, которые обратно пропорционально влияли на сложность - чем меньше эпизодов в распоряжении агента, тем быстрее ему требуется адаптироваться (к новой награде и/или к новой среде, соответственно).

\subsubsection{Эксперименты с редкой сменой наград и сред}

\begin{itemize}
  \item Gridworld 5x5
  \item $N_{s_0} = 100$
  \item $N_{rew} = 2$
  \item $N_{env} = 8$
\end{itemize}

Ниже представлены графики для двух разных сидов.

Сравнение агента с бейзлайнами:

% ![episode steps absolute](../../reports/v1_0/gridworld_transfer/gridworld_5x5_100_2_8_1337__steps.svg)
% ![episode steps log-relative](../../reports/v1_0/gridworld_transfer/gridworld_5x5_100_2_8_1337__steps_rel_htm_0.svg)
% ![episode steps absolute](../../reports/v1_0/gridworld_transfer/gridworld_5x5_100_2_8_42__steps.svg)
% ![episode steps log-relative](../../reports/v1_0/gridworld_transfer/gridworld_5x5_100_2_8_42__steps_rel_htm_0.svg)

Сравнение DQN жадного и $\epsilon$-жадного агента:

% ![episode steps absolute](../../reports/v1_0/gridworld_transfer/gridworld_5x5_100_2_8_1337__dqn__steps.svg)
% ![episode steps absolute](../../reports/v1_0/gridworld_transfer/gridworld_5x5_100_2_8_42__dqn__steps.svg)

Сравнение агента с разными горизонтами и размером списка целей:

% - горизонт планирования 1

%   ![episode steps absolute](../../reports/v1_0/gridworld_transfer/gridworld_5x5_100_2_8_1337__1__steps.svg)
%   ![episode steps log-relative](../../reports/v1_0/gridworld_transfer/gridworld_5x5_100_2_8_1337__1__steps_rel_htm_1_1g.svg)
%   ![episode steps absolute](../../reports/v1_0/gridworld_transfer/gridworld_5x5_100_2_8_42__1__steps.svg)
%   ![episode steps log-relative](../../reports/v1_0/gridworld_transfer/gridworld_5x5_100_2_8_42__1__steps_rel_htm_1_1g.svg)

% - горизонт планирования 2

%   ![episode steps absolute](../../reports/v1_0/gridworld_transfer/gridworld_5x5_100_2_8_1337__2__steps.svg)
%   ![episode steps log-relative](../../reports/v1_0/gridworld_transfer/gridworld_5x5_100_2_8_1337__2__steps_rel_htm_2_1g.svg)
%   ![episode steps absolute](../../reports/v1_0/gridworld_transfer/gridworld_5x5_100_2_8_42__2__steps.svg)
%   ![episode steps log-relative](../../reports/v1_0/gridworld_transfer/gridworld_5x5_100_2_8_42__2__steps_rel_htm_2_1g.svg)

% - горизонт планирования 4-8

%   ![episode steps absolute](../../reports/v1_0/gridworld_transfer/gridworld_5x5_100_2_8_1337__4-8__steps.svg)
%   ![episode steps absolute](../../reports/v1_0/gridworld_transfer/gridworld_5x5_100_2_8_1337__4-8__steps_rel_htm_4_1g.svg)
%   ![episode steps absolute](../../reports/v1_0/gridworld_transfer/gridworld_5x5_100_2_8_42__4-8__steps.svg)
%   ![episode steps absolute](../../reports/v1_0/gridworld_transfer/gridworld_5x5_100_2_8_42__4-8__steps_rel_htm_4_1g.svg)

\subsubsection{Эксперименты со средней частотой смены наград и сред}

\begin{itemize}
  \item Gridworld 5x5
  \item $N_{s_0} = 50$
  \item $N_{rew} = 4$
  \item $N_{env} = 4$
\end{itemize}

Сравнение агента с бейзлайнами:

% ![episode steps absolute](../../reports/v1_0/gridworld_transfer/gridworld_5x5_50_4_4_1337__steps.svg)
% ![episode steps log-relative](../../reports/v1_0/gridworld_transfer/gridworld_5x5_50_4_4_1337__steps_rel_htm_0.svg)

Сравнение DQN жадного и $\epsilon$-жадного агента:

% ![episode steps absolute](../../reports/v1_0/gridworld_transfer/gridworld_5x5_50_4_4_1337__dqn__steps.svg)
% ![episode steps absolute](../../reports/v1_0/gridworld_transfer/gridworld_5x5_50_4_4_1337__dqn__steps_rel_dqn_greedy.svg)

Сравнение агента с разными горизонтами и размером списка целей:

% - горизонт планирования 1

%   ![episode steps absolute](../../reports/v1_0/gridworld_transfer/gridworld_5x5_50_4_4_1337__1__steps.svg)

% - горизонт планирования 2

%   ![episode steps absolute](../../reports/v1_0/gridworld_transfer/gridworld_5x5_50_4_4_1337__2__steps.svg)

% - горизонт планирования 4-8

%   ![episode steps absolute](../../reports/v1_0/gridworld_transfer/gridworld_5x5_50_4_4_1337__4-8__steps.svg)
%   ![episode steps absolute](../../reports/v1_0/gridworld_transfer/gridworld_5x5_50_4_4_1337__4-8__steps_rel_htm_4_1g.svg)

% #### Эксперименты с частой сменой наград и сред

% - Gridworld 5x5
% - $N_{s_0} = 20$
% - $N_{rew} = 1$
% - $N_{env} = 20$

% Сравнение агента с бейзлайнами:

% ![episode steps absolute](../../reports/v1_0/gridworld_transfer/gridworld_5x5_20_1_20_1337__steps.svg)
% ![episode steps log-relative](../../reports/v1_0/gridworld_transfer/gridworld_5x5_20_1_20_1337__steps_rel_htm_0.svg)

% Сравнение DQN жадного и $\epsilon$-жадного агента:

% ![episode steps absolute](../../reports/v1_0/gridworld_transfer/gridworld_5x5_20_1_20_1337__dqn__steps.svg)

% Сравнение агента с разными горизонтами и размером списка целей:

% - горизонт планирования 1

%   ![episode steps absolute](../../reports/v1_0/gridworld_transfer/gridworld_5x5_20_1_20_1337__1__steps.svg)

% - горизонт планирования 2

%   ![episode steps absolute](../../reports/v1_0/gridworld_transfer/gridworld_5x5_20_1_20_1337__2__steps.svg)

% - горизонт планирования 4-8

%   ![episode steps absolute](../../reports/v1_0/gridworld_transfer/gridworld_5x5_20_1_20_1337__4-8__steps.svg)

% #### Эксперименты с ультрачастой сменой наград и сред

% Нужно?

% - Gridworld 5x5
% - $N_{s_0} = 1$
% - $N_{rew} = 1$
% - $N_{env} = 200$

% Сравнение агента с бейзлайнами:

% ![episode steps absolute](../../reports/v1_0/gridworld_transfer/gridworld_5x5_1_1_200_1337__steps.svg)

% Сравнение DQN жадного и $\epsilon$-жадного агента:

% ![episode steps absolute](../../reports/v1_0/gridworld_transfer/gridworld_5x5_1_1_200_1337__dqn__steps.svg)

% Сравнение агента с разными горизонтами и размером списка целей:

% - горизонт планирования 1

%   ![episode steps absolute](../../reports/v1_0/gridworld_transfer/gridworld_5x5_1_1_200_1337__1__steps.svg)

% - горизонт планирования 2

%   ![episode steps absolute](../../reports/v1_0/gridworld_transfer/gridworld_5x5_1_1_200_1337__2__steps.svg)

% - горизонт планирования 4-8

%   ![episode steps absolute](../../reports/v1_0/gridworld_transfer/gridworld_5x5_1_1_200_1337__4-8__steps.svg)

\section {Выводы}

Ключевые выводы для экспериментов с неизменной средой:

- Чем больше горизонт планирования, тем лучше играет агент
  - причем улучшение кумулятивно, т.е. каждый следующий +1 горизонта дает больший прирост, чем предыдущий
- Агент с планированием играет лучше, чем случайный
  - чем сложнее среда, тем менее заметен в относительных масштабах переход от horizon\_planning=0 к horizon\_planning=1
  - т.е. при фиксированном горизонте планирования с увеличением сложности среды результат в относительных масштабах начинает прижиматься к результатам случайного агента
- Агент с большим накопленным набором целей направленнее (=оптимальнее) исследует среду
  - предположительно потому, что менее вероятно использование случайной стратегии
  - в частных случаях это субоптимально - когда награда сразу достижима в рамках имеющегося горизонта планирования и ее положение не изменилось с предыдущего эпизода, т.к. агент с большим списком целей может выбрать для исследования псевдо-цель
  - вероятно, причина данного эффекта кроется в специальном виде использованных сред, начального положения агента и положения наград.
- Увеличение размера списка целей и величины горизонта планирования положительно сказываются на способности агента приспосабливаться к новому местоположению награды
- Увеличение горизонта планирования ухудшает производительность агента
  - однако это ухудшение происходит до некоторого момента, пока горизонт планирования не приблизится к длине оптимального пути к цели
- Агент учится и приспосабливается быстрее, чем DQN [в терминах числа эпизодов]
- Если горизонта планирования достаточно, то агент быстро учится решать задачу планирования идеально, достигая оптимальной стратегии
  - учится в ~1.5-2.5 быстрее DQN
  - но для любого фиксированного горизонта, если начать усложнять среду, то в какой-то момент DQN обгонит по результатам, потому что DQN решает задачу [почти всегда] оптимально, а htm агент - только в случае достаточного горизонта планирования
- Иногда htm agent сходится к субоптимальному решению
  - если горизонта планирования достаточно для субоптимального решения из начальной точки
  - и имеется несколько длинных узких "ходов" до цели, имеющих примерно одинаковую длину
  - тогда есть ненулевая вероятность, что агент не успеет запомнить путь через оптимальный коридор и будет планировать только через субоптимальный.
  - в такой ситуации eps-жадная стратегия DQN, наоборот, рано или поздно откроет оптимальное решение

Ключевые выводы для экспериментов со сменой сред:

- В вырожденном случае смены среды и награды на каждом эпизоде все агенты проигрывают случайному
  - DQN при этом играет значительно хуже HTM агента
- HTM агент в среднем быстрее чем DQN адаптируется к изменениям среды или в среде
  - благодаря этому обычно он ведет себя стабильнее
- Увеличение списка целей негативно сказывается на результатах агента
  - видимо, польза от выполнения псевдо-целей в среднем ниже штрафа
  - можно объяснить тем, что в среднем псевдо-цели не приближают агента к реальной и только увеличивают ожидаемое число шагов с использованием случайной стратегии

\section {Project specific terminology}

State-action (SA)

- Given trajectories : $\tau_i = (s_0, a_0, r_1, s_1, a_1, r_2, s_2, a_2, ...)$
- SA is a tuple $sa_t = (s_t, a_t)$
  - i.e. action $a_t$ taken from a state $s_t$
- TM learns agent's trajectories $\tau_i$
  - it operates with SA [encoded as] SDR sequences $(sa_0, sa_1, ... )$
  - which results that TM learns both
    - transitions $(s_{t-1}, a_{t-1}) \rightarrow s_t$
    - and which actions $a$ agent did from a state $s_t$
- TM then can make predictions
  - it can predict next state and all actions agent did from this state
  - each prediction is an SDR - a union (= bitwise OR) of pairs $(s, a^*)$
  - we give this union a special name (see below)

SA SDR Superposition

- Union (= bitwise OR) of any number of SA SDRs
- Rationale
  - if an agent saw transition $(s_{t-1}, a_{t-1}) \rightarrow s_t$ and did different actions $a_t$ from a state $s_t$
  - then prediction for a pair $(s_{t-1}, a_{t-1})$ will be a union of all seen pairs $(s_t, a_t)$
  - aka superposition of these SAs
- TM can work with SA SDR superpositions both as input and output
  - In general case (mentioned just before) TM's output is a superposition
  - TM accepts SA superpositions as input as well
    - it's possible because SA SDR superposition still is just an SDR
- *NB*: prediction process is called depolarization

\section {Encoding}

SA SDR encoder

- Encodes SA to SDR
- Uses separate state and action SDR encoders
  - Resulting SDRs are concatenated
- We consider only discrete environments
  - Hence both states and actions are discrete sets
- Each state or action can be represented as integer number
  - i.e. $s_t \in [0, |S|)$ and $a_t \in [0, |A|)$
  - to encode states and action we use Integer SDR encoder

Integer SDR encoder

% - Encodes integer number $x \in$ [0, `n_values`) to SDR
% - Parameters:
%   - `n_values` - a number of unique values
%   - `value_bits` - a number of bits used to encode each unique value
% - Resulting SDR has `total_bits` = `n_values` $\times$ `value_bits` bits
%   - they can be logically divided into `n_values` buckets of `value_bits` contiguous bits
%     - each bucket corresponds to a value from the range
%   - e.g. 2 $\rightarrow$ `0000 0000 1111` encoded by 3-by-4 integer encoder
%     - 3 buckets are separated by space to make it clear
%     - every value id encoded by 4 bits bucket
%     - note that buckets don't intersect
% - PROS
%   - Easy to encode/decode
%   - Easy to pretty print and read
%   - Hence easy to debug
%   - Different values have no intersection
%     - eliminates some unnecessary complexities
% - CONS
%   - Different values have no intersection
%     - narrows the use of SDRs
%     - doesn't use information about states similarity
%   - You have no direct control on sparsity, which is `1 / n_values`
%     - you may have problems with too low or too high sparsity levels
%     - we tested that TM works well with low sparsity
%       - which doesn't mean it has not negative effects at all
%     - just remember that sparsity may lead to problems

State SDR encoder

- For simple MDP environments we use integer encoding
- For complex environments encoding states [or observations] becomes tricky
  - of course, you still can enumerate all possible states and use Integer SDR encoder
  - but the number of all possible states grows very fast
    - in practice it works well only for small environments
  - also Integer SDR encoder cuts out information about states similarity, which may not be desirable
  - one possible solution is to encode every pixel [or grid cell] separately then concatenate results, which preserves information about similarity between states
- At the moment we use only simple MDP environments and hence integer encoding for states

\section{Planning alforithm details}

Let's consider an agent playing in some MDP environment. At some moment he is in a state, which we will denote as $s_0$, because it will be our starting point for planning.  
The agent also has a fixed set of goal states $\{g_i\}$, which he wants to reach.

The goal of the planning is to answer two questions. Is it possible to reach any of the goal states with at maximum of $n_max$ steps (i.e. actions)? And if it is possible, then what is the sequence of actions (i.e. policy) leading to that goal state?

Planning algorithm consists of 2 high-level steps: forward prediction and backtracking from the goal, which could be written in Python-like pseudocode as:

% ```python
% def plan_actions(initial_state):
%   reached_goals, active_segments_timeline = predict_to_goals(initial_state)
%   planned_actions = backtrack_from_goals(
%     reached_goals, active_segments_timeline
%   )

%   return planned_actions
% ```

\subsection{Step 1. Forward prediction}

We start from the state $s_0$ and want to predict all reachable next states $\{s_1^i\}$. To do this we could check all learned transitions $(s_0, a_0^i) \rightarrow s_1^i$. It's achievable simultaneously if we use $sa_0 = (s_0, \cup a^i)$ as a starting input for the agent's TM. $sa_0$ is then a SA SDR superposition and semantically means exactly a superposition of all possible state-action pairs $(s_0, a_0^i)$ which activate predictions for all desired transitions.

% __TBD: make simple MDP graph example__

As a prediction TM gives another SA SDR superposition $sa_1 = \cup (s_1, a_1^i)$ - a superposition of all next\_state-next\_action pairs. $sa_1$ can be further used as an input to TM, which gets us $sa_2$ prediction and so on.

As you see, for each step $k$, $sa_k$ contains a superposition of all states reachable in exactly $k$ actions (=moves). As a side note, the whole resulting process looks very similar to parallel breadth-first-search graph algorithm.

Recall that the goal of forward prediction is to find reachable goal states. Therefore at each step $k$ we look if any of the goal states $g_i$ are among the states in $sa_k$ superposition by checking their overlap $o = overlap(g_i, sa_k)$. We want to find only the closest goals, so if any of the goal states are matched, we stop forward prediction process and set these goal states as a result.

% Also during forward prediction we need to record active segments at each step $k$ denoted as `active_segments_timeline`. It's required by the second step of planning.

Note also, that cell activation and segment activation, despite named similar, refer to the different phases of TM workflow. The latter is a part of prediction, not activation - active segment depolarizes cell, i.e. predicts its future activation instead of activating it.

Below is a pseudocode for forward prediction phase.

% ```python
% def predict_to_reward(initial_state):
%   # initializes starting $sa_0$ as superposition with all possible actions
%   initial_sa = Sa(initial_state, AllActions)
%   proximal_input = encoder.encode(initial_sa)

%   reached_goals = []
%   active_segments_timeline = []
%   for i in range(max_steps):
%     # activates & depolarizes cells
%     active_cells, depolarized_cells = memory.process(
%       proximal_input, learn=False
%     )

%     # records active segments for $i$-th step
%     active_segments_timeline.append(
%       memory.active_segments(active_cells)
%     )

%     # gets $sa_{i+1}$ SA SDR superposition
%     proximal_input = columns_from_cells(depolarized_cells)

%     # gets list of matched goal states
%     reached_goals = [
%       goal_state
%       for goal_state in agent.goal_states
%       if is_matched(goal_state, proximal_input)
%     ]
%     if reached_goals:
%       break

%   return reached_goals, active_segments_timeline
% ```

\subsection{Step 2. Backtracking from goal states}

As a result of forward prediction phase we have a list of goal states reachable in $T$ actions along with recorded segments activation timeline. The goal of backtracking phase is to restore a correct sequence of actions leading to any of these goal states.

At high level view we start from any of the goal states and try moving backward [through time] to the initial state using segments activation timeline graph:

% ```python
% def backtrack_from_goals(reached_goal_states, active_segments_timeline):
%   T = len(active_segments_timeline)
%   # cells predicted to be active at timestep T, when the goal states were reached
%   depolarized_cells = active_segments_timeline[T-1].keys()

%   for goal_state in reached_goal_states:
%     # gets predicted cells related to the goal state
%     depolarized_goal_cells = intersect(goal_state, depolarized_cells)

%     # try backtracking from the goal state to the initial state
%     planning_successful, planned_actions = backtrack_from_state(
%         depolarized_goal_cells, T-1
%     )
%     if planning_successful:
%       return planned_actions
% ```

The key thing for this phase is moving backward in an implicitly built graph of active cells. Consider having chosen some goal state $g$. Let's start with the first step of backtracking - timestep $T-1$ - when the goal state $g$ was \textbf{predicted}.

Here we should pause a bit and expand the meaning ot "the goal state was predicted". Recall that we work with numeric MDP states encoded as SDR with Integer SDR encoder. A single value $x$ is decoded from an SDR $S$ means that $S$ has enough ON bits related to $x$, with "enough" is defined by \verb|activation_threshold| hyperparameter. So "the goal state $g$ was predicted" means that there were at least \verb|activation_threshold| ON bits related to it in a prediction.

From forward prediction phase we know the set of active cells at timestep $T-1$. And the fact that the goal state $g$ was predicted at that timestep creates a condition for backtracking - we have to make sure that it stays true. Obviously, the whole set of active cells at $T-1$ doesn't break this prediction condition.

But there's another condition - at each timestep there's only one transition $(s, a) \rightarrow s'$ allowed, because that's how MDP works. On the other hand the forward prediction process works with superpositions, and the set of active cells $sa_{T-1}$ may contain multiple different $(s, a)$ pairs, from which we may choose only those single pairs that don't break the prediction condition. The question is how to find these pairs?

The segments activations timeline allows us to take all active segments of the goal state $g$ cells at time $T-1$. Each active segment is associated with a set of active cells that activated it. By design each TM segment is supposed to recognize a single pattern, which is a state-action pair in our case.

\textbf{NB}: In general that's not true. A segment might recognize a set of patterns, and the most robust to the punishment during learning are those sets that have non-zero intersection, which minimizes expected punishment rate. It means, that each segment learns a cluster [or clusters] of patterns. But we can bound the segment's allowed size, which is a number of synapses, to fit no more than a single state-action pair size - then each segment will be learning to recognize only a single pattern. Still, this doesn't mean that all segment's synapses are always connected to the same state or action cells, although learning tends to fit them this way.

We might expect that the active cells sets $\{S_i^{T-1}\}$ associated with the active segments $\{A_i^{T-1}\}$ at time $T-1$ contain clusters, where each pair of active cells set has a high overlap score. Formally, we want to split $\{A_i^{T-1}\}$ into clusters $\{C_k^{T-1}| A_i, A_j \in C_k^{T-1}: overlap(A_i, A_j) \geq N_{cl}\}$ and keep only those that

At any step $k$ we have the set of cells that should be activated at $k+1$ timestep in order to reach the goal state. It means that some key portion of these cells should be predicted at timestep $k$ - at least `activation threshold` cells - which gives us

% - Given a set of depolarized cells `dep_cells` [at time $t$]
% - Call active segment as _potential_ if it induces enough depolarization among `dep_cells`
%   - _ehough_ means $\geq$ activation threshold
%   - i.e. a number of depolarized cells will be enough for subsequent depolarizations
% - Take presynaptic active cells of a potential segment as depolarized cells [at time $t-1$]

Backtracking algorithm

- Start from depolarized cells in reward = 1 columns [at time T]
- Recursively take backtracking steps for any potential segment until time 0 is reached

% Details

% - Given a set of depolarized cells `dep_cells` [at time $t$]
%   - Each depolarized cell have at least one active segment
%   - Each active segment is associated with a set of active presynaptic cells
%     - This set defines some SAR SDR
%     - i.e. each segment is conditioned on some correct SAR
% - I want to find all potential sets of active presynaptic cells that induce enough depolarization of `dep_cells`
%   - perfect case
%     - `dep_cells` conditioned on the same SAR
%     - i.e. this SAR induces full `dep_cells` depolarization
%   - perfect multiple case
%     - `dep_cells` conditioned on more than one SAR
%     - i.e. each of these SAR induces full `dep_cells` depolarization
% - real world relaxations
%   - some SARs don't induce enough depolarization
%     - i.e. $\lt$ threshold
%   - segments don't exactly match each other
%     - cells are conditioned on very similar SARs but not exactly the same

What is saved during this phase

% - active [presynaptic] cells at time $t$ are saved as `backtracking_SAR_conditions[t]`
% - they define condition on SAR at time $t-1$ to get desired depolarization at time $t$

% Второй шаг - рекурсивный бэктрекинг из столбцов, соответствующих награде, назад во времени:

% ```python
% def backtrack_from_reward(active_segments_timeline):
%     final_depolarized_cells = active_segments_timeline[T-1].keys()
%     depolarized_reward_cells = get_reward_cells(final_depolarized_cells)

%     return backtrack(depolarized_reward_cells, T-1)

% def backtrack(desired_depolarization: SparseSdr, t: int):
%     # presynaptic cells clusters, each can induce desired depolarization
%     cell_clusters = get_backtracking_candidate_clusters(
%         desired_depolarization, activation_threshold, t
%     )

%     for cluster in cell_clusters:
%         activation_timeline = backtrack(cluster, t-1)
%         if backtracking_succeeded:
%             activation_timeline.append(cluster)
%             return activation_timeline
% ```

% Откуда берутся кластеры-кандидаты:

% ```python
% def get_backtracking_candidate_clusters(
%         desired_depolarization: SparseSdr,
%         sufficient_activation_threshold: int,
%         t: int
% ):
%     # Active presynaptic cell clusters (clusterization by columns)
%     active_segments = active_segments_timeline[t][desired_depolarization]
%     candidate_clusters = merge_segments_into_clusters(active_segments)

%     # Keep clusters that induce sufficient depolarization among desired
%     for cluster in candidate_clusters:
%         count_induced_depolarization(cluster, desired_depolarization)

%         if induced_depolarization >= sufficient_activation_threshold:
%             backtracking_candidate_clusters.append(cluster)

%     return backtracking_candidate_clusters
% ```

% Последний шаг - предсказание награды по пути успешного бэктрекинга:

% ```python
% def check_activation_timeline_leads_to_reward(
%         initial_sar: Sar, activation_timeline
% ):
%     proximal_input = encoder.encode(initial_sar)

%     for i in range(T):
%         action = extract_action(activation_timeline[i])
%         proximal_input.action = action

%         active_cells, depolarized_cells = agent.process(
%             proximal_input, learn=False
%         )
%         proximal_input = columns_from_cells(depolarized_cells)

%         planned_actions.append(action)

%     if is_rewarding(proximal_input):
%         return planned_actions
% ```

% ### Step 1: Forward prediction

% Setting up

% - Given initial state $s_0$
% - To find a reward, I want to check all possible paths, starting from $s_0$
%   - Then SAR superposition  
%     (state = $s_0$, {all actions}, reward = 0)
%   - .. represents all possible beginnings of these paths
%   - So, encode it to SDR and set it as initial proximal input to TM

% Forward prediction

% - Given some SAR superposition as proximal input at time $t$
%   - induces cells activation
% - TM prediction: some SAR superposition at time $t+1$
%   - prediction is in form of depolarized cells
% - Save all active segments (discussed later in section)
% - Check if depolarized SAR superposition contains reward = 1
% - Continue forward prediction by taking predicted SAR superposition as input at $t+1$
%   - until reward is found
%   - or max forward steps are done

% A bit of details on TM work, step by step:

% - Given current proximal input [at time $t$]
% - It directly induces columns activation
% - Active columns [and depolarization from $t-1$] induce cells activation
%   - any depolarized cell in active columns becomes active
%   - or all cells become active if active column has no depolarized cells (bursting)
% - Cells activation induces segments activation
%   - number of active presynaptic cells $\geq$ activation threshold
% - Segments activation induces cells depolarization
%   - any active segment depolarizes its postsynaptic cell
% - Depolarized cells can be collapsed into depolarized columns
%   - any depolarized cell depolarizes its column
% - Depolarized columns define predicted SAR SDR superposition [at time $t+1$]

% What is saved during forward prediction phase:

% - At each time step $t$ there's a set of segments that are active
% - These active segments are saved into `active_segments_timeline[t]`
%   - as dictionary: `depolarized_cell` $\rightarrow$ `cell_active_segments_list`
%   - `cell_active_segments_list` - list of cell's active segments
%   - each active segment represented as a set of active presynaptic cells

% ### Step 2: Backtracking

% Backtracking step

% - Given a set of depolarized cells `dep_cells` [at time $t$]
% - Call active segment as _potential_ if it induces enough depolarization among `dep_cells`
%   - _ehough_ means $\geq$ activation threshold
%   - i.e. a number of depolarized cells will be enough for subsequent depolarizations
% - Take presynaptic active cells of a potential segment as depolarized cells [at time $t-1$]

% Backtracking algorithm

% - Start from depolarized cells in reward = 1 columns [at time T]
% - Recursively take backtracking steps for any potential segment until time 0 is reached

% Details

% - Given a set of depolarized cells `dep_cells` [at time $t$]
%   - Each depolarized cell have at least one active segment
%   - Each active segment is associated with a set of active presynaptic cells
%     - This set defines some SAR SDR
%     - i.e. each segment is conditioned on some correct SAR
% - I want to find all potential sets of active presynaptic cells that induce enough depolarization of `dep_cells`
%   - perfect case
%     - `dep_cells` conditioned on the same SAR
%     - i.e. this SAR induces full `dep_cells` depolarization
%   - perfect multiple case
%     - `dep_cells` conditioned on more than one SAR
%     - i.e. each of these SAR induces full `dep_cells` depolarization
% - real world relaxations
%   - some SARs don't induce enough depolarization
%     - i.e. $\lt$ threshold
%   - segments don't exactly match each other
%     - cells are conditioned on very similar SARs but not exactly the same

% What is saved during this phase

% - active [presynaptic] cells at time $t$ are saved as `backtracking_SAR_conditions[t]`
% - they define condition on SAR at time $t-1$ to get desired depolarization at time $t$

% ### Step 3: Re-check

% - Given active presynaptic cells at time $t$ needed to induce desired depolarization
% - Start forward prediction again from initial sar superposition
% - But now
%   - Check that SAR `backtracking_SAR_conditions[t]` is consistent with the current active SAR
%   - Extract action $a_t$ from `backtracking_SAR_conditions[t]`
%   - Action $a_t$ defines a path to make at time $t$
%   - Replace action superposition in proximal input with $a_t$
% - Check that given `backtracking_SAR_conditions` lead to reward

\end{document}

### 2020-09-26 htm_rl/htm_rl/common/int_sdr_encoder.py
Sequence encoder have been created.

@@ -4,6 +4,13 @@ from typing import Iterable, List
 from htm_rl.common.base_sar import Superposition
 from htm_rl.common.sdr import SparseSdr, sparse_to_dense
 from htm_rl.common.utils import isnone
+from typing import TypeVar, Generic
+from itertools import product, count
+
+import numpy as np
+
+ValueEncoder = TypeVar('ValueEncoder')
+SequenceEncoder = TypeVar('SequenceEncoder')
 
 
 @dataclass(frozen=True)
@@ -28,7 +35,31 @@ class BitRange:
         return range(self.l, self.r)
 
 
-class IntSdrEncoder:
+class BitRangePack:
+    """
+        Shorthand for representing contiguous index ranges [l1, r1), [l2, r2), ... of 1 bits.
+        Used to represent an active bucket during encoding. Use `unfold` method to get actual set of indices.
+    """
+    bit_ranges: List[BitRange]
+
+    def __init__(self, bit_ranges: List[BitRange]):
+        self.bit_ranges = bit_ranges
+
+    def shift(self, shift: int) -> 'BitRangePack':
+        new_bit_ranges = list()
+        for bit_range in self.bit_ranges:
+            new_bit_ranges.append(bit_range.shift(shift))
+        return BitRangePack(new_bit_ranges)
+
+    def unfold(self) -> Iterable[int]:
+        unfolded = list()
+        for bit_range in self.bit_ranges:
+            unfolded.extend(bit_range.unfold())
+
+        return unfolded
+
+
+class IntSdrEncoder(Generic[ValueEncoder]):
     """
     Encodes integer values from range [0, `n_values`) as SDR w/ `total_bits` bits.
     Every single value x from the range is encoded as `value_bits` contiguous 1 bits called
@@ -144,6 +175,84 @@ class IntSdrEncoder:
         return f'({name}: v{n_values} x b{value_bits})'
 
 
+class SequenceSdrEncoder(Generic[SequenceEncoder]):
+    """
+        Encodes sequence as SDR. Sequence should consist of numbers of types that have encoders.
+        Every single value from sequence is encoded by sdr encoders that correspond to
+        the sequence's values types.
+        Example:
+            TODO
+    """
+    name: str
+    encoders: List[ValueEncoder]
+    size: int
+    total_bits: int
+
+    def __init__(
+            self, name: str, encoders: List[ValueEncoder], size: int,
+            default_format: str = 'full'
+    ):
+        """
+            Initializes encoder.
+
+            :param name: a name for Sequence encoded
[…]

### 2020-09-26 htm_rl/htm_rl/envs/mymdp.py
Some bugs are fixed.

@@ -49,6 +49,7 @@ class GridWorld:
         self.world_size = world_size
         self.world_description = np.ones((world_size[0]+2, world_size[1]+2))
         self.world_description[1:-1, 1:-1] = np.array(world_description)
+        self.n_actions = 3
 
         if max_sight_range is None:
             max_sight_range = max(self.world_description.shape)
@@ -182,7 +183,7 @@ class GridWorld:
             return -1e-2
 
     def is_terminal(self):
-        if self.world_description[[self.agent_position['row'], self.agent_position['column']]] == 2:
+        if self.world_description[self.agent_position['row'], self.agent_position['column']] == 2:
             return True
         else:
             return False

### 2020-09-27 htm_rl/htm_rl/common/int_sdr_encoder.py
Semantic int encoder is added.

@@ -175,6 +175,89 @@ class IntSdrEncoder(Generic[ValueEncoder]):
         return f'({name}: v{n_values} x b{value_bits})'
 
 
+class IntSemanticSdrEncoder(IntSdrEncoder):
+    """
+    Encodes integer values from range [0, `n_values`) as SDR w/ `total_bits` bits.
+    Every single value x from the range is encoded as `value_bits` contiguous 1 bits called
+        a bucket. All bits outside corresponding bucket are set to zero.
+    To achieve semantic similarity use step parameter. Step is a half amount of bits that make up difference
+    of buckets of two adjacent integer values.
+
+
+    Example:
+        An encoder(n_values = 3, value_bits = 4, step=2) encodes values as:
+            0: 11110000
+            1: 00111100
+            2: 00001111
+    """
+
+    step: int
+    overlap: float
+
+    def __init__(self, name: str, n_values: int, value_bits: int, step: int, activation_threshold: int,
+                 default_format: str = 'full'):
+        super(IntSemanticSdrEncoder, self).__init__(name, n_values, value_bits, activation_threshold,
+                                                    default_format)
+        """
+        Initializes encoder.
+
+        :param name: a name for values encoded
+        :param n_values: defines a range [0, n_values) of values that can be encoded
+        :param value_bits: a number of bits in a bucket used to encode single value
+        :param step: a half number of bits that make up difference of buckets 
+            of two adjacent integer values.
+        :param activation_threshold: a number of 1 bits in a bucket, which are enough
+            to activate corresponding value, i.e. this value will be decoded.
+        """
+
+        self.step = step
+        self.total_bits = self.step * (self.n_values - 1) + self.value_bits
+        self.overlap = (self.value_bits - self.step) / self.value_bits
+
+    def value_activations(self, sparse_sdr: SparseSdr) -> Iterable[int]:
+        """
+           Counts number of activations in each value's bucket of bits in a given sparse SDR.
+           :param sparse_sdr: sparse SDR
+           :return: numpy array of activations - i-th elem is an activation of value i.
+        """
+
+        value_activations = np.zeros(self.n_values, dtype=np.int64)
+        for active_bit_index in sparse_sdr:
+            max_value = active_bit_index // self.step
+            min_value = (active_bit_index - self.value_bits + 1) // self.step
+
[…]

### 2020-09-28 htm_rl/htm_rl/common/int_sdr_encoder.py
Decoder bug.

@@ -309,7 +309,7 @@ class SequenceSdrEncoder(Generic[SequenceEncoder]):
         for i, encoder in enumerate(self.encoders):
             values.append(encoder.decode(
                 sparse_sdr[(self._shifts[i] <= sparse_sdr) &
-                           (sparse_sdr < self._shifts[i + 1])]
+                           (sparse_sdr < self._shifts[i + 1])] - self._shifts[i]
             )
             )
         return product(*values)

### 2020-09-28 reports/v1_0/paper.tex
add an empty skeleton for a future paper

\documentclass[a4paper]{article}
\usepackage{style}

\begin{document}

\title {Применение Temporal Memory фреймворка HTM в задачах обучения с подкреплением}
\author {Petr Kuderov}
\maketitle

\begin{abstract}
  TBD
\end{abstract}

\section{Введение}

Скелет
\begin{enumerate}
  \item abstract
  \item introduction
  \item problem in general
  \item related works
  \item problem setting and terminology: RL, HTM
  \item method
  \item testing
  \item results and outro
\end{enumerate}

Текущие вопросы
\begin{itemize}
  \item Цель: проверить применимость использования памяти агента на основе TM для моделирования среды (функции перехода)?
  \item Сужать постановку задачи до model based RL?
  \item Тестирование проводилось как для обычной задачи RL с неизменной наградой, так и для задачи meta-RL. Упоминать ли второе?
  \item Является ли метод планирования частью изначальной цели? Ведь именно ему посвящена статья.
\end{itemize}

\section{Related works}

\section{Постановка задачи}

Рассматривается

\section{Описание метода}

\section{Результаты тестирования}

\section{Выводы}

\end{document}

### 2020-10-05 htm_rl/htm_rl/agent/agent.py
Agent policy updated.

@@ -46,12 +46,17 @@ class Agent:
         self.memory.tm.reset()
         self._init_planning()
 
-    def make_step(self, state, reward, is_done, verbosity: int):
+    def make_step(self, state, reward, is_done, verbosity: int, info=None):
         trace(verbosity, 2, f'\nState: {state}; reward: {reward}')
         action = self._make_action(state, is_done, verbosity)
         trace(verbosity, 2, f'\nMake action: {action}')
 
-        self.memory.train(Sa(state, action), verbosity)
+        if info is None:
+            self.memory.train(Sa(state, action), verbosity)
+        else:
+            if (info['previous_action'] != action) or (reward > -100):
+                self.memory.train(Sa(state, action), verbosity)
+
         if reward > 0:
             self.planner.add_goal(state)
         return action
@@ -167,7 +172,7 @@ class AgentRunner:
         total_reward = 0.
         while step < self.max_steps and not done:
             state, reward, done, info = self.env.step(action)
-            action = self.agent.make_step(state, reward, done, self.verbosity)
+            action = self.agent.make_step(state, reward, done, self.verbosity, info)
             step += 1
             total_reward += reward
 

### 2020-10-05 htm_rl/htm_rl/envs/mymdp.py
Punishment is added.

@@ -75,6 +75,8 @@ class GridWorld:
                 1: rotate counterclockwise
                 2: step forward
         """
+        punish = 0
+
         if action == 0:
             self.agent_direction -= 1
             if self.agent_direction < 0:
@@ -88,27 +90,35 @@ class GridWorld:
                 # if not wall then go
                 if self.world_description[self.agent_position['row'], new_column] != 1:
                     self.agent_position['column'] = new_column
+                else:
+                    punish += 100
             elif self.agent_direction == 1:
                 # up
                 new_row = self.agent_position['row'] - 1
                 if self.world_description[new_row, self.agent_position['column']] != 1:
                     self.agent_position['row'] = new_row
+                else:
+                    punish += 100
             elif self.agent_direction == 2:
                 # left
                 new_column = self.agent_position['column'] - 1
                 if self.world_description[self.agent_position['row'], new_column] != 1:
                     self.agent_position['column'] = new_column
+                else:
+                    punish += 100
             elif self.agent_direction == 3:
                 # down
                 new_row = self.agent_position['row'] + 1
                 if self.world_description[new_row, self.agent_position['column']] != 1:
                     self.agent_position['row'] = new_row
+                else:
+                    punish += 100
             else:
                 raise ValueError
         else:
             raise ValueError
 
-        return self.observation(), self.reward(), self.is_terminal(), {}
+        return self.observation(), self.reward() - punish, self.is_terminal(), {'previous_action': action}
 
     def observation(self):
         # we see only cells that are in front of us and we can't see
@@ -118,7 +128,7 @@ class GridWorld:
 
         if self.agent_direction == 2:
             # for left
-            line = self.world_description[self.agent_position['row'], :self.agent_position['column']+1]
+            line = self.world_description[self.agent_position['row'], :self.agent_position['column']]
 
             # maze must have wall as border
             obstacle_pos = np.argwhere(line > 0)
@@ -130,7 +140,7 @@ class GridWorld:
 
         elif self.agent_direction == 0:
             # for right
-            line = self.world_description[self.agent_position['row'], self.agent_position['column']:]
+            line = self.world_description[self.agent_position['row'], self.agent_position['column']+1:]
 
             # maze must have wall as border
             obstacle_pos = np.argwhere(line > 0)
@@ -138,11 +148,11 @@ class GridWorld:
             # for right
[…]

### 2020-10-05 htm_rl/htm_rl/agent/planner.py
Skip some checking.

@@ -203,7 +203,7 @@ class Planner:
             desired_depolarization: SparseSdr, t: int,
             sufficient_activation_threshold: int, verbosity: int
     ):
-        if n_containing_segments < 5:   # obviously noise clusters
+        if n_containing_segments < 2:   # obviously noise clusters ??
             return
 
         active_cells = cell_cluster
@@ -215,13 +215,13 @@ class Planner:
                 active_cells, desired_depolarization, t
             )
 
-            if n_depolarized_cells != n_containing_segments:
-                # hypothesis: they should be equal every time, hence we never reach this branch!
-                trace(verbosity, 2, f'{n_containing_segments} -> {n_depolarized_cells}')
-                self.memory.print_cells(
-                    verbosity, 2, active_cells,
-                    f'n: {n_depolarized_cells} of {sufficient_activation_threshold}'
-                )
+            # if n_depolarized_cells != n_containing_segments:
+            #     # seems like n_depolarized_cells > n_containing_segments is normal situation
+            #     trace(verbosity, 2, f'{n_containing_segments} -> {n_depolarized_cells}')
+            #     self.memory.print_cells(
+            #         verbosity, 2, active_cells,
+            #         f'n: {n_depolarized_cells} of {sufficient_activation_threshold}'
+            #     )
 
         self.memory.print_cells(
             verbosity, 3, active_cells,

### 2020-10-05 htm_rl/htm_rl/common/int_sdr_encoder.py
encoders update

@@ -312,7 +312,7 @@ class SequenceSdrEncoder(Generic[SequenceEncoder]):
                            (sparse_sdr < self._shifts[i + 1])] - self._shifts[i]
             )
             )
-        return product(*values)
+        return list(product(*values))
 
     def format(self, sparse_sdr: SparseSdr, format_: str = None) -> str:
         """

### 2020-10-05 htm_rl/htm_rl/common/sa_sdr_encoder.py

@@ -125,3 +125,25 @@ class SaSdrEncoder:
     def _get_shifts(encoders) -> SaRelatedComposition:
         actions_shift = encoders.state.total_bits
         return SaRelatedComposition(0, actions_shift)
+
+
+def format_sa_superposition(sa_superposition: SaSuperposition) -> str:
+    """
+    Formats SA superposition
+    """
+    states = sa_superposition.state
+    actions = sa_superposition.action
+
+    format_ = 'SA superposition: \n ------------------ \n'
+
+    format_ += 'Possible states:\n'
+    for state in states:
+        format_ += str(state) + ' '
+
+    format_ += '\nPossible actions:\n'
+    for action in actions:
+        format_ += str(action) + ' '
+
+    format_ += '\n'
+
+    return format_

### 2020-10-05 reports/v1_0/paper.tex
init paper and eng report tex files

@@
 \section{Введение}
+
+Eng title in easychair: An example of biologically inspired memory-based method applied to Reinforcement Learning problem setting.
+
+Eng abstract:
+Hierarchical Temporary Memory (HTM) is a biologically inspired model of neocortex.
+Its key properties proved it to be useful useful for sequence learning and anomaly detection problems.
+While they also make it potentially useful for a model-based RL, this particular application of HTM has hardly been studied yet.
+
+We show a possible way to utilize HTM in RL setting to model an environment dynamics.
+We accompany it with an example of planning algorithm which is used to test learning abilities and find limitations of such a model.
 

### 2020-10-05 reports/v1_0/report_eng.tex
init paper and eng report tex files

--- 2020-09-25 reports/v1_0/report.tex
@@
 
-\title {HTM + RL: Project details}
+\title {!TBD! HTM + RL: Project details}
 \author {Petr Kuderov}
 \maketitle
+
+\section{TODO}
+
+\begin{itemize}
+  \item Где испытания на гридворлде с направлением взгляда агента? Если я их в итоге не использую в отчете, тогда убрать вообще упоминание о них.
+  \item Примеры сред перегенерировать с белой пустотой и черными стенами
+\end{itemize}
 
@@
 
-Переходы между состояниями однозначны, т.е. среда описывается детерминированным MDP. Когда агент пытается шагнуть "в стенку", он остается на месте.
-
+Когда агент пытается шагнуть "в стенку", он остается на месте.
+Переходы между состояниями однозначны - среда является детерминированной MDP.
 На каждом невыигрышном шаге агент получает небольшую отрицательную награду $-0.01$, на выигрышном - большую положительную $+1.0$.
@@
 Спланированную политику предоставляет планировщик.
-В случае, если он не может предоставить политику, применяется случайная.
+В случае, если он не может её предоставить, используется случайная политика.
 
@@
 
-Чтобы агент мог планировать, он наделен памятью для запоминания увиденных переходов между состояниями, которая эмулирует функцию динамики среды $f: (s, a) \rightarrow s'$.
-При этом память агента имеет некоторые особенности.
+Чтобы агент мог планировать, он наделен памятью для запоминания увиденных переходов между состояниями, которая эмулирует (или аппроксимирует) функцию динамики среды $f: (s, a) \rightarrow s'$.
+Память агента имеет некоторые особенности.
 
@@
 Это означает, что память может отвечать на запросы вида \textit{имея набор состояний $\{s_i\}$ и набор действий $\{a_j\}$, в какие состояния $\{s'_m\}$ могут привести все возможные пары $(s_i, a_j)$}?
-В ответ на такой запрос память возвращает суперпозицию следующих состояний.
+Ответом на такой запрос является суперпозиция следующих состояний.
 
-Вторая особенность - для каждого конкретного состояния $s'$ из суперпозиции в ответе на запрос к памяти есть возможность узнать $(s, a) = f^{-1}(s')$
+Вторая особенность - в ответе на запрос к памяти для каждого конкретного состояния $s'_m$ суперпозиции состояний есть возможность узнать, какие пары состояние-действие приводят в $s'_m$: $\{(s_i, a_j)\} = f^{-1}(s'_m)$.
 
-Третья особенность - память агента дополнительно снабжена подмодулем памяти о выигрышных состояниях.
+Третья особенность - память агента дополнительно снабжена подмодулем, запоминающим выигрышные состояния.
 На текущий момент этот подмодуль реализован с помощью наивного списка состояний, в которых агент получил положительную награду - мы называем этот подмодуль списком целей.
-Список целей имеет ограниченный размер S, являющийся гиперпараметром.
+Список целей имеет ограниченный размер S, являющийся гиперпараметром, и работает как кольцевая очередь.
 
@@
 Агент опирается на свой опыт и исходит из предположения, что награда может находиться в одном из выигрышных состояний предыдущих эпизодов.
-Поэтому состояния из списка целей являются приоритетными для исследования.
+Но даже если и нет, то само по себе направленное перемещение будет выгоднее случайного блуждания с точки зрения исследования среды.
+Поэтому состояния из списка целей являются приоритетными.
 
-Пусть в некоторый момент времени $t_0$ агент находится в состоянии $s_0$.
-Чтобы ответить на вопрос, есть ли достижимое состояние из списка целей в радиусе $N$ действий, начнем планировать из суперпозиции $(s_0, A)$, где $A$ - все множество доступных действий агента.
+Пусть в некоторый момент времени $t_0$ агент находится в состоянии $S_0 = s_0$.
+Чтобы ответить на вопрос, есть ли достижимое состояние из списка целей в радиусе $N$ действий, начнем планировать из суперпозиции $(S_0, A)$ (TODO: объяснить обозначения), где $A$ - все множество доступных действий агента.
 
-После первого шага предсказания мы будем иметь множество состояний $\{s_i\}^1$, в которые можно попасть из текущего за одно действие агента.
+После первого шага предсказания мы будем иметь множество состояний $S_1 = \{s^i\}$, в которые можно попасть из текущего за одно действие агента.
 Продолжим планировать из суперпозиции этих состояний так же по всем возможным действиям.
@@
 
-NB: первый этап может обнаружить сразу несколько достижимых целевых состояний в радиусе $T$ шагов.
-Второй этап проводится последовательно для каждого из этих состояний до первого успеха.
-Выбор очередности перебора целей в текущей реализации явно не определен.
+NB: первый этап может обнаружить сразу несколько достижимых целевых состояний в радиусе $T$ шагов. Выбирается какое-то одно из них, причем процедура выбора в текущей реализации явно не определена.
 
@@
 
-Итак, на текущий момент, планировщик уверен, что некоторое целевое состояние $s_T$может быть достигнуто за $T$ шагов, но не знает в точности как - на первом этапе мы работали с суперпозициями и использовали все возможные действия, поэтому какие конкретно действия привели к целевому состоянию, неизвестно.
+Итак, на текущий момент, планировщик уверен, что некоторое целевое состояние $s_T$ может быть достигнуто за $T$ шагов, но не знает в точности как - на первом этапе мы работали с суперпозициями и использовали все возможные действия, поэтому, какие конкретно действия привели к целевому состоянию, неизвестно.
 
-На момент окончания успешного планирования за $T$ шагов, имеем всю историю суперпозиций состояний на каждом из этих шагов $[s_0, \{s_i\}^1, .. \{s_i\}^t]$.
+В процессе проведения первого шага сохраняется информация о суперпозициях состояний на каждом из этих шагов $[S_0, S_1, .. S_T]$.
 
@@
 Здесь, правда, стоит уточнить, что $f^{-1}$ в общем случае возвращает не единственную пару $(s, a)$, а множество пар, и не всякая пара рекурсивно приведет нас обратно в начальную позицию (т.е.
-реально достижима из начальной позиции).
-Поэтому рекурсивно проверяется каждая из возможных пар до первого успешного возвращения в начало.
+реально достижима из начальной позиции за нужное число шагов).
+Поэтому рекурсивно проверяется каждая пар-претендентов до первого успешного возвращения в $s_0$.
 
@@
 Если процесс нахождения обратного пути завершился успешно, планировщик имеет политику действий на $T$ шагов вперед, которую агент безусловно выполняет.
-В случае успеха агент находит награду, и эпизод завершается, иначе агент исключает рассмотренное целевое состояние из списка целей до окончания текущего эпизода и пытается заново спланировать действия.
+Если в результате он находит награду, эпизод завершается, иначе агент исключает рассмотренное целевое состояние из списка целей до окончания текущего эпизода и пытается заново спланировать действия.
 
@@
 
-Тестирование агента содержало два различных набора экспериментов в соответствии с исследуемыми задачами transfer learning, отличающихся количеством сред в каждом из экспериментов.
+Тестирование агента содержало два различных набора экспериментов, отличающихся количеством сред в каждом из экспериментов.
 
@@
 
-\subsection{Тестирование на задаче transfer learning в одной среде}
+\subsection{Тестирование на задаче RL в фиксированной среде}
 
-В рамках одной среды проводилась серия испытаний по N эпизодов.
-В течение одного испытания из N эпизодов агент учился находить награду в некоторой ячейке.
-Между испытаниями местоположение награды изменялось.
-Исследовалась способность агентов адаптироваться к новому местоположению награды.
+Испытания проводились в трех заданных вручную средах-лабиринтах типа Gridworld без учета направления взгляда агента.
 
-Для этого набора экспериментов были выбраны 3 среды типа Gridworld без учета направления взгляда агента.
+В рамках испытания (т.е.
+на протяжении всего времени жизни агента) среда фиксирована, но местоположение награды изменялось каждые N эпизодов.
+Исследовалась способность агентов находить и адаптироваться к новому местоположению награды.
 
-Ниже к каждой из этих сред приведены графики результатов рассматриваемых агентов: а) награда за эпизод, б) количество шагов за эпизод и в) длительность выполнения одного эпизода (в секундах).
+Ниже к каждой из трёх сред приведены графики результатов рассматриваемых агентов: а) награда за эпизод, б) количество шагов за эпизод и в) длительность выполнения одного эпизода (в секундах).
 На всех графиках изображена скользящая средняя с окном, указанным в заголовке графика (например, \verb|MA = 20|)
@@
 
-\subsection{Тестирование на задаче transfer learning в нескольких средах}
+\subsection{Тестирование на задаче RL в сменяющихся средах}
 
-Эксперименты данного режима проводились на наборе различных сред-лабиринтов. Схема каждого эксперимента следующая:
+Испытание данного режима проводились на множестве случайно сгенерированных сред-лабиринтах типа GridWorld без учета направления взгляда агента. Схема каждого испытания следующая:
 
 \begin{itemize}
-  \item эксперимент проводится последовательно на $N_{env}$ средах
+  \item испытание проводится последовательно на $N_{env}$ средах
   \item при фиксированной среде последовательно рассматриваются $N_{rew}$ положений награды
   \item при фиксированном положении награды рассматривается $N_{s_0}$ начальных положений агента
-  \item в фиксированной конфигурации агент играет $N_{ep}$ эпизодов
+  \item в фиксированной конфигурации агент играет $N$ эпизодов
 \end{itemize}
 
-Исследовалась способность агента адаптироваться не только к новому местоположению награды в рамках одной среды, но и к новой среде.
+Таким образом испытание в общей сложности содержит $N_{env} \cdot N_{rew} \cdot N_{s_0} \cdot N$ эпизодов.
 
-Эксперименты тоже проводились на средах типа Gridworld без учета направления взгляда агента.
-В рамках каждого эксперимента все среды генерировались случайно на некотором фиксированном квадрате $n \times n$ (в основном $5 \times 5$ и $6 \times 6$).
+Исследовалась способность агента адаптироваться не только к новому местоположению награды в рамках одной среды, но и к новой среде целиком.
+
+Среды генерировались случайно на некотором фиксированном квадрате размера $n \times n$.
+Ииспытания проводились в средах $5 \times 5$, $6 \times 6$ и $8 \times 8$, но в отчете приведены результаты только в средах размера $5 \times 5$ - в остальных случаях результаты были схожими.
+
+Ключевым различием между испытаниями были суммарное число эпизодов с фиксированной наградой и средой, которые влияли на сложность обратно пропорционально - чем меньше эпизодов в распоряжении агента, тем быстрее ему требуется адаптироваться (к новой награде и/или к новой среде, соответственно).
 
@@
 %   ![gridworld 8x8 #2](../../reports/v1_0/gridworld_transfer/gridworld_8x8_10_1_10_1337_map_7_2032734714.svg)
-
-Ключевым различием между экспериментами были суммарное число эпизодов с фиксированной наградой и суммарное число эпизодов с фиксированной средой, которые обратно пропорционально влияли на сложность - чем меньше эпизодов в распоряжении агента, тем быстрее ему требуется адаптироваться (к новой награде и/или к новой среде, соответственно).
 
@@
 
-Let's consider an agent playing in some MDP environment. At some moment he is in a state, which we will denote as $s_0$, because it will be our starting point for planning.  
+Let's consider an agent playing in some MDP environment. At some point he is in a state, which we will denote as $s_0$, because it's used as a starting point for planning.  
 The agent also has a fixed set of goal states $\{g_i\}$, which he wants to reach.

### 2020-10-05 reports/v1_0/report.tex
report draft

@@
 
-\title {HTM + RL: Project details}
+\title {!TBD! HTM + RL: Project details}
 \author {Petr Kuderov}
 \maketitle
+
+\section{TODO}
+
+\begin{itemize}
+  \item Где испытания на гридворлде с направлением взгляда агента? Если я их в итоге не использую в отчете, тогда убрать вообще упоминание о них.
+  \item Примеры сред перегенерировать с белой пустотой и черными стенами
+  \item Блоки-сноски, помеченные NB, можно сделать в виде нормальных сносок.
+\end{itemize}
 
@@
 
-Переходы между состояниями однозначны, т.е. среда описывается детерминированным MDP. Когда агент пытается шагнуть "в стенку", он остается на месте.
-
+Когда агент пытается шагнуть "в стенку", он остается на месте.
+Переходы между состояниями однозначны - среда является детерминированной MDP.
 На каждом невыигрышном шаге агент получает небольшую отрицательную награду $-0.01$, на выигрышном - большую положительную $+1.0$.
 
-\section{Agent}
+\section{Агент}
 
@@
 
-Чтобы агент мог планировать, он наделен памятью для запоминания увиденных переходов между состояниями, которая эмулирует функцию динамики среды $f: (s, a) \rightarrow s'$.
-При этом память агента имеет некоторые особенности.
+\begin{function}
+  \SetAlgoNoLine
+  $s_0$: initial state \;
+  $g: $ reached goals \;
+  $act\_seg: $ active segments timeline \;
+  \quad \\
 
-Главная особенность - она способна работать с суперпозициями (=множествами).
+  $g, act\_seg \leftarrow PredictToGoal(s_0)$ \;
+  $plan \leftarrow BacktrackFromGoal(g, s_0, act\_seg)$ \;
+
+  \Return{$[a_0, a_1, .. a_T] = plan$}
+  \caption{PlanActions($s_0$)}
+\end{function}
+
+Чтобы агент мог планировать, он наделен памятью для запоминания увиденных переходов между состояниями, которая эмулирует (или аппроксимирует) функцию динамики среды $f: (s, a) \rightarrow s'$.
+Память агента имеет некоторые особенности.
+
+\subsection{Особенности устройства памяти агента}
+
+TODO поменять и разделить на 1) работает с SDR, их можно пересекать и объединять, 2) память устроена так, что не может, а работает с суперпозициями.
+Возможно еще стоит пока временно опустить слово суперпозиция.
+Кажется, что семантика отличия от объединения - в том, что объединяются именно цельные сущности.
+
+Главная особенность работы памяти агента - она способна работать с суперпозициями (=множествами).
 Это означает, что память может отвечать на запросы вида \textit{имея набор состояний $\{s_i\}$ и набор действий $\{a_j\}$, в какие состояния $\{s'_m\}$ могут привести все возможные пары $(s_i, a_j)$}?
-В ответ на такой запрос память возвращает суперпозицию следующих состояний.
+Ответом на такой запрос является суперпозиция следующих состояний $S = \cup s'_m$.
 
-Вторая особенность - для каждого конкретного состояния $s'$ из суперпозиции в ответе на запрос к памяти есть возможность узнать $(s, a) = f^{-1}(s')$
+Вторая особенность - в ответе на запрос к памяти для каждого конкретного состояния $s'_m$ суперпозиции состояний есть возможность узнать, какие пары состояние-действие приводят в $s'_m$: $\{(s_i, a_j)\} = f^{-1}(s'_m)$.
 
-Третья особенность - память агента дополнительно снабжена подмодулем памяти о выигрышных состояниях.
+Третья особенность - память агента дополнительно снабжена подмодулем, запоминающим выигрышные состояния.
 На текущий момент этот подмодуль реализован с помощью наивного списка состояний, в которых агент получил положительную награду - мы называем этот подмодуль списком целей.
-Список целей имеет ограниченный размер S, являющийся гиперпараметром.
+Список целей имеет ограниченный размер S, являющийся гиперпараметром, и работает как кольцевая очередь.
 
-На основе этих особенностей и построен алгоритм планировщика.
+На основе этих особенностей и построен алгоритм работы планировщика.
 
-\subsection{Планировщик. Шаг 1: планирование до цели}
+\subsection{TM в качестве памяти агента}
+
+Перед тем как перейти к описанию алгоритма планирования, разберем для начала, какими данными оперирует агент, и в каком формате они представлены.
+
+Пусть трактория движения агента - $\tau = (s_0, a_0, r_1, s_1, a_1, r_2, s_2, a_2, ...)$.
+Пара состояние-действие $(s_t, a_t)$ некоторого момента времени задает предусловие перехода в среде: $(s_t, a_t) \rightarrow s_{t+1}$.
+
+Память агента - это TM, которая запоминает следующие последовательности $[(s_0, a_0), (s_1, a_1), \dots]$.
+Сама TM устроена таким образом, что работает с разреженными векторами (SDR), поэтому пары состояние-действие требуется как-то закодировать в SDR.
+
+Обратите внимание, что TM в нашем случае учит
+\begin{itemize}
+  \item и переходы $(s, a) \rightarrow s'$,
+  \item и какие действия $\{a^i_{t+1}\}$ совершал агент, попав в некоторое состояние $s_{t+1}$.
+\end{itemize}
+
+Также обученная TM умеет делать предсказания - для текущей пары состояние-действие $(s_t, a_t)$ давать совокупность всех возможных продолжений $(s_{t+1}, a_{t+1})$.
+"Совокупность" выражается как побитовое объединение разреженных векторов, соответствующих этим парам.
+Такую совокупность мы называем суперпозицией состояний-действий.
+Cам же процесс предсказания называется деполяризацией и является естественным этапом цикла работы ТМ: активация-обучение-деполяризация-обучение-активация-обучение-деполяризация-\dots
+
+Т.к.
+TM работает с разреженными векторами в общем смысле, вектор, являющийся суперпозицией состояний-действий, является также валидным входным значением для ее активации [, означая совокупность составляющих этой суперпозиции].
+
+\subsubsection{Кодирование состояний и действий}
+
+И состояния, и действия в рамках рассматриваемых MDP - это целые числа из интервалов $[0, |S|)$ и $[0, |A|)$ соответственно.
+Каждый элемент пары кодируется независимо целочисленным кодировщиком - для состояний и действий он разный.
+Далее два полученных SDR вектора конкатенируются в итоговый SDR вектор.
+
+Целочисленный кодировщик работает с числами некоторого фиксированного интервала $[0, N)$ и каждому числу интервала ставит в соответствие одинаковое количество активных бит.
+Работу целочисленного кодировщика проще показать на примере.
+Возьмем для этой случай $N = 3$.
+Вот как кодируются числа 0, 1 и 2 соответственно:
+
+\begin{itemize}
+  \item $0 \rightarrow$ \verb|1111 0000 0000|
+  \item $1 \rightarrow$ \verb|0000 1111 0000|
+  \item $2 \rightarrow$ \verb|0000 0000 1111|
+\end{itemize}
+
+На примере видно, что итоговый вектор разбивается на $N$ групп одинаковой длины (buckets), и каждая группа соответствует только одному из значений интервала.
+Длина одной группы является гиперпараметром (в примере выше длина равна 4).
+
+Такой кодировщик обладает следующими свойствами:
+
+\begin{itemize}
+  \item Плюсы
+  \begin{itemize}
+    \item простой и быстрый алгоритм кодирования/декодирования
+    \item закодированные значения легкочитаемы во время дебага
+    \item разные значения не пересекаются, упрощая некоторые допущения в алгоритмах
+  \end{itemize}
+
+  \item Минусы
+  \begin{itemize}
+    \item разные значения не пересекаются, так что с точки зрения меры "близости" все значения не имеют ничего общего между собой
+    \item нет контроля над разреженностью закодированного вектора - она всегда равна $\frac{1}{N}$ и может оказаться вне удачных диапазонов.
+На протестированных нами примерах с низкой разреженностью ТМ показала себя хорошо.
+Тем не менее стоит помнить, что это потенциально проблемное место.
+  \end{itemize}
+\end{itemize}
+
+TODO упомянуть про кодирование состояний в сложных средах. State SDR encoder
+
+- For simple MDP environments we use integer encoding
+- For complex environments encoding states [or observations] becomes tricky
+  - of course, you still can enumerate all possible states and use Integer SDR encoder
+  - but the number of all possible states grows very fast
+    - in practice it works well only for small environments
+  - also Integer SDR encoder cuts out information about states similarity, which may not be desirable
+  - one possible solution is to encode every pixel [or grid cell] separately then concatenate results, which preserves information about similarity between states
+- At the moment we use only simple MDP environments and hence integer encoding for states
+
+TODO упомянуть про OR суперпозицию в виде примера. Пример суперпозиции двух векторов, кодирующих числа 2 и 5 отрезка [1, 6]:
+
+\begin{verbatim}
+  0000 1111 0000 0000 1111 0000
+  =
+  0000 1111 0000 0000 0000 0000
+  OR
+  0000 0000 0000 0000 1111 0000
+\end{verbatim}
+
+\subsection{Планировщик. Этап 1: планирование до цели}
+
+TODO proofread не забыл ли я что-то из англ версии
 
 Агент опирается на свой опыт и исходит из предположения, что награда может находиться в одном из выигрышных состояний предыдущих эпизодов.
-Поэтому состояния из списка целей являются приоритетными для исследования.
+Но даже если и нет, то само по себе направленное перемещение будет выгоднее случайного блуждания с точки зрения исследования среды.
+Поэтому состояния из списка целей являются приоритетными.
 
-Пусть в некоторый момент времени $t_0$ агент находится в состоянии $s_0$.
-Чтобы ответить на вопрос, есть ли достижимое состояние из списка целей в радиусе $N$ действий, начнем планировать из суперпозиции $(s_0, A)$, где $A$ - все множество доступных действий агента.
+Пусть в некоторый момент времени $t_0$ агент находится в состоянии $S_0 = s_0$.
+Чтобы ответить на вопрос, есть ли достижимое состояние из списка целей в радиусе $N$ действий, начнем планировать из суперпозиции $(S_0, A = \cup a^i)$ (TODO: объяснить обозначения), где $A$ - все множество доступных действий агента.
+TODO рисунок и аналогия с BFS
 
-После первого шага предсказания мы будем иметь множество состояний $\{s_i\}^1$, в которые можно попасть из текущего за одно действие агента.
+После первого шага предсказания мы будем иметь множество состояний $S_1 = \{s^i\}$, в которые можно попасть из текущего за одно действие агента.
 Продолжим планировать из суперпозиции этих состояний так же по всем возможным действиям.
@@
 
-NB: первый этап может обнаружить сразу несколько достижимых целевых состояний в радиусе $T$ шагов.
-Второй этап проводится последовательно для каждого из этих состояний до первого успеха.
-Выбор очередности перебора целей в текущей реализации явно не определен.
+В процессе проведения данного этапа на каждом шаге планировщик также сохраняет информацию о суперпозиции состояний, формируя их историю $[S_0, S_1, ..
+S_T]$.
+Более детально - для каждого шага сохраняются активации сегментов ячеек ТМ (TODO линк на что такое сегменты ТМ).
+Это позволяет не потерять информацию о зависимостях между переходами от одного шага к следующему - какие пары состояний-действий повлекли за собой активацию каких пар состояние-действие.
 
-\subsection{Планировщик. Шаг 2: нахождение обратного пути}
+TODO check different code styles
+\begin{function}
+  % \SetAlgoNoLine
+  \KwData{initial state $s_0$}
+  $sa_0$: state-action pair \;
+  $p_i$: proximal input \;
+  $act\_seg$: active segments timeline \;
+  \quad \\
 
-Итак, на текущий момент, планировщик уверен, что некоторое целевое состояние $s_T$может быть достигнуто за $T$ шагов, но не знает в точности как - на первом этапе мы работали с суперпозициями и использовали все возможные действия, поэтому какие конкретно действия привели к целевому состоянию, неизвестно.
+  $sa_0 \leftarrow (s_0, \cup a^i)$ \;
+  $p_0 \leftarrow Encode(sa_0)$ \;
 
-На момент окончания успешного планирования за $T$ шагов, имеем всю историю суперпозиций состояний на каждом из этих шагов $[s_0, \{s_i\}^1, .. \{s_i\}^t]$.
+  \ForEach {backward step $i$} {
+    $ActivateDepolarize(p_i)$ \tcc{Activate and depolarize TM cells}
 
-По целевому состоянию планировщик выбирает пару состояние-действие $(s_{T-1}, a_{T-1}) = f^{-1}(s_T)$, которые привели к нему.
-Таким образом для предпоследнего шага известно требуемое действие $a_{T-1}$ и состояние $s_{T-1}$, в которое нужно прийти.
+    $p_{i+1} \leftarrow ColumnsFromCells(depolarized\_cells)$ \tcc{Set proximal input for next step}
 
-Далее рекурсивно повторяем процедуру, на каждом шаге $t$ находя пару состояние-действие $(s_{t-1}, a_{t-1}) = f^{-1}(s_t)$, пока не вернемся в начальную позицию $s_0$.
+    $act\_seg[i] \leftarrow active\_segments$ \tcc{Record active segments for $i$-th step}
+    \quad \\
 
-Здесь, правда, стоит уточнить, что $f^{-1}$ в общем случае возвращает не единственную пару $(s, a)$, а множество пар, и не всякая пара рекурсивно приведет нас обратно в начальную позицию (т.е.
-реально достижима из начальной позиции).
-Поэтому рекурсивно проверяется каждая из возможных пар до первого успешного возвращения в начало.
+    $s, a \leftarrow Decode(p_{i+1})$ \;
+    $reached\_goal \leftarrow MatchGoals(s)$ \;
+    \If(){$reached\_goal \neq \emptyset$} {
+      \textbf{break}
+    }{}
+  }
 
-Заметим, что в процессе возвращения назад алгоритм формирует однозначную последовательность действий $[a_0, ..
-, a_{T-1}]$, которая приведет агента из начального состояния $s_0$ в искомое целевое состояние $s_T$.
+  \Return {$reached\_goal, act\_seg$} \;
+
+  \caption{PredictForward(initial state $s_0$)}
+\end{function}
+
+\textbf{NB}: первый этап может обнаружить сразу несколько достижимых целевых состояний в радиусе $T$ шагов.
+Выбирается какое-то одно из них, причем процедура выбора в текущей реализации явно не определена.
+
+\subsection{Планировщик. Этап 2: нахождение обратного пути}
+
+Итак, на момент окончания первого этапа планировщик "верит", что некоторое целевое состояние $g = s_T$ может быть достигнуто за $T$ шагов, но не знает в точности как.
+Логичным образом цель второго этапа - восстановить
+точную последовательность действий, приводящих из начального состояния $s_0$ в целевое $s_T$.
+
+Решать эту задачу предлагается с конца - двигаясь по графу истории активаций сегментов из конечного целевого состояния назад во времени, с целью попасть в начальное состояние на нулевом шаге.
+В связи с таким "движением обратно", данный этот этап назван бэктрекингом.
+
+TODO рисунок-схема
+
+\begin{function}
+  \SetAlgoNoLine
+  \KwData{$g = s_T$ - goal state; $act\_seg$ - active segments timeline}
+  \quad \\
+
+  $S_T$: state superposition at time $T$ \;
+  $s_T \leftarrow s_T \cap S_T$ \;
+  $(successful, actions) \leftarrow Backtrack(s_T, T-1, act\_seg)$ \;
+  \Return {$successful, actions$}
+  \caption{BacktrackFromGoal($g$, $act\_seg$)}
+\end{function}
+
+\subsubsection{Бэктрекинг: упрощенно}
+
+Сперва рассмотрим алгоритм бэктрекинга в немного упрощенном виде на уровне идеи, а в следующей секции будут описаны детали реализации.
+
+Напомним, что в процессе проведения первого шага сохраняется информация о суперпозициях состояний на каждом из этих шагов $[S_0, S_1, ..
+S_T]$.
+Также работа памяти планировщика устроена таким образом, что позволяет не только предсказывать следующее состояние $s' = f(s, a)$ на основе аппроксимированной функции перехода среды, но и обратно по целевому состоянию $s'$ определить пару состояние-действие $(s, a) = f^{-1}(s')$, которая привела к нему.
+
+Получается, имея конечное целевое состояние $g = s_T$, планировщик может определить, какая пара состояние-действие $(s_{T-1}, a_{T-1})$ привела в него.
+Следовательно, для шага $T-1$ известно требуемое действие $a_{T-1}$ и состояние $s_{T-1}$, в которое нужно прийти, а значит задача поиска пути из $s_0$ в $s_T$ рекурсивно сводится к задаче поиска пути в $s_{T-1}$.
+
+Повторяя процедуру, на каждом шаге $t$ находим пару состояние-действие $(s_{t-1}, a_{t-1}) = f^{-1}(s_t)$, пока не вернемся в начальную позицию $s_0$.
+
+Стоит уточнить, что $f^{-1}(s')$ в общем случае возвращает не единственную пару $(s, a)$, а множество пар, каждая из которых удовлетворяет условию $s' = f(s, a)$.
+Однако не всякая пара рекурсивно приведет нас обратно в начальную позицию именно в начальй момент времени, т.е.
+реально достижима из начальной позиции за нужное число шагов.
+Поэтому рекурсивно проверяется каждая из пар-претендентов до первого успешного возвращения в $s_0$.
+
+В результате работы бэктрекинга формируется однозначная последовательность действий $[a_0, ..
+, a_{T-1}]$, которая с точки зрения аппроксимированной функции перехода приведет агента из начального состояния $s_0$ в искомое целевое состояние $g = s_T$.
+
+TODO high-level backtracking algo
 
 Если процесс нахождения обратного пути завершился успешно, планировщик имеет политику действий на $T$ шагов вперед, которую агент безусловно выполняет.
-В случае успеха агент находит награду, и эпизод завершается, иначе агент исключает рассмотренное целевое состояние из списка целей до окончания текущего эпизода и пытается заново спланировать действия.
+Если после исполнения плана он находит награду, эпизод завершится, иначе агент исключает рассмотренное целевое состояние из списка целей до окончания текущего эпизода и пытается заново спланировать действия.
+
+\subsubsection{Бэктрекинг: детали реализации}
+
+TODO эту часть возможно стоит наверх туда, где описана работа памяти планировщика.
+
+Чтобы перейти к деталям бэктрекинга, напомним принцип работы TM планировщика.
+Она работает с последовательностями пар состояние-действие.
+Пары состояние-действие кодируются в SDR вектор и последовательно подаются на вход TM для запоминания.
+
+Сама ТМ представлена в виде набора столбцов ячеек, где каждый столбец соответствует одному биту входного вектора.
+Входной SDR вектор вызывает активацию столбцов, что приводит к активации ячеек столбца.
+При этом активируются только так называемые деполяризованные ячейки столбца, если таковые имеются.
+Иначе активируются все ячейки столбца (bursting).
+Получается такая контекстно-зависимая активация ячеек.
+В этой активации одновременно представлен и текущий "символ" последовательности, если смотреть с точки зрения активных столбцов (образно, вид анфас), и текущий префикс (или история) последовательности, если смотреть на уровне конфигурации активных ячеек (образно, вид в профиль).
+
+Активация ячеек ТМ делает возможным следующий промежуточный шаг ее работы - деполяризацию ячеек.
+Деполяризация формирует ожидание продолжения последовательности - какой SDR вектор будет подан на вход, а значит какие столбцы будут активированы в следующий момент времени.
+Ожидание выражается в переводе части ячеек этих столбцов в выделенное, деполяризованное, состояние.
+
+Процесс деполяризации происходит следующим образом.
+Все ячейки ТМ потенциально имеют связи друг с другом, и в процессе обучения некоторые из этих связей формируются в способные к активации (синапсы).
+Синапсы одной ячейки логически разделены на независимые группы - сегменты, где в разных сегментах могут независимо присутствовать одинаковые синапсы.
+Сегмент становится активным, когда количество активных пресинаптических ячеек больше некоторого порога.
+Активация сегмента ячейки в свою очередь ведет к ее деполяризации.
+
+Именно сегменты олицетворяют собой "ожидания" ячейки, формируя условия для её деполяризации, а впоследствии и активации.
+Получается следующая логика от лица ячейки: если в текущий момент времени один из моих сегментов распознал некоторый контекст, с большой вероятностью следует ожидать, что следом будет активирован и мой столбец, т.к.
+похоже на то, что похожую последовательность мы уже раньше наблюдали.
+В таком случае ячейка переходит в предсказательное деполяризованное состояние, и, если на следующем шаге ее столбец действительно оказывается активным, она тоже активируется.
+
+\textbf{NB}.
+Активный сегмент устанавливает двустороннюю связь между деполяризованной ячейкой и ячейками ее деполяризующими.
+Именно эта особенность будет использована для обращения функции перехода.
+Основная трудность в этой задачи заключается в том, как подняться от уровня ячеек до уровня состояний-действий.
+
+TODO схема активации и деполяризации
+
+Из первого этапа планирования известно, что искомая цель $g = s_T$ содержится в финальной предсказанной суперпозиции $S_T$, т.е.
+пересечение соответствующих им SDR векторов выше определенного порога, и можно говорить о том, что данная цель является частью суперпозиции.
+Далее под искомой целью будет пониматься уже именно их пересечение с суперпозицией $g = s_T \cap S_T$.
+
+Исходя из принципа работы планировщика на первом этапе, раз ячейка активна на шаге $t$, значит она была деполяризована на шаге $t-1$ - т.е.
+у нее имелся хотя бы один активный сегмент.
+Рассмотрим активные сегменты ячеек столбцов, соответствующих искомой цели $s_T$.
+Каждый такой сегмент задает условие деполяризации своей ячейки, и в нашем случае это некоторая пара состояние-действие.
+
+\begin{framed}
+  \textbf{NB}.
+В общем случае все не совсем так.
+Сегменты учатся распознавать наборы паттернов.
+Наиболее стойкие к наказанию в процессе обучения будут те наборы, которые имеют ненулевое пересечение паттернов между собой, что минимизирует ожидаемую частоту наказаний.
+Это означает, что каждый сегмент выучивает кластер или несколько кластеров близких между собой паттернов.
+Однако можно ограничить размеры сегментов, т.е.
+число допустимых в сегменте синапсов, до размера одного паттерна.
+В нашем случае это размер входного SDR вектора, или число столбцов.
+Тогда каждый сегмент сможет выучивать не больше одного очень плотно сгруппированного кластера паттернов.
+А, учитывая специфику кодирования пар состояние-действие, паттерн будет один с точностью до небольшой погрешности - одна единственная пара состояние-действие.
+\end{framed}
+
+Получается, множество активных сегментов, ответственных за деполяризацию целевого состояния, задает множество множеств активных ячеек, обеспечивающих её.
+Далее в ход идет предположение, что это множество множеств распадается на кластеры близких множеств, где каждый кластер соответствует конкретной паре состояние-действие, активных на $T-1$ шаге.
+Эти пары следом фильтруются на предмет того, достаточную ли деполяризацию каждая из них индуцирует.
+
+\textbf{NB}.
+Алгоритм кластеризации работает следующим образом.
+Стартуя с набора кластеров, соответствующих каждому из рассматриваемых сегментов, он итеративно склеивает существенно пересекающиеся кластеры пока это возможно.
+
+Прошедшие отбор пары являются кандидатами на искомый переход $(s_{T-1}, a_{T-1}) \rightarrow s_T$, поэтому каждый из соответствующих им кластеров ячеек рекурсивно проверяется бэктрекингом до первого успеха.
+Получается, на первом шаге входным условием для шага бэктрекинга являлась деполяризация ячеек целевого состояния, на втором - деполяризация кластера ячеек одного из рассматриваемых кандидатов и так далее.
+
+\begin{function}
+  \KwData{$d_t$: required cells depolarization at time $t$, $t$: timestep}
+  \If(){$t \leq 0$} {
+    \Return{True, $\emptyset$}
+  }{}
+
+  Init set of clusters $C = \{c_i\}$ as presynaptic cells of active segments of the $d_t$ cells. \;
+  Iteratively try to merge any two pair of sufficiently intersecting clusters \;
+  OR: \;
+  \While{$\exists c_i, c_j: match(c_i, c_j) \geq \Theta_{clust}$} {
+    $MergeClusters(i, j)$ \tcc{Merges them into one cluster $c = c_i \cup c_j$}
+  }
+  \ForEach{$c \in C$}{
+    Count how many of $d_t$ cells are depolarized by the cluster. \;
+    Keep in $C$ only clusters with sufficient induced depolarization. \;
+    OR: \;
+    \If() {$CountInducedDepolarization(c) < \Theta_{dep}$}{
+      $C.remove(c)$ \;
+    }
+  }
+  \quad \\
+
+  $backtracking\_candidates \leftarrow C$ \;
+  \ForEach{$c \in backtracking\_candidates$}{
+    $s_t, a_t \leftarrow Decode(ColumnsFromCells(c))$ \;
+
+    $successful, actions \leftarrow Backtrack(s_t, t-1, active\_segments)$ \;
+    \If() {successful} {
+      $actions.append(a_t)$ \;
+      \Return{True, actions}
+    }
+  }
+
+  \Return{False, $\emptyset$}
+  \caption{Backtrack($d_t$, $t$, $active\_segments$)}
+\end{function}
 
@@
 
-Сравнение агента с бейзлайнами:
-
-% ![episode steps absolute](../../reports/v1_0/multi_way_transfer/multi_way_v0__steps.svg)
-% ![episode steps log-relative](../../reports/v1_0/multi_way_transfer/multi_way_v0__steps_rel_htm_0.svg)
-
-Сравнение DQN жадного и $\epsilon$-жадного агента:
-
-% ![episode steps absolute](../../reports/v1_0/multi_way_transfer/multi_way_v0__dqn__steps.svg)
-% ![episode steps log-relative](../../reports/v1_0/multi_way_transfer/multi_way_v0__dqn__steps_rel_dqn_greedy.svg)
-
-Сравнение агента с разными горизонтами и размером списка целей:
-
-% - горизонт планирования 1
-
-%   ![episode steps absolute](../../reports/v1_0/multi_way_transfer/multi_way_v0__1__steps.svg)
-%   ![episode steps log-relative](../../reports/v1_0/multi_way_transfer/multi_way_v0__1__steps_rel_htm_1_1g.svg)
-
-% - горизонт планирования 2
-
-%   ![episode steps absolute](../../reports/v1_0/multi_way_transfer/multi_way_v0__2__steps.svg)
-%   ![episode steps log-relative](../../reports/v1_0/multi_way_transfer/multi_way_v0__2__steps_rel_htm_2_1g.svg)
-
-% - горизонт планирования 4
-
-%   ![episode steps absolute](../../reports/v1_0/multi_way_transfer/multi_way_v0__4__steps.svg)
-
 \subsubsection{\texttt{multi\_way\_v1}}
@@
 
-Сравнение агента с бейзлайнами:
-
-% ![episode steps absolute](../../reports/v1_0/multi_way_transfer/multi_way_v1__steps.svg)
-% ![episode steps log-relative](../../reports/v1_0/multi_way_transfer/multi_way_v1__steps_rel_htm_0.svg)
-
-Сравнение DQN жадного и $\epsilon$-жадного агента:
-
-% ![episode steps absolute](../../reports/v1_0/multi_way_transfer/multi_way_v1__dqn__steps.svg)
-% ![episode steps log-relative](../../reports/v1_0/multi_way_transfer/multi_way_v1__dqn__steps_rel_dqn_greedy.svg)
-
-Сравнение агента с разными горизонтами и размером списка целей:
-
-% - горизонт планирования 1
-
-%   ![episode steps absolute](../../reports/v1_0/multi_way_transfer/multi_way_v1__1__steps.svg)
-%   ![episode steps log-relative](../../reports/v1_0/multi_way_transfer/multi_way_v1__1__steps_rel_htm_1_1g.svg)
-
-% - горизонт планирования 2
-
-%   ![episode steps absolute](../../reports/v1_0/multi_way_transfer/multi_way_v1__2__steps.svg)
-%   ![episode steps log-relative](../../reports/v1_0/multi_way_transfer/multi_way_v1__2__steps_rel_htm_2_1g.svg)
-
-% - горизонт планирования 4
-
-%   ![episode steps absolute](../../reports/v1_0/multi_way_transfer/multi_way_v1__4__steps.svg)
-
-\subsubsection{\texttt{multi\_way\_v1}}
+\subsubsection{\texttt{multi\_way\_v2}}
 
@@
 
-Сравнение агента с бейзлайнами:
+\begin{figure}
+  \centering
+  \begin{minipage}{.32\linewidth}
+    \includesvg[width=\linewidth]{./multi_way_transfer/multi_way_v0__steps}
+    \subcaption{\texttt{multi\_way\_v0}}
+  \end{minipage}
+  \begin{minipage}{.32\linewidth}
+    \includesvg[width=\linewidth]{./multi_way_transfer/multi_way_v1__steps}
+    \subcaption{\texttt{multi\_way\_v1}}
+  \end{minipage}
+  \begin{minipage}{.32\linewidth}
+    \includesvg[width=\linewidth]{./multi_way_transfer/multi_way_v2__steps}
+    \subcaption{\texttt{multi\_way\_v2}}
+  \end{minipage}
+  \caption{Сравнение агента с бейзлайнами.}
+\end{figure}
 
-% ![episode steps absolute](../../reports/v1_0/multi_way_transfer/multi_way_v2__steps.svg)
-% ![episode steps log-relative](../../reports/v1_0/multi_way_transfer/multi_way_v2__steps_rel_htm_0.svg)
+\begin{figure}
+  \centering
+  \begin{minipage}{.98\linewidth}
+    \includesvg[width=\linewidth]{./multi_way_transfer/multi_way_v0__steps}
+    \subcaption{Сравнение агента с бейзлайнами}
+  \end{minipage}
 
-Сравнение DQN жадного и $\epsilon$-жадного агента:
+  \begin{minipage}{.49\linewidth}
+    \includesvg[width=\linewidth]{./multi_way_transfer/multi_way_v0__dqn__steps}
+    \subcaption{Сравнение жадного и $\epsilon$-жадного агента DQN!}
+  \end{minipage}
+  \begin{minipage}{.49\linewidth}
+    \includesvg[width=\linewidth]{./multi_way_transfer/multi_way_v0__1__steps}
+    \subcaption{Горизонт планирования: 1}
+  \end{minipage}
 
-% ![episode steps absolute](../../reports/v1_0/multi_way_transfer/multi_way_v2__dqn__steps.svg)
-% ![episode steps log-relative](../../reports/v1_0/multi_way_transfer/multi_way_v2__dqn__steps_rel_dqn_greedy.svg)
+  \begin{minipage}{.49\linewidth}
+    \includesvg[width=\linewidth]{./multi_way_transfer/multi_way_v0__2__steps}
+    \subcaption{Горизонт планирования: 2}
+  \end{minipage}
+  \begin{minipage}{.49\linewidth}
+    \includesvg[width=\linewidth]{./multi_way_transfer/multi_way_v0__4__steps}
+    \subcaption{Горизонт планирования: 4}
+  \end{minipage}
 
-Сравнение агента с разными горизонтами и размером списка целей:
+  \caption{ \texttt{multi\_way\_v0}: cравнение агента c бейзлайнами и между собой - с разными горизонтами планирования и размером списка целей.}
+\end{figure}
 
-% - горизонт планирования 2
+\subsubsection{Ключевые выводы для экспериментов с неизменной средой}
 
-%   ![episode steps absolute](../../reports/v1_0/multi_way_transfer/multi_way_v2__2__steps.svg)
-%   ![episode steps log-relative](../../reports/v1_0/multi_way_transfer/multi_way_v2__2__steps_rel_htm_2_1g.svg)
-
-% - горизонт планирования 4
-
-%   ![episode steps absolute](../../reports/v1_0/multi_way_transfer/multi_way_v2__4__steps.svg)
-
-% - горизонт планирования 8 и 12
-
-%   ![episode steps absolute](../../reports/v1_0/multi_way_transfer/multi_way_v2__8-12__steps.svg)
-
-\subsection{Тестирование на задаче transfer learning в нескольких средах}
-
-Эксперименты данного режима проводились на наборе различных сред-лабиринтов. Схема каждого эксперимента следующая:
+В результате проведенных экспериментов сделаны следующие выводы:
 
 \begin{itemize}
-  \item эксперимент проводится последовательно на $N_{env}$ средах
+  \item Чем больше горизонт планирования, тем лучше играет агент. Причем улучшение кумулятивно, т.е. каждые следующие +1 горизонта дают больший прирост, чем предыдущий.
+
+  \item Агент с планированием играет лучше, чем случайный.
+Чем сложнее среда, тем менее заметен в относительных масштабах переход от \verb|horizon_planning = 0| к \verb|horizon_planning = 1|.
+При фиксированном горизонте планирования с увеличением сложности среды результат в относительных масштабах начинает "приближаться" к результатам случайного агента.
+
+  \item Агент с большим накопленным набором целей направленнее (=оптимальнее) исследует среду.
+Мы предполагаем это потому, что менее вероятно использование агентом случайной стратегии.
+В частных случаях это субоптимально.
+Например, когда награда сразу достижима в рамках имеющегося горизонта планирования и ее положение не изменилось с предыдущего эпизода - тогда агент с большим списком целей может выбрать для исследования псевдо-цель.
+Вероятно, причина данного эффекта кроется в специальном виде использованных сред, начального положения агента и положения наград.
+
+  \item Увеличение размера списка целей и величины горизонта планирования положительно сказываются на способности агента приспосабливаться к новому местоположению награды.
+
+  \item Увеличение горизонта планирования ухудшает производительность агента.
+Однако это ухудшение происходит до некоторого момента, пока горизонт планирования не приблизится к длине оптимального пути к цели
+  \item Агент учится и приспосабливается быстрее, чем DQN [в терминах числа эпизодов].
+
+  \item Если горизонта планирования достаточно, то агент быстро учится решать задачу планирования идеально, достигая оптимальной стратегии.
+По нашим оценкам он учится примерно в ~1.5-2.5 быстрее DQN.
+Тем не менее, для любого фиксированного горизонта, если начать усложнять среду, то в какой-то момент DQN обгонит по результатам, потому что DQN решает задачу [почти всегда] оптимально, а htm агент - только в случае достаточного горизонта планирования.
+
+  \item Иногда htm agent сходится к субоптимальному решению.
+Например, если горизонта планирования достаточно для субоптимального решения из начальной точки и имеется несколько длинных узких "ходов" до цели, имеющих примерно одинаковую длину.
+Тогда есть ненулевая вероятность, что агент не успеет запомнить путь через оптимальный коридор и будет планировать только через субоптимальный.
+В такой ситуации $\epsilon$-жадная стратегия DQN, наоборот, рано или поздно откроет оптимальное решение.
+\end{itemize}
+
+\subsection{Тестирование на задаче RL в сменяющихся средах}
+
+Испытание данного режима проводились на множестве случайно сгенерированных сред-лабиринтах типа GridWorld без учета направления взгляда агента. Схема каждого испытания следующая:
+
+\begin{itemize}
+  \item испытание проводится последовательно на $N_{env}$ средах
   \item при фиксированной среде последовательно рассматриваются $N_{rew}$ положений награды
   \item при фиксированном положении награды рассматривается $N_{s_0}$ начальных положений агента
-  \item в фиксированной конфигурации агент играет $N_{ep}$ эпизодов
+  \item в фиксированной конфигурации агент играет $N$ эпизодов
 \end{itemize}
 
-Исследовалась способность агента адаптироваться не только к новому местоположению награды в рамках одной среды, но и к новой среде.
+Таким образом испытание в общей сложности содержит $N_{env} \cdot N_{rew} \cdot N_{s_0} \cdot N$ эпизодов.
 
-Эксперименты тоже проводились на средах типа Gridworld без учета направления взгляда агента.
-В рамках каждого эксперимента все среды генерировались случайно на некотором фиксированном квадрате $n \times n$ (в основном $5 \times 5$ и $6 \times 6$).
+Исследовалась способность агента адаптироваться не только к новому местоположению награды в рамках одной среды, но и к новой среде целиком.
 
-Примеры сгенерированных сред
+Среды генерировались случайно на некотором фиксированном квадрате размера $n \times n$.
+Ииспытания проводились в средах $5 \times 5$, $6 \times 6$ и $8 \times 8$, но в отчете приведены результаты только в средах размера $5 \times 5$ - в остальных случаях результаты были схожими.
 
-Легенда:
+Ключевым различием между испытаниями были суммарное число эпизодов с фиксированной наградой и средой, которые влияли на сложность обратно пропорционально - чем меньше эпизодов в распоряжении агента, тем быстрее ему требуется адаптироваться (к новой награде и/или к новой среде, соответственно).
 
-\begin{itemize}
-  \item темно фиолетовый - стены
-  \item желтый - начальная позиция агента
-  \item салатовый - награда
-\end{itemize}
-
-Примеры сред:
-
-% - 5x5
-%   ![gridworld 5x5 #1](../../reports/v1_0/gridworld_transfer/gridworld_5x5_1_1_200_1337_map_0_1173222464.svg)
-%   ![gridworld 5x5 #2](../../reports/v1_0/gridworld_transfer/gridworld_5x5_1_1_200_1337_map_2_1561234712.svg)
-
-% - 8x8
-%   ![gridworld 8x8 #1](../../reports/v1_0/gridworld_transfer/gridworld_8x8_10_1_10_1337_map_0_1173222464.svg)
-%   ![gridworld 8x8 #2](../../reports/v1_0/gridworld_transfer/gridworld_8x8_10_1_10_1337_map_7_2032734714.svg)
-
-Ключевым различием между экспериментами были суммарное число эпизодов с фиксированной наградой и суммарное число эпизодов с фиксированной средой, которые обратно пропорционально влияли на сложность - чем меньше эпизодов в распоряжении агента, тем быстрее ему требуется адаптироваться (к новой награде и/или к новой среде, соответственно).
+\begin{figure}
+  \centering
+  \begin{minipage}{.49\linewidth}
+    \includesvg[width=\linewidth]{./gridworld_transfer/gridworld_5x5_1_1_200_1337_map_0_1173222464}
+  \end{minipage}
+  \begin{minipage}{.49\linewidth}
+    \includesvg[width=\linewidth]{./gridworld_transfer/gridworld_5x5_1_1_200_1337_map_2_1561234712}
+  \end{minipage}
+  \caption{Примеры сгенерированных сред размера 5x5. Легенда: темно фиолетовый - стены, желтый - начальная позиция агента, салатовый - награда.}
+\end{figure}
 
@@
 
-Ниже представлены графики для двух разных сидов.
+\begin{figure}
+  \centering
+  \begin{minipage}{.49\linewidth}
+    \includesvg[width=\linewidth]{./gridworld_transfer/gridworld_5x5_100_2_8_1337__steps}
+  \end{minipage}
+  \begin{minipage}{.49\linewidth}
+    \includesvg[width=\linewidth]{./gridworld_transfer/gridworld_5x5_100_2_8_42__steps}
+  \end{minipage}
+  \caption{Сравнение агента с бейзлайнами в эксперименте с редкой сменой наград и сред.}
+\end{figure}
 
-Сравнение агента с бейзлайнами:
-
-% ![episode steps absolute](../../reports/v1_0/gridworld_transfer/gridworld_5x5_100_2_8_1337__steps.svg)
-% ![episode steps log-relative](../../reports/v1_0/gridworld_transfer/gridworld_5x5_100_2_8_1337__steps_rel_htm_0.svg)
-% ![episode steps absolute](../../reports/v1_0/gridworld_transfer/gridworld_5x5_100_2_8_42__steps.svg)
-% ![episode steps log-relative](../../reports/v1_0/gridworld_transfer/gridworld_5x5_100_2_8_42__steps_rel_htm_0.svg)
-
-Сравнение DQN жадного и $\epsilon$-жадного агента:
-
-% ![episode steps absolute](../../reports/v1_0/gridworld_transfer/gridworld_5x5_100_2_8_1337__dqn__steps.svg)
-% ![episode steps absolute](../../reports/v1_0/gridworld_transfer/gridworld_5x5_100_2_8_42__dqn__steps.svg)
-
-Сравнение агента с разными горизонтами и размером списка целей:
-
-% - горизонт планирования 1
-
-%   ![episode steps absolute](../../reports/v1_0/gridworld_transfer/gridworld_5x5_100_2_8_1337__1__steps.svg)
-%   ![episode steps log-relative](../../reports/v1_0/gridworld_transfer/gridworld_5x5_100_2_8_1337__1__steps_rel_htm_1_1g.svg)
-%   ![episode steps absolute](../../reports/v1_0/gridworld_transfer/gridworld_5x5_100_2_8_42__1__steps.svg)
-%   ![episode steps log-relative](../../reports/v1_0/gridworld_transfer/gridworld_5x5_100_2_8_42__1__steps_rel_htm_1_1g.svg)
-
-% - горизонт планирования 2
-
-%   ![episode steps absolute](../../reports/v1_0/gridworld_transfer/gridworld_5x5_100_2_8_1337__2__steps.svg)
-%   ![episode steps log-relative](../../reports/v1_0/gridworld_transfer/gridworld_5x5_100_2_8_1337__2__steps_rel_htm_2_1g.svg)
-%   ![episode steps absolute](../../reports/v1_0/gridworld_transfer/gridworld_5x5_100_2_8_42__2__steps.svg)
-%   ![episode steps log-relative](../../reports/v1_0/gridworld_transfer/gridworld_5x5_100_2_8_42__2__steps_rel_htm_2_1g.svg)
-
-% - горизонт планирования 4-8
-
-%   ![episode steps absolute](../../reports/v1_0/gridworld_transfer/gridworld_5x5_100_2_8_1337__4-8__steps.svg)
-%   ![episode steps absolute](../../reports/v1_0/gridworld_transfer/gridworld_5x5_100_2_8_1337__4-8__steps_rel_htm_4_1g.svg)
-%   ![episode steps absolute](../../reports/v1_0/gridworld_transfer/gridworld_5x5_100_2_8_42__4-8__steps.svg)
-%   ![episode steps absolute](../../reports/v1_0/gridworld_transfer/gridworld_5x5_100_2_8_42__4-8__steps_rel_htm_4_1g.svg)
+\begin{figure}
+  \centering
+  \begin{minipage}{.32\linewidth}
+    \includesvg[width=\linewidth]{./gridworld_transfer/gridworld_5x5_100_2_8_1337__dqn__steps}
+    \includesvg[width=\linewidth]{./gridworld_transfer/gridworld_5x5_100_2_8_42__dqn__steps}
+    \subcaption{Жадный и $\epsilon$-жадный DQN агент}
+  \end{minipage}
+  \begin{minipage}{.32\linewidth}
+    \includesvg[width=\linewidth]{./gridworld_transfer/gridworld_5x5_100_2_8_1337__2__steps}
+    \includesvg[width=\linewidth]{./gridworld_transfer/gridworld_5x5_100_2_8_42__2__steps}
+    \subcaption{Горизонт планирования 2}
+  \end{minipage}
+  \begin{minipage}{.32\linewidth}
+    \includesvg[width=\linewidth]{./gridworld_transfer/gridworld_5x5_100_2_8_1337__4-8__steps}
+    \includesvg[width=\linewidth]{./gridworld_transfer/gridworld_5x5_100_2_8_42__4-8__steps}
+    \subcaption{Горизонт планирования 4}
+  \end{minipage}
+  \caption{Результаты агентов в эксперименте с редкой сменой наград и сред.}
+\end{figure}
 
@@
 
-Сравнение агента с бейзлайнами:
+\begin{figure}
+  \centering
+  \begin{minipage}{.32\linewidth}
+    \includesvg[width=\linewidth]{./gridworld_transfer/gridworld_5x5_50_4_4_1337__steps}
+    \includesvg[width=\linewidth]{./gridworld_transfer/gridworld_5x5_50_4_4_1337__steps_rel_htm_0}
+  \end{minipage}
+  \begin{minipage}{.32\linewidth}
+    \includesvg[width=\linewidth]{./gridworld_transfer/gridworld_5x5_50_4_4_1337__dqn__steps}
+    \includesvg[width=\linewidth]{./gridworld_transfer/gridworld_5x5_50_4_4_1337__2__steps}
+  \end{minipage}
+  \begin{minipage}{.32\linewidth}
+    \includesvg[width=\linewidth]{./gridworld_transfer/gridworld_5x5_50_4_4_1337__1__steps}
+    \includesvg[width=\linewidth]{./gridworld_transfer/gridworld_5x5_50_4_4_1337__4-8__steps}
+  \end{minipage}
+  \caption{Результаты агентов в эксперименте со средней частотой смены наград и сред.}
+\end{figure}
 
-% ![episode steps absolute](../../reports/v1_0/gridworld_transfer/gridworld_5x5_50_4_4_1337__steps.svg)
-% ![episode steps log-relative](../../reports/v1_0/gridworld_transfer/gridworld_5x5_50_4_4_1337__steps_rel_htm_0.svg)
+\subsubsection{Эксперименты с частой сменой наград и сред}
 
-Сравнение DQN жадного и $\epsilon$-жадного агента:
+\begin{itemize}
+  \item Gridworld 5x5
+  \item $N_{s_0} = 20$
+  \item $N_{rew} = 1$
+  \item $N_{env} = 20$
+\end{itemize}
 
-% ![episode steps absolute](../../reports/v1_0/gridworld_transfer/gridworld_5x5_50_4_4_1337__dqn__steps.svg)
-% ![episode steps absolute](../../reports/v1_0/gridworld_transfer/gridworld_5x5_50_4_4_1337__dqn__steps_rel_dqn_greedy.svg)
+\begin{figure}
+  \centering
+  \begin{minipage}{.32\linewidth}
+    \includesvg[width=\linewidth]{./gridworld_transfer/gridworld_5x5_20_1_20_1337__steps}
+  \end{minipage}
+  \begin{minipage}{.32\linewidth}
+    \includesvg[width=\linewidth]{./gridworld_transfer/gridworld_5x5_20_1_20_1337__1__steps}
+  \end{minipage}
+  \begin{minipage}{.32\linewidth}
+    \includesvg[width=\linewidth]{./gridworld_transfer/gridworld_5x5_20_1_20_1337__2__steps}
+  \end{minipage}
+  \caption{Результаты агентов в эксперименте с частой сменой наград и сред.}
+\end{figure}
 
-Сравнение агента с разными горизонтами и размером списка целей:
+\subsubsection{Эксперименты с ультрачастой сменой наград и сред}
 
-% - горизонт планирования 1
+TODO Нужен ли этот эксперимент вообще?
 
-%   ![episode steps absolute](../../reports/v1_0/gridworld_transfer/gridworld_5x5_50_4_4_1337__1__steps.svg)
+\begin{itemize}
+  \item Gridworld 5x5
+  \item $N_{s_0} = 1$
+  \item $N_{rew} = 1$
+  \item $N_{env} = 200$
+\end{itemize}
 
-% - горизонт планирования 2
+\begin{figure}
+  \centering
+  \begin{minipage}{.49\linewidth}
+    \includesvg[width=\linewidth]{./gridworld_transfer/gridworld_5x5_1_1_200_1337__steps}
+  \end{minipage}
+  \begin{minipage}{.49\linewidth}
+    \includesvg[width=\linewidth]{./gridworld_transfer/gridworld_5x5_1_1_200_1337__2__steps}
+  \end{minipage}
+  \caption{Результаты агентов в эксперименте с ультрачастой сменой наград и сред.}
+\end{figure}
 
-%   ![episode steps absolute](../../reports/v1_0/gridworld_transfer/gridworld_5x5_50_4_4_1337__2__steps.svg)
+\subsubsection {Ключевые выводы для экспериментов со сменой сред}
 
-% - горизонт планирования 4-8
+HTM агент в среднем быстрее чем DQN адаптируется к изменениям среды или в среде. Благодаря этому обычно он ведет себя стабильнее.
 
-%   ![episode steps absolute](../../reports/v1_0/gridworld_transfer/gridworld_5x5_50_4_4_1337__4-8__steps.svg)
-%   ![episode steps absolute](../../reports/v1_0/gridworld_transfer/gridworld_5x5_50_4_4_1337__4-8__steps_rel_htm_4_1g.svg)
+Увеличение списка целей негативно сказывается на результатах агента.
+Получается, что польза от выполнения псевдо-целей в среднем ниже штрафа.
+Мы считаем это потому, что в среднем псевдо-цели не приближают агента к реальной и только увеличивают ожидаемое число шагов с использованием случайной стратегии.
 
-% #### Эксперименты с частой сменой наград и сред
+Данный результат оказался для нас неожиданным, т.к.
+расходится с выводами, полученными по результатам экспериментов без смены среды.
+Судя по всему, дело в специфике тех экспериментов - в них следование псевдо-целям всегда уменьшало расстояние до цели (кроме эксперимента \verb|multi_way_v2|, где это верно не всегда, но все же с достаточно большой вероятностью).
+Из-за этого мы наблюдали положительный эффект от увеличения списка подцелей.
 
-% - Gridworld 5x5
-% - $N_{s_0} = 20$
-% - $N_{rew} = 1$
-% - $N_{env} = 20$
-
-% Сравнение агента с бейзлайнами:
-
-% ![episode steps absolute](../../reports/v1_0/gridworld_transfer/gridworld_5x5_20_1_20_1337__steps.svg)
-% ![episode steps log-relative](../../reports/v1_0/gridworld_transfer/gridworld_5x5_20_1_20_1337__steps_rel_htm_0.svg)
-
-% Сравнение DQN жадного и $\epsilon$-жадного агента:
-
-% ![episode steps absolute](../../reports/v1_0/gridworld_transfer/gridworld_5x5_20_1_20_1337__dqn__steps.svg)
-
-% Сравнение агента с разными горизонтами и размером списка целей:
-
-% - горизонт планирования 1
-
-%   ![episode steps absolute](../../reports/v1_0/gridworld_transfer/gridworld_5x5_20_1_20_1337__1__steps.svg)
-
-% - горизонт планирования 2
-
-%   ![episode steps absolute](../../reports/v1_0/gridworld_transfer/gridworld_5x5_20_1_20_1337__2__steps.svg)
-
-% - горизонт планирования 4-8
-
-%   ![episode steps absolute](../../reports/v1_0/gridworld_transfer/gridworld_5x5_20_1_20_1337__4-8__steps.svg)
-
-% #### Эксперименты с ультрачастой сменой наград и сред
-
-% Нужно?
-
-% - Gridworld 5x5
-% - $N_{s_0} = 1$
-% - $N_{rew} = 1$
-% - $N_{env} = 200$
-
-% Сравнение агента с бейзлайнами:
-
-% ![episode steps absolute](../../reports/v1_0/gridworld_transfer/gridworld_5x5_1_1_200_1337__steps.svg)
-
-% Сравнение DQN жадного и $\epsilon$-жадного агента:
-
-% ![episode steps absolute](../../reports/v1_0/gridworld_transfer/gridworld_5x5_1_1_200_1337__dqn__steps.svg)
-
-% Сравнение агента с разными горизонтами и размером списка целей:
-
-% - горизонт планирования 1
-
-%   ![episode steps absolute](../../reports/v1_0/gridworld_transfer/gridworld_5x5_1_1_200_1337__1__steps.svg)
-
-% - горизонт планирования 2
-
-%   ![episode steps absolute](../../reports/v1_0/gridworld_transfer/gridworld_5x5_1_1_200_1337__2__steps.svg)
-
-% - горизонт планирования 4-8
-
-%   ![episode steps absolute](../../reports/v1_0/gridworld_transfer/gridworld_5x5_1_1_200_1337__4-8__steps.svg)
-
-\section {Выводы}
-
-Ключевые выводы для экспериментов с неизменной средой:
-
-- Чем больше горизонт планирования, тем лучше играет агент
-  - причем улучшение кумулятивно, т.е. каждый следующий +1 горизонта дает больший прирост, чем предыдущий
-- Агент с планированием играет лучше, чем случайный
-  - чем сложнее среда, тем менее заметен в относительных масштабах переход от horizon\_planning=0 к horizon\_planning=1
-  - т.е. при фиксированном горизонте планирования с увеличением сложности среды результат в относительных масштабах начинает прижиматься к результатам случайного агента
-- Агент с большим накопленным набором целей направленнее (=оптимальнее) исследует среду
-  - предположительно потому, что менее вероятно использование случайной стратегии
-  - в частных случаях это субоптимально - когда награда сразу достижима в рамках имеющегося горизонта планирования и ее положение не изменилось с предыдущего эпизода, т.к.
-агент с большим списком целей может выбрать для исследования псевдо-цель
-  - вероятно, причина данного эффекта кроется в специальном виде использованных сред, начального положения агента и положения наград.
-- Увеличение размера списка целей и величины горизонта планирования положительно сказываются на способности агента приспосабливаться к новому местоположению награды
-- Увеличение горизонта планирования ухудшает производительность агента
-  - однако это ухудшение происходит до некоторого момента, пока горизонт планирования не приблизится к длине оптимального пути к цели
-- Агент учится и приспосабливается быстрее, чем DQN [в терминах числа эпизодов]
-- Если горизонта планирования достаточно, то агент быстро учится решать задачу планирования идеально, достигая оптимальной стратегии
-  - учится в ~1.5-2.5 быстрее DQN
-  - но для любого фиксированного горизонта, если начать усложнять среду, то в какой-то момент DQN обгонит по результатам, потому что DQN решает задачу [почти всегда] оптимально, а htm агент - только в случае достаточного горизонта планирования
-- Иногда htm agent сходится к субоптимальному решению
-  - если горизонта планирования достаточно для субоптимального решения из начальной точки
-  - и имеется несколько длинных узких "ходов" до цели, имеющих примерно одинаковую длину
-  - тогда есть ненулевая вероятность, что агент не успеет запомнить путь через оптимальный коридор и будет планировать только через субоптимальный.
-  - в такой ситуации eps-жадная стратегия DQN, наоборот, рано или поздно откроет оптимальное решение
-
-Ключевые выводы для экспериментов со сменой сред:
-
-- В вырожденном случае смены среды и награды на каждом эпизоде все агенты проигрывают случайному
-  - DQN при этом играет значительно хуже HTM агента
-- HTM агент в среднем быстрее чем DQN адаптируется к изменениям среды или в среде
-  - благодаря этому обычно он ведет себя стабильнее
-- Увеличение списка целей негативно сказывается на результатах агента
-  - видимо, польза от выполнения псевдо-целей в среднем ниже штрафа
-  - можно объяснить тем, что в среднем псевдо-цели не приближают агента к реальной и только увеличивают ожидаемое число шагов с использованием случайной стратегии
-
-\section {Project specific terminology}
-
-State-action (SA)
-
-- Given trajectories : $\tau_i = (s_0, a_0, r_1, s_1, a_1, r_2, s_2, a_2, ...)$
-- SA is a tuple $sa_t = (s_t, a_t)$
-  - i.e. action $a_t$ taken from a state $s_t$
-- TM learns agent's trajectories $\tau_i$
-  - it operates with SA [encoded as] SDR sequences $(sa_0, sa_1, ... )$
-  - which results that TM learns both
-    - transitions $(s_{t-1}, a_{t-1}) \rightarrow s_t$
-    - and which actions $a$ agent did from a state $s_t$
-- TM then can make predictions
-  - it can predict next state and all actions agent did from this state
-  - each prediction is an SDR - a union (= bitwise OR) of pairs $(s, a^*)$
-  - we give this union a special name (see below)
-
-SA SDR Superposition
-
-- Union (= bitwise OR) of any number of SA SDRs
-- Rationale
-  - if an agent saw transition $(s_{t-1}, a_{t-1}) \rightarrow s_t$ and did different actions $a_t$ from a state $s_t$
-  - then prediction for a pair $(s_{t-1}, a_{t-1})$ will be a union of all seen pairs $(s_t, a_t)$
-  - aka superposition of these SAs
-- TM can work with SA SDR superpositions both as input and output
-  - In general case (mentioned just before) TM's output is a superposition
-  - TM accepts SA superpositions as input as well
-    - it's possible because SA SDR superposition still is just an SDR
-- *NB*: prediction process is called depolarization
-
-\section {Encoding}
-
-SA SDR encoder
-
-- Encodes SA to SDR
-- Uses separate state and action SDR encoders
-  - Resulting SDRs are concatenated
-- We consider only discrete environments
-  - Hence both states and actions are discrete sets
-- Each state or action can be represented as integer number
-  - i.e. $s_t \in [0, |S|)$ and $a_t \in [0, |A|)$
-  - to encode states and action we use Integer SDR encoder
-
-Integer SDR encoder
-
-% - Encodes integer number $x \in$ [0, `n_values`) to SDR
-% - Parameters:
-%   - `n_values` - a number of unique values
-%   - `value_bits` - a number of bits used to encode each unique value
-% - Resulting SDR has `total_bits` = `n_values` $\times$ `value_bits` bits
-%   - they can be logically divided into `n_values` buckets of `value_bits` contiguous bits
-%     - each bucket corresponds to a value from the range
-%   - e.g. 2 $\rightarrow$ `0000 0000 1111` encoded by 3-by-4 integer encoder
-%     - 3 buckets are separated by space to make it clear
-%     - every value id encoded by 4 bits bucket
-%     - note that buckets don't intersect
-% - PROS
-%   - Easy to encode/decode
-%   - Easy to pretty print and read
-%   - Hence easy to debug
-%   - Different values have no intersection
-%     - eliminates some unnecessary complexities
-% - CONS
-%   - Different values have no intersection
-%     - narrows the use of SDRs
-%     - doesn't use information about states similarity
-%   - You have no direct control on sparsity, which is `1 / n_values`
-%     - you may have problems with too low or too high sparsity levels
-%     - we tested that TM works well with low sparsity
-%       - which doesn't mean it has not negative effects at all
-%     - just remember that sparsity may lead to problems
-
-State SDR encoder
-
-- For simple MDP environments we use integer encoding
-- For complex environments encoding states [or observations] becomes tricky
-  - of course, you still can enumerate all possible states and use Integer SDR encoder
-  - but the number of all possible states grows very fast
-    - in practice it works well only for small environments
-  - also Integer SDR encoder cuts out information about states similarity, which may not be desirable
-  - one possible solution is to encode every pixel [or grid cell] separately then concatenate results, which preserves information about similarity between states
-- At the moment we use only simple MDP environments and hence integer encoding for states
-
-\section{Planning alforithm details}
-
-Let's consider an agent playing in some MDP environment. At some moment he is in a state, which we will denote as $s_0$, because it will be our starting point for planning.  
-The agent also has a fixed set of goal states $\{g_i\}$, which he wants to reach.
-
-The goal of the planning is to answer two questions.
-Is it possible to reach any of the goal states with at maximum of $n_max$ steps (i.e.
-actions)?
-And if it is possible, then what is the sequence of actions (i.e.
-policy) leading to that goal state?
-
-Planning algorithm consists of 2 high-level steps: forward prediction and backtracking from the goal, which could be written in Python-like pseudocode as:
-
-% ```python
-% def plan_actions(initial_state):
-%   reached_goals, active_segments_timeline = predict_to_goals(initial_state)
-%   planned_actions = backtrack_from_goals(
-%     reached_goals, active_segments_timeline
-%   )
-
-%   return planned_actions
-% ```
-
-\subsection{Step 1. Forward prediction}
-
-We start from the state $s_0$ and want to predict all reachable next states $\{s_1^i\}$.
-To do this we could check all learned transitions $(s_0, a_0^i) \rightarrow s_1^i$.
-It's achievable simultaneously if we use $sa_0 = (s_0, \cup a^i)$ as a starting input for the agent's TM.
-$sa_0$ is then a SA SDR superposition and semantically means exactly a superposition of all possible state-action pairs $(s_0, a_0^i)$ which activate predictions for all desired transitions.
-
-% __TBD: make simple MDP graph example__
-
-As a prediction TM gives another SA SDR superposition $sa_1 = \cup (s_1, a_1^i)$ - a superposition of all next\_state-next\_action pairs.
-$sa_1$ can be further used as an input to TM, which gets us $sa_2$ prediction and so on.
-
-As you see, for each step $k$, $sa_k$ contains a superposition of all states reachable in exactly $k$ actions (=moves).
-As a side note, the whole resulting process looks very similar to parallel breadth-first-search graph algorithm.
-
-Recall that the goal of forward prediction is to find reachable goal states.
-Therefore at each step $k$ we look if any of the goal states $g_i$ are among the states in $sa_k$ superposition by checking their overlap $o = overlap(g_i, sa_k)$.
-We want to find only the closest goals, so if any of the goal states are matched, we stop forward prediction process and set these goal states as a result.
-
-% Also during forward prediction we need to record active segments at each step $k$ denoted as `active_segments_timeline`. It's required by the second step of planning.
-
-Note also, that cell activation and segment activation, despite named similar, refer to the different phases of TM workflow.
-The latter is a part of prediction, not activation - active segment depolarizes cell, i.e.
-predicts its future activation instead of activating it.
-
-Below is a pseudocode for forward prediction phase.
-
-% ```python
-% def predict_to_reward(initial_state):
-%   # initializes starting $sa_0$ as superposition with all possible actions
-%   initial_sa = Sa(initial_state, AllActions)
-%   proximal_input = encoder.encode(initial_sa)
-
-%   reached_goals = []
-%   active_segments_timeline = []
-%   for i in range(max_steps):
-%     # activates & depolarizes cells
-%     active_cells, depolarized_cells = memory.process(
-%       proximal_input, learn=False
-%     )
-
-%     # records active segments for $i$-th step
-%     active_segments_timeline.append(
-%       memory.active_segments(active_cells)
-%     )
-
-%     # gets $sa_{i+1}$ SA SDR superposition
-%     proximal_input = columns_from_cells(depolarized_cells)
-
-%     # gets list of matched goal states
-%     reached_goals = [
-%       goal_state
-%       for goal_state in agent.goal_states
-%       if is_matched(goal_state, proximal_input)
-%     ]
-%     if reached_goals:
-%       break
-
-%   return reached_goals, active_segments_timeline
-% ```
-
-\subsection{Step 2. Backtracking from goal states}
-
-As a result of forward prediction phase we have a list of goal states reachable in $T$ actions along with recorded segments activation timeline.
-The goal of backtracking phase is to restore a correct sequence of actions leading to any of these goal states.
-
-At high level view we start from any of the goal states and try moving backward [through time] to the initial state using segments activation timeline graph:
-
-% ```python
-% def backtrack_from_goals(reached_goal_states, active_segments_timeline):
-%   T = len(active_segments_timeline)
-%   # cells predicted to be active at timestep T, when the goal states were reached
-%   depolarized_cells = active_segments_timeline[T-1].keys()
-
-%   for goal_state in reached_goal_states:
-%     # gets predicted cells related to the goal state
-%     depolarized_goal_cells = intersect(goal_state, depolarized_cells)
-
-%     # try backtracking from the goal state to the initial state
-%     planning_successful, planned_actions = backtrack_from_state(
-%         depolarized_goal_cells, T-1
-%     )
-%     if planning_successful:
-%       return planned_actions
-% ```
-
-The key thing for this phase is moving backward in an implicitly built graph of active cells.
-Consider having chosen some goal state $g$.
-Let's start with the first step of backtracking - timestep $T-1$ - when the goal state $g$ was \textbf{predicted}.
-
-Here we should pause a bit and expand the meaning ot "the goal state was predicted".
-Recall that we work with numeric MDP states encoded as SDR with Integer SDR encoder.
-A single value $x$ is decoded from an SDR $S$ means that $S$ has enough ON bits related to $x$, with "enough" is defined by \verb|activation_threshold| hyperparameter.
-So "the goal state $g$ was predicted" means that there were at least \verb|activation_threshold| ON bits related to it in a prediction.
-
-From forward prediction phase we know the set of active cells at timestep $T-1$.
-And the fact that the goal state $g$ was predicted at that timestep creates a condition for backtracking - we have to make sure that it stays true.
-Obviously, the whole set of active cells at $T-1$ doesn't break this prediction condition.
-
-But there's another condition - at each timestep there's only one transition $(s, a) \rightarrow s'$ allowed, because that's how MDP works.
-On the other hand the forward prediction process works with superpositions, and the set of active cells $sa_{T-1}$ may contain multiple different $(s, a)$ pairs, from which we may choose only those single pairs that don't break the prediction condition.
-The question is how to find these pairs?
-
-The segments activations timeline allows us to take all active segments of the goal state $g$ cells at time $T-1$.
-Each active segment is associated with a set of active cells that activated it.
-By design each TM segment is supposed to recognize a single pattern, which is a state-action pair in our case.
-
-\textbf{NB}: In general that's not true.
-A segment might recognize a set of patterns, and the most robust to the punishment during learning are those sets that have non-zero intersection, which minimizes expected punishment rate.
-It means, that each segment learns a cluster [or clusters] of patterns.
-But we can bound the segment's allowed size, which is a number of synapses, to fit no more than a single state-action pair size - then each segment will be learning to recognize only a single pattern.
-Still, this doesn't mean that all segment's synapses are always connected to the same state or action cells, although learning tends to fit them this way.
-
-We might expect that the active cells sets $\{S_i^{T-1}\}$ associated with the active segments $\{A_i^{T-1}\}$ at time $T-1$ contain clusters, where each pair of active cells set has a high overlap score.
-Formally, we want to split $\{A_i^{T-1}\}$ into clusters $\{C_k^{T-1}| A_i, A_j \in C_k^{T-1}: overlap(A_i, A_j) \geq N_{cl}\}$ and keep only those that
-
-At any step $k$ we have the set of cells that should be activated at $k+1$ timestep in order to reach the goal state.
-It means that some key portion of these cells should be predicted at timestep $k$ - at least `activation threshold` cells - which gives us
-
-% - Given a set of depolarized cells `dep_cells` [at time $t$]
-% - Call active segment as _potential_ if it induces enough depolarization among `dep_cells`
-%   - _ehough_ means $\geq$ activation threshold
-%   - i.e. a number of depolarized cells will be enough for subsequent depolarizations
-% - Take presynaptic active cells of a potential segment as depolarized cells [at time $t-1$]
-
-Backtracking algorithm
-
-- Start from depolarized cells in reward = 1 columns [at time T]
-- Recursively take backtracking steps for any potential segment until time 0 is reached
-
-% Details
-
-% - Given a set of depolarized cells `dep_cells` [at time $t$]
-%   - Each depolarized cell have at least one active segment
-%   - Each active segment is associated with a set of active presynaptic cells
-%     - This set defines some SAR SDR
-%     - i.e. each segment is conditioned on some correct SAR
-% - I want to find all potential sets of active presynaptic cells that induce enough depolarization of `dep_cells`
-%   - perfect case
-%     - `dep_cells` conditioned on the same SAR
-%     - i.e. this SAR induces full `dep_cells` depolarization
-%   - perfect multiple case
-%     - `dep_cells` conditioned on more than one SAR
-%     - i.e. each of these SAR induces full `dep_cells` depolarization
-% - real world relaxations
-%   - some SARs don't induce enough depolarization
-%     - i.e. $\lt$ threshold
-%   - segments don't exactly match each other
-%     - cells are conditioned on very similar SARs but not exactly the same
-
-What is saved during this phase
-
-% - active [presynaptic] cells at time $t$ are saved as `backtracking_SAR_conditions[t]`
-% - they define condition on SAR at time $t-1$ to get desired depolarization at time $t$
-
-% Второй шаг - рекурсивный бэктрекинг из столбцов, соответствующих награде, назад во времени:
-
-% ```python
-% def backtrack_from_reward(active_segments_timeline):
-%     final_depolarized_cells = active_segments_timeline[T-1].keys()
-%     depolarized_reward_cells = get_reward_cells(final_depolarized_cells)
-
-%     return backtrack(depolarized_reward_cells, T-1)
-
-% def backtrack(desired_depolarization: SparseSdr, t: int):
-%     # presynaptic cells clusters, each can induce desired depolarization
-%     cell_clusters = get_backtracking_candidate_clusters(
-%         desired_depolarization, activation_threshold, t
-%     )
-
-%     for cluster in cell_clusters:
-%         activation_timeline = backtrack(cluster, t-1)
-%         if backtracking_succeeded:
-%             activation_timeline.append(cluster)
-%             return activation_timeline
-% ```
-
-% Откуда берутся кластеры-кандидаты:
-
-% ```python
-% def get_backtracking_candidate_clusters(
-%         desired_depolarization: SparseSdr,
-%         sufficient_activation_threshold: int,
-%         t: int
-% ):
-%     # Active presynaptic cell clusters (clusterization by columns)
-%     active_segments = active_segments_timeline[t][desired_depolarization]
-%     candidate_clusters = merge_segments_into_clusters(active_segments)
-
-%     # Keep clusters that induce sufficient depolarization among desired
-%     for cluster in candidate_clusters:
-%         count_induced_depolarization(cluster, desired_depolarization)
-
-%         if induced_depolarization >= sufficient_activation_threshold:
-%             backtracking_candidate_clusters.append(cluster)
-
-%     return backtracking_candidate_clusters
-% ```
-
-% Последний шаг - предсказание награды по пути успешного бэктрекинга:
-
-% ```python
-% def check_activation_timeline_leads_to_reward(
-%         initial_sar: Sar, activation_timeline
-% ):
-%     proximal_input = encoder.encode(initial_sar)
-
-%     for i in range(T):
-%         action = extract_action(activation_timeline[i])
-%         proximal_input.action = action
-
-%         active_cells, depolarized_cells = agent.process(
-%             proximal_input, learn=False
-%         )
-%         proximal_input = columns_from_cells(depolarized_cells)
-
-%         planned_actions.append(action)
-
-%     if is_rewarding(proximal_input):
-%         return planned_actions
-% ```
-
-% ### Step 1: Forward prediction
-
-% Setting up
-
-% - Given initial state $s_0$
-% - To find a reward, I want to check all possible paths, starting from $s_0$
-%   - Then SAR superposition  
-%     (state = $s_0$, {all actions}, reward = 0)
-%   - .. represents all possible beginnings of these paths
-%   - So, encode it to SDR and set it as initial proximal input to TM
-
-% Forward prediction
-
-% - Given some SAR superposition as proximal input at time $t$
-%   - induces cells activation
-% - TM prediction: some SAR superposition at time $t+1$
-%   - prediction is in form of depolarized cells
-% - Save all active segments (discussed later in section)
-% - Check if depolarized SAR superposition contains reward = 1
-% - Continue forward prediction by taking predicted SAR superposition as input at $t+1$
-%   - until reward is found
-%   - or max forward steps are done
-
-% A bit of details on TM work, step by step:
-
-% - Given current proximal input [at time $t$]
-% - It directly induces columns activation
-% - Active columns [and depolarization from $t-1$] induce cells activation
-%   - any depolarized cell in active columns becomes active
-%   - or all cells become active if active column has no depolarized cells (bursting)
-% - Cells activation induces segments activation
-%   - number of active presynaptic cells $\geq$ activation threshold
-% - Segments activation induces cells depolarization
-%   - any active segment depolarizes its postsynaptic cell
-% - Depolarized cells can be collapsed into depolarized columns
-%   - any depolarized cell depolarizes its column
-% - Depolarized columns define predicted SAR SDR superposition [at time $t+1$]
-
-% What is saved during forward prediction phase:
-
-% - At each time step $t$ there's a set of segments that are active
-% - These active segments are saved into `active_segments_timeline[t]`
-%   - as dictionary: `depolarized_cell` $\rightarrow$ `cell_active_segments_list`
-%   - `cell_active_segments_list` - list of cell's active segments
-%   - each active segment represented as a set of active presynaptic cells
-
-% ### Step 2: Backtracking
-
-% Backtracking step
-
-% - Given a set of depolarized cells `dep_cells` [at time $t$]
-% - Call active segment as _potential_ if it induces enough depolarization among `dep_cells`
-%   - _ehough_ means $\geq$ activation threshold
-%   - i.e. a number of depolarized cells will be enough for subsequent depolarizations
-% - Take presynaptic active cells of a potential segment as depolarized cells [at time $t-1$]
-
-% Backtracking algorithm
-
-% - Start from depolarized cells in reward = 1 columns [at time T]
-% - Recursively take backtracking steps for any potential segment until time 0 is reached
-
-% Details
-
-% - Given a set of depolarized cells `dep_cells` [at time $t$]
-%   - Each depolarized cell have at least one active segment
-%   - Each active segment is associated with a set of active presynaptic cells
-%     - This set defines some SAR SDR
-%     - i.e. each segment is conditioned on some correct SAR
-% - I want to find all potential sets of active presynaptic cells that induce enough depolarization of `dep_cells`
-%   - perfect case
-%     - `dep_cells` conditioned on the same SAR
-%     - i.e. this SAR induces full `dep_cells` depolarization
-%   - perfect multiple case
-%     - `dep_cells` conditioned on more than one SAR
-%     - i.e. each of these SAR induces full `dep_cells` depolarization
-% - real world relaxations
-%   - some SARs don't induce enough depolarization
-%     - i.e. $\lt$ threshold
-%   - segments don't exactly match each other
-%     - cells are conditioned on very similar SARs but not exactly the same
-
-% What is saved during this phase
-
-% - active [presynaptic] cells at time $t$ are saved as `backtracking_SAR_conditions[t]`
-% - they define condition on SAR at time $t-1$ to get desired depolarization at time $t$
-
-% ### Step 3: Re-check
-
-% - Given active presynaptic cells at time $t$ needed to induce desired depolarization
-% - Start forward prediction again from initial sar superposition
-% - But now
-%   - Check that SAR `backtracking_SAR_conditions[t]` is consistent with the current active SAR
-%   - Extract action $a_t$ from `backtracking_SAR_conditions[t]`
-%   - Action $a_t$ defines a path to make at time $t$
-%   - Replace action superposition in proximal input with $a_t$
-% - Check that given `backtracking_SAR_conditions` lead to reward
+В вырожденном случае смены среды и награды на каждом эпизоде все агенты проигрывают случайному. При этом DQN играет значительно хуже HTM агента. TODO Почему?
 

### 2020-10-06 htm_rl/htm_rl/agent/agent.py
learn all, but stop episode if get impossible action during pretrain

@@ -51,11 +51,7 @@ class Agent:
         action = self._make_action(state, is_done, verbosity)
         trace(verbosity, 2, f'\nMake action: {action}')
 
-        if info is None:
-            self.memory.train(Sa(state, action), verbosity)
-        else:
-            if (info['previous_action'] != action) or (reward > -100):
-                self.memory.train(Sa(state, action), verbosity)
+        self.memory.train(Sa(state, action), verbosity)
 
         if reward > 0:
             self.planner.add_goal(state)
@@ -172,6 +168,9 @@ class AgentRunner:
         total_reward = 0.
         while step < self.max_steps and not done:
             state, reward, done, info = self.env.step(action)
+            # if pretrain, than learn only possible transitions
+            if self.agent.planner.planning_horizon == 0 and (reward <= (-100)):
+                break
             action = self.agent.make_step(state, reward, done, self.verbosity, info)
             step += 1
             total_reward += reward

### 2020-10-06 htm_rl/htm_rl/envs/mymdp.py
step number is added to observable state

@@ -59,7 +59,7 @@ class GridWorld:
             'column': agent_initial_position['column'] + 1}
 
         self.agent_position = converted_agent_position
-
+        self.step_number = 0
         self.agent_direction = agent_initial_direction
         self.max_sight_range = max_sight_range
         self.observable_state = None
@@ -117,6 +117,8 @@ class GridWorld:
                 raise ValueError
         else:
             raise ValueError
+
+        self.step_number += 1
 
         return self.observation(), self.reward() - punish, self.is_terminal(), {'previous_action': action}
 
@@ -182,9 +184,9 @@ class GridWorld:
         else:
             surface = line[stop_pos]
 
-        self.observable_state = {'distance': distance - 1, 'surface': surface}
+        self.observable_state = {'distance': distance - 1, 'surface': surface, 'step': self.step_number}
 
-        return distance - 1, surface
+        return distance - 1, surface, self.step_number
 
     def reward(self):
         if self.observable_state['surface'] == 2 and self.observable_state['distance'] == 0:
@@ -201,8 +203,9 @@ class GridWorld:
     def reset(self):
         self.agent_position = self._agent_initial_position.copy()
         self.agent_direction = self._agent_initial_direction
+        self.step_number = 0
         self.observation()
-        return self.observable_state['distance'], self.observable_state['surface']
+        return self.observable_state['distance'], self.observable_state['surface'], self.step_number
 
     def render(self):
         for i, row in enumerate(self.world_description):

### 2020-10-08 reports/v1_0/report.tex
WIP report: fixes, refinements

@@
 \begin{itemize}
-  \item Где испытания на гридворлде с направлением взгляда агента? Если я их в итоге не использую в отчете, тогда убрать вообще упоминание о них.
+  \item Убрать упоминание о разновидности гридворлда с учетом направления взгляда агента.
   \item Примеры сред перегенерировать с белой пустотой и черными стенами
   \item Блоки-сноски, помеченные NB, можно сделать в виде нормальных сносок.
+  \item Правильно проставить double quote marks через ``''
 \end{itemize}
@@
 
-Агент играет в среде GridWorld - лабиринт на квадратной сетке. Агент начинает игру в фиксированной точке старта, и ему требуется найти единственную награду, которая находится в фиксированном месте.
+Рассматривается классическая задача MDP - агент играет в детерминированной среде GridWorld, которая выглядит как лабиринт на квадратной сетке.
+Агент начинает игру в фиксированной точке старта, и ему требуется найти единственную награду, которая находится в фиксированном месте.
 
@@
 
-Текущая реализация агента содержит, грубо говоря, две политики - случайную и спланированную.
+В текущей реализации политика агента представляет собой смесь из двух политик - случайную и спланированную.
 Спланированную политику предоставляет планировщик.
-В случае, если он не может её предоставить, используется случайная политика.
+В случае, если он не может её предоставить, используется случайная.
 
-Таким образом основная логика агента содержится в реализации планировщика.
+Основная логика агента содержится именно в реализации планировщика.
 
-Задача планировщика - найти путь до награды, если она достижима за не более чем \verb|planning_horizon| шагов. Делается это в два этапа:
+Задача планировщика - найти путь до награды, если она достижима за не более чем \verb|planning_horizon| шагов. Эта задача решается в два этапа:
 
 \begin{enumerate}
-    \item Планирование до награды из текущего состояния
-    \item Нахождение обратного пути из состояния с наградой в текущее
+    \item Планирование до награды из текущего состояния.
+    \item Нахождение обратного пути из состояния с наградой в текущее.
 \end{enumerate}
@@
 
-Главная особенность работы памяти агента - она способна работать с суперпозициями (=множествами).
-Это означает, что память может отвечать на запросы вида \textit{имея набор состояний $\{s_i\}$ и набор действий $\{a_j\}$, в какие состояния $\{s'_m\}$ могут привести все возможные пары $(s_i, a_j)$}?
-Ответом на такой запрос является суперпозиция следующих состояний $S = \cup s'_m$.
+Главная особенность работы памяти агента - способность работать с объединениями "воспоминаний".
+Это означает, что память может отвечать на запросы вида \textit{имея набор состояний $S=\{s_i\}$ и набор действий $A=\{a_j\}$, в какие состояния $S'=\{s'_m\}$ могут привести все возможные пары $(s_i, a_j)$}?
+Ответом на такой запрос является объединение состояний в следующий момент времени $S' = \cup s'_m$.
 
-Вторая особенность - в ответе на запрос к памяти для каждого конкретного состояния $s'_m$ суперпозиции состояний есть возможность узнать, какие пары состояние-действие приводят в $s'_m$: $\{(s_i, a_j)\} = f^{-1}(s'_m)$.
+Вторая особенность - в ответе на запрос к памяти для каждого конкретного состояния $s'$ из множества предсказанных состояний $S'$ есть возможность узнать, какие пары состояние-действие приводят в $s'$: $\{(s, a)\} = f^{-1}(s')$.
+Обратите внимание, в общем случае в $s'$ может вести не единственная пара состояния-действия.
 
@@
 На текущий момент этот подмодуль реализован с помощью наивного списка состояний, в которых агент получил положительную награду - мы называем этот подмодуль списком целей.
-Список целей имеет ограниченный размер S, являющийся гиперпараметром, и работает как кольцевая очередь.
+Список целей имеет ограниченный размер, являющийся гиперпараметром, и работает как кольцевая очередь.
 
@@
 
-Перед тем как перейти к описанию алгоритма планирования, разберем для начала, какими данными оперирует агент, и в каком формате они представлены.
+Перед тем как перейти к описанию алгоритма планирования, опишем для начала, какими данными оперирует агент, и в каком формате они представлены.
 
 Пусть трактория движения агента - $\tau = (s_0, a_0, r_1, s_1, a_1, r_2, s_2, a_2, ...)$.
-Пара состояние-действие $(s_t, a_t)$ некоторого момента времени задает предусловие перехода в среде: $(s_t, a_t) \rightarrow s_{t+1}$.
+Тогда пара состояние-действие $(s_t, a_t)$ некоторого момента времени задает предусловие перехода в среде: $(s_t, a_t) \rightarrow s_{t+1}$.
 
-Память агента - это TM, которая запоминает следующие последовательности $[(s_0, a_0), (s_1, a_1), \dots]$.
-Сама TM устроена таким образом, что работает с разреженными векторами (SDR), поэтому пары состояние-действие требуется как-то закодировать в SDR.
+Память агента - это TM, которая запоминает последовательности вида $[(s_0, a_0), (s_1, a_1), \dots]$, сделанные на основе отыгранных траекторий в среде.
+Сама по себе TM работает с разреженными векторами (SDR), а значит пары состояние-действие перед подачей в ТМ требуется как-то закодировать (см.
+ссылка на подсекцию).
 
@@
 
-Также обученная TM умеет делать предсказания - для текущей пары состояние-действие $(s_t, a_t)$ давать совокупность всех возможных продолжений $(s_{t+1}, a_{t+1})$.
+Обученная TM умеет делать предсказания - для текущей пары состояние-действие $(s_t, a_t)$ давать совокупность всех возможных продолжений $(s_{t+1}, a_{t+1})$.
 "Совокупность" выражается как побитовое объединение разреженных векторов, соответствующих этим парам.
@@
 
-Т.к.
-TM работает с разреженными векторами в общем смысле, вектор, являющийся суперпозицией состояний-действий, является также валидным входным значением для ее активации [, означая совокупность составляющих этой суперпозиции].
+\begin{figure}
+  \includegraphics[width=1.\textwidth]{assets/memory_scheme.jpg}
+  \caption{TODO Набросок схемы.}
+\end{figure}
+
+Стоит отметить, что вектор, являющийся суперпозицией пар состояние-действие, тоже является валидным входным значением для ее активации [, означая совокупность составляющих этой суперпозиции].
+Эта особенность используется в процессе одного из шагов планирования.
 
@@
 
-И состояния, и действия в рамках рассматриваемых MDP - это целые числа из интервалов $[0, |S|)$ и $[0, |A|)$ соответственно.
-Каждый элемент пары кодируется независимо целочисленным кодировщиком - для состояний и действий он разный.
+В рамках рассматриваемых MDP все состояния $s \in S$ и действия $a \in A$ пронумерованы целыми числами в интервалах $[0, |S|)$ и $[0, |A|)$ соответственно.
+Каждый элемент пары состояние-действие кодируется независимо целочисленным кодировщиком - для состояний и действий он разный.
 Далее два полученных SDR вектора конкатенируются в итоговый SDR вектор.
@@
 Целочисленный кодировщик работает с числами некоторого фиксированного интервала $[0, N)$ и каждому числу интервала ставит в соответствие одинаковое количество активных бит.
-Работу целочисленного кодировщика проще показать на примере.
+Принцип его работы довольно очевиден и его проще показать на примере.
 Возьмем для этой случай $N = 3$.
@@
 
-На примере видно, что итоговый вектор разбивается на $N$ групп одинаковой длины (buckets), и каждая группа соответствует только одному из значений интервала.
+На примере видно, что итоговый вектор разбивается на $N$ групп одинаковой длины (buckets), и каждая группа активных бит соответствует только одному из значений интервала.
 Длина одной группы является гиперпараметром (в примере выше длина равна 4).
@@
 
-TODO упомянуть про кодирование состояний в сложных средах. State SDR encoder
+TODO упомянуть про кодирование состояний в сложных средах или выбросить следующий абзац на английском. State SDR encoder
 
@@
 
-Пусть в некоторый момент времени $t_0$ агент находится в состоянии $S_0 = s_0$.
-Чтобы ответить на вопрос, есть ли достижимое состояние из списка целей в радиусе $N$ действий, начнем планировать из суперпозиции $(S_0, A = \cup a^i)$ (TODO: объяснить обозначения), где $A$ - все множество доступных действий агента.
+Пусть в некоторый момент времени $t_0$ агент находится в состоянии $s_0$.
+Чтобы ответить на вопрос, есть ли достижимое состояние из списка целей в радиусе $N$ действий, он начинает планировать из суперпозиции $(S_0 = s_0, A = \cup a^i)$, где $A$ - все множество доступных действий агента.
+
 TODO рисунок и аналогия с BFS
 
-После первого шага предсказания мы будем иметь множество состояний $S_1 = \{s^i\}$, в которые можно попасть из текущего за одно действие агента.
-Продолжим планировать из суперпозиции этих состояний так же по всем возможным действиям.
-И так далее до тех пора пока либо не обнаружим хотя бы одно состояние из списка целей, либо не исчерпаем лимит горизонта планирования $N =$ \verb|planning_horizon|.
+После первого шага предсказания планировщик имеет множество состояний $S_1 = \{s^i\}$, в которые можно попасть из текущего за одно действие агента.
+Далее он продолжает планировать уже из суперпозиции $S_1$, т.е.
+попросту подав всю предсказанную на первом шаге суперпозицию на вход своей ТМ.
+Такой процесс предсказания и подачи всего предсказанного обратно на вход ТМ продолжается до тех пор, пока либо не будет обнаружено хотя бы одно состояние из списка целей, либо не будет исчерпан лимит горизонта планирования $N =$ \verb|planning_horizon|.
 
-Если цель найдена, то планировщик переходит к следующему этапу.
+Если цель найдена, то планировщик переходит к следующему этапу. В обратном случае планирование признается безуспешным, и агенту в текущем ходе остается воспользоваться случайной стратегией.
 
-В процессе проведения данного этапа на каждом шаге планировщик также сохраняет информацию о суперпозиции состояний, формируя их историю $[S_0, S_1, ..
+В процессе проведения этого этапа планировщик на каждом шаге сохраняет информацию об активных суперпозициях состояний, формируя их историю $[S_0, S_1, ..
 S_T]$.
-Более детально - для каждого шага сохраняются активации сегментов ячеек ТМ (TODO линк на что такое сегменты ТМ).
-Это позволяет не потерять информацию о зависимостях между переходами от одного шага к следующему - какие пары состояний-действий повлекли за собой активацию каких пар состояние-действие.
+Более детально - для каждого шага сохраняются активации сегментов ячеек ТМ (TODO линк на что такое сегменты ТМ далее).
+Это позволяет планировщику не потерять информацию о зависимостях между переходами от одного шага к следующему - какие пары состояний-действий $\{(s, a)\}$ повлекли за собой активацию каких пар $\{(s', a')\}$.
 
@@
 Получается такая контекстно-зависимая активация ячеек.
-В этой активации одновременно представлен и текущий "символ" последовательности, если смотреть с точки зрения активных столбцов (образно, вид анфас), и текущий префикс (или история) последовательности, если смотреть на уровне конфигурации активных ячеек (образно, вид в профиль).
+\textit{TODO выкинуть или написать понятнее В этой активации одновременно представлен и текущий "символ" последовательности, если смотреть с точки зрения активных столбцов (образно, вид анфас), и текущий префикс (или история) последовательности, если смотреть на уровне конфигурации активных ячеек (образно, вид в профиль).}
 
@@
 Деполяризация формирует ожидание продолжения последовательности - какой SDR вектор будет подан на вход, а значит какие столбцы будут активированы в следующий момент времени.
-Ожидание выражается в переводе части ячеек этих столбцов в выделенное, деполяризованное, состояние.
+Ожидание выражается в переводе части ячеек этих столбцов в выделенное - деполяризованное - состояние.
 
 Процесс деполяризации происходит следующим образом.
-Все ячейки ТМ потенциально имеют связи друг с другом, и в процессе обучения некоторые из этих связей формируются в способные к активации (синапсы).
+Все ячейки ТМ потенциально имеют связи друг с другом, и в процессе обучения некоторые из этих связей формируются в способные к активации синапсы.
 Синапсы одной ячейки логически разделены на независимые группы - сегменты, где в разных сегментах могут независимо присутствовать одинаковые синапсы.
@@
 Получается следующая логика от лица ячейки: если в текущий момент времени один из моих сегментов распознал некоторый контекст, с большой вероятностью следует ожидать, что следом будет активирован и мой столбец, т.к.
-похоже на то, что похожую последовательность мы уже раньше наблюдали.
+похоже на то, что похожую последовательность раньше уже наблюдалась.
 В таком случае ячейка переходит в предсказательное деполяризованное состояние, и, если на следующем шаге ее столбец действительно оказывается активным, она тоже активируется.
@@
 Именно эта особенность будет использована для обращения функции перехода.
-Основная трудность в этой задачи заключается в том, как подняться от уровня ячеек до уровня состояний-действий.
+Основная трудность в этой задаче заключается в том, как подняться от уровня ячеек до уровня состояний-действий.
 
@@
 
-\textbf{NB}.
-Алгоритм кластеризации работает следующим образом.
-Стартуя с набора кластеров, соответствующих каждому из рассматриваемых сегментов, он итеративно склеивает существенно пересекающиеся кластеры пока это возможно.
+\textbf{NB}. Алгоритм кластеризации, стартуя с набора кластеров, соответствующих каждому из рассматриваемых сегментов, итеративно склеивает существенно пересекающиеся кластеры пока это возможно.
 
@@
   \KwData{$d_t$: required cells depolarization at time $t$, $t$: timestep}
+  \quad \\
+
   \If(){$t \leq 0$} {
@@
 
+  \tcc{Finds clusters capable of depolarizing most of the $d_t$ cells}
   Init set of clusters $C = \{c_i\}$ as presynaptic cells of active segments of the $d_t$ cells. \;
-  Iteratively try to merge any two pair of sufficiently intersecting clusters \;
-  OR: \;
   \While{$\exists c_i, c_j: match(c_i, c_j) \geq \Theta_{clust}$} {
@@
   }
+  \quad \\
+
   \ForEach{$c \in C$}{
-    Count how many of $d_t$ cells are depolarized by the cluster. \;
-    Keep in $C$ only clusters with sufficient induced depolarization. \;
-    OR: \;
     \If() {$CountInducedDepolarization(c) < \Theta_{dep}$}{
@@
   $backtracking\_candidates \leftarrow C$ \;
+
+  \tcc{Recursively checks each candidate cluster - is it possible to "come back" to $s_0$}
+  \ForEach{$c \in backtracking\_candidates$}{
+    $s_t, a_t \leftarrow Decode(ColumnsFromCells(c))$ \;
+
+    $successful, actions \leftarrow Backtrack(s_t, t-1, active\_segments)$ \;
+    \If() {successful} {
+      $actions.append(a_t)$ \;
+      \Return{True, actions}
+    }
+  }
+
+  \Return{False, $\emptyset$}
+  \caption{Backtrack($d_t$, $t$, $active\_segments$)}
+\end{function}
+
+\begin{function}
+  \tcc{Другой вариант псевдокода бэктрекинга}
+  \KwData{$d_t$: required cells depolarization at time $t$, $t$: timestep}
+  \quad \\
+
+  \If(){$t \leq 0$} {
+    \Return{True, $\emptyset$}
+  }{}
+
+  Init set of clusters $C = \{c_i\}$ as presynaptic cells of active segments of the $d_t$ cells. \;
+  Iteratively merge any two sufficiently intersecting clusters \;
+  \quad \\
+
+  \ForEach{$c \in C$}{
+    Count how many of $d_t$ cells are depolarized by the cluster. \;
+    Keep in $C$ only clusters with sufficient induced depolarization. \;
+  }
+  \quad \\
+
+  $backtracking\_candidates \leftarrow C$ \;
+
+  \tcc{Recursively checks each candidate cluster - is it possible to "come back" to $s_0$}
   \ForEach{$c \in backtracking\_candidates$}{
@@
   \begin{minipage}{.32\linewidth}
-    \includesvg[width=\linewidth]{./multi_way_transfer/multi_way_v0__steps}
+    \includesvg[width=\linewidth]{./assets/multi_way_v0__steps}
     \subcaption{\texttt{multi\_way\_v0}}
@@
   \begin{minipage}{.32\linewidth}
-    \includesvg[width=\linewidth]{./multi_way_transfer/multi_way_v1__steps}
+    \includesvg[width=\linewidth]{./assets/multi_way_v1__steps}
     \subcaption{\texttt{multi\_way\_v1}}
@@
   \begin{minipage}{.32\linewidth}
-    \includesvg[width=\linewidth]{./multi_way_transfer/multi_way_v2__steps}
+    \includesvg[width=\linewidth]{./assets/multi_way_v2__steps}
     \subcaption{\texttt{multi\_way\_v2}}
@@
   \begin{minipage}{.98\linewidth}
-    \includesvg[width=\linewidth]{./multi_way_transfer/multi_way_v0__steps}
+    \includesvg[width=\linewidth]{./assets/multi_way_v0__steps}
     \subcaption{Сравнение агента с бейзлайнами}
@@
   \begin{minipage}{.49\linewidth}
-    \includesvg[width=\linewidth]{./multi_way_transfer/multi_way_v0__dqn__steps}
+    \includesvg[width=\linewidth]{./assets/multi_way_v0__dqn__steps}
     \subcaption{Сравнение жадного и $\epsilon$-жадного агента DQN!}
@@
   \begin{minipage}{.49\linewidth}
-    \includesvg[width=\linewidth]{./multi_way_transfer/multi_way_v0__1__steps}
+    \includesvg[width=\linewidth]{./assets/multi_way_v0__1__steps}
     \subcaption{Горизонт планирования: 1}
@@
   \begin{minipage}{.49\linewidth}
-    \includesvg[width=\linewidth]{./multi_way_transfer/multi_way_v0__2__steps}
+    \includesvg[width=\linewidth]{./assets/multi_way_v0__2__steps}
     \subcaption{Горизонт планирования: 2}
@@
   \begin{minipage}{.49\linewidth}
-    \includesvg[width=\linewidth]{./multi_way_transfer/multi_way_v0__4__steps}
+    \includesvg[width=\linewidth]{./assets/multi_way_v0__4__steps}
     \subcaption{Горизонт планирования: 4}
@@
   \begin{minipage}{.49\linewidth}
-    \includesvg[width=\linewidth]{./gridworld_transfer/gridworld_5x5_1_1_200_1337_map_0_1173222464}
+    \includesvg[width=\linewidth]{./assets/gridworld_5x5_1_1_200_1337_map_0_1173222464}
   \end{minipage}
   \begin{minipage}{.49\linewidth}
-    \includesvg[width=\linewidth]{./gridworld_transfer/gridworld_5x5_1_1_200_1337_map_2_1561234712}
+    \includesvg[width=\linewidth]{./assets/gridworld_5x5_1_1_200_1337_map_2_1561234712}
   \end{minipage}
@@
   \begin{minipage}{.49\linewidth}
-    \includesvg[width=\linewidth]{./gridworld_transfer/gridworld_5x5_100_2_8_1337__steps}
+    \includesvg[width=\linewidth]{./assets/gridworld_5x5_100_2_8_1337__steps}
   \end{minipage}
   \begin{minipage}{.49\linewidth}
-    \includesvg[width=\linewidth]{./gridworld_transfer/gridworld_5x5_100_2_8_42__steps}
+    \includesvg[width=\linewidth]{./assets/gridworld_5x5_100_2_8_42__steps}
   \end{minipage}
@@
   \begin{minipage}{.32\linewidth}
-    \includesvg[width=\linewidth]{./gridworld_transfer/gridworld_5x5_100_2_8_1337__dqn__steps}
-    \includesvg[width=\linewidth]{./gridworld_transfer/gridworld_5x5_100_2_8_42__dqn__steps}
+    \includesvg[width=\linewidth]{./assets/gridworld_5x5_100_2_8_1337__dqn__steps}
+    \includesvg[width=\linewidth]{./assets/gridworld_5x5_100_2_8_42__dqn__steps}
     \subcaption{Жадный и $\epsilon$-жадный DQN агент}
@@
   \begin{minipage}{.32\linewidth}
-    \includesvg[width=\linewidth]{./gridworld_transfer/gridworld_5x5_100_2_8_1337__2__steps}
-    \includesvg[width=\linewidth]{./gridworld_transfer/gridworld_5x5_100_2_8_42__2__steps}
+    \includesvg[width=\linewidth]{./assets/gridworld_5x5_100_2_8_1337__2__steps}
+    \includesvg[width=\linewidth]{./assets/gridworld_5x5_100_2_8_42__2__steps}
     \subcaption{Горизонт планирования 2}
@@
   \begin{minipage}{.32\linewidth}
-    \includesvg[width=\linewidth]{./gridworld_transfer/gridworld_5x5_100_2_8_1337__4-8__steps}
-    \includesvg[width=\linewidth]{./gridworld_transfer/gridworld_5x5_100_2_8_42__4-8__steps}
+    \includesvg[width=\linewidth]{./assets/gridworld_5x5_100_2_8_1337__4-8__steps}
+    \includesvg[width=\linewidth]{./assets/gridworld_5x5_100_2_8_42__4-8__steps}
     \subcaption{Горизонт планирования 4}
@@
   \begin{minipage}{.32\linewidth}
-    \includesvg[width=\linewidth]{./gridworld_transfer/gridworld_5x5_50_4_4_1337__steps}
-    \includesvg[width=\linewidth]{./gridworld_transfer/gridworld_5x5_50_4_4_1337__steps_rel_htm_0}
+    \includesvg[width=\linewidth]{./assets/gridworld_5x5_50_4_4_1337__steps}
+    \includesvg[width=\linewidth]{./assets/gridworld_5x5_50_4_4_1337__steps_rel_htm_0}
   \end{minipage}
   \begin{minipage}{.32\linewidth}
-    \includesvg[width=\linewidth]{./gridworld_transfer/gridworld_5x5_50_4_4_1337__dqn__steps}
-    \includesvg[width=\linewidth]{./gridworld_transfer/gridworld_5x5_50_4_4_1337__2__steps}
+    \includesvg[width=\linewidth]{./assets/gridworld_5x5_50_4_4_1337__dqn__steps}
+    \includesvg[width=\linewidth]{./assets/gridworld_5x5_50_4_4_1337__2__steps}
   \end{minipage}
   \begin{minipage}{.32\linewidth}
-    \includesvg[width=\linewidth]{./gridworld_transfer/gridworld_5x5_50_4_4_1337__1__steps}
-    \includesvg[width=\linewidth]{./gridworld_transfer/gridworld_5x5_50_4_4_1337__4-8__steps}
+    \includesvg[width=\linewidth]{./assets/gridworld_5x5_50_4_4_1337__1__steps}
+    \includesvg[width=\linewidth]{./assets/gridworld_5x5_50_4_4_1337__4-8__steps}
   \end{minipage}
@@
   \begin{minipage}{.32\linewidth}
-    \includesvg[width=\linewidth]{./gridworld_transfer/gridworld_5x5_20_1_20_1337__steps}
+    \includesvg[width=\linewidth]{./assets/gridworld_5x5_20_1_20_1337__steps}
   \end{minipage}
   \begin{minipage}{.32\linewidth}
-    \includesvg[width=\linewidth]{./gridworld_transfer/gridworld_5x5_20_1_20_1337__1__steps}
+    \includesvg[width=\linewidth]{./assets/gridworld_5x5_20_1_20_1337__1__steps}
   \end{minipage}
   \begin{minipage}{.32\linewidth}
-    \includesvg[width=\linewidth]{./gridworld_transfer/gridworld_5x5_20_1_20_1337__2__steps}
+    \includesvg[width=\linewidth]{./assets/gridworld_5x5_20_1_20_1337__2__steps}
   \end{minipage}
@@
   \begin{minipage}{.49\linewidth}
-    \includesvg[width=\linewidth]{./gridworld_transfer/gridworld_5x5_1_1_200_1337__steps}
+    \includesvg[width=\linewidth]{./assets/gridworld_5x5_1_1_200_1337__steps}
   \end{minipage}
   \begin{minipage}{.49\linewidth}
-    \includesvg[width=\linewidth]{./gridworld_transfer/gridworld_5x5_1_1_200_1337__2__steps}
+    \includesvg[width=\linewidth]{./assets/gridworld_5x5_1_1_200_1337__2__steps}
   \end{minipage}

### 2020-10-27 watcher/fly.py
add form's difference for cells

@@ -15,7 +15,7 @@ if __name__ == '__main__':
     drawer = Drawer('data/'+args.name, win)
 
     while not win.should_close():
-        drawer.update_instance_cubes_color_array()
+        drawer.update()
         # the main application loop
         while not drawer.window.should_close() and not drawer.window.next and not drawer.window.previous:
             drawer.process()

### 2020-10-27 watcher/utils/drawer.py

@@ -15,13 +15,8 @@ class Drawer:
         with open(path, "r") as outfile:
             self.data_base = json.load(outfile)
         self.window = window
-        self.current = 0
         self.active_cell = -1
-
-        self.instance_array_cube = []
-        self.instance_array_cube_color = []
-        self.offset = 0
-        self.distance = 4
+        self.current = 0
 
         self.cells_in_column = self.data_base['params']['cells_per_column']
         self.num_of_cells = self.data_base['params']['num_cells']
@@ -29,8 +24,6 @@ class Drawer:
         self.start_point = [0, 0, 0]
         self.size = int(np.sqrt(self.num_of_cells / self.cells_in_column))
         self.pick = Picker(self.window.width, self.window.height, self.num_of_cells)
-        self.textDrawer = TextDrawer()
-        self.textDrawer.load_font('fonts/calibri.ttf', 40 * 64, 1 * 64)
 
         # color settings
         self.background_color = [0.98, 1.0, 0.98]
@@ -41,43 +34,31 @@ class Drawer:
         self.picked_color = [0., 0., 0.9]
         self.default_color = [0.8, 0.8, 0.8]
         self.bright_off = 2
-        self.bright_less = 0.
+        self.bright_less = 0.4
 
-        # shaders
-        vertex_src = """
-        # version 330
+        self.textDrawer = TextDrawer()
+        self.textDrawer.load_font('fonts/calibri.ttf', 40 * 64, 1 * 64)
 
-        layout(location = 0) in vec3 a_position;
-        layout(location = 1) in vec3 a_offset;
-        layout(location = 2) in vec3 a_color;
-
-        uniform mat4 model;
-        uniform mat4 projection;
-        uniform mat4 view;
-
-        out vec3 v_color;
-
-        void main()
-        {
-            vec3 final_pos = a_position + a_offset;
-            gl_Position =  projection * view * model * vec4(final_pos, 1.0f);
-            v_color = a_color;
+        params = {
+            'data_base': self.data_base,
+            'cells_in_column': self.cells_in_column,
+            'num_of_cells': self.num_of_cells,
+            'start_point': self.start_point,
+            'size': self.size,
+            'background_color': self.background_color,
+            'active_cells_color': self.active_cells_color,
+            'predictive_cells_color': self.predictive_cells_color,
+            'active_and_predictive_cells_color': self.active_and_predictive_cells_color,
[…]

### 2020-10-27 htm_rl/htm_rl/envs/gridworld_pomdp.py
State id is added.

@@ -38,7 +38,9 @@ class GridWorld:
         max_sight_range:
             int max number of cells that agent can see on line
         observable_vars:
-            ordered subset of {'distance', 'surface', 'relative_row', 'relative_column', 'relative_direction'}
+            ordered subset of {'distance', 'surface', 'relative_row',
+                                'relative_column', 'relative_direction',
+                                'state_id'}
             as list, output order will be the same as in observable_vars
     """
 
@@ -196,11 +198,15 @@ class GridWorld:
         if relative_direction < 0:
             relative_direction += 4
 
+        state_id = (self.agent_position['column'] - 1 + (self.agent_position['row'] - 1) * self.world_size[1]
+                    + self.agent_direction * self.world_size[0] * self.world_size[1])
+
         self.observable_state = {'distance': distance - 1,
                                  'surface': surface,
                                  'relative_row': relative_coordinates[0],
                                  'relative_column': relative_coordinates[1],
-                                 'relative_direction': relative_direction}
+                                 'relative_direction': relative_direction,
+                                 'state_id': state_id}
 
         if self.observable_vars is not None:
             filtered_observation = dict()

### 2020-10-27 watcher/fly.py
configure initial camera view

@@ -1,5 +1,5 @@
 from utils.drawer import Drawer
-from utils.window import Window
+
 import argparse
 
 if __name__ == '__main__':
@@ -11,15 +11,14 @@ if __name__ == '__main__':
         width, height = 1280, 720
     else:
         width, height = args.size.split(',')
-    win = Window(int(width), int(height))
-    drawer = Drawer('data/'+args.name, win)
+    drawer = Drawer('data/'+args.name, [int(width), int(height)])
 
-    while not win.should_close():
+    while not drawer.window.should_close():
         drawer.update()
         # the main application loop
         while not drawer.window.should_close() and not drawer.window.next and not drawer.window.previous:
             drawer.process()
-        if drawer.window.next and drawer.current + 1 < len(drawer.data_base.keys()): drawer.current = drawer.current + 1
+        if drawer.window.next and drawer.current + 2 < len(drawer.data_base.keys()): drawer.current = drawer.current + 1
         if drawer.window.previous and drawer.current > 0: drawer.current = drawer.current - 1
         drawer.window.next = False
         drawer.window.previous = False

### 2020-10-27 watcher/utils/camera.py

@@ -1,16 +1,15 @@
 from pyrr import Vector3, vector, vector3, matrix44
-from math import sin, cos, radians
+from math import sin, cos, radians, sqrt, atan, pi
 
 class Camera:
-    def __init__(self):
-        self.camera_pos = Vector3([80., 30.0, 80.0])
-        self.camera_front = Vector3([-1.0, -1.0, -1.0])
-        self.camera_up = Vector3([0.0, 1.0, 0.0])
-        self.camera_right = Vector3([1.0, 0.0, 0.0])
+    def __init__(self, target):
+        r = sqrt(target[0] ** 2 + target[2] ** 2)
 
+        self.camera_pos = Vector3([-2*target[0], 3*target[1], -2*target[2]])
         self.mouse_sensitivity = 0.25
-        self.jaw = -90
-        self.pitch = 0
+        self.jaw = 45
+        self.pitch = - atan(5*target[1]/(6*r)) * 180 / pi
+        self.update_camera_vectors()
 
     def get_view_matrix(self):
         return matrix44.create_look_at(self.camera_pos, self.camera_pos + self.camera_front, self.camera_up)
@@ -23,10 +22,10 @@ class Camera:
         self.pitch += yoffset
 
         if constrain_pitch:
-            if self.pitch > 45:
-                self.pitch = 45
-            if self.pitch < -45:
-                self.pitch = -45
+            if self.pitch > 90:
+                self.pitch = 90
+            if self.pitch < -90:
+                self.pitch = -90
 
         self.update_camera_vectors()
 

### 2020-10-27 watcher/utils/drawer.py

@@ -6,15 +6,16 @@ import numpy as np
 
 from utils.picker import Picker
 from utils.text import TextDrawer
+from utils.window import Window
 
 
 class Drawer:
 
-    def __init__(self, path, window):
+    def __init__(self, path, win_params):
         # data
         with open(path, "r") as outfile:
             self.data_base = json.load(outfile)
-        self.window = window
+
         self.active_cell = -1
         self.current = 0
 
@@ -23,6 +24,10 @@ class Drawer:
 
         self.start_point = [0, 0, 0]
         self.size = int(np.sqrt(self.num_of_cells / self.cells_in_column))
+        distance = 5.
+
+        target = (distance * self.size / 2., 3 * distance * self.cells_in_column / 2., distance * self.size / 2.)
+        self.window = Window(win_params[0], win_params[1], target)
         self.pick = Picker(self.window.width, self.window.height, self.num_of_cells)
 
         # color settings
@@ -55,7 +60,7 @@ class Drawer:
             'bright_off': self.bright_off,
             'bright_less': self.bright_less,
             'draw_cells': [None],
-            'distance': 3,
+            'distance': distance,
             'win_width': self.window.width,
             'win_height': self.window.height
         }

### 2020-10-27 watcher/utils/window.py

@@ -5,10 +5,10 @@ from utils.camera import Camera
 
 class Window():
 
-    def __init__(self, width, height):
+    def __init__(self, width, height, cam_target):
         self.width = width
         self.height = height
-        self.cam = Camera()
+        self.cam = Camera(cam_target)
 
         # control variables
         self.left = False

### 2020-10-27 htm_rl/htm_rl/envs/gridworld_pomdp.py
Gridworld state representation modifications.

@@ -3,10 +3,10 @@ import numpy as np
 legend = {
     'map':
         {
-             0: ' ',
-             1: "#",
-             2: '*'
-         },
+            0: ' ',
+            1: "#",
+            2: '*'
+        },
     'agent':
         {
             0: '>',
@@ -40,8 +40,15 @@ class GridWorld:
         observable_vars:
             ordered subset of {'distance', 'surface', 'relative_row',
                                 'relative_column', 'relative_direction',
-                                'state_id'}
+                                'state_id',
+                                'flatten_index',
+                                'row',
+                                'column',
+                                'direction'}
             as list, output order will be the same as in observable_vars
+            'state_id' correspond to mdp framework, unique id for every state.
+        one_value_state:
+            bool convert tuple of state to one value
     """
 
     def __init__(self, world_description,
@@ -49,13 +56,16 @@ class GridWorld:
                  agent_initial_position: dict,
                  agent_initial_direction=0,
                  max_sight_range=None,
-                 observable_vars=None):
+                 observable_vars=None,
+                 one_value_state=False):
 
+        self.one_value_state = one_value_state
         self.observable_vars = observable_vars
         self.world_size = world_size
-        self.world_description = np.ones((world_size[0]+2, world_size[1]+2))
-        self.world_description[1:-1, 1:-1] = np.array(world_description)
+        self.world_description = np.ones((world_size[0] + 2, world_size[1] + 2), dtype=np.int)
+        self.world_description[1:-1, 1:-1] = np.array(world_description, dtype=np.int)
         self.n_actions = 3
+        self.n_states = world_size[0] * world_size[1] * 4
 
         if max_sight_range is None:
             max_sight_range = max(self.world_description.shape)
@@ -70,6 +80,23 @@ class GridWorld:
         self.max_sight_range = max_sight_range
         self.observable_state = None
         self.filtered_observation = None
+
+        self.dimensions = {
+            'distance': max(self.world_size),
+            'surface': 3,
+            'relative_row': 2 * (self.world_size[0] + 1),
+            'relative_column': 2 * (self.world_size[1] + 1),
+            'relative_direction': 4,
+            'state_id': self.n_states,
+            'flatten_index': self.world_size[0] * self.world_size[1],
+            'row': self.world_size[0] + 1,
+            'column': self.world_size[1] + 1,
+            'direction': 4
+        }
+
+        self.n_obs_states = 1
+        for key in self.observable_vars:
[…]

### 2020-10-28 htm_rl/htm_rl/agent/agent.py
WIP mcts

@@ -6,6 +6,7 @@ import numpy as np
 import matplotlib.pyplot as plt
 from tqdm import trange
 
+from htm_rl.agent.mcts_actor_critic import MctsActorCritic
 from htm_rl.agent.memory import Memory
 from htm_rl.agent.planner import Planner
 from htm_rl.agent.train_eval import RunStats, RunResultsProcessor
@@ -23,7 +24,9 @@ class Agent:
     _use_cooldown: bool
     _planning_horizon: int
 
-    def __init__(self, memory: Memory, planner: Planner, n_actions, use_cooldown=False):
+    def __init__(self, memory: Memory, planner: Planner, n_actions, use_cooldown=False,
+            mcts_actor_critic: MctsActorCritic = None
+    ):
         self.memory = memory
         self.planner = planner
         self._n_actions = n_actions
@@ -31,6 +34,7 @@ class Agent:
         self._planning_horizon = self.planner.planning_horizon
         self.set_planning_horizon(self._planning_horizon)
         self._goal_state = None
+        self._mcts_actor_critic = mcts_actor_critic
 
     def set_planning_horizon(self, planning_horizon: int):
         self.planner.planning_horizon = planning_horizon
@@ -45,9 +49,23 @@ class Agent:
     def reset(self):
         self.memory.tm.reset()
         self._init_planning()
+        if self._mcts_actor_critic is not None:
+            self._mcts_actor_critic.reset()
+            ac = self._mcts_actor_critic
+            q = np.round(ac.cell_value / ac.cell_visited_count, 2)
+            value_bit_buckets = (q[i: i + 8] for i in range(0, q.size, 8))
+            q_str = '\n'.join(
+                ' '.join(map(str, value_bits)) for value_bits in value_bit_buckets
+            )
+            trace(3, 1, q_str)
+            trace(3, 1, '='*20)
 
     def make_step(self, state, reward, is_done, verbosity: int):
         trace(verbosity, 2, f'\nState: {state}; reward: {reward}')
+        if self._mcts_actor_critic is not None:
+            encoded_state = self.memory.encoder.encode(Sa(state, None))
+            self._mcts_actor_critic.add_step(encoded_state, max(reward, 0))
+
         action = self._make_action(state, is_done, verbosity)
         trace(verbosity, 2, f'\nMake action: {action}')
 

### 2020-10-28 htm_rl/htm_rl/agent/mcts_actor_critic.py

@@ -25,8 +25,8 @@ class MctsActorCritic:
         self.online_learning = online
         self._rng = np.random.default_rng(seed)
 
-        self._cell_visited_count = np.zeros(cells_sdr_size, dtype=np.int)
-        self._cell_value = np.full(cells_sdr_size, 1e-5, dtype=np.float)
+        self._cell_visited_count = np.full(cells_sdr_size, 1e-5, dtype=np.float)
+        self._cell_value = np.zeros(cells_sdr_size, dtype=np.float)
         self._cell_eligibility_trace = np.zeros(cells_sdr_size, dtype=np.float)
         self._total_steps = 0
 

### 2020-10-28 htm_rl/htm_rl/config.py

@@ -13,6 +13,7 @@ from htm_rl.agent.agent import Agent, AgentRunner, TransferLearningExperimentRun
 from htm_rl.agent.legacy_agent import LegacyAgent
 from htm_rl.agent.legacy_memory import LegacyMemory
 from htm_rl.agent.legacy_planner import LegacyPlanner
+from htm_rl.agent.mcts_actor_critic import MctsActorCritic
 from htm_rl.agent.memory import Memory
 from htm_rl.agent.planner import Planner
 from htm_rl.agent.train_eval import RunResultsProcessor
@@ -174,6 +175,7 @@ def register_classes(yaml: YAML):
         GridworldMapGenerator,
         TransferLearningExperimentRunner2,
         RunResultsProcessor,
+        MctsActorCritic,
     ]
 
     constructor: SafeConstructor = yaml.constructor

### 2020-10-28 watcher/utils/camera.py
add winner cells

@@ -5,9 +5,9 @@ class Camera:
     def __init__(self, target):
         r = sqrt(target[0] ** 2 + target[2] ** 2)
 
-        self.camera_pos = Vector3([-2*target[0], 3*target[1], -2*target[2]])
+        self.camera_pos = Vector3([4*target[0], 3*target[1], 4*target[2]])
         self.mouse_sensitivity = 0.25
-        self.jaw = 45
+        self.jaw = 225
         self.pitch = - atan(5*target[1]/(6*r)) * 180 / pi
         self.update_camera_vectors()
 

### 2020-10-28 watcher/utils/drawer.py

@@ -34,6 +34,7 @@ class Drawer:
         self.background_color = [0.98, 1.0, 0.98]
         self.active_cells_color = [0.9, 0., 0.]
         self.predictive_cells_color = [0., 0.9, 0.]
+        self.winner_cells_color = [0.99, 0.06, 0.75]
         self.active_and_predictive_cells_color = [0.9, 0.9, 0.]
         self.active_segments_color = [0., 0., 0.]
         self.picked_color = [0., 0., 0.9]
@@ -53,6 +54,7 @@ class Drawer:
             'background_color': self.background_color,
             'active_cells_color': self.active_cells_color,
             'predictive_cells_color': self.predictive_cells_color,
+            'winner_cells_color': self.winner_cells_color,
             'active_and_predictive_cells_color': self.active_and_predictive_cells_color,
             'active_segments_color': self.active_segments_color,
             'picked_color': self.picked_color,
@@ -213,11 +215,11 @@ class Drawer:
         active_predictive = []
         not_colored = []
         for cell in range(self.num_of_cells):
-            if self.data_base[str(self.current)][str(cell)]['St'] == 1:
+            if abs(self.data_base[str(self.current)][str(cell)]['St']) == 1:
                 active.append(cell)
-            elif self.data_base[str(self.current)][str(cell)]['St'] == 2:
+            elif abs(self.data_base[str(self.current)][str(cell)]['St']) == 2:
                 predictive.append(cell)
-            elif self.data_base[str(self.current)][str(cell)]['St'] == 3:
+            elif abs(self.data_base[str(self.current)][str(cell)]['St']) == 3:
                 active_predictive.append(cell)
             else:
                 not_colored.append(cell)
@@ -280,7 +282,7 @@ class Drawer:
                                       backColor=self.active_segments_color)
         else:
             if self.active_cell >= 0:
-                self.textDrawer.draw_text('Not connected cells', (self.window.width - left, self.window.width - 310),
+                self.textDrawer.draw_text('Not connected cells', (self.window.width - left, self.window.width - 380),
                                           (self.window.width, self.window.width), scale=(1.0, 1.),
                                           foreColor=np.array(self.default_color) * self.bright_less,
                                           backColor=np.array(self.default_color) * self.bright_less)
@@ -292,11 +294,15 @@ class Drawer:
                                       (self.window.width, self.window.width), scale=(1.0, 1.),
                                       foreColor=self.predictive_cells_color,
                                       backColor=self.predictive_cells_color)
[…]

### 2020-10-28 watcher/writer.py

@@ -13,6 +13,7 @@ class Writer():
     def write(self):
         active_cells = self.tm.getActiveCells().sparse
         predictive_cells = self.tm.getPredictiveCells().sparse
+        winner_cells = self.tm.getWinnerCells().sparse
         active_segments = self.tm.getActiveSegments()
 
         cells = {}
@@ -22,6 +23,7 @@ class Writer():
             if cell in active_cells: cell_dict['St'] = 1
             if cell in predictive_cells: cell_dict['St'] = 2
             if cell in active_cells and cell in predictive_cells: cell_dict['St'] = 3
+            if cell in winner_cells: cell_dict['St'] = - cell_dict['St']
             segments = self.tm.connections.segmentsForCell(cell)
             dict_segments = {}
             for segment in segments:

### 2020-10-28 watcher/utils/drawer.py
add permanence for synapses and text feature

@@ -17,6 +17,7 @@ class Drawer:
             self.data_base = json.load(outfile)
 
         self.active_cell = -1
+        self.active_connection = -1
         self.current = 0
 
         self.cells_in_column = self.data_base['params']['cells_per_column']
@@ -228,40 +229,29 @@ class Drawer:
     def update(self):
         self.cells_active.draw_cells, self.cells_predictive.draw_cells, self.cells_active_predictive.draw_cells, self.cells_not.draw_cells = self.cells_type()
 
-        self.cells_active.update(self.current,
-                          self.window.show_segments,
-                          self.window.show_active_segments,
-                          self.active_cell)
-        self.cells_predictive.update(self.current,
-                          self.window.show_segments,
-                          self.window.show_active_segments,
-                          self.active_cell)
-        self.cells_active_predictive.update(self.current,
-                                     self.window.show_segments,
-                                     self.window.show_active_segments,
-                                     self.active_cell)
-        self.cells_not.update(self.current,
-                                     self.window.show_segments,
-                                     self.window.show_active_segments,
-                                     self.active_cell)
+        self.update_color()
 
     def update_color(self):
-        self.cells_active.update_instance_cubes_color_array(self.current,
-                                                     self.window.show_segments,
-                                                     self.window.show_active_segments,
-                                                     self.active_cell)
-        self.cells_predictive.update_instance_cubes_color_array(self.current,
-                                                     self.window.show_segments,
-                                                     self.window.show_active_segments,
-                                                     self.active_cell)
-        self.cells_active_predictive.update_instance_cubes_color_array(self.current,
-                                                                self.window.show_segments,
-                                                                self.window.show_active_segments,
-                                                                self.active_cell)
-        self.cells_not.update_instance_cubes_color_array(self.current,
-                                                                self.window.show_segments,
-                                                                self.window.show_active_segments,
-                                                                self.active_cell)
+        self.cells_active.update(self.current,
+                                 self.window.show_segments,
+                                 self.window.show_active_segments,
+                                 self.active_cell,
+                                 self.active_connection)
+        self.cells_predictive.update(self.current,
+                                     self.window.show_segments,
[…]

### 2020-10-28 watcher/utils/window.py

@@ -24,6 +24,7 @@ class Window():
         self.follow_cursor = False
 
         self.picker = False
+        self.picker_right = False
         self.lastX = self.width / 2
         self.lastY = self.height / 2
         self.mouseX = 0
@@ -113,6 +114,8 @@ class Window():
     def mouse_button_callback(self, window, button, action, mods):
         if button == glfw.MOUSE_BUTTON_LEFT and action == glfw.PRESS:
             self.picker = True
+        if button == glfw.MOUSE_BUTTON_RIGHT and action == glfw.PRESS:
+            self.picker_right = True
         if button == glfw.MOUSE_BUTTON_MIDDLE and action == glfw.PRESS:
             self.follow_cursor = True
         elif button == glfw.MOUSE_BUTTON_MIDDLE and action == glfw.RELEASE:

### 2020-10-28 watcher/writer.py

@@ -10,13 +10,13 @@ class Writer():
                   'num_cells': self.num_cells}
         self.data = {'params': params}
 
-    def write(self):
+    def write(self, text='Watcher'):
         active_cells = self.tm.getActiveCells().sparse
         predictive_cells = self.tm.getPredictiveCells().sparse
         winner_cells = self.tm.getWinnerCells().sparse
         active_segments = self.tm.getActiveSegments()
 
-        cells = {}
+        cells = {'text': text}
         for cell in range(self.num_cells):
             cell_dict = {}
             cell_dict['St'] = 0
@@ -29,8 +29,9 @@ class Writer():
             for segment in segments:
                 dict_segment = {}
                 dict_segment['St'] = 1 if segment in active_segments else 0
-                dict_segment['Ce'] = [self.tm.connections.presynapticCellForSynapse(synapse) for synapse
-                                      in self.tm.connections.synapsesForSegment(segment)]
+                dict_segment['Ce'] = {self.tm.connections.presynapticCellForSynapse(synapse):
+                                       self.tm.connections.permanenceForSynapse(synapse) for synapse
+                                      in self.tm.connections.synapsesForSegment(segment)}
                 dict_segments[segment] = dict_segment
             cell_dict['Se'] = dict_segments
             cells[cell] = cell_dict

### 2020-10-29 reports/v1_0/paper_eng/paper_eng.tex
pre-final draft of the paper

\documentclass[runningheads]{llncs}
\usepackage{style_eng}
\bibliographystyle{splncs04}

\begin{document}

\title {Planning with Hierarchical Temporal Memory for deterministic Markov Decision Problem}
\author {Petr Kuderov \inst{1} \and Aleksandr I. Panov \inst{1}}

\authorrunning{P. Kuderov, A. Panov}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
\institute{Moscow Institute of Physics and Technology, Russia}

\maketitle

\begin{abstract}
  Among the key problems in artificial intelligence is sequential decision making, which can be formalized as Markov Decision Process (MDP). One approach to solve it, called model-based Reinforcement Learning (RL), combines learning the model of the environment and the global policy. Having a good model of the environment opens up such properties as data efficiency and targeted exploration.

  While most of the memory-based approaches are based on using Artificial Neural Networks (ANNs), in our work we instead draw the ideas from Hierarchical Temporary Memory (HTM) framework, which is based on human-like memory model. We utilize it to build an agent's memory that learns the environment dynamics. We also accompany it with an example of planning algorithm, that enables the agent to solve RL tasks.
\end{abstract}

\keywords{Model-based reinforcement learning \and Markov Decision Process \and planning \and Hierarchical Temporal Memory \and Sparse Distributed Representation.}

\section{Introduction}

Reinforcement Learning (RL) seek to find methods that enable robots or virtual agents to act autonomously in order to solve human-defined tasks. It requires from such methods to learn a policy for sequential decision-making problems, which are commonly formalized as MDP optimization \cite{Puterman_1994}. Recently RL has achieved notable landmarks in various task domains such as classical board games (chess \cite{Campbell_Hoane_Hsu_2002_DeepBlue}, Go \cite{Silver_Go}, poker \cite{Brown_Sandholm_2017_Poker}) and visually rich computer games (Atari \cite{Mnih_2015_Atari}, Dota 2 \cite{openai_2019_dota}). At the core of some of the methods used, especially for board games, were planning algorithms, which rely on the knowledge of the environment dynamics.

In general case, though, the model of the environment is unknown, and there are two common opposite approaches in RL to deal with it: a model-free and a model-based. The former approach doesn't explicitly learn and make use of the knowledge of the environment. Instead, it directly learns the global policy from interactions with the environment. While model-free approach has many successful examples (\cite{Mnih_2015_Atari,haarnoja_2018_sac,schulman_2017_ppo}), among its disadvantages are data inefficiency, as it requires high amount of interactions with the environment, and struggle for precise lookahead.

Model-based RL can address issues of both planning and model-free RL by first learning the environment dynamics, and then to exploit it for more efficient learning of the agent's policy (\cite{sutton_barto_2018}). Typically, the model is represented by MDP and consists of two components: a state transition model and a reward model. After the model has been learned, MDP planning method such as MCTS (\cite{Coulom_2007_mcts}) can be applied to infer the optimal policy. Many model-based RL approaches build a model that operates at the pixel-level (\cite{kaiser_2020_modelbased}), which is hungry for computational resources and attentive to unimportant details. Another approach is to try building a latent-space model. For example, a latent-space model can represent an abstract MDP that's equivalent to the real one in terms of state values, i.e. the cumulative reward of the planned trajectory (\cite{silver_2020_muzero}).

To model the environment it's common to use ANNs (\cite{silver_2020_muzero,Ha_Schmidhuber_2018_worldmodels}). It's also very tempting to take more inspiration from human brains, especially for how our memory functions (\cite{Hassabis_2017_neuro}). One example of such human-like memory model is Hierarchical Temporal Memory (HTM) framework (\cite{George_Hawkins_2009}). This framework utilizes discrete binary sparse distributed representations (SDR) (\cite{ahmad_2016_sdr}), which enables more efficient data processing, high noise tolerance and a natural way to set similarity metric between data embeddings. It also provides a model of the sequential temporal memory (TM) (\cite{hawkins_TM}). Learning in TM isn't based on the gradient descend, and typically it learns with a faster rate compared to common ANNs.

In our work we present a simplified memory-based framework based on the ideas of HTM that can be used to model the environment dynamics, supplied with the planning algorithm to infer the optimal global policy. We study applicability of the presented memory model and viability of the planning method in different maze environments.

\begin{figure}
  \centering
  \begin{minipage}{.49\linewidth}
    \includesvg[width=\linewidth]{gridworld_5x5_1_1_200_1337_map_0_1173222464}
  \end{minipage}
  \begin{minipage}{.49\linewidth}
    \includesvg[width=\linewidth]{gridworld_5x5_1_1_200_1337_map_2_1561234712}
  \end{minipage}
  \caption{Example of generated maze environments of 5x5 and 8x8 sizes. Walls - dark purple, starting agent's position - yellow, rewarding goal state - salad.} \label{fig_maze_examples}
\end{figure}

\section{Background}

\subsection{Reinforcement Learning}

In this paper we consider the classic decision-making problem with an agent operating in an environment, which is formalized as Markov Decision Process (MDP). As a starting point in the research we simplify the problem assuming that the environment is fully observable and deterministic. We also consider only environments with the distinctive goal-oriented tasks, where an agent is expected to reach desirable goal states in a limited time.

Therefore, MDP is defined as a tuple $M = (\St, \A, R, P, s_0, s_g)$, where $\St = \{1, \dots, s_n\}$ is the state space of the environment, $\A = \{1, \dots, a_m\}$ is the action space available to an agent, and $s_0$ and $s_g$ are the initial and the goal state distributions respectively. At each time step $t$, an agent observes a state $s_t \in \St$ and selects an action $a_t \in \A$. As a result, it receives the reward $r_t \in R(s_t, a_t)$ and transitions to the new state $s_{t+1} = P(s_t, a_t)$. An agent's policy $\pi: S \rightarrow A$ is a mapping from a state to an action. The agent's goal is to maximize the expected return $\E \left[ \sum_{t=0}^{\infty} r_t \right]$.

We don't use discount factor $\gamma$ in the return calculation. Instead, we assume that the reward function is designed the way that it highly rewards reaching the goal states and slightly punishes in any other states, which should reflect the goal-oriented nature of the environments that were taken into consideration.

\subsection{Temporal Memory}

Temporal Memory (TM) from Hierarchical Temporal Memory (HTM) framework is a model of the sequential memory. It works with sparse distributed representations (SDR), which are sparse binary vectors, typically high-dimensional. Therefore, the data encoding scheme should be chosen too.

In our work both actions and states are represented as integer numbers from a fixed range (although, different for states and actions). Thus, we use a simple encoding scheme, where numbers from some range $[0, N)$ are mapped to non-overlapping equally-sized sets of active bits of the output SDR vector. For example for $N = 3$ numbers 0, 1 and 2 are encoded correspondingly as following:

$$
\begin{aligned}
  0 & \rightarrow \verb|1111 0000 0000| \\
  1 & \rightarrow \verb|0000 1111 0000| \\
  2 & \rightarrow \verb|0000 0000 1111| \\
\end{aligned}
$$

As you see, resulting vector is split into buckets of bits, and the size of buckets is a hyperparameter (in the example above it's 4).

For SDR vectors a union operation is defined as a bitwise OR applied to the corresponding vectors. Also we define a similarity metric, which is the number of the same one-bits of the vectors in consideration.

Temporal Memory model not only provides an algorithm to learn sequences of SDR vectors $[v_0, v_1, v_2, \dots ]$ but also to make predictions about what SDR vector $v'$ comes next based on the past experience. The prediction is represented also as an SDR vector, which is a union of SDR vectors $v' = \cup v_{o_i}$, where each vector $v_{o_i}$ represents a single separate expected outcome. Notice that as the union of SDR vectors is also an SDR vector, TM model allows you to work simultaneously with the unions of sequences as well.

\section{Planning with Temporal Memory}

In our method we train an agent to reach rewarding goal states $s_g$. An agent is supplied with the memory module to learn the environment's dynamics and a planning algorithm which heavily relies on the learned model. The planning algorithm constructs a plan of actions $p = [a_t, a_{t+1}, \dots, a_{t+h-1}]$ that leads to the desired state $s_g = s_{t+h}$ from the current state $s_t$. In the case when the planning algorithm fails to provide a plan we switch the agent to the exploration strategy $\pi_r$ in order to improve the learned model. As the result, at each timestep an agent operates either according to a constructed plan $p$ or according to the random uniformly-distributed exploration policy $\pi_r$.

\subsection{Memory module}

The agent's memory module consists of two submodules. One submodule, called \textit{transition memory}, learns the transition model $f: \St \times \A \rightarrow \St$ of the environment, while the other, called \textit{goal states memory} keeps track of the rewarding goal states $s_g$.

Consider an agent's trajectory $\tau = [s_0, a_0, r_1, s_1, a_1, r_2, \dots, r_{T-1}, s_T]$. Each pair $(s_t, a_t)$ at the moment $t$ defines the precondition to the transition in the environment $(s_t, a_t) \rightarrow s_{t+1}$. Therefore, we made transition memory to learn the sequences of such transitions $Tr = [(s_0, a_0), (s_1, a_1), \dots]$ from the trajectories that the agent produces. The learning is performed in an online fashion, i.e. at each timestep.

The transition memory submodule is based on a Temporal Memory.Recall that TM works with sparse distributed representations (SDR). To encode each state-action pair $(s, a)$ we use two separate integer encoders for states and actions respectively to obtain binary sparse vectors. The states encoder maps integer numbers from $[0, |\St|)$ range, and the actions encoder maps integer numbers from $[0, |\A|)$ range. As most of the time our method works with SDR, for simplicity further on we use the same notation $s$ and $a$ for their corresponding SDR too. To work with pairs $(s, a)$ we just concatenate their SDR vectors together into one.

For any given state-action pair $(s, a)$ the transition memory module can give a prediction on what state $s'$ is coming next.
Precisely, it allows prediction queries of the form:

\begin{quote}
  \textit{given a superposition \footnotemark of states $S = \cup s_i$ and a superposition of actions $A = \cup a_j$, what superposition of next states $S'= \cup s'_m$ all possible state-action pairs $(s_i, a_j)$ lead to?}
\end{quote}

\footnotetext{TM can work with unions of SDR vectors. When SDR vectors in a union are semantically meaningful, e.g. when they represent states or actions, we call such unions \textit{superpositions} of states or actions respectively \footnotemark.}

\footnotetext{As you see, in our current encoding scheme the different states or actions are encoded with non-overlapping bits which makes them distinguishable from each other in a superposition. Whilst it's true, the current planning algorithm only requires a separation of the state bits from the action bits in a state-action pair SDR. This is done intentionally. Non-overlapping scheme makes things simpler to read and debug, but cannot keep the similarity between states (or observations). For the future work we plan to experiment with more complex environments and therefore to switch to overlapping mappings.}

In some specific circumstances the transition memory submodule can be used backward - as a reverse of an environment transition function $f$. This feature becomes available after getting the result to a prediction query and only for the states $s'_m$ that match the result $S' = \cup s'_m$. For each such state $s'$ it can provide which state-action pairs from the original query lead to $s'$: $\{(s, a)\} = f^{-1}(s')$. Notice that in general case it's possible that \textit{a multiple} state-action pairs lead to $s'$.

\subsection{Planning algorithm}

Planning process is split into two stages (see function ~\ref{func_PlanActions}):

\begin{enumerate}
  \item Forward planning from the current state until reward is reached.
  \item Deducing the backward path from the reward state to the current.
\end{enumerate}

\begin{function}
  \caption{PlanActions($s_0$)}\label{func_PlanActions}
  \SetAlgoNoLine
  $s_0$: initial state \;
  $g: $ reached goal \;
  $H: $ Temporal Memory active segments history \;
  \quad \\

  $g, H \leftarrow PredictToGoal(s_0)$ \;
  $plan \leftarrow BacktrackFromGoal(g, s_0, H)$ \;

  \Return{$[a_0, a_1, .. a_T] = plan$}
\end{function}

\subsubsection{Forward planning}

Agent uses its experience and expects the reward to be in one of the previous rewarding states. But even if it is wrong, we expect that direct goal-based exploration is more optimal than random walking. Hence it keeps the goal list and gives such goal states a priority.

Consider that at some timestep $t_0$ the agent is in state $s_0$. To find out if there is a reachable goal state in a radius of $N$ actions, it start planning from the superposition $(S_0 = s_0, A = \cup a^i)$, where $A$ is a union of all available actions (we use superscripts for indices here as subscripts are used to denote timestep). It is similar to making all possible actions simultaneously in parallel. To achieve this planner uses the memory and gets a superposition of next state-action pairs.

After the first planning step the planner has a superposition of the next states $S_1 = \cup s^i$, reachable in one agent's action. To get all reachable states in two agent's action, planner makes another prediction step - now from the superposition $(S_1, A)$, and so on (see function ~\ref{func_PredictToGoal}).

\begin{function}
  \caption{PredictToGoal($s_0$)} \label{func_PredictToGoal}
  \SetNoFillComment
  \KwData{$s_0$ - initial state}

  $sa_0 \leftarrow Encode((s_0, \cup a^i))$ \tcp*{Init starting superposition}
  $H \leftarrow []$ \tcp*{Init active segments history}

  \ForEach {timestep $i$} {
    $memory.ActivatePredict(sa_i)$ \;
    \quad \\

    $H[i] \leftarrow memory.active\_segments$ \;
    $sa_{i+1} \leftarrow PredictionFromSegments(H[i])$ \;
    \quad \\

    $s, a \leftarrow Decode(sa_{i+1})$ \tcp*{Both $s$ and $a$ are superpositions!}
    $reached\_goal \leftarrow MatchGoals(s)$ \;
    \If(){$reached\_goal \neq \emptyset$} {
      \Return {$reached\_goal, H$} \;
    }{}
  }
  \Return{$\emptyset$}
\end{function}

This process is continued until either some state from the goal list is reached, or planner exceeds its planning horizon limit. In the former case planner proceeds to the next state of planning process. In the latter, agent uses its random strategy to make a move before it tries to plan again. To test if the goal state is in superposition we intersect their sparse vectors and sum active bits, which is the same as dot product of such vectors. If they match is over some predefined threshold, the state is considered presented in this superposition.

During this stage planner collects a history of active superpositions $H = [S_0, S_1, \dots S_T]$. In more details, he keeps the track of Temporal Memory cells' segments activations. Each segment is a pair $(p, \{c_i\})$, where $p$ - index of predicted cell and $\{c_i\}$ is a set of cell indices that induce prediction of $p$. This information allows planner to use inversed environment transition function to get which $\{(s, a)\}$ has lead to a state $s'$ by tracking which cells predict $s'$ cells.

It's possible to find multiple reachable goals at the end of this stage. In our current implementation we use the first matched from the goal list.

\subsubsection{Backtracking}

After the first stage planner ``believes'' that some goal state $g = s_T$ can be reached in $T$ steps. The problem is that he doesn't know how to get there as until now he worked with superpositions. So the main goal of this stage is to restore the sequence of actions leading from $s_0$ to $s_T$. Further on this sequence of action is meant to be the agent's plan in environment.

Planner solves the problem from the end. It moves backward in the history of segments activations graph, trying to get to the starting state $s_0$ at the first timestep. Due to its ``backward moving'' nature we call this stage \textit{backtracking} (see function ~\ref{func_BacktrackFromGoal}).

\begin{function}
  \caption{BacktrackFromGoal($g$, $H$)} \label{func_BacktrackFromGoal}
  % \SetAlgoNoLine
  \KwData{$g = s_T$ - goal state; $H$ - active segments history}

  $T \leftarrow |H| + 1$ \tcp*{As the last recorded transition is for $T-1 \rightarrow T$}
  $S_T \leftarrow PredictionFromSegments(H[T-1])$ \tcp*{State superposition at $T$}
  $s_T \leftarrow s_T \cap S_T$ \tcp*{Keeps only matched bits}
  \quad \\
  $is\_successful, actions \leftarrow Backtrack(s_T, T-1, H)$ \;
  \Return {$is\_successful, actions$}
\end{function}

Recall that planner keeps track of all state superpositions $H = [S_0, S_1, \dots S_T]$ it has reached on each step $t$, and for any state $s'$ from each of these superpositions $S_i$ the memory can answer which state-action pairs has lead to it.

Therefore, given the reached goal state $g = s_T$ planner can get which state-action pair $(s_{T-1}, a_{T-1})$ has lead to it. Hence he knows for $T-1$ step which action is required to get from $s_{T-1}$ to the goal state $s_T$. And also he can recursively reduce the $s_0 \leadsto s_T$ pathfinding problem to the  $s_0 \leadsto s_{T-1}$. Planner repeats this procedure and for each timestep $t$ sequentially finds state-action pairs $(s_{t-1}, a_{t-1}) = f^{-1}(s_t)$ until he returns to the initial state at the first timestep (see ~\ref{func_Backtrack} function).

In general case $f^{-1}(s_{t+1})$ gives multiple pairs $(s_t, a_t)$, and each leads to $s_{t+1}$. But not every pair is actually reachable in exactly $t$ steps from the $s_0$, i.e. not from every pair you can backtrack to the initial state at time $0$. Hence planner sequentially tests each pair until he find the first successful.

As the result of the backtracking stage planner forms a plan of actions $plan = [a_0, a_1, \dots a_{T-1}]$ which he ``believes'' [in terms of his approximation of the environment dynamics] will lead the agent from the initial state $s_0$ to the target state $g = s_T$.

\begin{function}
  \caption{Backtrack($p_t$, $t$, $H$)} \label{func_Backtrack}
  \KwData{$p_t$ - indices of required predicted cells at time $t$, $t$ - current timestep, $H$ - history of active segments}

  \If(){$t \leq 0$} {
    \Return{True, $[]$}
  }{}

  $P \leftarrow H[t][p_t]$ \tcp*{List of cell sets, each predicts one of $p_t$ cells}
  $C \leftarrow P$ \tcp*{Initializes cell clusters - candidates for backtracking}
  \quad \\

  \tcp{Iteratively merges any two sufficiently matching cell clusters}
  \While{$\exists c_i, c_j \in C: match(c_i, c_j) \geq \Theta$} {
    $C[c_i] = c_i \cup c_j$ \;
    $C[c_j].remove()$ \;
  }
  \quad \\

  \ForEach{$c \in C$}{
    \tcp{Filters by the amount of $p_t$ cells cluster $c$ predicts}
    $p = PredictionFromCluster(c, H[t])$ \;
    \If(){$match(p, p_t) \le \Omega$}{
      \textbf{continue}
    }{}
    \quad \\

    \tcp{Recursively checks candidate cluster $c$}
    $s_t, a_t \leftarrow Decode(ColumnsFromCells(c))$ \;

    $is\_successful, actions \leftarrow Backtrack(s_t, t-1, H)$ \;
    \If() {is\_successful} {
      $actions.append(a_t)$ \;
      \Return{True, actions}
    }
  }

  \Return{False, $[]$}
\end{function}

After successful backtracking the planner has a policy for the next $T$ steps, which an agent uses. And if it ends up in a rewarding state, an episode finishes. Otherwise it continues and the planner removes that state from its goal list for the rest of the episode.

\section{Experiments}

In our experiments we decided to use classic GridWorld environments, which are represented as mazes on a square grid (see Figure ~\ref{fig_maze_examples}). Each state can be defined with the position of an agent, i.e. in which cell it stands. Thus, state space $\St$ consists of all possible agent's positions. We enumerate all positions (and therefore states) with integer numbers. An agent starts from the fixed state $s_0$, and its goal is to find a single reward placed at some fixed position $s_g$. Each environments has deterministic transition function $P: \St \times \A \rightarrow \St$. The action space consists of 4 actions $\A = \{ 0, 1, 2, 3 \}$ defining agent's move directions: 0 - east, 1 - north, 2 - west, 3 - south. Each action moves an agent to the adjacent grid cell in the corresponding direction, except when the agent tries to move into the maze wall - in this case it stays in the current cell.

When an agent reaches the goal state $s_g$ it is rewarded with high positive value. Any other steps it is slightly punished. In our experiments we used $+1.0$ for the goal reward and $-0.01$ for the punishment. In each testing environment the optimal path from the starting point to the goal was less than 20 steps. Such reward function makes possible easy episode outcomes differentiation, depending on whether or not the goal state has been reached during it, and if the goal has been reached then how fast.

All experiments were divided into two groups. Each experiment from the first group was held in a fixed environment, while each experiment from the second group had an environment being changed every $N_{ep}$ episodes.

In our experiments we tested ability to model the environment and utilize learned model to plan actions that lead to the goal states. We compared our method's performance with the baselines: random strategy and DQN (\cite{Mnih_2015_Atari}). We also compared performance for the different planning horizons and the number of simultaneously tracked goals in the goal tracking memory.

\subsection{Experiments in fixed environments}

For this experiment setup we used a number of handcrafted mazes. Each experiment had a fixed maze and the starting point $s_0$, but the place of the reward was being changed every $N$ episodes. We tested an agent's ability to find the goal state and how it adapts to the new goal position.

We present results for two mazes: \texttt{multi\_way\_v0} and \texttt{multi\_way\_v2} (see Figure ~\ref{fig_comparison_mw} and ~\ref{fig_detailed_mw0}). Each experiment in \texttt{multi\_way\_v0} had 4 different rewarding goal positions being sequentially changed every 100 episodes. Each experiment in \texttt{multi\_way\_v2} had 6 rewarding goal positions being changed every 200 episodes. Schematic view of these mazes is provided in Appendix ~\ref{apx_handcrafted_mazes}.

\begin{figure}
  \centering
  \begin{minipage}{.49\linewidth}
    \includesvg[width=\linewidth]{multi_way_v0__steps}
    \subcaption{\texttt{multi\_way\_v0}}
  \end{minipage}
  \begin{minipage}{.49\linewidth}
    \includesvg[width=\linewidth]{multi_way_v2__steps}
    \subcaption{\texttt{multi\_way\_v2}}
  \end{minipage}
  \caption{Comparison of the episode duration (timesteps) for the handcrafted mazes. \texttt{htm\_X\_Yg} is our agent with the planning horizon X and Y tracked last goals, \texttt{htm\_0} - random policy, \texttt{dqn\_greedy} - greedy DQN agent. Results are plotted with the moving average MA = 20 steps.} \label{fig_comparison_mw}
\end{figure}

\begin{figure}
  \centering
  \begin{minipage}{.49\linewidth}
    \includesvg[width=\linewidth]{multi_way_v0__dqn__steps}
    \subcaption{Greedy and $\epsilon$-greedy DQN agent}
  \end{minipage}
  \begin{minipage}{.49\linewidth}
    \includesvg[width=\linewidth]{multi_way_v0__1__steps}
    \subcaption{Planning horizon H = 1}
  \end{minipage}

  \begin{minipage}{.49\linewidth}
    \includesvg[width=\linewidth]{multi_way_v0__2__steps}
    \subcaption{Planning horizon H = 2}
  \end{minipage}
  \begin{minipage}{.49\linewidth}
    \includesvg[width=\linewidth]{multi_way_v0__4__steps}
    \subcaption{Planning horizon H = 4}
  \end{minipage}

  \caption{Detailed results for \texttt{multi\_way\_v0}. a)Comparison with DQN: blue - greedy policy, yellow - $\epsilon$-greedy policy, green - our method with planning horizon 4 and the last 1 goal tracked; b-d) comparison between tracking all goal states and only the last goal for the different planning horizon. Results are plotted with the moving average MA = 20 steps.} \label{fig_detailed_mw0}
\end{figure}

In the tested environments our method performed better than random policy. But the more complicated an environment was the less distinctive was the difference between the $H = 0$ and $H = 1$. Our method performs better with the increase of the planning horizon $H$. Each increment of the planning horizon is cumulative, i.e. each next increment is more advantageous.

Regarding the goal tracking memory size, we found that increase of the size leads to better exploration. We think that's because an agent less frequently uses random strategy and therefore moves less chaotically. However, this behavior may be suboptimal. For example when the rewarding goal state is reachable from the current position in less than $H$ actions but there's a fake goal state in the opposite direction that is closer to the agent. We think that these results are highly biased due to the special type of handcrafted mazes and initial conditions.

We also found that an agent learns and adapts faster than DQN agent (in terms of the number of episodes required to steadily reach the goal during the episode). And longer planning horizon leads to a better adaptation to the changed rewarding position.

When the planning horizon is big enough to reach the goal state from the starting position, then the agent learns the optimal policy very fast. According to our experiments, in this case it learns $\sim$1.5-2.5 times faster than DQN. But for any fixed planning horizon if we start rising an environment complexity, then at some point DQN starts to perform better than our method. That's because DQN always finds an optimal path, and our agent finds it only when it has enough planning horizon.

\subsection{Experiments in changing environments}

In this experiment setup we used a number of randomly generated GridWorld mazes of fixed size (examples provided in Figure ~\ref{fig_maze_examples}). Each experiment has the following scheme. It has been made sequentially in $N_{env}$ randomly generated mazes. For a fixed maze the rewarding goal position was changed sequentially $N_{rew}$ times. For a fixed rewarding goal position the initial position was changed sequentially $N_{s_0}$ times. And for a whole configuration being fixed agent played $N$ episodes. So, each experiment had $N_{env} \cdot N_{rew} \cdot N_{s_0} \cdot N$ total episodes. We tested an agent's ability not only to find the rewarding goal and adapt to its new position, but also how the agent adapts to the whole new maze.

We tested agent in mazes generated on $5 \times 5$, $6 \times 6$ and $8 \times 8$ square grids. In this section we present only the results for the $5 \times 5$ mazes, but for the other tested sizes the results are similar. The key difference in complexity of each experimental setup is the total number of the episodes with fixed reward and maze. The less episodes has agent to adapt, the more agile it's required to be.

We present results for the following experimental setups (Figure 4):

\begin{enumerate}
  \item Experimental setup with the \textbf{rare} change of reward and maze: $N_{s_0} = 100$, $N_{rew} = 2$, $N_{env} = 8$.

  \item Experimental setup with the \textbf{frequent} change of reward and maze: $N_{s_0} = 20$, $N_{rew} = 1$, $N_{env} = 20$.
\end{enumerate}

\begin{figure}
  \centering
  \begin{minipage}{.49\linewidth}
    \includesvg[width=\linewidth]{gridworld_5x5_100_2_8_1337__steps}
    \subcaption{Comparison with baselines}
    \includesvg[width=\linewidth]{gridworld_5x5_100_2_8_1337__dqn__steps}
    \subcaption{Greedy and $\epsilon$-greedy DQN agent}
    \includesvg[width=\linewidth]{gridworld_5x5_100_2_8_1337__2__steps}
    \subcaption{Planning horizon: 2}
    \includesvg[width=\linewidth]{gridworld_5x5_100_2_8_1337__4-8__steps}
    \subcaption{Planning horizon: 4-8}
  \end{minipage}
  \begin{minipage}{.49\linewidth}
    \includesvg[width=\linewidth]{gridworld_5x5_20_1_20_1337__steps}
    \subcaption{Comparison with baselines}
    \includesvg[width=\linewidth]{gridworld_5x5_20_1_20_1337__dqn__steps}
    \subcaption{Greedy and $\epsilon$-greedy DQN agent}
    \includesvg[width=\linewidth]{gridworld_5x5_20_1_20_1337__2__steps}
    \subcaption{Planning horizon: 2}
    \includesvg[width=\linewidth]{gridworld_5x5_20_1_20_1337__4-8__steps}
    \subcaption{Planning horizon: 4-8}
  \end{minipage}

  \caption{Results for experiments in changing environments. Left row: results for the setup with rare changes; right row: results for the setup with frequent changes.}
\end{figure}

From this set of experiments we conclude that in average our agent adapts to changes faster than DQN. Which also makes it perform more stable.

In contradiction to our results for the experiments in the fixed handcrafted environments we found that the increase of the number of the tracked goals makes our method to perform worse.

\section{Conclusion}

We introduced a novel memory-based method that combines learning the model of the environment with the planning algorithm to infer a local policy. We showed that under certain conditions (enough planning horizon) the resulting global policy is optimal or at least very close to it. We also showed that in such cases, compared to DQN, our method learns with the comparable speed or slightly faster. The planning algorithm with any non-zero planning horizon performs better than random strategy. However, for small planning horizons and hard tasks results are sufficiently degraded and not that far from the pure random strategy.

To model the environment dynamics we used human-like model of the memory called HTM. We showed that it's capable to quickly and reliably learn the transition model of the visited states in just a few episodes. As far as we know it is the first time HTM model was used for the model-based RL method. So, even though we applied it to model very simple deterministic environments with perfectly distinguishable states the results are promising, and we look forward to adapt our method to more complex environments.

\bibliography{citations}

\appendix

\section{Appendix}
\subsection{Handcrafted mazes} \label{apx_handcrafted_mazes}

The notation we use in the schematic views of the environments: \verb|-| - empty cell, \verb|#| - wall, \verb|@| - agent, \verb|X| - reward.

\subsubsection{multi\_way\_v0}

Each experiment had 4 different rewarding places being sequentially changed every 100 episodes:

\begin{verbatim}
  #####   #####   #####   #####
  #@--#   #@-X#   #@--#   #@--#
  #-#-#   #-#-#   #-#-#   #-#X#
  #--X#   #---#   #X--#   #---#
  #####   #####   #####   #####
\end{verbatim}

\subsubsection{multi\_way\_v2}

Each experiment had 6 rewarding places being changed every 200 episodes:

\begin{verbatim}
  #########  #########  #########  #########  #########  #########
  #---#####  #-X-#####  #---#####  #---#####  #---#####  #---#####
  #-#-#####  #-#-#####  #-#-#####  #-#-#####  #-#-#####  #-#-#####
  #--@---##  #--@---##  #--@--X##  #--@---##  #X-@---##  #--@---##
  ###--####  ###--####  ###--####  ###--####  ###--####  ###--####
  ###-#--X#  ###-#---#  ###-#---#  ###-#---#  ###-#---#  ###-#--X#
  ###---###  ###---###  ###---###  ###---###  ###---###  ###---###
  ###-#####  ###-#####  ###-#####  ###X#####  ###-#####  ###-#####
  #########  #########  #########  #########  #########  #########
\end{verbatim}

\end{document}

### 2020-10-29 reports/v1_0/paper_eng/paper_eng.tex
fix citations

@@
 \begin{abstract}
-  Among the key problems in artificial intelligence is sequential decision making, which can be formalized as Markov Decision Process (MDP).
+  Sequential decision making is among the key problems in Artificial Intelligence.
+It can be formalized as Markov Decision Process (MDP).
 One approach to solve it, called model-based Reinforcement Learning (RL), combines learning the model of the environment and the global policy.
@@
 
-Model-based RL can address issues of both planning and model-free RL by first learning the environment dynamics, and then to exploit it for more efficient learning of the agent's policy (\cite{sutton_barto_2018}).
+Model-based RL can address issues of both planning and model-free RL by first learning the environment dynamics, and then to exploit it for more efficient learning of the agent's policy (\cite{moerland_2020_modelbased}).
 Typically, the model is represented by MDP and consists of two components: a state transition model and a reward model.
@@
 One example of such human-like memory model is Hierarchical Temporal Memory (HTM) framework (\cite{George_Hawkins_2009}).
-This framework utilizes discrete binary sparse distributed representations (SDR) (\cite{ahmad_2016_sdr}), which enables more efficient data processing, high noise tolerance and a natural way to set similarity metric between data embeddings.
+This framework utilizes discrete binary sparse distributed representations (SDR) (\cite{Cui_Ahmad_Hawkins_2017_sdr}), which enables more efficient data processing, high noise tolerance and a natural way to set similarity metric between data embeddings.
 It also provides a model of the sequential temporal memory (TM) (\cite{hawkins_TM}).
@@
 As the result, at each timestep an agent operates either according to a constructed plan $p$ or according to the random uniformly-distributed exploration policy $\pi_r$.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=0.8\linewidth]{agent_scheme}
+  \caption{Schematic view of our agent and its training loop.} \label{fig_agent_scheme}
+\end{figure}
 

### 2020-10-29 reports/v1_0/paper_eng/Example.tex
WIP: add new paper template, not adapted yet

\documentclass[a4paper,twoside]{article}

\usepackage{epsfig}
\usepackage{subcaption}
\usepackage{calc}
\usepackage{amssymb}
\usepackage{amstext}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{multicol}
\usepackage{pslatex}
\usepackage{apalike}
\usepackage{SCITEPRESS}     % Please add other packages that you may need BEFORE the SCITEPRESS.sty package.

\begin{document}

\title{Authors' Instructions: Preparation of Camera-Ready Contributions to SCITEPRESS Proceedings}

\author{\authorname{First Author Name\sup{1}\orcidAuthor{0000-0000-0000-0000}, Second Author Name\sup{1}\orcidAuthor{0000-0000-0000-0000} and Third Author Name\sup{2}\orcidAuthor{0000-0000-0000-0000}}
\affiliation{\sup{1}Institute of Problem Solving, XYZ University, My Street, MyTown, MyCountry}
\affiliation{\sup{2}Department of Computing, Main University, MySecondTown, MyCountry}
\email{\{f\_author, s\_author\}@ips.xyz.edu, t\_author@dc.mu.edu}
}

\keywords{The paper must have at least one keyword. The text must be set to 9-point font size and without the use of bold or italic font style. For more than one keyword, please use a comma as a separator. Keywords must be titlecased.}

\abstract{The abstract should summarize the contents of the paper and should contain at least 70 and at most 200 words. The text must be set to 9-point font size.}

\onecolumn \maketitle \normalsize \setcounter{footnote}{0} \vfill

\section{\uppercase{Introduction}}
\label{sec:introduction}

\noindent Your paper will be part of the conference proceedings
therefore we ask that authors follow the guidelines explained in
this example in order to achieve the highest quality possible
\cite{Smith98}.

Be advised that papers in a technically unsuitable form will be
returned for retyping. After returned the manuscript must be
appropriately modified.

\section{\uppercase{Manuscript Preparation}}

\noindent We strongly encourage authors to use this document for the
preparation of the camera-ready. Please follow the instructions
closely in order to make the volume look as uniform as possible
\cite{Moore99}.

Please remember that all the papers must be in English and without
orthographic errors.

Do not add any text to the headers (do not set running heads) and
footers, not even page numbers, because text will be added
electronically.

For a best viewing experience the used font must be Times New
Roman, except on special occasions, such as program code
\ref{subsubsec:program_code}.

\vfill
\subsection{Manuscript Setup}

\noindent The template is composed by a set of 7 files, in the
following 2 groups:\\
\noindent {\bf Group 1.} To format your paper you will need to copy
into your working directory, but NOT edit, the following 4 files:
\begin{verbatim}
  - apalike.bst
  - apalike.sty
  - article.cls
  - scitepress.sty
\end{verbatim}

\noindent {\bf Group 2.} Additionally, you may wish to copy and edit
the following 3 example files:
\begin{verbatim}
  - example.bib
  - example.tex
  - scitepress.eps
\end{verbatim}

\subsection{Page Setup}

The paper size must be set to A4 (210x297 mm). The document
margins must be the following:

\begin{itemize}
    \item Top: 3,3 cm;
    \item Bottom: 4,2 cm;
    \item Left: 2,6 cm;
    \item Right: 2,6 cm.
\end{itemize}

It is advisable to keep all the given values because any text or
material outside the aforementioned margins will not be printed.

\vfill
\subsection{First Section}

This section must be in one column.

\subsubsection{Title and Subtitle}

Use the command \textit{$\backslash$title} and follow the given structure in "example.tex". The title and subtitle must be with initial letters
capitalized (titlecased). The separation between the title and subtitle is done by adding a colon ":" just before the subtitle beginning. In the title or subtitle, words like "is", "or", "then", etc. should not be capitalized unless they are the first word of the title or subtitle. No formulas or special characters of any form or language are allowed in the title or subtitle.

\subsubsection{Authors and Affiliations}

Use the command \textit{$\backslash$author} and follow the given structure in "example.tex". Please note that the name of each author must start with its first name.

\subsubsection{Keywords}

Use the command \textit{$\backslash$keywords} and follow the given structure in "example.tex". Each paper must have at least one keyword. If more than one is specified, please use a comma as a separator. The sentence must end with a period.

\subsubsection{Abstract}

Use the command \textit{$\backslash$abstract} and follow the given structure in "example.tex".
Each paper must have an abstract up to 200 words. The sentence
must end with a period.

\subsection{Second Section}

Files "example.tex" and "example.bib" show how to create a paper
with a corresponding list of references.

This section must be in two columns.

Each column must be 7,5-centimeter wide with a column spacing
of 0,8-centimeter.

The section text must be set to 10-point.

Section, subsection and sub-subsection first paragraph should not
have the first line indent.

To remove the paragraph indentation (only necessary for the
sections), use the command \textit{$\backslash$noindent} before the
paragraph first word.

If you use other style files (.sty) you MUST include them in the
final manuscript zip file.

\vfill
\subsubsection{Section Titles}

The heading of a section title should be in all-capitals.

Example: \textit{$\backslash$section\{FIRST TITLE\}}

\subsubsection{Subsection Titles}

The heading of a subsection title must be with initial letters
capitalized (titlecased).

Words like "is", "or", "then", etc. should not be capitalized unless
they are the first word of the subsection title.

Example: \textit{$\backslash$subsection\{First Subtitle\}}

\subsubsection{Sub-Subsection Titles}

The heading of a sub subsection title should be with initial letters
capitalized (titlecased).

Words like "is", "or", "then", etc should not be capitalized unless
they are the first word of the sub subsection title.

Example: \textit{$\backslash$subsubsection\{First Subsubtitle\}}

\subsubsection{Tables}

Tables must appear inside the designated margins or they may span
the two columns.

Tables in two columns must be positioned at the top or bottom of the
page within the given margins. To span a table in two columns please add an asterisk (*) to the table \textit{begin} and \textit{end} command.

Example: \textit{$\backslash$begin\{table*\}}

\hspace*{1.5cm}\textit{$\backslash$end\{table*\}}\\

Tables should be centered and should always have a caption
positioned above it. The font size to use is 9-point. No bold or
italic font style should be used.

The final sentence of a caption should end with a period.

\begin{table}[h]
\caption{This caption has one line so it is
centered.}\label{tab:example1} \centering
\begin{tabular}{|c|c|}
  \hline
  Example column 1 & Example column 2 \\
  \hline
  Example text 1 & Example text 2 \\
  \hline
\end{tabular}
\end{table}

\begin{table}[h]
\vspace{-0.2cm}
\caption{This caption has more than one line so it has to be
justified.}\label{tab:example2} \centering
\begin{tabular}{|c|c|}
  \hline
  Example column 1 & Example column 2 \\
  \hline
  Example text 1 & Example text 2 \\
  \hline
\end{tabular}
\end{table}

Please note that the word "Table" is spelled out.

\vfill
\subsubsection{Figures}

Please produce your figures electronically, and integrate them into
your document and zip file.

Check that in line drawings, lines are not interrupted and have a
constant width. Grids and details within the figures must be clearly
readable and may not be written one on top of the other.

Figure resolution should be at least 300 dpi.

Figures must appear inside the designated margins or they may span
the two columns.

Figures in two columns must be positioned at the top or bottom of
the page within the given margins. To span a figure in two columns please add an asterisk (*) to the figure \textit{begin} and \textit{end} command.

Example: \textit{$\backslash$begin\{figure*\}}

\hspace*{1.5cm}\textit{$\backslash$end\{figure*\}}

Figures should be centered and should always have a caption
positioned under it. The font size to use is 9-point. No bold or
italic font style should be used.

\begin{figure}[!h]
  \centering
   {\epsfig{file = SCITEPRESS.eps, width = 5.5cm}}
  \caption{This caption has one line so it is centered.}
  \label{fig:example1}
 \end{figure}

\begin{figure}[!h]
  \vspace{-0.2cm}
  \centering
   {\epsfig{file = SCITEPRESS.eps, width = 5.5cm}}
  \caption{This caption has more than one line so it has to be justified.}
  \label{fig:example2}
  \vspace{-0.1cm}
\end{figure}

The final sentence of a caption should end with a period.

Please note that the word "Figure" is spelled out.

\subsubsection{Equations}

Equations should be placed on a separate line, numbered and
centered.\\The numbers accorded to equations should appear in
consecutive order inside each section or within the contribution,
with the number enclosed in brackets and justified to the right,
starting with the number 1.

Example:

\begin{equation}\label{eq1}
    a=b+c
\end{equation}

\subsubsection{Algorithms and Listings}

Algorithms and Listings captions should have font size 9-point, no bold or
italic font style should be used and the final sentence of a caption should end with a period.
Captions with one line should be centered and if it has more than one line it should be set to justified.

\vfill
\subsubsection{Program Code}\label{subsubsec:program_code}

Program listing or program commands in text should be set in
typewriter form such as Courier New.

Example of a Computer Program in Pascal:

\begin{small}
\begin{verbatim}
 Begin
     Writeln('Hello World!!');
 End.
\end{verbatim}
\end{small}

The text must be aligned to the left and in 9-point type.

\subsubsection{Reference Text and Citations}

References and citations should follow the APA (Author, date)
System Convention (see the References section in the compiled
manuscript). As example you may consider the citation
\cite{Smith98}. Besides that, all references should be cited in the
text. No numbers with or without brackets should be used to list the
references.

References should be set to 9-point. Citations should be 10-point
font size.

You may check the structure of "example.bib" before constructing the
references.

For more instructions about the references and citations usage
please see the appropriate link at the conference website.

\section{\uppercase{Copyright Form}}

\noindent For the mutual benefit and protection of Authors and
Publishers, it is necessary that Authors provide formal written
Consent to Publish and Transfer of Copyright before publication of
the Book. The signed Consent ensures that the publisher has the
Author's authorization to publish the Contribution.

The copyright form is located on the authors' reserved area.

The form should be completed and signed by one author on
behalf of all the other authors.

\section{\uppercase{Conclusions}}
\label{sec:conclusion}

\noindent Please note that ONLY the files required to compile your paper should be submitted. Previous versions or examples MUST be removed from the compilation directory before submission.

We hope you find the information in this template useful in the preparation of your submission.

\vfill
\section*{\uppercase{Acknowledgements}}

\noindent If any, should be placed before the references section
without numbering. To do so please use the following command:
\textit{$\backslash$section*\{ACKNOWLEDGEMENTS\}}

\bibliographystyle{apalike}
{\small
\bibliography{example}}

\section*{\uppercase{Appendix}}

\noindent If any, the appendix should appear directly after the
references without numbering, and not on a new page. To do so please use the following command:
\textit{$\backslash$section*\{APPENDIX\}}

\end{document}

### 2020-10-29 reports/v1_0/paper_eng/paper_eng.tex
adapt new paper template, fix planning section

@@
-\documentclass[runningheads]{llncs}
+\documentclass[a4paper,twoside]{article}
+
+\usepackage{epsfig}
+\usepackage{subcaption}
+\usepackage{calc}
+\usepackage{amssymb}
+\usepackage{amstext}
+\usepackage{amsmath}
+\usepackage{amsthm}
+\usepackage{multicol}
+\usepackage{pslatex}
+\usepackage{apalike}
+\usepackage{epspdfconversion}
+\usepackage{mathptmx}
+
 \usepackage{style_eng}
-\bibliographystyle{splncs04}
+
+\usepackage{SCITEPRESS}     % Please add other packages that you may need BEFORE the SCITEPRESS.sty package.
+
+\bibliographystyle{apalike}
 
@@
 
-\title {Planning with Hierarchical Temporal Memory for deterministic Markov Decision Problem}
-\author {Petr Kuderov \inst{1} \and Aleksandr I. Panov \inst{1}}
+\title {Planning with Hierarchical Temporal Memory for Deterministic Markov Decision Problem}
 
-\authorrunning{P. Kuderov, A. Panov}
-% First names are abbreviated in the running head.
-% If there are more than two authors, 'et al.' is used.
-%
-\institute{Moscow Institute of Physics and Technology, Russia}
+\author{
+  \authorname{
+    Petr Kuderov\sup{1}\orcidAuthor{0000-0001-7055-3426}, Aleksandr I. Panov\sup{1}\orcidAuthor{0000-0002-9747-3837}
+  }
+  \affiliation{
+    \sup{1}Moscow Institute of Physics and Technology, Russia
+  }
+  \email{
+    kuderov.pv@phystech.edu, panov.ai@mipt.ru
+  }
+}
 
-\maketitle
+\keywords{Model-based reinforcement learning, Markov Decision Process, planning, Hierarchical Temporal Memory, Sparse Distributed Representation.}
 
-\begin{abstract}
+\abstract{
 Sequential decision making is among the key problems in Artificial Intelligence.
@@
 Having a good model of the environment opens up such properties as data efficiency and targeted exploration.
-
 While most of the memory-based approaches are based on using Artificial Neural Networks (ANNs), in our work we instead draw the ideas from Hierarchical Temporary Memory (HTM) framework, which is based on human-like memory model.
 We utilize it to build an agent's memory that learns the environment dynamics.
-We also accompany it with an example of planning algorithm, that enables the agent to solve RL tasks.
-\end{abstract}
+We also accompany it with an example of planning algorithm, that enables the agent to solve RL tasks.}
 
-\keywords{Model-based reinforcement learning \and Markov Decision Process \and planning \and Hierarchical Temporal Memory \and Sparse Distributed Representation.}
+\onecolumn \maketitle \normalsize \setcounter{footnote}{0}
+\vfill
 
-\section{Introduction}
+\section{\uppercase{Introduction}}
 
@@
 It requires from such methods to learn a policy for sequential decision-making problems, which are commonly formalized as MDP optimization \cite{Puterman_1994}.
-Recently RL has achieved notable landmarks in various task domains such as classical board games (chess \cite{Campbell_Hoane_Hsu_2002_DeepBlue}, Go \cite{Silver_Go}, poker \cite{Brown_Sandholm_2017_Poker}) and visually rich computer games (Atari \cite{Mnih_2015_Atari}, Dota 2 \cite{openai_2019_dota}).
+Recently RL has achieved notable landmarks in various task domains such as classical board games chess \cite{Campbell_Hoane_Hsu_2002_DeepBlue}, Go \cite{Silver_Go}, poker \cite{Brown_Sandholm_2017_Poker} and visually rich computer games like Atari \cite{Mnih_2015_Atari} and Dota 2 \cite{openai_2019_dota}.
 At the core of some of the methods used, especially for board games, were planning algorithms, which rely on the knowledge of the environment dynamics.
@@
 Instead, it directly learns the global policy from interactions with the environment.
-While model-free approach has many successful examples (\cite{Mnih_2015_Atari,haarnoja_2018_sac,schulman_2017_ppo}), among its disadvantages are data inefficiency, as it requires high amount of interactions with the environment, and struggle for precise lookahead.
+While model-free approach has many successful examples \cite{Mnih_2015_Atari,haarnoja_2018_sac,schulman_2017_ppo}, among its disadvantages are data inefficiency, as it requires high amount of interactions with the environment, and struggle for precise lookahead.
 
-Model-based RL can address issues of both planning and model-free RL by first learning the environment dynamics, and then to exploit it for more efficient learning of the agent's policy (\cite{moerland_2020_modelbased}).
+Model-based RL can address issues of both planning and model-free RL by first learning the environment dynamics, and then to exploit it for more efficient learning of the agent's policy \cite{moerland_2020_modelbased}.
 Typically, the model is represented by MDP and consists of two components: a state transition model and a reward model.
-After the model has been learned, MDP planning method such as MCTS (\cite{Coulom_2007_mcts}) can be applied to infer the optimal policy.
-Many model-based RL approaches build a model that operates at the pixel-level (\cite{kaiser_2020_modelbased}), which is hungry for computational resources and attentive to unimportant details.
+After the model has been learned, MDP planning method such as MCTS \cite{Coulom_2007_mcts} can be applied to infer the optimal policy.
+Many model-based RL approaches build a model that operates at the pixel-level \cite{kaiser_2020_modelbased}, which is hungry for computational resources and attentive to unimportant details.
 Another approach is to try building a latent-space model.
 For example, a latent-space model can represent an abstract MDP that's equivalent to the real one in terms of state values, i.e.
-the cumulative reward of the planned trajectory (\cite{silver_2020_muzero}).
+the cumulative reward of the planned trajectory \cite{silver_2020_muzero}.
 
-To model the environment it's common to use ANNs (\cite{silver_2020_muzero,Ha_Schmidhuber_2018_worldmodels}).
-It's also very tempting to take more inspiration from human brains, especially for how our memory functions (\cite{Hassabis_2017_neuro}).
-One example of such human-like memory model is Hierarchical Temporal Memory (HTM) framework (\cite{George_Hawkins_2009}).
-This framework utilizes discrete binary sparse distributed representations (SDR) (\cite{Cui_Ahmad_Hawkins_2017_sdr}), which enables more efficient data processing, high noise tolerance and a natural way to set similarity metric between data embeddings.
-It also provides a model of the sequential temporal memory (TM) (\cite{hawkins_TM}).
+To model the environment it's common to use ANNs \cite{silver_2020_muzero,Ha_Schmidhuber_2018_worldmodels}.
+It's also very tempting to take more inspiration from human brains, especially for how our memory functions \cite{Hassabis_2017_neuro}.
+One example of such human-like memory model is Hierarchical Temporal Memory (HTM) framework \cite{George_Hawkins_2009}.
+This framework utilizes discrete binary sparse distributed representations (SDR) \cite{Cui_Ahmad_Hawkins_2017_sdr}, which enables more efficient data processing, high noise tolerance and a natural way to set similarity metric between data embeddings.
+It also provides a model of the sequential temporal memory (TM) \cite{hawkins_TM}.
 Learning in TM isn't based on the gradient descend, and typically it learns with a faster rate compared to common ANNs.
@@
 
-\section{Background}
+\section{\uppercase{Background}}
 
@@
 As a result, it receives the reward $r_t \in R(s_t, a_t)$ and transitions to the new state $s_{t+1} = P(s_t, a_t)$.
-An agent's policy $\pi: S \rightarrow A$ is a mapping from a state to an action.
+An agent's policy $\pi: \St \rightarrow \A$ is a mapping from a state to an action.
 The agent's goal is to maximize the expected return $\E \left[ \sum_{t=0}^{\infty} r_t \right]$.
@@
 
-\subsection{Temporal Memory}
+\subsection{Temporal Memory} \label{sec_TM}
 
@@
 For SDR vectors a union operation is defined as a bitwise OR applied to the corresponding vectors.
-Also we define a similarity metric, which is the number of the same one-bits of the vectors in consideration.
+Also we define a similarity metric, which is the number of the same one-bits of the vectors in consideration, which is, again, the same as the dot product of such vectors.
 
@@
 
-\section{Planning with Temporal Memory}
+\section{\uppercase{Planning with Temporal Memory}}
 
@@
   \centering
-  \includegraphics[width=0.8\linewidth]{agent_scheme}
+  \includegraphics[width=\linewidth]{agent_scheme}
   \caption{Schematic view of our agent and its training loop.} \label{fig_agent_scheme}
@@
 
-\subsection{Memory module}
+\subsection{Memory module} \label{sec_MemoryModule}
 
 The agent's memory module consists of two submodules.
-One submodule, called \textit{transition memory}, learns the transition model $f: \St \times \A \rightarrow \St$ of the environment, while the other, called \textit{goal states memory} keeps track of the rewarding goal states $s_g$.
+One submodule, called \textit{transition memory}, learns the transition model $f: \St \times \A \rightarrow \St$ of the environment, while the other, called \textit{goal states memory}, keeps track of the rewarding goal states $s_g$.
 
-Consider an agent's trajectory $\tau = [s_0, a_0, r_1, s_1, a_1, r_2, \dots, r_{T-1}, s_T]$.
+Consider a trajectory of an agent $\tau = [s_0, a_0, r_0, s_1, a_1, r_1, \dots, r_{T-1}, s_T]$.
 Each pair $(s_t, a_t)$ at the moment $t$ defines the precondition to the transition in the environment $(s_t, a_t) \rightarrow s_{t+1}$.
@@
 
-The transition memory submodule is based on a Temporal Memory.Recall that TM works with sparse distributed representations (SDR).
+The transition memory submodule is based on a Temporal Memory.
+Recall that TM works with sparse distributed representations (SDRs).
 To encode each state-action pair $(s, a)$ we use two separate integer encoders for states and actions respectively to obtain binary sparse vectors.
 The states encoder maps integer numbers from $[0, |\St|)$ range, and the actions encoder maps integer numbers from $[0, |\A|)$ range.
-As most of the time our method works with SDR, for simplicity further on we use the same notation $s$ and $a$ for their corresponding SDR too.
-To work with pairs $(s, a)$ we just concatenate their SDR vectors together into one.
+As most of the time our method works with SDR, for simplicity further on we use the same notation $s$ and $a$ for state's and action's corresponding SDR too.
+To work with pairs $(s, a)$ we just concatenate their respective SDR vectors together into one.
 
-For any given state-action pair $(s, a)$ the transition memory module can give a prediction on what state $s'$ is coming next.
+For any given state-action pair $(s, a)$ the transition memory submodule can give a prediction on what state $s'$ is coming next.
 Precisely, it allows prediction queries of the form:
@@
 
-\footnotetext{TM can work with unions of SDR vectors.
-When SDR vectors in a union are semantically meaningful, e.g.
-when they represent states or actions, we call such unions \textit{superpositions} of states or actions respectively \footnotemark.}
+  \footnotetext{In the Sec.~\ref{sec_TM} we explained that TM can work with unions of SDR vectors.
+When SDR vectors in a union aren't just random vectors, when they represent states or actions, we call such unions \textit{superpositions} of states or actions respectively.
+Note that once a superposition vector is made you cannot simply split it back into the origin SDR vectors of the union.
+\footnotemark}
 
-\footnotetext{As you see, in our current encoding scheme the different states or actions are encoded with non-overlapping bits which makes them distinguishable from each other in a superposition.
+\footnotetext{However, our current encoding scheme maps the different states or actions with non-overlapping bits, which keeps them easily distinguishable from each other even in a superposition.
 Whilst it's true, the current planning algorithm only requires a separation of the state bits from the action bits in a state-action pair SDR.
 This is done intentionally.
-Non-overlapping scheme makes things simpler to read and debug, but cannot keep the similarity between states (or observations).
-For the future work we plan to experiment with more complex environments and therefore to switch to overlapping mappings.}
+Non-overlapping scheme makes things simpler to read and debug.
+The downside is it cannot keep the similarity between states (or observations).
+That's why for the future work, when we plan to experiment with more complex environments, transition to overlapping mapping seems inevitable.}
 
-In some specific circumstances the transition memory submodule can be used backward - as a reverse of an environment transition function $f$.
+Under certain conditions the transition memory submodule can be used backward - as an inverse of an environment transition function $f$.
 This feature becomes available after getting the result to a prediction query and only for the states $s'_m$ that match the result $S' = \cup s'_m$.
@@
 
-Planning process is split into two stages (see function ~\ref{func_PlanActions}):
+The goal of planning is to find a plan of actions $p = [a_t, a_{t+1}, \dots, a_{t+h-1}]$ that leads to the desired state $s_g = s_{t+h}$ from the current state $s_t$.
+We restrict resulting plan to have no more than $H$ actions.
+Therefore, $H$ defines planning horizon of an agent.
 
-\begin{enumerate}
-  \item Forward planning from the current state until reward is reached.
-  \item Deducing the backward path from the reward state to the current.
-\end{enumerate}
-
-\begin{function}
-  \caption{PlanActions($s_0$)}\label{func_PlanActions}
-  \SetAlgoNoLine
-  $s_0$: initial state \;
-  $g: $ reached goal \;
-  $H: $ Temporal Memory active segments history \;
-  \quad \\
-
-  $g, H \leftarrow PredictToGoal(s_0)$ \;
-  $plan \leftarrow BacktrackFromGoal(g, s_0, H)$ \;
-
-  \Return{$[a_0, a_1, .. a_T] = plan$}
-\end{function}
+Planning process is split into two stages.
+It starts with forward planning from the current state $s_t$ until a rewarding goal state $s_g$ is reached.
+At this stage the planning algorithm operates with superpositions and cannot track the exact path to the goal state.
+Because of that we switch it to backtracking stage, during which the exact path from the reached goal state back to the starting state $s_t$ is deduced.
 
@@
 
-Agent uses its experience and expects the reward to be in one of the previous rewarding states.
-But even if it is wrong, we expect that direct goal-based exploration is more optimal than random walking.
-Hence it keeps the goal list and gives such goal states a priority.
+In our method an agent uses its experience and expects the reward to be in one of the rewarding goals states from previous episodes.
+But even if that's not true, we still expect that the direct goal-based exploration could be more optimal than pure random walking.
+Hence we set the agent to keep track of previous rewarding states using its goal tracking memory submodule.
+We give such goal states a priority.
 
-Consider that at some timestep $t_0$ the agent is in state $s_0$.
-To find out if there is a reachable goal state in a radius of $N$ actions, it start planning from the superposition $(S_0 = s_0, A = \cup a^i)$, where $A$ is a union of all available actions (we use superscripts for indices here as subscripts are used to denote timestep).
-It is similar to making all possible actions simultaneously in parallel.
-To achieve this planner uses the memory and gets a superposition of next state-action pairs.
+Consider that at some timestep $t_0$ an agent is in state $s_0$.
+To find out if there is a reachable goal state in a radius of $H$ actions, we start planning from the state-action superposition $(S_0 = s_0, A = \cup a^i)$ \footnotemark, where $A$ is a superposition of all available actions (see Alg.~\ref{alg_ForwardPlanning}).
+One step of the planning is just the transition memory submodule making a prediction.
+This gives us a superposition of the state-action pairs expected to go next.
+Predicting for a superposition $(S_0 = s_0, A = \cup a^i)$ is similar to making all possible actions from the state $s_0$ simultaneously in parallel and getting the result of all possible outcomes entangled together into a single superposition SDR vector.
+However, we still can split the resulting superposition into the pair of states superposition and actions superpositions, because it is just a concatenation of independent parts.
 
-After the first planning step the planner has a superposition of the next states $S_1 = \cup s^i$, reachable in one agent's action.
-To get all reachable states in two agent's action, planner makes another prediction step - now from the superposition $(S_1, A)$, and so on (see function ~\ref{func_PredictToGoal}).
+\footnotetext{We use superscripts for action indices here to avoid messing with subscripts denoting the timestep.}
 
-\begin{function}
-  \caption{PredictToGoal($s_0$)} \label{func_PredictToGoal}
+After the first planning step we have a superposition of the next states $S_1 = \cup s_1^i$, reachable in one agent's action.
+To get all reachable states in two agent's action, we repeat prediction step, now from the superposition $(S_1, A)$.
+Again, it's similar to making all possible actions from each of the next states in $S_1$.
+
+\begin{algorithm}
+  \caption{Forward Planning} \label{alg_ForwardPlanning}
   \SetNoFillComment
   \KwData{$s_0$ - initial state}
-
-  $sa_0 \leftarrow Encode((s_0, \cup a^i))$ \tcp*{Init starting superposition}
-  $H \leftarrow []$ \tcp*{Init active segments history}
-
-  \ForEach {timestep $i$} {
-    $memory.ActivatePredict(sa_i)$ \;
   \quad \\
 
-    $H[i] \leftarrow memory.active\_segments$ \;
-    $sa_{i+1} \leftarrow PredictionFromSegments(H[i])$ \;
+  $sa_0 \gets (s_0, A)$ \tcp{starting superposition}
+  $\Theta \gets []$ \tcp{active segments history}
+
+  \For{$i \gets 0, \dots, H-1$} {
+    $memory.ActivatePredict(sa_i)$ \;
+    $\Theta[i] \gets memory.active\_segments$ \;
     \quad \\
 
-    $s, a \leftarrow Decode(sa_{i+1})$ \tcp*{Both $s$ and $a$ are superpositions!}
-    $reached\_goal \leftarrow MatchGoals(s)$ \;
+    $(S_i, A_i) \gets PredictionFromSegments(\Theta[i])$\;
+    $sa_{i+1} \gets (S_i, A)$ \;
+    \quad \\
+
+    $reached\_goal \gets MatchGoals(S_i)$ \;
     \If(){$reached\_goal \neq \emptyset$} {
-      \Return {$reached\_goal, H$} \;
+      \Return {$reached\_goal, \Theta$} \;
     }{}
@@
   \Return{$\emptyset$}
-\end{function}
+\end{algorithm}
 
-This process is continued until either some state from the goal list is reached, or planner exceeds its planning horizon limit.
-In the former case planner proceeds to the next state of planning process.
-In the latter, agent uses its random strategy to make a move before it tries to plan again.
-To test if the goal state is in superposition we intersect their sparse vectors and sum active bits, which is the same as dot product of such vectors.
-If they match is over some predefined threshold, the state is considered presented in this superposition.
+This prediction process continues until either one of the tracked goal states is reached, or planning horizon limit is exceeded.
+In the former case the planning algorithm proceeds to the next stage.
+In the latter, we cancel it, and an agent has to use random strategy to make a move before it tries to plan again.
+To test if any of the tracked goal states is reached at $i$-th planning step, we calculate similarity between each of their SDR vectors and $i$-th superposition $S_i$.
+The goal state is considered presented in $S_i$ superposition, and therefore reached during the forward planning, if its similirity is over some predefined threshold.
 
-During this stage planner collects a history of active superpositions $H = [S_0, S_1, \dots S_T]$.
-In more details, he keeps the track of Temporal Memory cells' segments activations.
-Each segment is a pair $(p, \{c_i\})$, where $p$ - index of predicted cell and $\{c_i\}$ is a set of cell indices that induce prediction of $p$.
-This information allows planner to use inversed environment transition function to get which $\{(s, a)\}$ has lead to a state $s'$ by tracking which cells predict $s'$ cells.
+During this stage we also collect a history of active superpositions $\Theta = [S_0, S_1, \dots S_h]$.
+In more details, we keep the track of Temporal Memory cells' segments activations.
+Each segment is a pair $(p, \{c_i\})$, where $p$ - an index of the predicted cell and $\{c_i\}$ is a set of cell indices that induce prediction of $p$.
+This information allows us to provide the transition memory submodule with an inversed transition function $f^{-1}$ discussed in Sec.~\ref{sec_MemoryModule}.
+Thus, the transition memory submodule can answer which $\{(s, a)\}$ has lead to a state $s'$ by tracking which cells predicted $s'$ cells.
 
-It's possible to find multiple reachable goals at the end of this stage. In our current implementation we use the first matched from the goal list.
+It's possible to find multiple reachable goals at the end of this stage. In our current implementation we use the first matched from the set of tracked goals.
 
@@
 
-After the first stage planner ``believes'' that some goal state $g = s_T$ can be reached in $T$ steps.
-The problem is that he doesn't know how to get there as until now he worked with superpositions.
-So the main goal of this stage is to restore the sequence of actions leading from $s_0$ to $s_T$.
-Further on this sequence of action is meant to be the agent's plan in environment.
+After the first stage the planning algorithm ``believes'' that some goal state $g = s_h$ can be reached in $h$ steps.
+The problem is that it doesn't know how to get there because until now it worked with superpositions.
+Therefore, the main goal of this stage is to restore the sequence of actions leading from $s_0$ to $s_h$.
+Further on this sequence of action is meant to be the agent's plan in the environment.
 
-Planner solves the problem from the end.
-It moves backward in the history of segments activations graph, trying to get to the starting state $s_0$ at the first timestep.
-Due to its ``backward moving'' nature we call this stage \textit{backtracking} (see function ~\ref{func_BacktrackFromGoal}).
+We solve the problem from the end.
+We move backward in the history of segments activations graph, trying to get to the starting state $s_0$ at the first timestep.
+Due to its ``backward moving'' nature we call this stage \textit{backtracking}.
 
-\begin{function}
-  \caption{BacktrackFromGoal($g$, $H$)} \label{func_BacktrackFromGoal}
-  % \SetAlgoNoLine
-  \KwData{$g = s_T$ - goal state; $H$ - active segments history}
-
-  $T \leftarrow |H| + 1$ \tcp*{As the last recorded transition is for $T-1 \rightarrow T$}
-  $S_T \leftarrow PredictionFromSegments(H[T-1])$ \tcp*{State superposition at $T$}
-  $s_T \leftarrow s_T \cap S_T$ \tcp*{Keeps only matched bits}
-  \quad \\
-  $is\_successful, actions \leftarrow Backtrack(s_T, T-1, H)$ \;
-  \Return {$is\_successful, actions$}
-\end{function}
-
-Recall that planner keeps track of all state superpositions $H = [S_0, S_1, \dots S_T]$ it has reached on each step $t$, and for any state $s'$ from each of these superpositions $S_i$ the memory can answer which state-action pairs has lead to it.
-
-Therefore, given the reached goal state $g = s_T$ planner can get which state-action pair $(s_{T-1}, a_{T-1})$ has lead to it.
-Hence he knows for $T-1$ step which action is required to get from $s_{T-1}$ to the goal state $s_T$.
-And also he can recursively reduce the $s_0 \leadsto s_T$ pathfinding problem to the  $s_0 \leadsto s_{T-1}$.
-Planner repeats this procedure and for each timestep $t$ sequentially finds state-action pairs $(s_{t-1}, a_{t-1}) = f^{-1}(s_t)$ until he returns to the initial state at the first timestep (see ~\ref{func_Backtrack} function).
+Recall that we kept track of all state superpositions $\Theta = [S_0, S_1, \dots S_h]$ the planning algorithm has reached at each step.
+For any state $s'$ from each of these superpositions $S_i$ the transition memory submodule can answer which state-action pairs has lead to it.
+Therefore, given the reached goal state $g = s_h$ we can infer which state-action pair $(s_{h-1}, a_{h-1})$ has lead to it.
+Hence we know for $(h-1)$-th step which action is required to get from $s_{h-1}$ to the desired state $s_h$.
+And also we can recursively reduce the $s_0 \leadsto s_h$ pathfinding problem to the  $s_0 \leadsto s_{h-1}$.
+We can repeat this procedure and for each timestep $t$ sequentially find state-action pairs $(s_{t-1}, a_{t-1}) = f^{-1}(s_t)$ until we return to the initial state at the first timestep.
+See Alg.~\ref{alg_Backtracking} for a more detailed and formal description of the backtracking recursive algorithm.
 
 In general case $f^{-1}(s_{t+1})$ gives multiple pairs $(s_t, a_t)$, and each leads to $s_{t+1}$.
-But not every pair is actually reachable in exactly $t$ steps from the $s_0$, i.e.
-not from every pair you can backtrack to the initial state at time $0$.
-Hence planner sequentially tests each pair until he find the first successful.
+But not every pair is actually reachable in exactly $t$ steps from the $s_0$, which means that not from every pair you can backtrack to the initial state at time $0$.
+Hence we sequentially test each pair until we find the first successful.
 
-As the result of the backtracking stage planner forms a plan of actions $plan = [a_0, a_1, \dots a_{T-1}]$ which he ``believes'' [in terms of his approximation of the environment dynamics] will lead the agent from the initial state $s_0$ to the target state $g = s_T$.
+As the result of the backtracking stage the planning algorithm forms a plan of actions $p = [a_t, a_{t+1}, \dots a_{t+h}]$ which is ``believed'' to lead the agent from the current state $s_t$ to the target state $g = s_{t+h}$.
 
-\begin{function}
-  \caption{Backtrack($p_t$, $t$, $H$)} \label{func_Backtrack}
-  \KwData{$p_t$ - indices of required predicted cells at time $t$, $t$ - current timestep, $H$ - history of active segments}
+\begin{algorithm}
+  \caption{Backtracking} \label{alg_Backtracking}
+  \KwData{$t$ - current timestep, $p_t$ - cells SDR that is required to be predicted at time $t$, $\Theta$ - a history of active segments}
+  \quad \\
 
@@
 
-  $P \leftarrow H[t][p_t]$ \tcp*{List of cell sets, each predicts one of $p_t$ cells}
-  $C \leftarrow P$ \tcp*{Initializes cell clusters - candidates for backtracking}
+  \tcp{Inits candidates for backtracking with the list of cell SDRs, each predicting one of $p_t$ cells}
+  $C \gets \Theta[t][p_t]$ \;
   \quad \\
 
-  \tcp{Iteratively merges any two sufficiently matching cell clusters}
-  \While{$\exists c_i, c_j \in C: match(c_i, c_j) \geq \Theta$} {
+  \tcp{Iteratively unites any two sufficiently similar cell SDRs}
+  \While{$\exists c_i, c_j \in C: match(c_i, c_j) \geq \Omega_1$} {
     $C[c_i] = c_i \cup c_j$ \;
@@
   \ForEach{$c \in C$}{
-    \tcp{Filters by the amount of $p_t$ cells cluster $c$ predicts}
-    $p = PredictionFromCluster(c, H[t])$ \;
-    \If(){$match(p, p_t) \le \Omega$}{
+    \tcp{Filters by the amount of $p_t$ cells that $c$ predicts}
+    $p = PredictionFromSegment(c, \Theta[t])$ \;
+    \If(){$match(p, p_t) \le \Omega_2$}{
       \textbf{continue}
@@
 
-    \tcp{Recursively checks candidate cluster $c$}
-    $s_t, a_t \leftarrow Decode(ColumnsFromCells(c))$ \;
+    \tcp{Recursively checks candidate $c$}
+    $s_t, a_t \leftarrow ColumnsFromCells(c)$ \;
 
-    $is\_successful, actions \leftarrow Backtrack(s_t, t-1, H)$ \;
+    $is\_successful, plan \leftarrow Backtrack(s_t, t-1, \Theta)$ \;
     \If() {is\_successful} {
-      $actions.append(a_t)$ \;
-      \Return{True, actions}
+      $plan.append(Decode(a_t))$ \;
+      \Return{True, plan}
     }
@@
   \Return{False, $[]$}
-\end{function}
+\end{algorithm}
 
-After successful backtracking the planner has a policy for the next $T$ steps, which an agent uses.
-And if it ends up in a rewarding state, an episode finishes.
-Otherwise it continues and the planner removes that state from its goal list for the rest of the episode.
+After successful backtracking the planning algorithm returns a policy for the next $h$ steps, which an agent uses.
+If the agent ends up in a rewarding state, an episode finishes.
+Otherwise it continues and the goal tracking memory submodule removes that goal state from the tracking set for the rest of the episode.
 
-\section{Experiments}
+\section{\uppercase{Experiments}}
 
@@
 
-\section{Conclusion}
+\section{\uppercase{Conclusions}}
 
@@
 
-\bibliography{citations}
+{\small\bibliography{citations}}
 
-\appendix
+\section*{\uppercase{Appendix}}
 
-\section{Appendix}
 \subsection{Handcrafted mazes} \label{apx_handcrafted_mazes}
@@
 
-\subsubsection{multi\_way\_v0}
-
-Each experiment had 4 different rewarding places being sequentially changed every 100 episodes:
+\texttt{multi\_way\_v0}. Each experiment had 4 different rewarding places being sequentially changed every 100 episodes:
 
@@
 
-\subsubsection{multi\_way\_v2}
-
-Each experiment had 6 rewarding places being changed every 200 episodes:
+\texttt{multi\_way\_v2}. Each experiment had 6 rewarding places being changed every 200 episodes:
 
 \begin{verbatim}
-  #########  #########  #########  #########  #########  #########
-  #---#####  #-X-#####  #---#####  #---#####  #---#####  #---#####
-  #-#-#####  #-#-#####  #-#-#####  #-#-#####  #-#-#####  #-#-#####
-  #--@---##  #--@---##  #--@--X##  #--@---##  #X-@---##  #--@---##
-  ###--####  ###--####  ###--####  ###--####  ###--####  ###--####
-  ###-#--X#  ###-#---#  ###-#---#  ###-#---#  ###-#---#  ###-#--X#
-  ###---###  ###---###  ###---###  ###---###  ###---###  ###---###
-  ###-#####  ###-#####  ###-#####  ###X#####  ###-#####  ###-#####
-  #########  #########  #########  #########  #########  #########
+  #########  #########  #########
+  #---#####  #-X-#####  #---#####
+  #-#-#####  #-#-#####  #-#-#####
+  #--@---##  #--@---##  #--@--X##
+  ###--####  ###--####  ###--####
+  ###-#--X#  ###-#---#  ###-#---#
+  ###---###  ###---###  ###---###
+  ###-#####  ###-#####  ###-#####
+  #########  #########  #########
+
+  #########  #########  #########
+  #---#####  #---#####  #---#####
+  #-#-#####  #-#-#####  #-#-#####
+  #--@---##  #X-@---##  #--@---##
+  ###--####  ###--####  ###--####
+  ###-#---#  ###-#---#  ###-#--X#
+  ###---###  ###---###  ###---###
+  ###X#####  ###-#####  ###-#####
+  #########  #########  #########
 \end{verbatim}

### 2020-11-01 reports/v1_0/paper_eng/paper_eng.tex

--- 2020-10-29 reports/v1_0/paper_eng/paper_eng.tex
@@
 \begin{abstract}
-  Among the key problems in artificial intelligence is sequential decision making, which can be formalized as Markov Decision Process (MDP).
-One approach to solve it, called model-based Reinforcement Learning (RL), combines learning the model of the environment and the global policy.
-Having a good model of the environment opens up such properties as data efficiency and targeted exploration.
+  Hierarchical Temporary Memory (HTM) is a biologically inspired model of neocortex.
+Its key properties proved it to be useful useful for sequence learning and anomaly detection problems.
+While they also make it potentially useful for a model-based RL, this particular application of HTM has hardly been studied yet.
 
-  While most of the memory-based approaches are based on using Artificial Neural Networks (ANNs), in our work we instead draw the ideas from Hierarchical Temporary Memory (HTM) framework, which is based on human-like memory model.
-We utilize it to build an agent's memory that learns the environment dynamics.
-We also accompany it with an example of planning algorithm, that enables the agent to solve RL tasks.
+  We show a possible way to utilize HTM in RL setting to model an environment dynamics.
+We accompany it with an example of planning algorithm which is used to test learning abilities and find limitations of such a model.
 \end{abstract}
 
-\keywords{Model-based reinforcement learning \and Markov Decision Process \and planning \and Hierarchical Temporal Memory \and Sparse Distributed Representation.}
+\keywords{First keyword  \and Second keyword \and Another keyword.}
 
@@
 
-Reinforcement Learning (RL) seek to find methods that enable robots or virtual agents to act autonomously in order to solve human-defined tasks.
-It requires from such methods to learn a policy for sequential decision-making problems, which are commonly formalized as MDP optimization \cite{Puterman_1994}.
-Recently RL has achieved notable landmarks in various task domains such as classical board games (chess \cite{Campbell_Hoane_Hsu_2002_DeepBlue}, Go \cite{Silver_Go}, poker \cite{Brown_Sandholm_2017_Poker}) and visually rich computer games (Atari \cite{Mnih_2015_Atari}, Dota 2 \cite{openai_2019_dota}).
-At the core of some of the methods used, especially for board games, were planning algorithms, which rely on the knowledge of the environment dynamics.
+Reinforcement Learning (RL) has recently got a lot of ...
+A great part of its success was due to advances in application of Artificial Neural Networks (ANN).
+Although ANNs model is very popular and successful at the moment, it's built on top of very simplified model of a neuron.
 
-In general case, though, the model of the environment is unknown, and there are two common opposite approaches in RL to deal with it: a model-free and a model-based.
-The former approach doesn't explicitly learn and make use of the knowledge of the environment.
-Instead, it directly learns the global policy from interactions with the environment.
-While model-free approach has many successful examples (\cite{Mnih_2015_Atari,haarnoja_2018_sac,schulman_2017_ppo}), among its disadvantages are data inefficiency, as it requires high amount of interactions with the environment, and struggle for precise lookahead.
+There are more sophiscticated models of neuron [TODO links], which are mostly used by neurophysiologists and serve their needs - to reverse engineer internal physical structure and operational model of neuron.
+These models are very computationally expensive compared to Artificial neuron, so they are rarely used by Machine Learning practitioners as a basis for complex neural networks to solve real ML tasks.
 
-Model-based RL can address issues of both planning and model-free RL by first learning the environment dynamics, and then to exploit it for more efficient learning of the agent's policy (\cite{sutton_barto_2018}).
-Typically, the model is represented by MDP and consists of two components: a state transition model and a reward model.
-After the model has been learned, MDP planning method such as MCTS (\cite{Coulom_2007_mcts}) can be applied to infer the optimal policy.
-Many model-based RL approaches build a model that operates at the pixel-level (\cite{kaiser_2020_modelbased}), which is hungry for computational resources and attentive to unimportant details.
-Another approach is to try building a latent-space model.
-For example, a latent-space model can represent an abstract MDP that's equivalent to the real one in terms of state values, i.e.
-the cumulative reward of the planned trajectory (\cite{silver_2020_muzero}).
+Still, it's very tempting to take inspiration from [human] brains and how they function as, in the end, this's what AI researchers are trying to recreate [TODO Hassabis paper on inspiration].
 
-To model the environment it's common to use ANNs (\cite{silver_2020_muzero,Ha_Schmidhuber_2018_worldmodels}).
-It's also very tempting to take more inspiration from human brains, especially for how our memory functions (\cite{Hassabis_2017_neuro}).
-One example of such human-like memory model is Hierarchical Temporal Memory (HTM) framework (\cite{George_Hawkins_2009}).
-This framework utilizes discrete binary sparse distributed representations (SDR) (\cite{ahmad_2016_sdr}), which enables more efficient data processing, high noise tolerance and a natural way to set similarity metric between data embeddings.
-It also provides a model of the sequential temporal memory (TM) (\cite{hawkins_TM}).
-Learning in TM isn't based on the gradient descend, and typically it learns with a faster rate compared to common ANNs.
+HTM neuron model is on of the models aiming the middle point between neurophysiological plausibility, simplicity and computational efficiency.
+TODO more about it, it's applications and TM.
+EDIT \textit{This model works mostly with discrete [binary] sparse signals.
+And it gives the model some cool features - the natural way to set similarity metric between these vector, resilience to the noise, computational efficency and so on.
+This model also proposes some basic Neural Network building blocks, like sequential Temporal Memory, which is somewhat similar to what we have in mainstream ML.
+It doesn't use backpropagation for learning, and it doesn't need to set up very small learning rate, so in the end TM learns much faster than common ANNs.}
 
-In our work we present a simplified memory-based framework based on the ideas of HTM that can be used to model the environment dynamics, supplied with the planning algorithm to infer the optimal global policy.
-We study applicability of the presented memory model and viability of the planning method in different maze environments.
+In our work we studied if and how Temporal Memory could be reliably used to model an environment an agent's opeates in - how an environment changes over time and how it reacts to agent's actions.
+The agent's own inner model of the environment allows it to plan ahead, to see the outcomes of its actions and therefore to identify desirable or dangerous moves.
 
-\begin{figure}
-  \centering
-  \begin{minipage}{.49\linewidth}
-    \includesvg[width=\linewidth]{gridworld_5x5_1_1_200_1337_map_0_1173222464}
-  \end{minipage}
-  \begin{minipage}{.49\linewidth}
-    \includesvg[width=\linewidth]{gridworld_5x5_1_1_200_1337_map_2_1561234712}
-  \end{minipage}
-  \caption{Example of generated maze environments of 5x5 and 8x8 sizes. Walls - dark purple, starting agent's position - yellow, rewarding goal state - salad.} \label{fig_maze_examples}
-\end{figure}
+The trial and error strategy can be very expensive in the real world.
+For example, you couldn't be able to afford breaking the car a thousand times to teach it how to bypass an obstacle, especially if the obstacle is human.
+Planning allows an agent to spend less time acting in the real world by spending more time learning in its own internal simulation, like in a "dream".
 
-\section{Background}
+We provide an example of planning strategy to test both:
 
-\subsection{Reinforcement Learning}
+\begin{itemize}
+  \item The quality of memory - how good it approximates the environment's dynamics?
+  \item The viability of the planning method - does this method allow an agent to solve tasks and how much it helps?
+\end{itemize}
 
-In this paper we consider the classic decision-making problem with an agent operating in an environment, which is formalized as Markov Decision Process (MDP).
-As a starting point in the research we simplify the problem assuming that the environment is fully observable and deterministic.
-We also consider only environments with the distinctive goal-oriented tasks, where an agent is expected to reach desirable goal states in a limited time.
+We also studied how the planning method's parameters affect the performance of an agent.
 
-Therefore, MDP is defined as a tuple $M = (\St, \A, R, P, s_0, s_g)$, where $\St = \{1, \dots, s_n\}$ is the state space of the environment, $\A = \{1, \dots, a_m\}$ is the action space available to an agent, and $s_0$ and $s_g$ are the initial and the goal state distributions respectively.
-At each time step $t$, an agent observes a state $s_t \in \St$ and selects an action $a_t \in \A$.
-As a result, it receives the reward $r_t \in R(s_t, a_t)$ and transitions to the new state $s_{t+1} = P(s_t, a_t)$.
-An agent's policy $\pi: S \rightarrow A$ is a mapping from a state to an action.
-The agent's goal is to maximize the expected return $\E \left[ \sum_{t=0}^{\infty} r_t \right]$.
+We restricted tasks to be held in very simple deterministic MDPs with perfectly distinguishable environment states.
+In this case Temporal Memory proved to learn fast and reliably the model of the environment.
+We also found out that our planning method mostly helps an agent to solve problems, and in special cases it allows an agent to solve them in optimal manner (or nearly optimal).
 
-We don't use discount factor $\gamma$ in the return calculation.
-Instead, we assume that the reward function is designed the way that it highly rewards reaching the goal states and slightly punishes in any other states, which should reflect the goal-oriented nature of the environments that were taken into consideration.
+\section{Problem statement}
 
-\subsection{Temporal Memory}
+Consider classic Markov Decision Problem (MDP) - when agent plays in deterministic GridWorld environment, which is a maze on a square grid.
+Agent starts in the fixed point, and its goal is to find a single reward placed at some fixed point.
 
-Temporal Memory (TM) from Hierarchical Temporal Memory (HTM) framework is a model of the sequential memory.
-It works with sparse distributed representations (SDR), which are sparse binary vectors, typically high-dimensional.
-Therefore, the data encoding scheme should be chosen too.
+Agent can move to the adjacent grid cells in horizontal and vertical directions.
+Each action move is enumerated from 0 to 3 counter-clockwise starting from east.
+When agent tries to move into the maze wall, it stays at the same place.
 
-In our work both actions and states are represented as integer numbers from a fixed range (although, different for states and actions).
-Thus, we use a simple encoding scheme, where numbers from some range $[0, N)$ are mapped to non-overlapping equally-sized sets of active bits of the output SDR vector.
-For example for $N = 3$ numbers 0, 1 and 2 are encoded correspondingly as following:
+We consider only deterministic environments. When agent finds the goal it is rewarded with high positive value $+1$, otherwise each unsuccessful step it is slightly punished with $-0.01$ value.
 
-$$
-\begin{aligned}
-  0 & \rightarrow \verb|1111 0000 0000| \\
-  1 & \rightarrow \verb|0000 1111 0000| \\
-  2 & \rightarrow \verb|0000 0000 1111| \\
-\end{aligned}
-$$
+\section{Method description}
+In our implementation an agent's policy is a mixture of two: random and planned.
+Planned policy is provided by the corresponding part of an agent called \textit{planner}.
+If planner cannot provide a plan, agent degrades to a fully random exploration strategy.
 
-As you see, resulting vector is split into buckets of bits, and the size of buckets is a hyperparameter (in the example above it's 4).
+So, the primary part of our agent is a planning algorithm.
 
-For SDR vectors a union operation is defined as a bitwise OR applied to the corresponding vectors.
-Also we define a similarity metric, which is the number of the same one-bits of the vectors in consideration.
-
-Temporal Memory model not only provides an algorithm to learn sequences of SDR vectors $[v_0, v_1, v_2, \dots ]$ but also to make predictions about what SDR vector $v'$ comes next based on the past experience.
-The prediction is represented also as an SDR vector, which is a union of SDR vectors $v' = \cup v_{o_i}$, where each vector $v_{o_i}$ represents a single separate expected outcome.
-Notice that as the union of SDR vectors is also an SDR vector, TM model allows you to work simultaneously with the unions of sequences as well.
-
-\section{Planning with Temporal Memory}
-
-In our method we train an agent to reach rewarding goal states $s_g$.
-An agent is supplied with the memory module to learn the environment's dynamics and a planning algorithm which heavily relies on the learned model.
-The planning algorithm constructs a plan of actions $p = [a_t, a_{t+1}, \dots, a_{t+h-1}]$ that leads to the desired state $s_g = s_{t+h}$ from the current state $s_t$.
-In the case when the planning algorithm fails to provide a plan we switch the agent to the exploration strategy $\pi_r$ in order to improve the learned model.
-As the result, at each timestep an agent operates either according to a constructed plan $p$ or according to the random uniformly-distributed exploration policy $\pi_r$.
-
-\subsection{Memory module}
-
-The agent's memory module consists of two submodules.
-One submodule, called \textit{transition memory}, learns the transition model $f: \St \times \A \rightarrow \St$ of the environment, while the other, called \textit{goal states memory} keeps track of the rewarding goal states $s_g$.
-
-Consider an agent's trajectory $\tau = [s_0, a_0, r_1, s_1, a_1, r_2, \dots, r_{T-1}, s_T]$.
-Each pair $(s_t, a_t)$ at the moment $t$ defines the precondition to the transition in the environment $(s_t, a_t) \rightarrow s_{t+1}$.
-Therefore, we made transition memory to learn the sequences of such transitions $Tr = [(s_0, a_0), (s_1, a_1), \dots]$ from the trajectories that the agent produces.
-The learning is performed in an online fashion, i.e.
-at each timestep.
-
-The transition memory submodule is based on a Temporal Memory.Recall that TM works with sparse distributed representations (SDR).
-To encode each state-action pair $(s, a)$ we use two separate integer encoders for states and actions respectively to obtain binary sparse vectors.
-The states encoder maps integer numbers from $[0, |\St|)$ range, and the actions encoder maps integer numbers from $[0, |\A|)$ range.
-As most of the time our method works with SDR, for simplicity further on we use the same notation $s$ and $a$ for their corresponding SDR too.
-To work with pairs $(s, a)$ we just concatenate their SDR vectors together into one.
-
-For any given state-action pair $(s, a)$ the transition memory module can give a prediction on what state $s'$ is coming next.
-Precisely, it allows prediction queries of the form:
-
-\begin{quote}
-  \textit{given a superposition \footnotemark of states $S = \cup s_i$ and a superposition of actions $A = \cup a_j$, what superposition of next states $S'= \cup s'_m$ all possible state-action pairs $(s_i, a_j)$ lead to?}
-\end{quote}
-
-\footnotetext{TM can work with unions of SDR vectors.
-When SDR vectors in a union are semantically meaningful, e.g.
-when they represent states or actions, we call such unions \textit{superpositions} of states or actions respectively \footnotemark.}
-
-\footnotetext{As you see, in our current encoding scheme the different states or actions are encoded with non-overlapping bits which makes them distinguishable from each other in a superposition.
-Whilst it's true, the current planning algorithm only requires a separation of the state bits from the action bits in a state-action pair SDR.
-This is done intentionally.
-Non-overlapping scheme makes things simpler to read and debug, but cannot keep the similarity between states (or observations).
-For the future work we plan to experiment with more complex environments and therefore to switch to overlapping mappings.}
-
-In some specific circumstances the transition memory submodule can be used backward - as a reverse of an environment transition function $f$.
-This feature becomes available after getting the result to a prediction query and only for the states $s'_m$ that match the result $S' = \cup s'_m$.
-For each such state $s'$ it can provide which state-action pairs from the original query lead to $s'$: $\{(s, a)\} = f^{-1}(s')$.
-Notice that in general case it's possible that \textit{a multiple} state-action pairs lead to $s'$.
-
-\subsection{Planning algorithm}
+The main goal of a planner is to find a path to the reward. This path should not take more than \verb|planning_horizon| steps, which defines planning horizon of an agent.
 
@@
 
-\subsubsection{Forward planning}
+To support planning capability agent is provided with the memory module to remember environment transitions that it saw.
+In other words it learns environment dynamics by approximating its transition function $f: (s, a) \rightarrow s'$.
+Hence given any state and action it could make a prediction on what state is coming next.
+
+Agent's memory has some particular features which we describe before heading to the planning algorithm details.
+
+\subsection{Agent's memory}
+
+The agent's memory has an ability to work with unions of what it has ``remembered''.
+By working with union we mean working with a set as a whole without an ability to split it into its separate elements.
+In our case the memory can make a prediction about next states for a union of state-action pairs.
+Precisely, it allows queries of the form:
+
+\begin{quote}
+  \textit{given a union of states $S=\cup s_i$ and a union of actions $A=\cup a_j$, the union of which next states $S'= \cup s'_m$ all possible state-action pairs $(s_i, a_j)$ lead to?}
+\end{quote}
+
+In some specific circumstances the memory can be used backward - as a reverse of an environment transition function $f$.
+This feature becomes available after answering a prediction query and only for the states $s'_m$ from the result $S' = \cup s'_m$.
+For each such state $s'$ the memory can answer which state-action pairs from the original query lead to it: $\{(s, a)\} = f^{-1}(s')$.
+Note that in general case it's possible that \textit{a multiple} state-action pairs lead to $s'$.
+
+Also the memory module has a submodule to keep track of rewarding states that agent has ever met.
+We call it a \textit{goal list}.
+This list has restricted maximum size, which is a hyperparameter.
+At the moment it's implemented as a naive circular queue (with traits of a set, so that it only keeps unique elements).
+
+These features make a basis for the planning algorithm.
+
+\subsubsection{Agent's memory implementation details}
+
+Consider an agent's trajectory $\tau = (s_0, a_0, r_1, s_1, a_1, r_2, s_2, a_2, ...)$.
+Then each pair $(s_t, a_t)$ at some moment $t$ defines precondition to the environment transition: $(s_t, a_t) \rightarrow s_{t+1}$.
+
+Under the hood the agent's memory implementation is based on HTM Temporal Memory (TM) ~\cite{hawkins:TM}. TM learns sequences $[(s_0, a_0), (s_1, a_1), \dots]$ from the trajectories the agent plays.
+
+For each state-action pair $(s, a)$ the memory can answer which state-action pairs $(s', a')$ it expects to see next.
+The answer is provided as a union of these pairs.
+We call such unions \textit{state-action superpositions}.
+Note that it not only learns transitions $(s, a) \rightarrow s'$, but also what actions $\{a'_i\}$ agents did after getting to the next state $s'$.
+
+The memory works with sparse binary vectors. And the union operation for binary sparse vectors is defined naturally as a bitwise OR operation.
+
+Both state and action are encoded independently as sparse binary vectors and then two resulting vectors are concatenated together.
+Encoding scheme has each integer number from the range $[0, N)$ mapped to the fixed number of active bits.
+For example for $N = 3$ numbers 0, 1 and 2 are encoded correspondingly as following:
+
+\begin{itemize}
+  \item $0 \rightarrow$ \verb|1111 0000 0000|
+  \item $1 \rightarrow$ \verb|0000 1111 0000|
+  \item $2 \rightarrow$ \verb|0000 0000 1111|
+\end{itemize}
+
+As you see, resulting vector is split into buckets of bits. The size of buckets is a hyperparameter (in an example above it's 4).
+
+And an example of superposition of numbers 1 and 4 (from the $[0, 5]$ range) is OR applied to two corresponding sparse vectors:
+
+\begin{verbatim}
+  0000 1111 0000 0000 1111 0000
+  =
+  0000 1111 0000 0000 0000 0000
+  OR
+  0000 0000 0000 0000 1111 0000
+\end{verbatim}
+
+As you see, in our current encoding scheme different states and actions could be distinguished from each other in a superposition.
+But the planning algorithm only requires a separation of state superpositions from action superpositions, because for the future work with more complex environments encoding scheme is planned to change.
+
+As superposition of state-action pairs is also a valid binary sparse vector, the memory can naturally work with them as an input to prediction query as well.
+In this case it makes prediction for a multiple different environment transitions simultaneously.
+The result of prediction is also a superposition, because even for a single transition agent could do multiple different actions from the next state $s'$.
+
+\subsection{Planner. Stage 1: forward planning to a goal}
 
@@
 In the latter, agent uses its random strategy to make a move before it tries to plan again.
+
 To test if the goal state is in superposition we intersect their sparse vectors and sum active bits, which is the same as dot product of such vectors.
@@
 
-\subsubsection{Backtracking}
+\subsection{Planner. Stage 2: deducing the backward path}
 
@@
 
-\section{Experiments}
+\section{Results}
 
-In our experiments we decided to use classic GridWorld environments, which are represented as mazes on a square grid (see Figure ~\ref{fig_maze_examples}).
-Each state can be defined with the position of an agent, i.e.
-in which cell it stands.
-Thus, state space $\St$ consists of all possible agent's positions.
-We enumerate all positions (and therefore states) with integer numbers.
-An agent starts from the fixed state $s_0$, and its goal is to find a single reward placed at some fixed position $s_g$.
-Each environments has deterministic transition function $P: \St \times \A \rightarrow \St$.
-The action space consists of 4 actions $\A = \{ 0, 1, 2, 3 \}$ defining agent's move directions: 0 - east, 1 - north, 2 - west, 3 - south.
-Each action moves an agent to the adjacent grid cell in the corresponding direction, except when the agent tries to move into the maze wall - in this case it stays in the current cell.
+The main goal of this work was to prove that the memory built upon Temporal Memory from the HTM framework is able to reliably model the environment's dynamics and therefore it allows an agent to make successful plans with it.
+We compare the method's performance with baselines: random walk and DQN[TODO link].
 
-When an agent reaches the goal state $s_g$ it is rewarded with high positive value.
-Any other steps it is slightly punished.
-In our experiments we used $+1.0$ for the goal reward and $-0.01$ for the punishment.
-In each testing environment the optimal path from the starting point to the goal was less than 20 steps.
-Such reward function makes possible easy episode outcomes differentiation, depending on whether or not the goal state has been reached during it, and if the goal has been reached then how fast.
+We also investigated, how planning horizon and the size of the goal list affects agent's performance.
 
-All experiments were divided into two groups.
-Each experiment from the first group was held in a fixed environment, while each experiment from the second group had an environment being changed every $N_{ep}$ episodes.
+Our experiments were split into two sets. The first set was in a fixed environment and the second - in a changing environment.
 
-In our experiments we tested ability to model the environment and utilize learned model to plan actions that lead to the goal states.
-We compared our method's performance with the baselines: random strategy and DQN (\cite{Mnih_2015_Atari}).
-We also compared performance for the different planning horizons and the number of simultaneously tracked goals in the goal tracking memory.
+For an implementation of the random walk agent we used our agent with the zeroed planning horizon. We also used both greedy and $\epsilon$-greedy learned DQN strateies in our comparisons.
 
@@
 
-For this experiment setup we used a number of handcrafted mazes.
-Each experiment had a fixed maze and the starting point $s_0$, but the place of the reward was being changed every $N$ episodes.
-We tested an agent's ability to find the goal state and how it adapts to the new goal position.
+In this experiment setup we used a number of handcrafted GridWorld mazes.
+Each experiment has a fixed maze environment, but the place of the reward was being changed each $N$ episodes.
+We tested an agent's ability to find the reward and how it adapts to its new position.
 
-We present results for two mazes: \texttt{multi\_way\_v0} and \texttt{multi\_way\_v2} (see Figure ~\ref{fig_comparison_mw} and ~\ref{fig_detailed_mw0}).
-Each experiment in \texttt{multi\_way\_v0} had 4 different rewarding goal positions being sequentially changed every 100 episodes.
-Each experiment in \texttt{multi\_way\_v2} had 6 rewarding goal positions being changed every 200 episodes.
-Schematic view of these mazes is provided in Appendix ~\ref{apx_handcrafted_mazes}.
+Below is the notation used in the environment schemes below:
+\begin{itemize}
+    \item \verb|-| empty cell,
+    \item \verb|#| wall,
+    \item \verb|@| agent,
+    \item \verb|X| reward
+\end{itemize}
+
+\texttt{multi\_way\_v0} was the smallest environment. Each experiment had 4 different rewarding places being sequentially changed each 100 episodes:
+
+\begin{verbatim}
+  #####   #####   #####   #####
+  #@--#   #@-X#   #@--#   #@--#
+  #-#-#   #-#-#   #-#-#   #-#X#
+  #--X#   #---#   #X--#   #---#
+  #####   #####   #####   #####
+\end{verbatim}
+
+\texttt{multi\_way\_v2} was the largest of tested environments. Each experiment had 6 rewarding places being changed each 200 episodes:
+
+\begin{verbatim}
+  #########  #########  #########  #########  #########  #########
+  #---#####  #-X-#####  #---#####  #---#####  #---#####  #---#####
+  #-#-#####  #-#-#####  #-#-#####  #-#-#####  #-#-#####  #-#-#####
+  #--@---##  #--@---##  #--@--X##  #--@---##  #X-@---##  #--@---##
+  ###--####  ###--####  ###--####  ###--####  ###--####  ###--####
+  ###-#--X#  ###-#---#  ###-#---#  ###-#---#  ###-#---#  ###-#--X#
+  ###---###  ###---###  ###---###  ###---###  ###---###  ###---###
+  ###-#####  ###-#####  ###-#####  ###X#####  ###-#####  ###-#####
+  #########  #########  #########  #########  #########  #########
+\end{verbatim}
 
@@
   \end{minipage}
-  \caption{Comparison of the episode duration (timesteps) for the handcrafted mazes.
-\texttt{htm\_X\_Yg} is our agent with the planning horizon X and Y tracked last goals, \texttt{htm\_0} - random policy, \texttt{dqn\_greedy} - greedy DQN agent.
-Results are plotted with the moving average MA = 20 steps.} \label{fig_comparison_mw}
+  \caption{Comparison with the baselines.
+\texttt{htm\_X\_Yg} is our agent with the planning horizon X and the goal list size Y, \texttt{htm\_0} - random walk agent, \texttt{dqn\_greedy} - dqn agent.} \label{fig_comparison_mw}
 \end{figure}
+
+The results in these two environments are presented on Figure ~\ref{fig_comparison_mw}.
 
@@
     \includesvg[width=\linewidth]{multi_way_v0__dqn__steps}
-    \subcaption{Greedy and $\epsilon$-greedy DQN agent}
+    \subcaption{Comparison with greedy and $\epsilon$-greedy DQN agent}
   \end{minipage}
@@
     \includesvg[width=\linewidth]{multi_way_v0__1__steps}
-    \subcaption{Planning horizon H = 1}
+    \subcaption{Planning horizon = 1}
   \end{minipage}
@@
     \includesvg[width=\linewidth]{multi_way_v0__2__steps}
-    \subcaption{Planning horizon H = 2}
+    \subcaption{Planning horizon = 2}
   \end{minipage}
@@
     \includesvg[width=\linewidth]{multi_way_v0__4__steps}
-    \subcaption{Planning horizon H = 4}
+    \subcaption{Planning horizon = 3}
   \end{minipage}
 
-  \caption{Detailed results for \texttt{multi\_way\_v0}.
-a)Comparison with DQN: blue - greedy policy, yellow - $\epsilon$-greedy policy, green - our method with planning horizon 4 and the last 1 goal tracked; b-d) comparison between tracking all goal states and only the last goal for the different planning horizon.
-Results are plotted with the moving average MA = 20 steps.} \label{fig_detailed_mw0}
+  \caption{Detailed results for \texttt{multi\_way\_v0}: comparison with DQN and with itself for different goal list sizes and planning horizons.}
 \end{figure}
 
-In the tested environments our method performed better than random policy.
-But the more complicated an environment was the less distinctive was the difference between the $H = 0$ and $H = 1$.
-Our method performs better with the increase of the planning horizon $H$.
-Each increment of the planning horizon is cumulative, i.e.
-each next increment is more advantageous.
+As the results of this set of experiments we concluded:
 
-Regarding the goal tracking memory size, we found that increase of the size leads to better exploration.
-We think that's because an agent less frequently uses random strategy and therefore moves less chaotically.
-However, this behavior may be suboptimal.
-For example when the rewarding goal state is reachable from the current position in less than $H$ actions but there's a fake goal state in the opposite direction that is closer to the agent.
-We think that these results are highly biased due to the special type of handcrafted mazes and initial conditions.
+\begin{itemize}
+  \item The longer planning horizon the better agent performs. And each increment is cumulative, i.e. each next increment is more advantageous.
 
-We also found that an agent learns and adapts faster than DQN agent (in terms of the number of episodes required to steadily reach the goal during the episode).
-And longer planning horizon leads to a better adaptation to the changed rewarding position.
+  \item Agent with planning performs better than fully random.
+But the more complicated an environment the less distinctive the difference between the absence of planning and \verb|horizon_planning = 1|.
+Even more, for a fixed planning horizon results are getting closer with the rising the size and complexity of an environment.
 
-When the planning horizon is big enough to reach the goal state from the starting position, then the agent learns the optimal policy very fast.
-According to our experiments, in this case it learns $\sim$1.5-2.5 times faster than DQN.
-But for any fixed planning horizon if we start rising an environment complexity, then at some point DQN starts to perform better than our method.
-That's because DQN always finds an optimal path, and our agent finds it only when it has enough planning horizon.
+  \item Agent with the longer goal list size better explores the environment.
+We think that's because agent less frequently uses random strategy.
+But in some cases this behavior is suboptimal.
+For example, when the rewarding goal state is reachable from the current position, but there's a fake goal state closer to the agent in the opposite direction.
+We think that these results are highly biased due to the special type of handcrafted mazes.
+
+  \item Agent with longer goal list and planning horizon better adapts to the changed reward position.
+  \item Agent learns and adapts faster than DQN agent [in terms of the required number of episodes].
+  \item If planning horizon is big enough to solve find a path from the starting position, then agent learns optimal policy very fast.
+In our experiments it learns ~1.5-2.5 times faster than DQN.
+But for any fixed planning horizon if we start rising an environment complexity at some point DQN start to perform better than our agent.
+That's because DQN always find an optimal path and our agent only when it has enough planning horizon.
+\end{itemize}
 
@@
 
-In this experiment setup we used a number of randomly generated GridWorld mazes of fixed size (examples provided in Figure ~\ref{fig_maze_examples}).
-Each experiment has the following scheme.
-It has been made sequentially in $N_{env}$ randomly generated mazes.
-For a fixed maze the rewarding goal position was changed sequentially $N_{rew}$ times.
-For a fixed rewarding goal position the initial position was changed sequentially $N_{s_0}$ times.
-And for a whole configuration being fixed agent played $N$ episodes.
-So, each experiment had $N_{env} \cdot N_{rew} \cdot N_{s_0} \cdot N$ total episodes.
-We tested an agent's ability not only to find the rewarding goal and adapt to its new position, but also how the agent adapts to the whole new maze.
+In this experiment setup we used a number of randomly generated GridWorld mazes of fixed size. Each experiment has a following scheme:
 
-We tested agent in mazes generated on $5 \times 5$, $6 \times 6$ and $8 \times 8$ square grids.
-In this section we present only the results for the $5 \times 5$ mazes, but for the other tested sizes the results are similar.
-The key difference in complexity of each experimental setup is the total number of the episodes with fixed reward and maze.
-The less episodes has agent to adapt, the more agile it's required to be.
+\begin{itemize}
+  \item it's made sequentially in $N_{env}$ randomly generated mazes
+  \item for a fixed maze the reward position is changed sequentially $N_{rew}$ times
+  \item for a fixed reward position the initial position is changed sequentially $N_{s_0}$ times
+  \item for a fixed whole configuration agent plays $N$ episodes.
+\end{itemize}
+
+So each experiment has $N_{env} \cdot N_{rew} \cdot N_{s_0} \cdot N$ total episodes.
+We tested an agent's ability not only to find reward and adapt to its new position, but also how it adapts to the whole new maze.
+
+We tested agent in mazes generated on $5 \times 5$, $6 \times 6$ and $8 \times 8$ square grids. Here we present only results for $5 \times 5$ mazes, but for other sizes the results are similar.
+
+\begin{figure}
+  \centering
+  \begin{minipage}{.49\linewidth}
+    \includesvg[width=\linewidth]{gridworld_5x5_1_1_200_1337_map_0_1173222464}
+  \end{minipage}
+  \begin{minipage}{.49\linewidth}
+    \includesvg[width=\linewidth]{gridworld_5x5_1_1_200_1337_map_2_1561234712}
+  \end{minipage}
+  \caption{Example of generated mazes of 5x5 and 8x8 sizes. Legend: dark purple - walls, yellow - starting agent's position, salad - reward.}
+\end{figure}
+
+The key difference in complexity of each experimental setup is the total number of episodes with fixed reward and maze. Less episodes has agent to adapt, the more agile it's required to be.
 
@@
 \begin{enumerate}
-  \item Experimental setup with the \textbf{rare} change of reward and maze: $N_{s_0} = 100$, $N_{rew} = 2$, $N_{env} = 8$.
+  \item Experimental setup with \textbf{rare} change of reward and maze
+  \begin{itemize}
+    \item Gridworld 5x5
+    \item $N_{s_0} = 100$
+    \item $N_{rew} = 2$
+    \item $N_{env} = 8$
+  \end{itemize}
 
-  \item Experimental setup with the \textbf{frequent} change of reward and maze: $N_{s_0} = 20$, $N_{rew} = 1$, $N_{env} = 20$.
+  \item Experimental setup with \textbf{frequent} change of reward and maze
+  \begin{itemize}
+    \item Gridworld 5x5
+    \item $N_{s_0} = 20$
+    \item $N_{rew} = 1$
+    \item $N_{env} = 20$
+  \end{itemize}
 \end{enumerate}
@@
 
-In contradiction to our results for the experiments in the fixed handcrafted environments we found that the increase of the number of the tracked goals makes our method to perform worse.
+In contradiction to our results for the experiments in fixed handcrafted environments we found that longer goal list makes our agent to perform worse.
 
@@
 
-We introduced a novel memory-based method that combines learning the model of the environment with the planning algorithm to infer a local policy.
-We showed that under certain conditions (enough planning horizon) the resulting global policy is optimal or at least very close to it.
-We also showed that in such cases, compared to DQN, our method learns with the comparable speed or slightly faster.
-The planning algorithm with any non-zero planning horizon performs better than random strategy.
-However, for small planning horizons and hard tasks results are sufficiently degraded and not that far from the pure random strategy.
-
-To model the environment dynamics we used human-like model of the memory called HTM.
-We showed that it's capable to quickly and reliably learn the transition model of the visited states in just a few episodes.
-As far as we know it is the first time HTM model was used for the model-based RL method.
-So, even though we applied it to model very simple deterministic environments with perfectly distinguishable states the results are promising, and we look forward to adapt our method to more complex environments.
-
 \bibliography{citations}
 
-\appendix
-
-\section{Appendix}
-\subsection{Handcrafted mazes} \label{apx_handcrafted_mazes}
-
-The notation we use in the schematic views of the environments: \verb|-| - empty cell, \verb|#| - wall, \verb|@| - agent, \verb|X| - reward.
-
-\subsubsection{multi\_way\_v0}
-
-Each experiment had 4 different rewarding places being sequentially changed every 100 episodes:
-
-\begin{verbatim}
-  #####   #####   #####   #####
-  #@--#   #@-X#   #@--#   #@--#
-  #-#-#   #-#-#   #-#-#   #-#X#
-  #--X#   #---#   #X--#   #---#
-  #####   #####   #####   #####
-\end{verbatim}
-
-\subsubsection{multi\_way\_v2}
-
-Each experiment had 6 rewarding places being changed every 200 episodes:
-
-\begin{verbatim}
-  #########  #########  #########  #########  #########  #########
-  #---#####  #-X-#####  #---#####  #---#####  #---#####  #---#####
-  #-#-#####  #-#-#####  #-#-#####  #-#-#####  #-#-#####  #-#-#####
-  #--@---##  #--@---##  #--@--X##  #--@---##  #X-@---##  #--@---##
-  ###--####  ###--####  ###--####  ###--####  ###--####  ###--####
-  ###-#--X#  ###-#---#  ###-#---#  ###-#---#  ###-#---#  ###-#--X#
-  ###---###  ###---###  ###---###  ###---###  ###---###  ###---###
-  ###-#####  ###-#####  ###-#####  ###X#####  ###-#####  ###-#####
-  #########  #########  #########  #########  #########  #########
-\end{verbatim}
-
 \end{document}

### 2020-11-01 reports/v1_0/report.tex

@@
 
-Рассматриваются два близких варианта GridWorld - с учетом направления взгляда агента и без.
-
-В первом случае в каждый момент времени агент находится в некоторой клетке лабиринта, указывая лицом одно из 4х направлений: восток, север, запад, юг.
-Из текущего состояния он может совершить следующие действия: шагнуть прямо (в соседнюю клетку), повернуться на 90 градусов по или против часовой стрелки.
-По умолчанию, в наших экспериментах одно из направлений поворота, по часовой стрелке, было запрещено.
-
-Во втором случае агент не имеет направления взгляда и может шагать непосредственно в любом из четырех направлений (восток, север, запад, юг), т.е. в одну из четырех соседних клеток.
+Агент может перемещаться в любую из соседних клеток, пронумерованных против часовой стрелки, начиная с востока: восток, север, запад, юг.
 
@@
 Переходы между состояниями однозначны - среда является детерминированной MDP.
-На каждом невыигрышном шаге агент получает небольшую отрицательную награду $-0.01$, на выигрышном - большую положительную $+1.0$.
+На каждом невыигрышном шаге агент получает небольшую отрицательную награду $-0.01$, на выигрышном - большую положительную $+1$.
+
+\begin{framed}
+  Изначально рассматривался еще один вариант GridWorld - с учетом направления взгляда агента.
+
+  В этом случае в каждый момент времени агент находится в некоторой клетке лабиринта, указывая лицом одно из 4х направлений: восток, север, запад, юг.
+Доступны следующие действия: шагнуть прямо (в соседнюю клетку), повернуться на 90 градусов по или против часовой стрелки.
+В некоторых экспериментах одно из направлений поворота - по часовой стрелке - было запрещено.
+
+  Такой вариант среды имеет некоторые, иногда предпочтительные, особенности: а) минимальное пространство действий (можно обойтись двумя действиями), б) усложнение задачи поиска награды в минимальных средах из-за более длинной последовательности действий до неё.
+Последнее удобно использовать для сравнения работы агента со случайной стратегией, т.к.
+даже в маленьких лабиринтах случайная стратегия выдает очень плохие результаты, и разница может быть хорошо заметна.
+Это в свою очередь важно на ранних быстрых циклах экспериментов.
+\end{framed}
 
@@
 
-В текущей реализации политика агента представляет собой смесь из двух политик - случайную и спланированную.
-Спланированную политику предоставляет планировщик.
+В текущей реализации стратегия агента представляет собой смесь из двух стратегий: случайной и спланированной.
+Спланированную стратегию предоставляет планировщик.
 В случае, если он не может её предоставить, используется случайная.
@@
 
-Задача планировщика - найти путь до награды, если она достижима за не более чем \verb|planning_horizon| шагов. Эта задача решается в два этапа:
+Задача планировщика - найти путь до награды, и этот путь не должен быть длиннее, чем \verb|planning_horizon| шагов. Последнее в свою очередь задает горизонт планирования.
+
+Процесс планирования разделен на два этапа:
 

### 2020-11-01 watcher/setup.py
Add files via upload
Setup file for watcher is added.

@@ -1,0 +1,12 @@
+from setuptools import setup, find_packages
+
+setup(
+    name='watcher',
+    version='1.0',
+    description='TM visualization tool',
+    url='https://github.com/cog-isa/htm-rl/watcher',
+    author='Artem Latyshev',
+    author_email='https://github.com/Arkol7',
+    license='MIT',
+    packages=find_packages(),
+)

### 2020-11-03 htm_rl/htm_rl/config.py
refine making reports for paper

@@ -106,7 +106,10 @@ class TestRunner:
 
         if aggregate:
             aggregate_file_masks = self.config['aggregate_masks']
-            run_results_processor.aggregate_results(aggregate_file_masks, report_name_suffix, silent_run)
+            for_paper = self.config['paper']
+            run_results_processor.aggregate_results(
+                aggregate_file_masks, report_name_suffix, silent_run, for_paper
+            )
 
 
 class TagProxies:

### 2020-11-03 htm_rl/htm_rl/make_gridworld_transfer_reports.sh

@@ -1,75 +1,32 @@
 #!/bin/zsh
 
-# env examples
-python ./run_mdp_tests.py -c ./experiments/gridworld_transfer/gridworld_5x5_1_1_200_1337.yml -d 3
-rm ./experiments/gridworld_transfer/gridworld_5x5_1_1_200_1337_map_1_*.svg
-python ./run_mdp_tests.py -c ./experiments/gridworld_transfer/gridworld_8x8_10_1_10_1337.yml -d 8
-rm ./experiments/gridworld_transfer/gridworld_8x8_10_1_10_1337_map_1_*.svg
-rm ./experiments/gridworld_transfer/gridworld_8x8_10_1_10_1337_map_2_*.svg
-rm ./experiments/gridworld_transfer/gridworld_8x8_10_1_10_1337_map_3_*.svg
-rm ./experiments/gridworld_transfer/gridworld_8x8_10_1_10_1337_map_4_*.svg
-rm ./experiments/gridworld_transfer/gridworld_8x8_10_1_10_1337_map_5_*.svg
-rm ./experiments/gridworld_transfer/gridworld_8x8_10_1_10_1337_map_6_*.svg
+# # env examples
+# python ./run_mdp_tests.py -c ./experiments/paper/gridworld_5x5_1_1_200_1337.yml -d 3
+# rm ./experiments/paper/gridworld_5x5_1_1_200_1337_map_1_*.svg
+# python ./run_mdp_tests.py -c ./experiments/paper/gridworld_8x8_10_1_10_1337.yml -d 8
+# rm ./experiments/paper/gridworld_8x8_10_1_10_1337_map_1_*.svg
+# rm ./experiments/paper/gridworld_8x8_10_1_10_1337_map_2_*.svg
+# rm ./experiments/paper/gridworld_8x8_10_1_10_1337_map_3_*.svg
+# rm ./experiments/paper/gridworld_8x8_10_1_10_1337_map_4_*.svg
+# rm ./experiments/paper/gridworld_8x8_10_1_10_1337_map_5_*.svg
+# rm ./experiments/paper/gridworld_8x8_10_1_10_1337_map_6_*.svg
 
 
 # gridworld_5x5_100_2_8_1337
[…]

### 2020-11-03 htm_rl/htm_rl/make_multi_way_transfer_reports.sh

@@ -1,45 +1,20 @@
 #!/bin/zsh
 
 # multi_way_v0
-python ./run_mdp_tests.py -c ./experiments/multi_way_transfer/multi_way_v0.yml -g htm_0 dqn_greedy "htm_*_1g" -s
+python ./run_mdp_tests.py -c ./experiments/paper/multi_way_v0.yml -g htm_0 dqn_greedy "htm_*_1g" -sp
 
-python ./run_mdp_tests.py -c ./experiments/multi_way_transfer/multi_way_v0.yml -g htm_1_1g htm_1 -n 1 -s
+python ./run_mdp_tests.py -c ./experiments/paper/multi_way_v0.yml -g htm_1_1g htm_1 -n 1 -sp
 
-python ./run_mdp_tests.py -c ./experiments/multi_way_transfer/multi_way_v0.yml -g htm_2_1g htm_2 -n 2 -s
+python ./run_mdp_tests.py -c ./experiments/paper/multi_way_v0.yml -g htm_2_1g htm_2 -n 2 -sp
 
-python ./run_mdp_tests.py -c ./experiments/multi_way_transfer/multi_way_v0.yml -g htm_4_1g htm_4 -n 4 -s
-
-python ./run_mdp_tests.py -c ./experiments/multi_way_transfer/multi_way_v0.yml -g dqn_greedy dqn_eps htm_4_1g -n dqn -s
-
-rm ./experiments/multi_way_transfer/*__*__rewards*
-rm ./experiments/multi_way_transfer/*__*__times*
-
-
-# multi_way_v1
-python ./run_mdp_tests.py -c ./experiments/multi_way_transfer/multi_way_v1.yml -g htm_0 dqn_greedy "htm_*_1g" -s
-
-python ./run_mdp_tests.py -c ./experiments/multi_way_transfer/multi_way_v1.yml -g htm_1_1g htm_1 -n 1 -s
-
-python ./run_mdp_tests.py -c ./experiments/multi_way_transfer/multi_way_v1.yml -g htm_2_1g htm_2 -n 2 -s
-
-python ./run_mdp_tests.py -c ./experiments/multi_way_transfer/multi_way_v1.yml -g htm_4_1g htm_4 -n 4 -s
-
-python ./run_mdp_tests.py -c ./experiments/multi_way_transfer/multi_way_v1.yml -g dqn_greedy dqn_eps htm_4_1g -n dqn -s
-
[…]

### 2020-11-03 htm_rl/htm_rl/run_mdp_tests.py

@@ -19,6 +19,7 @@ def main():
     parser.add_argument('-n', '--name', dest='report_name', default='')
     parser.add_argument('-d', '--dry', dest='dry', default=None, type=int)
     parser.add_argument('-s', '--silent', dest='silent', action='store_true', default=False)
+    parser.add_argument('-p', '--paper', dest='paper', action='store_true', default=False)
     args = parser.parse_args()
 
     config_path = Path(args.config)
@@ -27,6 +28,7 @@ def main():
     config['test_dir'] = test_dir
 
     config['silent'] = args.silent
+    config['paper'] = args.paper
 
     dry_run = args.dry is not None
     config['dry_run'] = dry_run

### 2020-11-03 reports/v1_0/paper_eng/paper_eng.tex
final version of paper

--- 2020-10-29 reports/v1_0/paper_eng/paper_eng.tex
@@
 
-We present results for two mazes: \texttt{multi\_way\_v0} and \texttt{multi\_way\_v2} (see Figure ~\ref{fig_comparison_mw} and ~\ref{fig_detailed_mw0}).
+We present results for two mazes: \texttt{multi\_way\_v0} and \texttt{multi\_way\_v2} (see Fig.~\ref{fig_comparison_mw} and Fig.~\ref{fig_detailed_mw0}).
 Each experiment in \texttt{multi\_way\_v0} had 4 different rewarding goal positions being sequentially changed every 100 episodes.
@@
   \centering
-  \begin{minipage}{.49\linewidth}
+  \begin{minipage}{.95\linewidth}
     \includesvg[width=\linewidth]{multi_way_v0__steps}
     \subcaption{\texttt{multi\_way\_v0}}
-  \end{minipage}
-  \begin{minipage}{.49\linewidth}
     \includesvg[width=\linewidth]{multi_way_v2__steps}
@@
   \end{minipage}
-  \caption{Comparison of the episode duration (timesteps) for the handcrafted mazes.
-\texttt{htm\_X\_Yg} is our agent with the planning horizon X and Y tracked last goals, \texttt{htm\_0} - random policy, \texttt{dqn\_greedy} - greedy DQN agent.
-Results are plotted with the moving average MA = 20 steps.} \label{fig_comparison_mw}
-\end{figure}
 
-\begin{figure}
-  \centering
-  \begin{minipage}{.49\linewidth}
-    \includesvg[width=\linewidth]{multi_way_v0__dqn__steps}
-    \subcaption{Greedy and $\epsilon$-greedy DQN agent}
-  \end{minipage}
-  \begin{minipage}{.49\linewidth}
-    \includesvg[width=\linewidth]{multi_way_v0__1__steps}
-    \subcaption{Planning horizon H = 1}
-  \end{minipage}
-
-  \begin{minipage}{.49\linewidth}
-    \includesvg[width=\linewidth]{multi_way_v0__2__steps}
-    \subcaption{Planning horizon H = 2}
-  \end{minipage}
-  \begin{minipage}{.49\linewidth}
-    \includesvg[width=\linewidth]{multi_way_v0__4__steps}
-    \subcaption{Planning horizon H = 4}
-  \end{minipage}
-
-  \caption{Detailed results for \texttt{multi\_way\_v0}.
-a)Comparison with DQN: blue - greedy policy, yellow - $\epsilon$-greedy policy, green - our method with planning horizon 4 and the last 1 goal tracked; b-d) comparison between tracking all goal states and only the last goal for the different planning horizon.
-Results are plotted with the moving average MA = 20 steps.} \label{fig_detailed_mw0}
+  \caption{Episode durations (timesteps) for the handcrafted mazes.
+Our agent performed consistently better than random policy, and with enough planning horizon it performed better than DQN agent.
+Notice, how fast the agent with planning horizon $8$ learns the optimal path compared to the DQN agent.} \label{fig_comparison_mw}
 \end{figure}
@@
 In the tested environments our method performed better than random policy.
-But the more complicated an environment was the less distinctive was the difference between the $H = 0$ and $H = 1$.
+But the more complicated an environment is the less distinctive is the difference between the $H = 0$ and $H = 1$.
 Our method performs better with the increase of the planning horizon $H$.
@@
 We think that's because an agent less frequently uses random strategy and therefore moves less chaotically.
-However, this behavior may be suboptimal.
+However, sometimes this behavior is suboptimal.
 For example when the rewarding goal state is reachable from the current position in less than $H$ actions but there's a fake goal state in the opposite direction that is closer to the agent.
@@
 
+\begin{figure}
+  \centering
+  \begin{minipage}{.95\linewidth}
+    \includesvg[width=\linewidth]{multi_way_v0__1__steps}
+    \subcaption{Planning horizon H = 1}
+    \includesvg[width=\linewidth]{multi_way_v0__2__steps}
+    \subcaption{Planning horizon H = 2}
+    \includesvg[width=\linewidth]{multi_way_v0__4__steps}
+    \subcaption{Planning horizon H = 4}
+  \end{minipage}
+
+  \caption{Detailed results for \texttt{multi\_way\_v0}. We compared two cases for the different planning horizons: tracking 16 rewarding goal states and only the last goal.} \label{fig_detailed_mw0}
+\end{figure}
+
 When the planning horizon is big enough to reach the goal state from the starting position, then the agent learns the optimal policy very fast.
-According to our experiments, in this case it learns $\sim$1.5-2.5 times faster than DQN.
+According to our experiments, in this case it learns approximately 1.5-2.5 times faster than DQN.
 But for any fixed planning horizon if we start rising an environment complexity, then at some point DQN starts to perform better than our method.
@@
 
-In this experiment setup we used a number of randomly generated GridWorld mazes of fixed size (examples provided in Figure ~\ref{fig_maze_examples}).
+In this experiment setup we used a number of randomly generated GridWorld mazes of fixed size (see Fig.~\ref{fig_maze_examples}).
 Each experiment has the following scheme.
@@
 
-We tested agent in mazes generated on $5 \times 5$, $6 \times 6$ and $8 \times 8$ square grids.
-In this section we present only the results for the $5 \times 5$ mazes, but for the other tested sizes the results are similar.
-The key difference in complexity of each experimental setup is the total number of the episodes with fixed reward and maze.
-The less episodes has agent to adapt, the more agile it's required to be.
-
-We present results for the following experimental setups (Figure 4):
-
-\begin{enumerate}
-  \item Experimental setup with the \textbf{rare} change of reward and maze: $N_{s_0} = 100$, $N_{rew} = 2$, $N_{env} = 8$.
-
-  \item Experimental setup with the \textbf{frequent} change of reward and maze: $N_{s_0} = 20$, $N_{rew} = 1$, $N_{env} = 20$.
-\end{enumerate}
-
 \begin{figure}
   \centering
-  \begin{minipage}{.49\linewidth}
+  \begin{minipage}{.9\linewidth}
     \includesvg[width=\linewidth]{gridworld_5x5_100_2_8_1337__steps}
@@
   \end{minipage}
-  \begin{minipage}{.49\linewidth}
+  \caption{Episode durations for the experiment with rarely changed environments.
+a) Our agent with the planning horizon 8 performed the best.
+Each change of the environment configuration there's a spiked degradation of performance for DQN agent, while our agents quickly relearn new position.
+b) According to our results in this task $\epsilon$-greedy DQN agent adapts much faster.
+c-d) Comparison for the different planning horizon values and the number of simultaneously tracked goals.
+} \label{fig_rare_changes}
+\end{figure}
+
+We tested agent in mazes generated on $5 \times 5$, $6 \times 6$ and $8 \times 8$ square grids.
+In this section we present only the results for the $5 \times 5$ mazes, but for the other tested sizes the results are similar.
+The key difference in complexity of each experimental setup is the total number of the episodes with fixed reward and maze.
+The less episodes has agent to adapt, the more agile it's required to be.
+
+We present results for the following experimental setups:
+
+\begin{enumerate}
+  \item Experimental setup with the \textbf{rare} change of reward and maze: $N_{s_0} = 100$, $N_{rew} = 2$, $N_{env} = 8$ (see Fig.~\ref{fig_rare_changes})
+
+  \item Experimental setup with the \textbf{frequent} change of reward and maze: $N_{s_0} = 20$, $N_{rew} = 1$, $N_{env} = 20$ (see Fig.~\ref{fig_frequent_changes}).
+\end{enumerate}
+
+\begin{figure}
+  \centering
+  \begin{minipage}{.9\linewidth}
     \includesvg[width=\linewidth]{gridworld_5x5_20_1_20_1337__steps}
@@
 
-  \caption{Results for experiments in changing environments. Left row: results for the setup with rare changes; right row: results for the setup with frequent changes.}
+  \caption{Episode durations for the experiment with frequently changed environments.
+a) Our method performed more stable with small planning horizons (1-4), while with planning horizon 8 it performed either very good or very bad.
+DQN agent results were worse than of random policy - the frequency of changes was too high for it to adapt in time.
+b) Unlike greedy DQN agent $\epsilon$-greedy agent performed at least comparable to our method.
+c-d) Comparison for the different planning horizon values and the number of simultaneously tracked goals.
+Larger goal tracking memory size degraded agent's performance.
+} \label{fig_frequent_changes}
 \end{figure}

### 2020-11-05 htm_rl/htm_rl/agent/memory.py
Watcher is incorporated to memory.

@@ -12,6 +12,8 @@ from htm_rl.common.sdr import SparseSdr
 from htm_rl.common.utils import trace
 from htm_rl.htm_plugins.temporal_memory import TemporalMemory
 
+from watcher.writer import Writer
+
 
 class Memory:
     tm: TemporalMemory
@@ -24,7 +26,8 @@ class Memory:
             sdr_formatter: Callable[[SparseSdr], str],
             sa_superposition_formatter: Callable[[SaSuperposition], str],
             collect_anomalies: bool = False,
-            start_indicator: Sa = None
+            start_indicator: Sa = None,
+            output_file: str = None
     ):
         self.start_indicator = start_indicator
         self.tm = tm
@@ -33,6 +36,12 @@ class Memory:
         self.format_sa_superposition = sa_superposition_formatter
         self.anomalies = [] if collect_anomalies else None
         self._proximal_input_sdr = SDR(self.tm.n_columns)
+
+        self.output_file = output_file
+        if self.output_file is not None:
+            self.writer = Writer(self.tm)
+        else:
+            self.writer = None
 
     def reset(self):
         self.tm.reset()
@@ -203,6 +212,13 @@ class Memory:
             sa_superposition = self.encoder.decode(proximal_input)
             trace(verbosity, req_level, self.format_sa_superposition(sa_superposition))
 
+    def print_tm_state(self, verbosity: int, req_level: int,  label: str):
+        if (self.output_file is None) or (verbosity < req_level):
+            return
+
+        self.writer.write(label)
+        self.writer.save(self.output_file)
+
     def save_tm_state(self):
         """Saves TM state."""
         return pickle.dumps(self.tm)
@@ -210,3 +226,5 @@ class Memory:
     def restore_tm_state(self, tm_dump):
         """Restores saved TM state."""
         self.tm = pickle.loads(tm_dump)
+
+

### 2020-11-05 htm_rl/htm_rl/agent/planner.py
Watcher prints are added to planner.

@@ -66,10 +66,12 @@ class Planner:
         reached_goals = []
         active_segments_timeline = []
 
-        for _ in range(self.planning_horizon):
+        for i in range(self.planning_horizon):
             active_cells, depolarized_cells = self.memory.process(
                 proximal_input, learn=False, verbosity=verbosity
             )
+
+            self.memory.print_tm_state(verbosity, 1, f'forward planning step: {i}')
 
             active_segments_t = self.memory.active_segments(active_cells)
             active_segments_timeline.append(active_segments_t)

### 2020-11-06 htm_rl/htm_rl/agent/mcts_agent.py
init mcts agent and planner

@@ -1,0 +1,244 @@
+from collections import deque
+from itertools import islice
+from typing import Any, List
+
+import numpy as np
+import matplotlib.pyplot as plt
+from tqdm import trange
+
+from htm_rl.agent.mcts_actor_critic import MctsActorCritic
+from htm_rl.agent.memory import Memory
+from htm_rl.agent.planner import Planner
+from htm_rl.agent.train_eval import RunStats, RunResultsProcessor
+from htm_rl.common.base_sa import Sa
+from htm_rl.common.utils import timed, trace
+from htm_rl.envs.gridworld_map_generator import GridworldMapGenerator
+from htm_rl.envs.gridworld_mdp import GridworldMdp
+from htm_rl.envs.mdp import Mdp
+
+
+class MctsAgent:
+    memory: Memory
+    planner: Planner
+    _n_actions: int
+    _planning_horizon: int
+
+    def __init__(
+            self, memory: Memory, planner: Planner,
+            mcts_actor_critic: MctsActorCritic, n_actions
+    ):
+        self.memory = memory
+        self.planner = planner
+        self._mcts_actor_critic = mcts_actor_critic
+        self._n_actions = n_actions
+        self._planning_horizon = self.planner.planning_horizon
+        self._planning_enabled = self._planning_horizon > 0
+        self._planned_actions = deque(maxlen=self._planning_horizon + 1)
+        self._goal_state = None
+
+    def set_planning_horizon(self, planning_horizon: int):
+        self.planner.planning_horizon = planning_horizon
+        self._planning_horizon = planning_horizon
+        self._planning_enabled = planning_horizon > 0
+        self._init_planning()
+
+    def reset(self):
+        self.memory.tm.reset()
+        self._init_planning()
+        if self._mcts_actor_critic is not None:
+            self._mcts_actor_critic.reset()
+            ac = self._mcts_actor_critic
+            q = np.round(ac.cell_value / ac.cell_visited_count, 2)
+            value_bit_buckets = (q[i: i + 8] for i in range(0, q.size, 8))
+            q_str = '\n'.join(
+                ' '.join(map(str, value_bits)) for value_bits in value_bit_buckets
+            )
+            trace(3, 1, q_str)
+            trace(3, 1, '=' * 20)
+
+    def make_step(self, state, reward, is_done, verbosity: int):
+        trace(verbosity, 2, f'\nState: {state}; reward: {reward}')
+        if self._mcts_actor_critic is not None:
[…]

### 2020-11-06 htm_rl/htm_rl/agent/mcts_planner.py

@@ -1,0 +1,402 @@
+from collections import deque
+from typing import List, Mapping, Set, Tuple, Deque
+
+from htm_rl.agent.memory import Memory
+from htm_rl.common.base_sa import Sa
+from htm_rl.common.int_sdr_encoder import IntSdrEncoder
+from htm_rl.common.sa_sdr_encoder import SaSdrEncoder
+from htm_rl.common.sdr import SparseSdr
+from htm_rl.common.utils import range_reverse, trace
+
+
+# noinspection PyPep8Naming
+
+
+class MctsPlanner:
+    memory: Memory
+    planning_horizon: int
+    encoder: SaSdrEncoder
+    episode_goal_memory: 'CircularSet'
+    inter_episode_goal_memory: 'CircularSet'
+
+    # segments[time][cell] = segments = [ segment ] = [ [presynaptic_cell] ]
+    _active_segments_timeline: List[Mapping[int, List[Set[int]]]]
+    _clusters_merging_threshold: int
+
+    def __init__(self, memory: Memory, planning_horizon: int, goal_memory_size: int):
+        self.memory = memory
+        self.encoder = memory.encoder
+        self.planning_horizon = planning_horizon
+        self.episode_goal_memory = CircularSet(goal_memory_size)
+        self.inter_episode_goal_memory = CircularSet(goal_memory_size)
+
+        self._active_segments_timeline = []
+        self._initial_tm_state = None
+
+        # TODO: WARN! Handcrafted threshold
+        self._clusters_merging_threshold = self.memory.tm.activation_threshold - 2
+
+    def plan_actions(self, initial_sa: Sa, verbosity: int):
+        planned_actions, goal_state = [], None
+        if self.planning_horizon == 0 or not self.episode_goal_memory:
+            return planned_actions, goal_state
+
+        # TODO: check if it should be there
+        if initial_sa.state in self.episode_goal_memory:
+            self.episode_goal_memory.remove(initial_sa.state)
+
+        trace(verbosity, 2, '\n======> Planning')
+
+        # saves initial TM state at the start of planning.
+        self._initial_tm_state = self.memory.save_tm_state()
+
+        reached_goals = self._predict_to_goals(initial_sa, verbosity)
+        planned_actions, goal_state = self._backtrack(reached_goals, verbosity)
+
+        trace(verbosity, 2, '<====== Planning complete')
+        return planned_actions, goal_state
+
[…]

### 2020-11-06 htm_rl/htm_rl/agent/memory.py
WIP mcts planning: draft for planner

@@ -116,8 +116,12 @@ class Memory:
         cpc = self.tm.cells_per_column
         left, right = columns_range
         # to cells range
-        left, right = left * cpc, right * cpc
-        return [cell for cell in cells if left <= cell < right]
+        cells_range = BitRange(left * cpc, right * cpc)
+        return [
+            cell
+            for cell in cells
+            if cell in cells_range
+        ]
 
     def filter_cells_by_columns(self, cells: SparseSdr, columns: SparseSdr) -> SparseSdr:
         columns = frozenset(columns)

### 2020-11-06 htm_rl/htm_rl/config.py

@@ -14,6 +14,8 @@ from htm_rl.agent.legacy_agent import LegacyAgent
 from htm_rl.agent.legacy_memory import LegacyMemory
 from htm_rl.agent.legacy_planner import LegacyPlanner
 from htm_rl.agent.mcts_actor_critic import MctsActorCritic
+from htm_rl.agent.mcts_agent import MctsAgent, MctsAgentRunner
+from htm_rl.agent.mcts_planner import MctsPlanner
 from htm_rl.agent.memory import Memory
 from htm_rl.agent.planner import Planner
 from htm_rl.agent.train_eval import RunResultsProcessor
@@ -175,6 +177,7 @@ def register_classes(yaml: YAML):
         GridworldMapGenerator,
         TransferLearningExperimentRunner2,
         RunResultsProcessor,
+        MctsAgent, MctsAgentRunner, MctsPlanner,
         MctsActorCritic,
     ]
 

### 2020-12-18 reports/v1_0/paper_eng/paper_eng.tex
wip: edit introduction and conclusions, polish figures

@@
 \usepackage{subcaption}
+\captionsetup[sub]{skip=1pt}
 \usepackage{calc}
@@
   \authorname{
-    Petr Kuderov\sup{1}\orcidAuthor{0000-0001-7055-3426}, Aleksandr I. Panov\sup{1}\orcidAuthor{0000-0002-9747-3837}
+    Petr Kuderov\sup{1}\orcidAuthor{0000-0001-7055-3426}, Aleksandr I. Panov\sup{1,2}\orcidAuthor{0000-0002-9747-3837}
   }
   \affiliation{
-    \sup{1}Moscow Institute of Physics and Technology, Russia
+    \sup{1}Moscow Institute of Physics and Technology, Moscow, Russia\\
+    \sup{2}Artificial Intelligence Research Institute, Federal Research Center ``Computer Science and Control'' of the Russian Academy of Sciences, Moscow, Russia
   }
-  \email{
-    kuderov.pv@phystech.edu, panov.ai@mipt.ru
-  }
+  \email{kuderov.pv@phystech.edu}
 }
@@
 Having a good model of the environment opens up such properties as data efficiency and targeted exploration.
-While most of the memory-based approaches are based on using Artificial Neural Networks (ANNs), in our work we instead draw the ideas from Hierarchical Temporary Memory (HTM) framework, which is based on human-like memory model.
+While most of the memory-based approaches are based on using Artificial Neural Networks (ANNs), in our work we instead draw the ideas from Hierarchical Temporal Memory (HTM) framework, which is based on human-like memory model.
 We utilize it to build an agent's memory that learns the environment dynamics.
@@
 
-Reinforcement Learning (RL) seek to find methods that enable robots or virtual agents to act autonomously in order to solve human-defined tasks.
+Reinforcement Learning (RL) seeks to find methods that enable robots or virtual agents to act autonomously in order to solve human-defined tasks.
 It requires from such methods to learn a policy for sequential decision-making problems, which are commonly formalized as MDP optimization \cite{Puterman_1994}.
-Recently RL has achieved notable landmarks in various task domains such as classical board games chess \cite{Campbell_Hoane_Hsu_2002_DeepBlue}, Go \cite{Silver_Go}, poker \cite{Brown_Sandholm_2017_Poker} and visually rich computer games like Atari \cite{Mnih_2015_Atari} and Dota 2 \cite{openai_2019_dota}.
+Recently, RL has achieved notable landmarks in various task domains such as classical board games chess \cite{Campbell_Hoane_Hsu_2002_DeepBlue}, Go \cite{Silver_Go}, poker \cite{Brown_Sandholm_2017_Poker} and visually rich computer games like Atari \cite{Mnih_2015_Atari} and Dota 2 \cite{openai_2019_dota}.
 At the core of some of the methods used, especially for board games, were planning algorithms, which rely on the knowledge of the environment dynamics.
+However, the model of the environment is unknown in general case, and there are two common opposite approaches in RL to deal with it: a model-free and a model-based.
 
-In general case, though, the model of the environment is unknown, and there are two common opposite approaches in RL to deal with it: a model-free and a model-based.
-The former approach doesn't explicitly learn and make use of the knowledge of the environment.
+Model-free approach doesn't explicitly learn and make use of the knowledge of the environment.
 Instead, it directly learns the global policy from interactions with the environment.
-While model-free approach has many successful examples \cite{Mnih_2015_Atari,haarnoja_2018_sac,schulman_2017_ppo}, among its disadvantages are data inefficiency, as it requires high amount of interactions with the environment, and struggle for precise lookahead.
+While model-free approach has many successful examples among its disadvantages are data inefficiency as it requires high amount of interactions with the environment and struggles for precise lookahead \cite{Mnih_2015_Atari,haarnoja_2018_sac,schulman_2017_ppo}.
 
-Model-based RL can address issues of both planning and model-free RL by first learning the environment dynamics, and then to exploit it for more efficient learning of the agent's policy \cite{moerland_2020_modelbased}.
+Model-based RL can address the issues of both planning and model-free RL by first learning the environment dynamics and then to exploit it for more efficient learning of the agent's policy \cite{moerland_2020_modelbased}.
 Typically, the model is represented by MDP and consists of two components: a state transition model and a reward model.
-After the model has been learned, MDP planning method such as MCTS \cite{Coulom_2007_mcts} can be applied to infer the optimal policy.
-Many model-based RL approaches build a model that operates at the pixel-level \cite{kaiser_2020_modelbased}, which is hungry for computational resources and attentive to unimportant details.
-Another approach is to try building a latent-space model.
-For example, a latent-space model can represent an abstract MDP that's equivalent to the real one in terms of state values, i.e.
-the cumulative reward of the planned trajectory \cite{silver_2020_muzero}.
+After the model has been learned, MDP planning method, such as MCTS \cite{Coulom_2007_mcts}, can be applied to infer the optimal policy.
+One model-based RL approach is to build a model that operates on the raw input data level \cite{Gorodetskiy2020,kaiser_2020_modelbased}.
+For visually rich environments this may be expensive in terms of computations and also may lead to a model attentive to unimportant details.
+Another approach is to build a more compact latent-space model instead, for example, to represent an abstract MDP equivalent to the real one in terms of state values \cite{silver_2020_muzero}.
 
-To model the environment it's common to use ANNs \cite{silver_2020_muzero,Ha_Schmidhuber_2018_worldmodels}.
-It's also very tempting to take more inspiration from human brains, especially for how our memory functions \cite{Hassabis_2017_neuro}.
-One example of such human-like memory model is Hierarchical Temporal Memory (HTM) framework \cite{George_Hawkins_2009}.
-This framework utilizes discrete binary sparse distributed representations (SDR) \cite{Cui_Ahmad_Hawkins_2017_sdr}, which enables more efficient data processing, high noise tolerance and a natural way to set similarity metric between data embeddings.
-It also provides a model of the sequential temporal memory (TM) \cite{hawkins_TM}.
+To model the environment it's common to use Artificial Neural Networks (ANNs) \cite{silver_2020_muzero,Ha_Schmidhuber_2018_worldmodels}.
+But it's also very tempting to take more inspiration from human brains, especially for how our memory functions \cite{Hassabis_2017_neuro}.
+Hierarchical Temporal Memory (HTM) framework \cite{George_Hawkins_2009} is an example of such human-like memory model.
+This framework utilizes discrete binary sparse distributed representations (SDR) \cite{Cui_Ahmad_Hawkins_2017_sdr} that enables more efficient data processing, high noise tolerance and a natural way to set similarity metric between data embeddings.
+HTM also provides a model of the sequential temporal memory (TM) \cite{hawkins_TM}.
 Learning in TM isn't based on the gradient descend, and typically it learns with a faster rate compared to common ANNs.
@@
 In our work we present a simplified memory-based framework based on the ideas of HTM that can be used to model the environment dynamics, supplied with the planning algorithm to infer the optimal global policy.
+Some attempts to extend the HTM model have been made before \cite{Skrynnik2016,Daylidyonok2018,Nugamanov2020}, but the full implementation of the planning subsystem with consideration for reinforcement signal was not implemented in other works.
 We study applicability of the presented memory model and viability of the planning method in different maze environments.
@@
 
-In this paper we consider the classic decision-making problem with an agent operating in an environment, which is formalized as Markov Decision Process (MDP).
-As a starting point in the research we simplify the problem assuming that the environment is fully observable and deterministic.
-We also consider only environments with the distinctive goal-oriented tasks, where an agent is expected to reach desirable goal states in a limited time.
+In this paper we consider the classic decision-making problem with an agent operating in an environment that is formalized as Markov Decision Process (MDP).
+As a starting point in our research we simplify the problem assuming that the environment is fully observable and deterministic.
+We also consider only environments with the distinctive goal-oriented tasks, where an agent is expected to reach desirable goal state in a limited time.
 
@@
 We don't use discount factor $\gamma$ in the return calculation.
-Instead, we assume that the reward function is designed the way that it highly rewards reaching the goal states and slightly punishes in any other states, which should reflect the goal-oriented nature of the environments that were taken into consideration.
+Instead, we assume that the reward function is designed the way that it highly rewards reaching the goal states and slightly punishes reaching any other states - it should reflect the goal-oriented nature of the environments that were taken into consideration.
 
@@
 Temporal Memory (TM) from Hierarchical Temporal Memory (HTM) framework is a model of the sequential memory.
-It works with sparse distributed representations (SDR), which are sparse binary vectors, typically high-dimensional.
+It works with sparse distributed representations (SDR), i.e.
+sparse binary vectors, which are typically high-dimensional.
 Therefore, the data encoding scheme should be chosen too.
+In \ref{sec_MemoryModule} we describe how agent's memory is organized on top of TM, so this section mostly covers the TM's input.
 
@@
 
-Temporal Memory model not only provides an algorithm to learn sequences of SDR vectors $[v_0, v_1, v_2, \dots ]$ but also to make predictions about what SDR vector $v'$ comes next based on the past experience.
-The prediction is represented also as an SDR vector, which is a union of SDR vectors $v' = \cup v_{o_i}$, where each vector $v_{o_i}$ represents a single separate expected outcome.
-Notice that as the union of SDR vectors is also an SDR vector, TM model allows you to work simultaneously with the unions of sequences as well.
+Temporal Memory model provides algorithms not only to learn sequences of SDR vectors $[v_0, v_1, v_2, \dots ]$ but also to make predictions about what SDR vector $v'$ comes next based on the past experience.
+The prediction is represented as an SDR vector too - it's a union of SDR vectors $v' = \cup v_{o_i}$, where each vector $v_{o_i}$ represents a single separate expected outcome.
+Notice that as the union of SDR vectors is also an SDR vector, TM model allows working simultaneously with the unions of sequences as well.
 
@@
 
-In our method we train an agent to reach rewarding goal states $s_g$.
-An agent is supplied with the memory module to learn the environment's dynamics and a planning algorithm which heavily relies on the learned model.
+In our method we train an agent to reach a rewarding goal state $s_g$.
+Agent isn't explicitly provided with the goal it should reach, but it's allowed to remember epxerienced rewarding states.
+An agent is supplied with the memory module to learn the environment's dynamics and a planning algorithm which heavily relies on the learned model (see Fig.
+\ref{fig_agent_scheme}).
 The planning algorithm constructs a plan of actions $p = [a_t, a_{t+1}, \dots, a_{t+h-1}]$ that leads to the desired state $s_g = s_{t+h}$ from the current state $s_t$.
-In the case when the planning algorithm fails to provide a plan we switch the agent to the exploration strategy $\pi_r$ in order to improve the learned model.
-As the result, at each timestep an agent operates either according to a constructed plan $p$ or according to the random uniformly-distributed exploration policy $\pi_r$.
+In case if the planning algorithm fails to provide a plan the agent is switched to the exploration strategy $\pi_r$ in order to improve the learned model.
+As the result at each timestep an agent operates either according to a constructed plan $p$ or according to the random uniformly-distributed exploration policy $\pi_r$.
 
@@
 The agent's memory module consists of two submodules.
-One submodule, called \textit{transition memory}, learns the transition model $f: \St \times \A \rightarrow \St$ of the environment, while the other, called \textit{goal states memory}, keeps track of the rewarding goal states $s_g$.
+One submodule, called \textit{transition memory}, learns the transition model $f: \St \times \A \rightarrow \St$ of the environment, while the other, called \textit{goal states memory}, keeps track of the past rewarding goal states $s_g$.
 
@@
 The transition memory submodule is based on a Temporal Memory.
-Recall that TM works with sparse distributed representations (SDRs).
 To encode each state-action pair $(s, a)$ we use two separate integer encoders for states and actions respectively to obtain binary sparse vectors.
@@
 To work with pairs $(s, a)$ we just concatenate their respective SDR vectors together into one.
-
 For any given state-action pair $(s, a)$ the transition memory submodule can give a prediction on what state $s'$ is coming next.
@@
 \begin{quote}
-  \textit{given a superposition \footnotemark of states $S = \cup s_i$ and a superposition of actions $A = \cup a_j$, what superposition of next states $S'= \cup s'_m$ all possible state-action pairs $(s_i, a_j)$ lead to?}
+  \textit{given a superposition of states $S = \cup s_i$ and a superposition of actions $A = \cup a_j$, what superposition of next states $S'= \cup s'_m$ all possible state-action pairs $(s_i, a_j)$ lead to?}
 \end{quote}
 
-  \footnotetext{In the Sec.~\ref{sec_TM} we explained that TM can work with unions of SDR vectors.
-When SDR vectors in a union aren't just random vectors, when they represent states or actions, we call such unions \textit{superpositions} of states or actions respectively.
-Note that once a superposition vector is made you cannot simply split it back into the origin SDR vectors of the union.
-\footnotemark}
+Here by \textit{superpositions} of states (or actions) we mean a union of their respective SDR vectors.
+By calling it not just a union we emphasize that the SDR vectors defining a union are meaningful.
+Note also that once a superposition SDR vector is obtained you cannot simply split it back into the origins formed it\footnotemark.
 
-\footnotetext{However, our current encoding scheme maps the different states or actions with non-overlapping bits, which keeps them easily distinguishable from each other even in a superposition.
-Whilst it's true, the current planning algorithm only requires a separation of the state bits from the action bits in a state-action pair SDR.
-This is done intentionally.
-Non-overlapping scheme makes things simpler to read and debug.
-The downside is it cannot keep the similarity between states (or observations).
-That's why for the future work, when we plan to experiment with more complex environments, transition to overlapping mapping seems inevitable.}
+\footnotetext{However, encoding scheme used maps different states or actions to non-overlapping sets of bits.
+That keeps them easily distinguishable from each other even in a superposition.
+Non-overlapping scheme makes things simpler to read and debug, but it has a crucial downside too - it cannot preserve possible similarity between states (or observations).
+So we made the planning algorithm only to require a separation of the state bits from the action bits in a state-action pair SDR $(s, a)$, which means it respects that origin SDRs cannot be induced from a superposition.
+In future works we plan experimenting with more complex environments and visually rich observations, therefore the transition to overlapping encoding scheme seems inevitable, hence it has already been reflected in the planning algorithm.}
 
-Under certain conditions the transition memory submodule can be used backward - as an inverse of an environment transition function $f$.
+Under certain conditions the transition memory submodule can also be used backward - as an inverse of an environment transition function $f$.
 This feature becomes available after getting the result to a prediction query and only for the states $s'_m$ that match the result $S' = \cup s'_m$.
@@
 We restrict resulting plan to have no more than $H$ actions.
-Therefore, $H$ defines planning horizon of an agent.
+Therefore, $H$ defines a planning horizon of an agent.
 
 Planning process is split into two stages.
-It starts with forward planning from the current state $s_t$ until a rewarding goal state $s_g$ is reached.
+It starts with the forward planning from the current state $s_t$ until a rewarding goal state $s_g$ is reached.
 At this stage the planning algorithm operates with superpositions and cannot track the exact path to the goal state.
-Because of that we switch it to backtracking stage, during which the exact path from the reached goal state back to the starting state $s_t$ is deduced.
+Because of that it proceeds then to the backtracking stage, during which the exact path from the reached goal state back to the starting state $s_t$ is deduced.
 
@@
 
-In our method an agent uses its experience and expects the reward to be in one of the rewarding goals states from previous episodes.
-But even if that's not true, we still expect that the direct goal-based exploration could be more optimal than pure random walking.
+In our method an agent uses its experience and expects the reward to be in one of the rewarding goals states from the previous episodes.
+But even if that's not true, we still expect that the direct goal-based exploration could be more optimal for an agent than pure random walking.
 Hence we set the agent to keep track of previous rewarding states using its goal tracking memory submodule.
@@
 
-In our experiments we decided to use classic GridWorld environments, which are represented as mazes on a square grid (see Figure ~\ref{fig_maze_examples}).
+In our experiments we decided to use classic GridWorld environments that are represented as mazes on a square grid (see Figure ~\ref{fig_maze_examples}).
 Each state can be defined with the position of an agent, i.e.
@@
 
+In our experiments we tested ability to model the environment and utilize learned model to plan actions that lead to the goal states.
 All experiments were divided into two groups.
 Each experiment from the first group was held in a fixed environment, while each experiment from the second group had an environment being changed every $N_{ep}$ episodes.
-
-In our experiments we tested ability to model the environment and utilize learned model to plan actions that lead to the goal states.
 We compared our method's performance with the baselines: random strategy and DQN (\cite{Mnih_2015_Atari}).
-We also compared performance for the different planning horizons and the number of simultaneously tracked goals in the goal tracking memory.
+We also compared performance for different planning horizons and the number of simultaneously tracked goals in the goal tracking memory.
 
@@
 
-For this experiment setup we used a number of handcrafted mazes.
+This experiment setup was based on a number of handcrafted mazes.
 Each experiment had a fixed maze and the starting point $s_0$, but the place of the reward was being changed every $N$ episodes.
 We tested an agent's ability to find the goal state and how it adapts to the new goal position.
+
+\begin{figure}
+  \centering
+  \begin{minipage}{\linewidth}
+    \includesvg[width=\linewidth]{multi_way_v0__steps}
+    \subcaption{}
+    \vspace*{12pt}
+
+    \includesvg[width=\linewidth]{multi_way_v2__steps}
+    \subcaption{}
+    \vspace*{6pt}
+  \end{minipage}
+
+  \caption{Episode durations (in steps) for handcrafted mazes \texttt{multi\_way\_v0} (a) and \texttt{multi\_way\_v2} (b).
+Our agent performed consistently better than random policy, and with enough planning horizon it performed better than DQN agent.
+Notice, how fast the agent with planning horizon $8$ learns the optimal path compared to the DQN agent.} \label{fig_comparison_mw}
+\end{figure}
 
@@
   \centering
-  \begin{minipage}{.95\linewidth}
-    \includesvg[width=\linewidth]{multi_way_v0__steps}
-    \subcaption{\texttt{multi\_way\_v0}}
-    \includesvg[width=\linewidth]{multi_way_v2__steps}
-    \subcaption{\texttt{multi\_way\_v2}}
+  \begin{minipage}{\linewidth}
+    \includesvg[width=\linewidth]{multi_way_v0__1__steps}
+    \subcaption{Planning horizon H = 1}
+    \vspace*{10pt}
+
+    \includesvg[width=\linewidth]{multi_way_v0__2__steps}
+    \subcaption{Planning horizon H = 2}
+    \vspace*{6pt}
   \end{minipage}
 
-  \caption{Episode durations (timesteps) for the handcrafted mazes.
-Our agent performed consistently better than random policy, and with enough planning horizon it performed better than DQN agent.
-Notice, how fast the agent with planning horizon $8$ learns the optimal path compared to the DQN agent.} \label{fig_comparison_mw}
+  \caption{Episode durations (in steps) for the agent on \texttt{multi\_way\_v0} depending on how many rewarding goal states it can track - last 16 or just the last one - grouped by the planning horizon length.} \label{fig_detailed_mw0}
 \end{figure}
@@
 
-\begin{figure}
-  \centering
-  \begin{minipage}{.95\linewidth}
-    \includesvg[width=\linewidth]{multi_way_v0__1__steps}
-    \subcaption{Planning horizon H = 1}
-    \includesvg[width=\linewidth]{multi_way_v0__2__steps}
-    \subcaption{Planning horizon H = 2}
-    \includesvg[width=\linewidth]{multi_way_v0__4__steps}
-    \subcaption{Planning horizon H = 4}
-  \end{minipage}
-
-  \caption{Detailed results for \texttt{multi\_way\_v0}. We compared two cases for the different planning horizons: tracking 16 rewarding goal states and only the last goal.} \label{fig_detailed_mw0}
-\end{figure}
-
-When the planning horizon is big enough to reach the goal state from the starting position, then the agent learns the optimal policy very fast.
+When the planning horizon is large enough to reach the goal state from the starting position, then the agent learns the optimal policy very fast.
 According to our experiments, in this case it learns approximately 1.5-2.5 times faster than DQN.
@@
 Each experiment has the following scheme.
-It has been made sequentially in $N_{env}$ randomly generated mazes.
-For a fixed maze the rewarding goal position was changed sequentially $N_{rew}$ times.
-For a fixed rewarding goal position the initial position was changed sequentially $N_{s_0}$ times.
-And for a whole configuration being fixed agent played $N$ episodes.
-So, each experiment had $N_{env} \cdot N_{rew} \cdot N_{s_0} \cdot N$ total episodes.
-We tested an agent's ability not only to find the rewarding goal and adapt to its new position, but also how the agent adapts to the whole new maze.
+Agent faces sequentially $N_{env}$ randomly generated mazes.
+For a fixed maze the rewarding goal position being changes sequentially $N_{rew}$ times.
+For a fixed rewarding goal position the initial position changes sequentially $N_{s_0}$ times.
+And for a whole configuration being fixed agent plays $N$ episodes.
+Thus each experiment has $N_{env} \cdot N_{rew} \cdot N_{s_0} \cdot N$ episodes in total.
+We have tested an agent's ability not only to find the rewarding goal and adapt to its new position, but also how the agent adapts to the whole new maze.
 
@@
   \centering
-  \begin{minipage}{.9\linewidth}
+  \begin{minipage}{\linewidth}
     \includesvg[width=\linewidth]{gridworld_5x5_100_2_8_1337__steps}
-    \subcaption{Comparison with baselines}
-    \includesvg[width=\linewidth]{gridworld_5x5_100_2_8_1337__dqn__steps}
-    \subcaption{Greedy and $\epsilon$-greedy DQN agent}
+    \subcaption{}
+    \vspace*{10pt}
+
+    % \includesvg[width=\linewidth]{gridworld_5x5_100_2_8_1337__dqn__steps}
+    % \subcaption{Greedy and $\epsilon$-greedy DQN agent}
+    % \vspace*{10pt}
+
     \includesvg[width=\linewidth]{gridworld_5x5_100_2_8_1337__2__steps}
-    \subcaption{Planning horizon: 2}
-    \includesvg[width=\linewidth]{gridworld_5x5_100_2_8_1337__4-8__steps}
-    \subcaption{Planning horizon: 4-8}
+    \subcaption{}
+    % \vspace*{10pt}
+
+    % \includesvg[width=\linewidth]{gridworld_5x5_100_2_8_1337__4-8__steps}
+    % \subcaption{}
+    \vspace*{4pt}
   \end{minipage}
   \caption{Episode durations for the experiment with rarely changed environments.
-a) Our agent with the planning horizon 8 performed the best.
-Each change of the environment configuration there's a spiked degradation of performance for DQN agent, while our agents quickly relearn new position.
-b) According to our results in this task $\epsilon$-greedy DQN agent adapts much faster.
-c-d) Comparison for the different planning horizon values and the number of simultaneously tracked goals.
-} \label{fig_rare_changes}
+a) Comparison with the baselines.
+Our agent with the planning horizon 8 performed the best.
+Each change of the environment configuration there's a spiked degradation of the performance for DQN agent, while our agents quickly relearn new position.
+b) Performance comparison for the different number of simultaneously tracked goals; planning horizon lengths are 4 and 8.} \label{fig_rare_changes}
 \end{figure}
@@
   \centering
-  \begin{minipage}{.9\linewidth}
+  \begin{minipage}{.95\linewidth}
     \includesvg[width=\linewidth]{gridworld_5x5_20_1_20_1337__steps}
-    \subcaption{Comparison with baselines}
-    \includesvg[width=\linewidth]{gridworld_5x5_20_1_20_1337__dqn__steps}
-    \subcaption{Greedy and $\epsilon$-greedy DQN agent}
+    \subcaption{}
+    \vspace*{10pt}
+
+    % \includesvg[width=\linewidth]{gridworld_5x5_20_1_20_1337__dqn__steps}
+    % \subcaption{Greedy and $\epsilon$-greedy DQN agent}
+    % \vspace*{10pt}
+
     \includesvg[width=\linewidth]{gridworld_5x5_20_1_20_1337__2__steps}
-    \subcaption{Planning horizon: 2}
-    \includesvg[width=\linewidth]{gridworld_5x5_20_1_20_1337__4-8__steps}
-    \subcaption{Planning horizon: 4-8}
+    \subcaption{}
+    % \vspace*{10pt}
+
+    % \includesvg[width=\linewidth]{gridworld_5x5_20_1_20_1337__4-8__steps}
+    % \subcaption{}
+    \vspace*{4pt}
   \end{minipage}
@@
   \caption{Episode durations for the experiment with frequently changed environments.
-a) Our method performed more stable with small planning horizons (1-4), while with planning horizon 8 it performed either very good or very bad.
-DQN agent results were worse than of random policy - the frequency of changes was too high for it to adapt in time.
-b) Unlike greedy DQN agent $\epsilon$-greedy agent performed at least comparable to our method.
-c-d) Comparison for the different planning horizon values and the number of simultaneously tracked goals.
-Larger goal tracking memory size degraded agent's performance.
-} \label{fig_frequent_changes}
+a) Comparison with the baselines.
+Our method performed more stable with small planning horizons (1-4), while with planning horizon 8 it performed either very good or very bad.
+DQN agent results were worse than of random policy - the frequency of changes was too high for it to adapt before the next change.
+b) Comparison for the different number of simultaneously tracked goals; planning horizon lengths are 4 and 8.} \label{fig_frequent_changes}
 \end{figure}
@@
 We showed that under certain conditions (enough planning horizon) the resulting global policy is optimal or at least very close to it.
-We also showed that in such cases, compared to DQN, our method learns with the comparable speed or slightly faster.
+We also found that in such cases, compared to DQN, our method learns with the comparable speed or slightly faster in terms of the number of episodes required.
 The planning algorithm with any non-zero planning horizon performs better than random strategy.
-However, for small planning horizons and hard tasks results are sufficiently degraded and not that far from the pure random strategy.
+However, for the small planning horizons and hard tasks results are sufficiently degraded and close to a pure random strategy.
 
 To model the environment dynamics we used human-like model of the memory called HTM.
-We showed that it's capable to quickly and reliably learn the transition model of the visited states in just a few episodes.
+We found that it's capable to quickly and reliably learn the transition model of the visited states in just a few episodes.
 As far as we know it is the first time HTM model was used for the model-based RL method.
 So, even though we applied it to model very simple deterministic environments with perfectly distinguishable states the results are promising, and we look forward to adapt our method to more complex environments.
+
+Our experiments showed us that the naive scaling of our method to environments with the larger state space is limited due to limitations for planning horizon increase.
+As for now we see two potential paths to remedy this problem.
+The first is to learn the model in a compact latent-space.
+The second is to make use of hierarchical approach by building a hierarchy of models and planners that operate in a different scales of time and space.
+Also the fact that the backing up strategy for a planner is purely random and not based on the agent's experience makes it a good target for further improvements too.
+
+\section*{\uppercase{Acknowledgements}}
+The reported study was supported by RFBR, research Projects No. 17-29-07051 and No. 18-29-22047.
 

### 2020-12-19 reports/v1_0/paper_eng/paper_eng.tex
fix figure captions and positioning

@@
   \end{minipage}
-  \caption{Example of generated maze environments of 5x5 and 8x8 sizes. Walls - dark purple, starting agent's position - yellow, rewarding goal state - salad.} \label{fig_maze_examples}
+  \caption{Two examples of generated 5x5 maze environments. Walls - dark purple, starting agent's position - yellow, rewarding goal state - salad.} \label{fig_maze_examples}
 \end{figure}
@@
   \caption{Episode durations (in steps) for handcrafted mazes \texttt{multi\_way\_v0} (a) and \texttt{multi\_way\_v2} (b).
-Our agent performed consistently better than random policy, and with enough planning horizon it performed better than DQN agent.
-Notice, how fast the agent with planning horizon $8$ learns the optimal path compared to the DQN agent.} \label{fig_comparison_mw}
+Results are shown with the moving average 20.
+Our agent performed consistently better than pure random policy, and with enough planning horizon it performed better than DQN agent.
+Notice, how fast the agent with the planning horizon $8$ learns an optimal path compared to the DQN agent.} \label{fig_comparison_mw}
 \end{figure}
@@
 
-  \caption{Episode durations (in steps) for the agent on \texttt{multi\_way\_v0} depending on how many rewarding goal states it can track - last 16 or just the last one - grouped by the planning horizon length.} \label{fig_detailed_mw0}
+  \caption{Episode durations (in steps) for the agent on \texttt{multi\_way\_v0} depending on how many rewarding goal states it can track - last 16 or just the last one - grouped by the planning horizon length.
+Results are shown with the moving average 20.} \label{fig_detailed_mw0}
 \end{figure}
@@
 
-    % \includesvg[width=\linewidth]{gridworld_5x5_100_2_8_1337__dqn__steps}
-    % \subcaption{Greedy and $\epsilon$-greedy DQN agent}
-    % \vspace*{10pt}
-
     \includesvg[width=\linewidth]{gridworld_5x5_100_2_8_1337__2__steps}
     \subcaption{}
-    % \vspace*{10pt}
-
-    % \includesvg[width=\linewidth]{gridworld_5x5_100_2_8_1337__4-8__steps}
-    % \subcaption{}
     \vspace*{4pt}
@@
   \caption{Episode durations for the experiment with rarely changed environments.
+Results are shown with the moving average 40.
 a) Comparison with the baselines.
 Our agent with the planning horizon 8 performed the best.
-Each change of the environment configuration there's a spiked degradation of the performance for DQN agent, while our agents quickly relearn new position.
-b) Performance comparison for the different number of simultaneously tracked goals; planning horizon lengths are 4 and 8.} \label{fig_rare_changes}
+Notice, how an each change of the environment configuration results in a spiked degradation of the performance for DQN agent, while our agent quickly relearns new position.
+b) Performance comparison for the different number of simultaneously tracked goals - the last goal or last 4; planning horizon length = 2.
+} \label{fig_rare_changes}
 \end{figure}
@@
 The less episodes has agent to adapt, the more agile it's required to be.
+
+\begin{figure}
+  \centering
+  \begin{minipage}{\linewidth}
+    \includesvg[width=\linewidth]{gridworld_5x5_20_1_20_1337__steps}
+    \subcaption{}
+    \vspace*{10pt}
+
+    \includesvg[width=\linewidth]{gridworld_5x5_20_1_20_1337__2__steps}
+    \subcaption{}
+    \vspace*{4pt}
+  \end{minipage}
+
+  \caption{Episode durations for the experiment with frequently changed environments.
+Results are shown with the moving average 40.
+a) Comparison with the baselines.
+Our method performed more stable with small planning horizons (1-4), while with planning horizon 8 it performed either very good or very bad.
+DQN agent results were worse than of random policy - the frequency of changes was too high for it to adapt before the next change.
+b) Comparison for the different number of simultaneously tracked goals; planning horizon length = 2.} \label{fig_frequent_changes}
+\end{figure}
 
@@
 \end{enumerate}
-
-\begin{figure}
-  \centering
-  \begin{minipage}{.95\linewidth}
-    \includesvg[width=\linewidth]{gridworld_5x5_20_1_20_1337__steps}
-    \subcaption{}
-    \vspace*{10pt}
-
-    % \includesvg[width=\linewidth]{gridworld_5x5_20_1_20_1337__dqn__steps}
-    % \subcaption{Greedy and $\epsilon$-greedy DQN agent}
-    % \vspace*{10pt}
-
-    \includesvg[width=\linewidth]{gridworld_5x5_20_1_20_1337__2__steps}
-    \subcaption{}
-    % \vspace*{10pt}
-
-    % \includesvg[width=\linewidth]{gridworld_5x5_20_1_20_1337__4-8__steps}
-    % \subcaption{}
-    \vspace*{4pt}
-  \end{minipage}
-
-  \caption{Episode durations for the experiment with frequently changed environments.
-a) Comparison with the baselines.
-Our method performed more stable with small planning horizons (1-4), while with planning horizon 8 it performed either very good or very bad.
-DQN agent results were worse than of random policy - the frequency of changes was too high for it to adapt before the next change.
-b) Comparison for the different number of simultaneously tracked goals; planning horizon lengths are 4 and 8.} \label{fig_frequent_changes}
-\end{figure}
 

### 2020-12-19 reports/v1_0/talk_eng/slides.tex
paper: draft for slides

\documentclass[default]{beamer}
\usepackage{slides_style}
\setbeamertemplate{navigation symbols}{}
\usetheme{Boadilla}

\begin{document}

\title[]{
	Planning with Hierarchical Temporal Memory for Deterministic Markov Decision Problem
}
\author[Petr Kuderov]{
    Petr Kuderov, Aleksandr I. Panov
}
\institute[MIPT]{
    Moscow Institute of Physics and Technology, Moscow, Russia\\
	kuderov.pv@phystech.edu
}
\date[]{Feb 2021}

\begin{frame}
	\titlepage
	\centering
	\includegraphics[height=20pt]{mipt_logo} \hspace{10pt}
	\includegraphics[height=20pt]{cdsl_logo} \hspace{10pt}
\end{frame}

\begin{frame}{Introduction and motivation}
	\begin{itemize}
		\item why we need a model
		\item why we draw inspirations from human memory models
	\end{itemize}
\end{frame}

\begin{frame}{Research question}
	- Check applicability of TM
  - Build planning method
\end{frame}

\begin{frame}{HTM and TM}
	SDR, prediction and backward prediction	
\end{frame}

\begin{frame}{Agent's architecture}
	- figure
	- explain how it works on a high level	
\end{frame}

\begin{frame}{Planning, stages}
	- the goal of planning
	- two stages, what their goals
	- forward planning: goal reachability and keep track of predictions
	- backtracking: find the exact path to the goal
\end{frame}

\begin{frame}{Forward planning}
	explain how it's done
\end{frame}

\begin{frame}{Backtracking}
	explain recursive algo
\end{frame}

\begin{frame}{Results. Experimental setup 1}
	- explain setup
	- figures
	- results	
\end{frame}

\begin{frame}{Results. Experimental setup 2}
	- explain setup
	- figures
	- results	
\end{frame}

\begin{frame}{Conclusions}
	regarding research questions
\end{frame}

\begin{frame}{Limitations and possible improvements}
	- deal with planning horizon: two ways
	- regading random strategy	
\end{frame}

\end{document}

### 2020-12-19 reports/v1_0/paper_eng/paper_eng.tex
polishing paper text

@@
 Reinforcement Learning (RL) seeks to find methods that enable robots or virtual agents to act autonomously in order to solve human-defined tasks.
-It requires from such methods to learn a policy for sequential decision-making problems, which are commonly formalized as MDP optimization \cite{Puterman_1994}.
+It requires from such methods to learn a policy for sequential decision-making problems  which are commonly formalized as MDP optimization \cite{Puterman_1994}.
 Recently, RL has achieved notable landmarks in various task domains such as classical board games chess \cite{Campbell_Hoane_Hsu_2002_DeepBlue}, Go \cite{Silver_Go}, poker \cite{Brown_Sandholm_2017_Poker} and visually rich computer games like Atari \cite{Mnih_2015_Atari} and Dota 2 \cite{openai_2019_dota}.
@@
 Instead, it directly learns the global policy from interactions with the environment.
-While model-free approach has many successful examples among its disadvantages are data inefficiency as it requires high amount of interactions with the environment and struggles for precise lookahead \cite{Mnih_2015_Atari,haarnoja_2018_sac,schulman_2017_ppo}.
+While model-free approach has many successful examples, among its disadvantages are data inefficiency as it requires high amount of interactions with the environment and struggles for precise lookahead \cite{Mnih_2015_Atari,haarnoja_2018_sac,schulman_2017_ppo}.
 
@@
 After the model has been learned, MDP planning method, such as MCTS \cite{Coulom_2007_mcts}, can be applied to infer the optimal policy.
-One model-based RL approach is to build a model that operates on the raw input data level \cite{Gorodetskiy2020,kaiser_2020_modelbased}.
+One model-based RL approach is to build model that operates on the raw input data level \cite{Gorodetskiy2020,kaiser_2020_modelbased}.
 For visually rich environments this may be expensive in terms of computations and also may lead to a model attentive to unimportant details.
 Another approach is to build a more compact latent-space model instead, for example, to represent an abstract MDP equivalent to the real one in terms of state values \cite{silver_2020_muzero}.
+While it can help dealing with drawbacks of the former approach, it also requires additional efforts to ensure that sufficiently good latent-space mapping is learned.
+To model the environment it's common to use Artificial Neural Networks (ANNs) \cite{silver_2020_muzero,Ha_Schmidhuber_2018_worldmodels}.
+It's also very tempting to take more inspiration from human brains, especially for how our memory functions \cite{Hassabis_2017_neuro}.
 
-To model the environment it's common to use Artificial Neural Networks (ANNs) \cite{silver_2020_muzero,Ha_Schmidhuber_2018_worldmodels}.
-But it's also very tempting to take more inspiration from human brains, especially for how our memory functions \cite{Hassabis_2017_neuro}.
 Hierarchical Temporal Memory (HTM) framework \cite{George_Hawkins_2009} is an example of such human-like memory model.
-This framework utilizes discrete binary sparse distributed representations (SDR) \cite{Cui_Ahmad_Hawkins_2017_sdr} that enables more efficient data processing, high noise tolerance and a natural way to set similarity metric between data embeddings.
-HTM also provides a model of the sequential temporal memory (TM) \cite{hawkins_TM}.
-Learning in TM isn't based on the gradient descend, and typically it learns with a faster rate compared to common ANNs.
+This framework utilizes discrete binary sparse distributed representations (SDR) that enable more efficient data processing, high noise tolerance and a natural way to set similarity metric between data embeddings \cite{Cui_Ahmad_Hawkins_2017_sdr}.
+HTM also provides a model of the sequential temporal memory (TM) \cite{hawkins_TM,Ahmad_Lavin_2017_data_streams}.
+Learning in TM is not based on gradient descend, and typically it learns with a faster rate compared to common ANNs \cite{Cui_Ahmad_Hawkins_2016_seq_learn}.
 
-In our work we present a simplified memory-based framework based on the ideas of HTM that can be used to model the environment dynamics, supplied with the planning algorithm to infer the optimal global policy.
+In our work we present a simplified memory-based framework based on the ideas of HTM that can be used to model the environment dynamics.
+Great part of our contribution is that we also supply it with the planning algorithm that enables an agent to infer the optimal global policy.
 Some attempts to extend the HTM model have been made before \cite{Skrynnik2016,Daylidyonok2018,Nugamanov2020}, but the full implementation of the planning subsystem with consideration for reinforcement signal was not implemented in other works.
@@
 
-Temporal Memory (TM) from Hierarchical Temporal Memory (HTM) framework is a model of the sequential memory.
+Temporal Memory (TM) from Hierarchical Temporal Memory (HTM) framework is a model of sequential memory.
 It works with sparse distributed representations (SDR), i.e.
 sparse binary vectors, which are typically high-dimensional.
-Therefore, the data encoding scheme should be chosen too.
-In \ref{sec_MemoryModule} we describe how agent's memory is organized on top of TM, so this section mostly covers the TM's input.
+In general, an input data is not represented as SDR vectors.
+Therefore, the data encoding scheme should be chosen too, and this section mostly covers questions related TM's input.
+Further in \ref{sec_MemoryModule} we describe how agent's memory is organized on top of TM.
 
-In our work both actions and states are represented as integer numbers from a fixed range (although, different for states and actions).
-Thus, we use a simple encoding scheme, where numbers from some range $[0, N)$ are mapped to non-overlapping equally-sized sets of active bits of the output SDR vector.
-For example for $N = 3$ numbers 0, 1 and 2 are encoded correspondingly as following:
+In our work both actions and states are represented as integer numbers from a fixed range (although, different for states and for actions).
+Thus, we use simple encoding scheme, where numbers from a range $[0, N)$ are mapped to non-overlapping equally-sized sets of active bits of the output SDR vector.
+For example for $N = 3$ numbers from the range $[0, 3)$ are encoded correspondingly as following:
 
@@
 
-As you see, resulting vector is split into buckets of bits, and the size of buckets is a hyperparameter (in the example above it's 4).
+As you see, resulting vector is split into buckets of 4 bits. In general the size of buckets is a hyperparameter (we mostly used 8 or 10).
 
-For SDR vectors a union operation is defined as a bitwise OR applied to the corresponding vectors.
-Also we define a similarity metric, which is the number of the same one-bits of the vectors in consideration, which is, again, the same as the dot product of such vectors.
+A union operation is defined for SDR vectors - a bitwise OR applied to the corresponding vectors.
+We also define similarity metric, which is the number of the same one-bits of the vectors in consideration.
+Or it can be seen as the dot product of such vectors, which among other things draws parallels between semantics and geometry, e.g.
+non-related SDR vectors are orthogonal.
 
@@
 Each experiment in \texttt{multi\_way\_v0} had 4 different rewarding goal positions being sequentially changed every 100 episodes.
-Each experiment in \texttt{multi\_way\_v2} had 6 rewarding goal positions being changed every 200 episodes.
+Each experiment in \texttt{multi\_way\_v2} had 6 rewarding goal positions being changed every 100 episodes.
 Schematic view of these mazes is provided in Appendix ~\ref{apx_handcrafted_mazes}.
@@
 
-\texttt{multi\_way\_v2}. Each experiment had 6 rewarding places being changed every 200 episodes:
+\texttt{multi\_way\_v2}. Each experiment had 6 rewarding places being changed every 100 episodes:
 

### 2021-01-31 reports/v1_0/paper_eng/paper_eng.tex

@@
 In our work we present a simplified memory-based framework based on the ideas of HTM that can be used to model the environment dynamics.
-Great part of our contribution is that we also supply it with the planning algorithm that enables an agent to infer the optimal global policy.
+One of the main parts of our contribution is that we also supply it with the planning algorithm that enables an agent to infer the optimal global policy.
 Some attempts to extend the HTM model have been made before \cite{Skrynnik2016,Daylidyonok2018,Nugamanov2020}, but the full implementation of the planning subsystem with consideration for reinforcement signal was not implemented in other works.
@@
 We introduced a novel memory-based method that combines learning the model of the environment with the planning algorithm to infer a local policy.
-We showed that under certain conditions (enough planning horizon) the resulting global policy is optimal or at least very close to it.
+It is shown that under certain conditions (enough planning horizon) the resulting global policy is optimal or at least very close to it.
 We also found that in such cases, compared to DQN, our method learns with the comparable speed or slightly faster in terms of the number of episodes required.
-The planning algorithm with any non-zero planning horizon performs better than random strategy.
+The planning algorithm with any non-zero planning horizon performs substantially better than random strategy.
 However, for the small planning horizons and hard tasks results are sufficiently degraded and close to a pure random strategy.
@@
 
-Our experiments showed us that the naive scaling of our method to environments with the larger state space is limited due to limitations for planning horizon increase.
-As for now we see two potential paths to remedy this problem.
-The first is to learn the model in a compact latent-space.
-The second is to make use of hierarchical approach by building a hierarchy of models and planners that operate in a different scales of time and space.
+Our experiments showed that the naive scaling of our method to environments with the larger state space is bounded due to limitations for planning horizon increase.
+As for now we see two potential paths to relieve the limits.
+One is to learn the model in a compact latent-space.
+Another is to make use of hierarchical approach by building a hierarchy of models and planners that operate in a different scales of time and space.
 Also the fact that the backing up strategy for a planner is purely random and not based on the agent's experience makes it a good target for further improvements too.

### 2021-01-31 htm_rl/htm_rl/experimental/list_sdr_encoder.py
fix changelog links; remove outdated `experimental` folder

@@ -1,88 +1,0 @@
-# from typing import List, NamedTuple, Iterable
-#
-# from htm_rl.minigrid.sar import SuperpositionList
-# from htm_rl.common.int_sdr_encoder import IntSdrEncoder, BitRange
-# from htm_rl.common.sdr import SparseSdr
-# from htm_rl.common.utils import isnone
-#
-# Dim2d = NamedTuple('Dim2d', (('rows', int), ('cols', int)))
-#
-#
-# """
-# !!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!
-# Outdated and probably doesn't work at all.
-# Kept for future needs. Will be adapted or deleted when the time is come.
-# !!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!
-# """
-#
-#
-# class ListSdrEncoder:
-#     value_bits: int
-#     total_bits: int
-#     activation_threshold: int
-#
-#     _elem_encoder: IntSdrEncoder
-#     _n_elems: int
-#     _n_dim: Dim2d
-#     _shifts: List[int]
-#
-#     def __init__(self, elem_encoder: IntSdrEncoder, n_dim: Dim2d):
-#         n_rows, n_cols = n_dim
-#         n_elems = n_rows * n_cols
-#
-#         self.value_bits = n_elems * elem_encoder.value_bits
-#         self.total_bits = n_elems * elem_encoder.total_bits
-#         self.activation_threshold = n_elems * elem_encoder.activation_threshold
-#
-#         self._elem_encoder = elem_encoder
-#         self._n_elems = n_elems
-#         self._n_dim = n_dim
-#         self._shifts = self._get_shifts(elem_encoder, n_elems)
-#
-#     def encode(self, values) -> Iterable[BitRange]:
-#         values = isnone(values, [])
-#         return (
-#             BitRange(bit_range.l + shift, bit_range.r + shift)
-#             for x, shift in zip(values, self._shifts)
-#             for bit_range in self._elem_encoder.encode(x)
-#         )
-#
-#     def decode(self, indices: SparseSdr) -> SuperpositionList:
-#         elem_sdrs = self.split_to_elems(indices)
-#         return list(map(self._elem_encoder.decode, elem_sdrs))
-#
-#     def split_to_elems(self, indices: SparseSdr) -> List[SparseSdr]:
-#         elem_buckets = [[] for _ in range(self._n_elems)]
-#         for ind in indices:
[…]

### 2021-01-31 htm_rl/htm_rl/experimental/run_gridworld_investigation_test.py

@@ -1,243 +1,0 @@
-# import random
-#
-# import matplotlib.pyplot as plt
-# import numpy as np
-#
-# from htm_rl.agent import Agent
-# from htm_rl.experimental.list_sdr_encoder import Dim2d
-# from htm_rl.envs.minigrid import make_minigrid
-# from htm_rl.mdp.sar import Sar, SarSuperpositionFormatter
-# from htm_rl.mdp.sar_sdr_encoder import SarSdrEncoder
-# from htm_rl.agent.planner import Planner
-# from htm_rl.common.int_sdr_encoder import IntSdrEncoderShortFormat as IntSdrEncoder
-# from htm_rl.htm_plugins.temporal_memory import TemporalMemory
-#
-#
-# """
-# !!!!!!!!!!!!!!!!!!!!!!!!!!!
-# It's outdated and doesn't work RN as for now minigrid is not a main target for testing.
-#
-# Don't delete as it contains some useful code. It will be updated at some point in future.
-# !!!!!!!!!!!!!!!!!!!!!!!!!!!
-#
-#
-# Here you can test agent+planner on easy MDP variations of minigrid envs.
-# """
-#
-#
-# def render_env(env, render: bool, pause: float = None):
-#     if not render:
-#         return
-#
-#     plt.imshow(env.render('rgb_array'))
-#     if pause is None or pause < .01:
-#         plt.show(block=True)
-#     else:
-#         plt.show(block=False)
-#         plt.pause(pause)
-#
-#
-# def print_debug_sar(sar, encoder, sar_formatter):
-#     indices = encoder.encode(sar)
-#     print(encoder.format_sar_superposition(indices))
-#     sar_superposition = encoder.decode(indices)
-#     print(sar_formatter.format_sar_superposition(sar_superposition))
-#
-#
-# def print_proximal_input(proximal_input, encoder, print_enabled):
-#     if print_enabled:
-#         print(encoder.format_sar_superposition(proximal_input))
-#
-# def get_all_observations(actions, env, render, pause):
-#     observation = env.reset()
-#     observations = [observation]
-#     for action in actions:
-#         render_env(env, render, pause)
-#
-#         next_observation, reward, done, info = env.step(action)
-#         observation = next_observation
-#         observations.append(observation)
-#
-#     return observations
-#
-#
-# def get_observations_mapping(observations):
-#     observations = np.array(observations)
-#     observations = np.unique(observations, axis=0)
[…]

### 2021-01-31 htm_rl/htm_rl/experimental/run_gridworld_test.py

@@ -1,148 +1,0 @@
-# import random
-#
-# import matplotlib.pyplot as plt
-# import numpy as np
-#
-# from htm_rl.agent import Agent
-# from htm_rl.experimental.list_sdr_encoder import ListSdrEncoder, Dim2d
-# from htm_rl.envs.minigrid import make_minigrid
-# from htm_rl.agent.planner import Planner
-# from htm_rl.minigrid.sar import Sar, SarSuperpositionFormatter
-# from htm_rl.minigrid.sar_sdr_encoder import SarSdrEncoder
-# from htm_rl.common.int_sdr_encoder import IntSdrEncoder
-# from htm_rl.htm_plugins.temporal_memory import TemporalMemory
-#
-#
-# """
-# !!!!!!!!!!!!!!!!!!!!!!!!!!!
-# It's outdated and doesn't work RN as for now minigrid is not a main target for testing.
-#
-# Don't delete as it contains some useful code. It will be updated at some point in future.
-# !!!!!!!!!!!!!!!!!!!!!!!!!!!
-#
-#
-# Here you can test agent+planner on easy POMDP minigrid envs.
-# """
-#
-#
-# def render_env(env, render: bool, pause: float = None):
-#     if not render:
-#         return
-#
-#     plt.imshow(env.render('rgb_array'))
-#     if pause is None or pause < .01:
-#         plt.show(block=True)
-#     else:
-#         plt.show(block=False)
-#         plt.pause(.1)
-#
-#
-# def print_debug_sar(sar, encoder, sar_formatter):
-#     indices = encoder.encode(sar)
-#     print(encoder.format_sar_superposition(indices))
-#     sar_superposition = encoder.decode(indices)
-#     print(sar_formatter.format_sar_superposition(sar_superposition))
-#
-#
-# def train_for(n_steps, observation, reward, a_ind, print_enabled):
-#     reward_reached = 0
-#     done = False
-#     for _ in range(n_steps):
-#         if np.random.rand() < 2:
-#             action = np.random.choice(3)
-#         else:
-#             action = actions[a_ind % len(actions)]
-#         a_ind += 1
-#
-#         render_env(env, render, pause)
-#         sar = Sar(observation, action, reward)
-#         proximal_input = encoder.encode(sar)
-#         agent.train_one_step(proximal_input, print_enabled)
-#         # print_debug_sar(sar, encoder, sar_formatter)
-#
[…]

### 2021-01-31 htm_rl/htm_rl/experimental/tests.py

@@ -1,112 +1,0 @@
-from inspect import ismethod
-
-import numpy as np
-from htm.bindings.sdr import SDR
-#
-# from mdp_planner import DataEncoder, DataMultiEncoder, TemporalMemory, HtmAgent
-#
-#
-# class TestDataEncoder:
-#     def __init__(self):
-#         self.encoder = DataEncoder('-', n_vals=2, value_bits=3, activation_threshold=2)
-#
-#     def test_encode_dense(self):
-#         result = self.encoder.encode_dense(1)
-#         expected = np.array([0, 0, 0, 1, 1, 1], dtype=np.int8)
-#         assert np.array_equal(result, expected)
-#
-#     def test_encode_sparse(self):
-#         arr_sparse = self.encoder.encode_sparse(1)
-#         assert arr_sparse == [3, 4, 5]
-#
-#     def test_str_from_dense(self):
-#         arr_dense = np.array([0, 0, 0, 1, 1, 1], dtype=np.int8)
-#         res = self.encoder.str_from_dense(arr_dense)
-#         assert res == '000 111'
-#
-#     def test_decode_dense(self):
-#         decoded = self.encoder.decode_dense(np.array([0, 1, 1, 0, 1, 0]))
-#         assert decoded == [0]
-#
-#         decoded = self.encoder.decode_dense(np.array([0, 1, 1, 1, 1, 1]))
-#         assert decoded == [0, 1]
-#
-#     def test_decode_sparse(self):
-#         decoded = self.encoder.decode_sparse([1, 2, 4])
-#         assert decoded == [0]
-#
-#         decoded = self.encoder.decode_sparse([1, 2, 3, 4, 5])
-#         assert decoded == [0, 1]
-#
-#     def test_to_str(self):
-#         assert str(self.encoder) == 'DataEncoder("-", v2 x b3)'
-#
-#
-# class TestDataMultiEncoder:
-#     def __init__(self):
-#         self.encoder = DataMultiEncoder((
-#             DataEncoder('1', 2, 4),
-#             DataEncoder('2', 3, 2)
-#         ))
-#
-#     def test_encode_dense(self):
-#         result = self.encoder.encode_dense((0, 2))
-#         expected = np.array([1, 1, 1, 1, 0, 0, 0, 0, 0, 0, 0, 0, 1, 1], dtype=np.int8)
-#         assert np.array_equal(result, expected)
-#
-#     def test_encode_sparse(self):
-#         result = self.encoder.encode_sparse((1, 1))
[…]

### 2021-01-31 htm_rl/htm_rl/common/utils.py
init map generator

@@ -25,6 +25,14 @@ def trace(verbosity: int, req_level: int, str_to_print: str = ''):
         print(str_to_print)
 
 
+def clip(x, high):
+    if x >= high:
+        return high - 1
+    if x < 0:
+        return 0
+    return x
+
+
 def timed(f):
     @wraps(f)
     def wrap(*args, **kw):

### 2021-01-31 htm_rl/htm_rl/envs/biogwlab/map_generator.py

@@ -1,0 +1,332 @@
+from itertools import product
+
+import numpy as np
+import matplotlib.pyplot as plt
+from htm_rl.common.utils import timed, trace, clip
+
+
+class BioGWLabEnvGenerator:
+    verbosity: int
+
+    def __init__(self):
+        self.verbosity = 2
+
+    def generate(self):
+        size = 10
+        seed = 11
+        density = .25
+        n_areas = 5
+        obstacle_generator = ObstacleGenerator(size, density, self.verbosity)
+        obstacle_mask, _ = obstacle_generator.generate(seed)
+
+        areas_generator = AreasGenerator(size, n_areas, self.verbosity)
+        areas_map = areas_generator.generate(seed)
+
+        wall_colorer = WallColorGenerator()
+        wall_colors = wall_colorer.color_walls(obstacle_mask, areas_map, seed)
+
+        food_generator = FoodGenerator(self.verbosity)
+        food_items, food_map, food_mask = food_generator.generate(areas_map, obstacle_mask, seed)
+        # self._plot_food_map(food_map, obstacle_mask)
+
+        scent_generator = ScentGenerator(self.verbosity)
+        food_scents = scent_generator.generate(food_items, obstacle_mask)
+        # self._plot_food_scent_map(food_scents, 0)
+
+        self._plot_final_map(areas_map, obstacle_mask, wall_colors, food_map)
+
+    def _plot_final_map(self, areas_map, obstacle_mask, wall_colors, food_map):
+        final_map = areas_map + wall_colors.max(initial=0)*2 + 4
+        m = final_map.max()
+        final_map[food_map == 0] = m + 5
+        final_map[food_map == 2] = m + 10
+        final_map[food_map == 1] = -5
+        final_map[food_map == 3] = -10
+        final_map[obstacle_mask] = wall_colors[obstacle_mask]
+        trace_image(self.verbosity, 1, final_map)
+
+    def _plot_food_scent_map(self, food_scents, channel):
+        food_scent = food_scents.sum(axis=-1)
+        trace_image(3, 1, food_scent[:, :, channel])
+
+    def _plot_food_map(self, food_items, obstacle_mask):
+        food_items += 2
+        food_items[obstacle_mask] = 0
+        trace_image(3, 1, food_items)
+
+
+class ScentGenerator:
+    BEANS_SCENT_WEIGHT =            [.9, .4, .1, .0, .5, .3]
[…]

### 2021-02-01 htm_rl/htm_rl/envs/biogwlab/dynamics.py
add env wrapper and env dynamics

@@ -1,0 +1,197 @@
+from typing import Tuple, List, Optional
+
+import numpy as np
+
+from htm_rl.common.utils import isnone, clip
+
+
+class BioGwLabEnvState:
+    directions = [(1, 0, 'right'), (0, -1, 'down'), (-1, 0, 'left'), (0, 1, 'up')]
+
+    size: int
+    seed: int
+    n_area_types: int
+    n_wall_colors: int
+    n_food_types: int
+    n_scent_channels: int
+    obstacle_mask: np.ndarray
+    areas_map: np.ndarray
+    wall_colors: np.ndarray
+    food_items: List[Tuple[int, int, int]]
+    food_map: np.ndarray
+    food_mask: np.ndarray
+    food_scents: np.ndarray
+    food_scent: np.ndarray
+
+    agent_initial_position: Optional[Tuple[int, int]]
+    agent_initial_direction: Optional[int]
+    agent_position: Optional[Tuple[int, int]]
+    agent_direction: Optional[int]
+
+    def __init__(
+            self, size: int, seed: int, n_area_types: int,
+            n_wall_colors: int, n_food_types: int,
+            n_scent_channels: int, obstacle_mask: np.ndarray,
+            areas_map: np.ndarray, wall_colors: np.ndarray,
+            food_items: List[Tuple[int, int, int]], food_mask: np.ndarray,
+            food_map: np.ndarray, food_scents: np.ndarray
+    ):
+        self.size = size
+        self.seed = seed
+        self.n_area_types = n_area_types
+        self.n_wall_colors = n_wall_colors
+        self.n_food_types = n_food_types
+        self.n_scent_channels = n_scent_channels
+        self.obstacle_mask = obstacle_mask
+        self.areas_map = areas_map
+        self.wall_colors = wall_colors
+        self.food_items = food_items
+        self.food_map = food_map
+        self.food_mask = food_mask
+        self.food_scents = food_scents
+        self.food_scent = food_scents.sum(axis=-1)
+
+        self.agent_initial_position = None
+        self.agent_initial_direction = None
+        self.agent_position = None
+        self.agent_direction = None
+
+
+class BioGwLabDynamics:
+    actions = {
+        'stay': {
+            'cost': .3
+        },
+        'move': {
+            'cost': .7,
+        },
+        'turn': {
+            'cost': .5,
+        }
+    }
+    food_rewards = [1., -1., 5., -5.]
+    time_cost = -.01
+
+    def __init__(self):
[…]

### 2021-02-01 htm_rl/htm_rl/envs/biogwlab/map_generator.py

@@ -1,39 +1,54 @@
 from itertools import product
 
-import numpy as np
 import matplotlib.pyplot as plt
+import numpy as np
+
 from htm_rl.common.utils import timed, trace, clip
+from htm_rl.envs.biogwlab.dynamics import BioGwLabEnvState
 
 
-class BioGWLabEnvGenerator:
+class BioGwLabEnvGenerator:
+    size: int
+    density: float
+    n_areas: int
+    seed: int
     verbosity: int
 
-    def __init__(self):
+    def __init__(self, size, density, n_areas, seed):
+        self.size = size
+        self.density = density
+        self.n_areas = n_areas
+        self.seed = seed
         self.verbosity = 2
 
     def generate(self):
-        size = 10
-        seed = 11
-        density = .25
-        n_areas = 5
+        size = self.size
+        seed = self.seed
+        density = self.density
+        n_areas = self.n_areas
         obstacle_generator = ObstacleGenerator(size, density, self.verbosity)
         obstacle_mask, _ = obstacle_generator.generate(seed)
 
         areas_generator = AreasGenerator(size, n_areas, self.verbosity)
-        areas_map = areas_generator.generate(seed)
+        areas_map, n_area_types = areas_generator.generate(seed)
 
         wall_colorer = WallColorGenerator()
-        wall_colors = wall_colorer.color_walls(obstacle_mask, areas_map, seed)
+        wall_colors, n_wall_colors = wall_colorer.color_walls(obstacle_mask, areas_map, seed)
 
         food_generator = FoodGenerator(self.verbosity)
-        food_items, food_map, food_mask = food_generator.generate(areas_map, obstacle_mask, seed)
-        # self._plot_food_map(food_map, obstacle_mask)
+        food_items, food_map, food_mask, n_food_types = food_generator.generate(areas_map, obstacle_mask, seed)
+        self._plot_food_map(food_map, obstacle_mask)
 
         scent_generator = ScentGenerator(self.verbosity)
-        food_scents = scent_generator.generate(food_items, obstacle_mask)
-        # self._plot_food_scent_map(food_scents, 0)
+        food_scents, n_scent_channels = scent_generator.generate(food_items, obstacle_mask)
+        self._plot_food_scent_map(food_scents, 0)
 
         self._plot_final_map(areas_map, obstacle_mask, wall_colors, food_map)
+        return BioGwLabEnvState(
[…]

### 2021-02-01 htm_rl/htm_rl/envs/biogwlab/dynamics.py
add representer wrappers, observation wrapper

@@ -1,60 +1,11 @@
-from typing import Tuple, List, Optional
+from itertools import product
+from typing import Tuple
 
 import numpy as np
+from numpy.random._generator import Generator
 
 from htm_rl.common.utils import isnone, clip
-
-
-class BioGwLabEnvState:
-    directions = [(1, 0, 'right'), (0, -1, 'down'), (-1, 0, 'left'), (0, 1, 'up')]
-
-    size: int
-    seed: int
-    n_area_types: int
-    n_wall_colors: int
-    n_food_types: int
-    n_scent_channels: int
-    obstacle_mask: np.ndarray
-    areas_map: np.ndarray
-    wall_colors: np.ndarray
-    food_items: List[Tuple[int, int, int]]
-    food_map: np.ndarray
-    food_mask: np.ndarray
-    food_scents: np.ndarray
-    food_scent: np.ndarray
-
-    agent_initial_position: Optional[Tuple[int, int]]
-    agent_initial_direction: Optional[int]
-    agent_position: Optional[Tuple[int, int]]
-    agent_direction: Optional[int]
-
-    def __init__(
-            self, size: int, seed: int, n_area_types: int,
-            n_wall_colors: int, n_food_types: int,
-            n_scent_channels: int, obstacle_mask: np.ndarray,
-            areas_map: np.ndarray, wall_colors: np.ndarray,
-            food_items: List[Tuple[int, int, int]], food_mask: np.ndarray,
-            food_map: np.ndarray, food_scents: np.ndarray
-    ):
-        self.size = size
-        self.seed = seed
-        self.n_area_types = n_area_types
-        self.n_wall_colors = n_wall_colors
-        self.n_food_types = n_food_types
-        self.n_scent_channels = n_scent_channels
-        self.obstacle_mask = obstacle_mask
-        self.areas_map = areas_map
-        self.wall_colors = wall_colors
-        self.food_items = food_items
-        self.food_map = food_map
-        self.food_mask = food_mask
-        self.food_scents = food_scents
-        self.food_scent = food_scents.sum(axis=-1)
-
-        self.agent_initial_position = None
-        self.agent_initial_direction = None
-        self.agent_position = None
-        self.agent_direction = None
+from htm_rl.envs.biogwlab.env_state import BioGwLabEnvState
 
 
 class BioGwLabDynamics:
@@ -93,12 +44,12 @@ class BioGwLabDynamics:
         return reward
 
     def turn(self, turn_direction, state):
[…]

### 2021-02-01 htm_rl/htm_rl/envs/biogwlab/env_state.py

@@ -1,0 +1,68 @@
+from typing import Tuple, List, Optional
+
+import numpy as np
+
+
+class BioGwLabEnvState:
+    directions = [(1, 0, 'right'), (0, -1, 'down'), (-1, 0, 'left'), (0, 1, 'up')]
+
+    size: int
+    seed: int
+    n_area_types: int
+    n_wall_colors: int
+    n_food_types: int
+    n_scent_channels: int
+    obstacle_mask: np.ndarray
+    areas_map: np.ndarray
+    wall_colors: np.ndarray
+    food_items: List[Tuple[int, int, int]]
+    food_map: np.ndarray
+    food_mask: np.ndarray
+    food_scents: np.ndarray
+    food_scent: np.ndarray
+    scent: np.ndarray
+
+    agent_initial_position: Optional[Tuple[int, int]]
+    agent_initial_direction: Optional[int]
+    agent_position: Optional[Tuple[int, int]]
+    agent_direction: Optional[int]
+
+    def __init__(
+            self, size: int, seed: int, n_area_types: int,
+            n_wall_colors: int, n_food_types: int,
+            n_scent_channels: int, obstacle_mask: np.ndarray,
+            areas_map: np.ndarray, wall_colors: np.ndarray,
+            food_items: List[Tuple[int, int, int]], food_mask: np.ndarray,
+            food_map: np.ndarray, food_scents: np.ndarray
+    ):
+        self.size = size
+        self.seed = seed
+        self.n_area_types = n_area_types
+        self.n_wall_colors = n_wall_colors
+        self.n_food_types = n_food_types
+        self.n_scent_channels = n_scent_channels
+        self.obstacle_mask = obstacle_mask
+        self.areas_map = areas_map
+        self.wall_colors = wall_colors
+        self.food_items = food_items
+        self.food_map = food_map
+        self.food_mask = food_mask
+        self.food_scents = food_scents
+        self.food_scent = self.get_food_scent(food_scents)
+
+        self.agent_initial_position = None
+        self.agent_initial_direction = None
+        self.agent_position = None
+        self.agent_direction = None
+
+    def get_food_scent(self, food_scents):
+        food_scent = food_scents.sum(axis=-1)
+        normalize_factors = (
+            food_scent
+                .reshape((food_scent.shape[0], -1))
+                .sum(axis=-1)
+                .reshape((-1, 1, 1))
+        )
+        food_scent /= normalize_factors
+        return food_scent
+

### 2021-02-01 htm_rl/htm_rl/envs/biogwlab/map_generator.py

@@ -4,7 +4,7 @@ import matplotlib.pyplot as plt
 import numpy as np
 
 from htm_rl.common.utils import timed, trace, clip
-from htm_rl.envs.biogwlab.dynamics import BioGwLabEnvState
+from htm_rl.envs.biogwlab.env_state import BioGwLabEnvState
 
 
 class BioGwLabEnvGenerator:
@@ -62,7 +62,14 @@ class BioGwLabEnvGenerator:
 
     def _plot_food_scent_map(self, food_scents, channel):
         food_scent = food_scents.sum(axis=-1)
-        trace_image(self.verbosity, 3, food_scent[:, :, channel])
+        food_scent /= (
+            food_scent
+                .reshape((food_scent.shape[0], -1))
+                .sum(axis=-1)
+                .reshape((-1, 1, 1))
+        )
+
+        trace_image(self.verbosity, 2, food_scent[channel, :, :])
 
     def _plot_food_map(self, food_map, obstacle_mask):
         food_map = food_map.copy()
@@ -93,12 +100,15 @@ class ScentGenerator:
 
     def generate(self, food_items, obstacle_mask):
         size = obstacle_mask.shape[0]
-        food_scents = np.zeros((size, size, self.n_scent_channels, len(food_items)), dtype=np.float)
+        food_scents = np.zeros((self.n_scent_channels, size, size, len(food_items)), dtype=np.float)
         for k, (i, j, food_item) in enumerate(food_items):
             for _i, _j in product(range(size), range(size)):
                 d = np.sqrt(self._dist(i, j, _i, _j))
                 d = max(1., d)
-                food_scents[_i, _j, :, k] = self.scent_weights[food_item] / d**1.5
+                food_scents[:, _i, _j, k] = self.scent_weights[food_item] / d**1.5
+
+        # zeroes scent for obstacle cells
+        food_scents *= ~obstacle_mask.reshape((1,)+obstacle_mask.shape + (1,))
 
         return food_scents, self.n_scent_channels
 
@@ -137,7 +147,7 @@ class FoodGenerator:
             size=n_foods-self.n_food_types
         )
         foods = list(range(self.n_food_types)) + list(foods)
-        trace(self.verbosity, 2, str(np.unique(foods, return_counts=True)[1]))
+        trace(self.verbosity, 2, f'Food: {np.unique(foods, return_counts=True)[1]}')
 
         food_mask = np.zeros_like(areas_map, dtype=np.bool)
         food_items = []

### 2021-02-01 htm_rl/htm_rl/envs/biogwlab/dynamics.py
fix observation wrapper

@@ -8,7 +8,7 @@ from htm_rl.common.utils import isnone, clip
 from htm_rl.envs.biogwlab.env_state import BioGwLabEnvState
 
 
-class BioGwLabDynamics:
+class BioGwLabEnvDynamics:
     actions = {
         'stay': {
             'cost': .3
@@ -96,12 +96,14 @@ class BioGwLabEnv:
     actions = ['stay', 'move', 'turn left', 'turn right']
 
     state: BioGwLabEnvState
-    dynamics: BioGwLabDynamics
+    dynamics: BioGwLabEnvDynamics
 
-    def __init__(self, state: BioGwLabEnvState, dynamics: BioGwLabDynamics):
+    def __init__(self, state: BioGwLabEnvState, dynamics: BioGwLabEnvDynamics):
         self.state = state
         self.dynamics = dynamics
         self.rnd = np.random.default_rng(seed=state.seed)
+
+        self.generate_initial_position()
 
     def generate_initial_position(self):
         position = self._get_random_empty_position()
@@ -205,7 +207,7 @@ class BioGwLabStateScentRepresenter:
         scent_map = np.zeros((channels, size, size), dtype=np.int8)
         for channel in range(channels):
             scent = state.food_scent[channel].ravel()
-            n_cells = scent.size()
+            n_cells = scent.size
             n_active = int(.2 * n_cells)
             activations = self.rnd.choice(n_cells, p=scent, size=n_active)
             scent_map[channel, np.divmod(activations, size)] = 1
@@ -221,7 +223,7 @@ class BioGwLabEnvRepresentationWrapper(BioGwLabEnv):
 
     visual_representation: np.ndarray
 
-    def __init__(self, state: BioGwLabEnvState, dynamics: BioGwLabDynamics):
+    def __init__(self, state: BioGwLabEnvState, dynamics: BioGwLabEnvDynamics):
         super().__init__(state, dynamics)
 
         self.visual_representer = BioGwLabStateVisualRepresenter(self.state)
@@ -252,21 +254,21 @@ class BioGwLabEnvRepresentationWrapper(BioGwLabEnv):
 
 
 class BioGwLabEnvObservationWrapper(BioGwLabEnvRepresentationWrapper):
-    i_transformation = [1, -1, -1, 1]
-    j_transformation = [-1, -1, 1, 1]
+    i_transformation = [1, -1, -1, 1, ]
[…]

### 2021-02-01 htm_rl/htm_rl/envs/biogwlab/map_generator.py

@@ -58,7 +58,7 @@ class BioGwLabEnvGenerator:
         final_map[food_map == 1] = -5
         final_map[food_map == 3] = -10
         final_map[obstacle_mask] = wall_colors[obstacle_mask]
-        trace_image(self.verbosity, 2, final_map)
+        trace_image(self.verbosity, 3, final_map)
 
     def _plot_food_scent_map(self, food_scents, channel):
         food_scent = food_scents.sum(axis=-1)
@@ -69,7 +69,7 @@ class BioGwLabEnvGenerator:
                 .reshape((-1, 1, 1))
         )
 
-        trace_image(self.verbosity, 2, food_scent[channel, :, :])
+        trace_image(self.verbosity, 3, food_scent[channel, :, :])
 
     def _plot_food_map(self, food_map, obstacle_mask):
         food_map = food_map.copy()

### 2021-02-01 htm_rl/htm_rl/envs/biogwlab/dynamics.py
fix scent generation

@@ -204,13 +204,13 @@ class BioGwLabStateScentRepresenter:
     def generate_scent_map(self, state: BioGwLabEnvState):
         channels = state.n_scent_channels
         size = state.size
-        scent_map = np.zeros((channels, size, size), dtype=np.int8)
+        scent_map = np.zeros((size, size, channels), dtype=np.int8)
         for channel in range(channels):
-            scent = state.food_scent[channel].ravel()
+            scent = state.food_scent[:, :, channel].ravel()
             n_cells = scent.size
             n_active = int(.2 * n_cells)
             activations = self.rnd.choice(n_cells, p=scent, size=n_active)
-            scent_map[channel, np.divmod(activations, size)] = 1
+            scent_map[np.divmod(activations, size), channel] = 1
         return scent_map
 
     def init_outer_cells(self, repr):
@@ -276,13 +276,12 @@ class BioGwLabEnvObservationWrapper(BioGwLabEnvRepresentationWrapper):
         vis_observation = self._clip_observation(
             vis_repr, self.view_rect, self.visual_representer.init_outer_cells
         )
-        # scent_observation = self._clip_observation(
-        #     scent_repr, self.scent_rect, self.scent_representer.init_outer_cells
-        # )
-        #
-        # observation = np.concatenate(vis_observation, scent_observation, axis=None)
-        # return observation
-        return vis_observation
+        scent_observation = self._clip_observation(
+            scent_repr, self.scent_rect, self.scent_representer.init_outer_cells
+        )
+
+        observation = np.concatenate((vis_observation, scent_observation), axis=None)
+        return observation
 
     def _clip_observation(self, full_repr, obs_rect, init_obs):
         state = self.state

### 2021-02-01 htm_rl/htm_rl/envs/biogwlab/env_state.py

@@ -59,9 +59,9 @@ class BioGwLabEnvState:
         food_scent = food_scents.sum(axis=-1)
         normalize_factors = (
             food_scent
-                .reshape((food_scent.shape[0], -1))
-                .sum(axis=-1)
-                .reshape((-1, 1, 1))
+                .reshape((-1, food_scent.shape[-1]))
+                .sum(axis=0)
+                .reshape((1, 1, -1))
         )
         food_scent /= normalize_factors
         return food_scent

### 2021-02-01 htm_rl/htm_rl/envs/biogwlab/map_generator.py

@@ -64,12 +64,12 @@ class BioGwLabEnvGenerator:
         food_scent = food_scents.sum(axis=-1)
         food_scent /= (
             food_scent
-                .reshape((food_scent.shape[0], -1))
-                .sum(axis=-1)
-                .reshape((-1, 1, 1))
+                .reshape((-1, food_scent.shape[-1]))
+                .sum(axis=0)
+                .reshape((1, 1, -1))
         )
 
-        trace_image(self.verbosity, 3, food_scent[channel, :, :])
+        trace_image(self.verbosity, 3, food_scent[:, :, channel])
 
     def _plot_food_map(self, food_map, obstacle_mask):
         food_map = food_map.copy()
@@ -100,15 +100,15 @@ class ScentGenerator:
 
     def generate(self, food_items, obstacle_mask):
         size = obstacle_mask.shape[0]
-        food_scents = np.zeros((self.n_scent_channels, size, size, len(food_items)), dtype=np.float)
+        food_scents = np.zeros((size, size, self.n_scent_channels, len(food_items)), dtype=np.float)
         for k, (i, j, food_item) in enumerate(food_items):
             for _i, _j in product(range(size), range(size)):
                 d = np.sqrt(self._dist(i, j, _i, _j))
                 d = max(1., d)
-                food_scents[:, _i, _j, k] = self.scent_weights[food_item] / d**1.5
+                food_scents[_i, _j, :, k] = self.scent_weights[food_item] / d**1.5
 
         # zeroes scent for obstacle cells
-        food_scents *= ~obstacle_mask.reshape((1,)+obstacle_mask.shape + (1,))
+        food_scents *= ~obstacle_mask.reshape(obstacle_mask.shape + (1, 1,))
 
         return food_scents, self.n_scent_channels
 

### 2021-02-01 htm_rl/htm_rl/agent/mcts_agent.py
wip: add agent runner and complete pipeline, doesn't work correctly

@@ -1,6 +1,8 @@
+from itertools import islice
 from typing import Any
 
 import numpy as np
+from numpy.random._generator import Generator
 from tqdm import trange
 
 from htm_rl.agent.mcts_actor_critic import MctsActorCritic
@@ -9,6 +11,8 @@ from htm_rl.agent.memory import Memory
 from htm_rl.agent.train_eval import RunStats, RunResultsProcessor
 from htm_rl.common.base_sa import Sa
 from htm_rl.common.utils import timed, trace
+from htm_rl.envs.biogwlab.dynamics import BioGwLabEnv, BioGwLabEnvDynamics, BioGwLabEnvObservationWrapper
+from htm_rl.envs.biogwlab.map_generator import BioGwLabEnvGenerator
 from htm_rl.envs.gridworld_pomdp import GridworldPomdp
 
 
@@ -164,3 +168,143 @@ class MctsAgentRunner:
 
     def store_results(self, run_results_processor: RunResultsProcessor):
         run_results_processor.store_result(self.train_stats, f'{self.name}')
+
+
+class MctsAgentRunner2:
+    agent: MctsAgent
+    env: Any
+    n_episodes: int
+    max_steps: int
+    pretrain: int
+    verbosity: int
+    train_stats: RunStats
+    name: str
+
+    def __init__(self, agent, env, n_episodes, max_steps, pretrain, verbosity):
+        self.agent = agent
+        self.env = env
+        self.n_episodes = n_episodes
+        self.max_steps = max_steps
+        self.pretrain = pretrain
+        self.verbosity = verbosity
+        self.train_stats = RunStats()
+        self.name = '???'
+
+    def run(self):
+        trace(self.verbosity, 1, '============> RUN MCTS AGENT')
+        planning_horizon = self.agent.planner.planning_horizon
+        if self.pretrain > 0:
+            self.agent.set_planning_horizon(0)
+
+        for ep in trange(self.n_episodes):
+            if 0 < self.pretrain == ep:
+                self.agent.set_planning_horizon(planning_horizon)
+
+            (steps, reward), elapsed_time = self.run_episode()
+            self.train_stats.append_stats(steps, reward, elapsed_time)
+            trace(self.verbosity, 2, '')
+        trace(self.verbosity, 1, '<============')
+
+    def run_iterable(self, iterable):
+        trace(self.verbosity, 1, '============> RUN MCTS AGENT')
[…]

### 2021-02-01 htm_rl/htm_rl/agent/train_eval.py

@@ -97,8 +97,8 @@ class RunResultsProcessor:
 
         ma = self.moving_average
         # self._plot_figure(df_times, f'episode execution time, sec, MA={ma}', report_name, 'times')
-        # self._plot_figure(df_rewards, f'episode reward, MA={ma}', report_name, 'rewards')
-        self._plot_figure(df_steps, f'episode duration, steps, MA={ma}', report_name, 'steps')
+        self._plot_figure(df_rewards, f'episode reward, MA={ma}', report_name, 'rewards')
+        # self._plot_figure(df_steps, f'episode duration, steps, MA={ma}', report_name, 'steps')
         # self._plot_figure(
         #     df_steps_rel,
         #     f'episode duration in steps log-relative to `{rel_col}`, MA={ma}', report_name,

### 2021-02-01 htm_rl/htm_rl/common/random_sdr_encoder_sp.py

@@ -53,8 +53,11 @@ class RandomSdrEncoderSp:
         self.value_bits = int(sparsity * total_bits)
         self.activation_threshold = int(.8 * self.value_bits)
 
-    def encode(self, x: int) -> BitSparseArr:
-        sparse_sdr = list(self.random_encoder.encode(x).unfold())
+    def encode(self, x) -> BitSparseArr:
+        if self.random_encoder is not None:
+            sparse_sdr = list(self.random_encoder.encode(x).unfold())
+        else:
+            sparse_sdr = list(np.nonzero(x)[0])
         sparse_sdr = self.spatial_pooler.encode(sparse_sdr)
         return BitSparseArr(sparse_sdr)
 

### 2021-02-01 htm_rl/htm_rl/config.py

@@ -14,7 +14,7 @@ from htm_rl.agent.legacy_agent import LegacyAgent
 from htm_rl.agent.legacy_memory import LegacyMemory
 from htm_rl.agent.legacy_planner import LegacyPlanner
 from htm_rl.agent.mcts_actor_critic import MctsActorCritic
-from htm_rl.agent.mcts_agent import MctsAgent, MctsAgentRunner
+from htm_rl.agent.mcts_agent import MctsAgent, MctsAgentRunner, TransferLearningExperimentRunner3, MctsAgentRunner2
 from htm_rl.agent.mcts_agent_q import MctsAgentQ
 from htm_rl.agent.mcts_planner import MctsPlanner
 from htm_rl.agent.mcts_planner_q import MctsPlannerQ
@@ -28,6 +28,7 @@ from htm_rl.common.random_sdr_encoder_sp import RandomSdrEncoderSp
 from htm_rl.common.sa_sdr_encoder import SaSdrEncoder
 from htm_rl.common.sar_sdr_encoder import SarSdrEncoder
 from htm_rl.common.vector_sdr_encoder import VectorSdrEncoder
+from htm_rl.envs.biogwlab.map_generator import BioGwLabEnvGenerator
 from htm_rl.envs.gridworld_map_generator import GridworldMapGenerator
 from htm_rl.envs.mdp import (
     SarSuperpositionFormatter, PovBasedGridworldMdpGenerator, Mdp, SaSuperpositionFormatter,
@@ -188,10 +189,13 @@ def register_classes(yaml: YAML):
         TransferLearningExperimentRunner,
         GridworldMapGenerator,
         TransferLearningExperimentRunner2,
+        TransferLearningExperimentRunner3,
         RunResultsProcessor,
         MctsAgent, MctsAgentRunner, MctsPlanner,
         MctsAgentQ, MctsPlannerQ,
         MctsActorCritic,
+        MctsAgentRunner2,
+        BioGwLabEnvGenerator,
     ]
 
     constructor: SafeConstructor = yaml.constructor

### 2021-02-01 htm_rl/htm_rl/envs/biogwlab/dynamics.py

@@ -27,7 +27,7 @@ class BioGwLabEnvDynamics:
         pass
 
     def is_terminal(self, state):
-        return False
+        return state.n_foods <= 0
 
     def stay(self, state):
         i, j = state.agent_position
@@ -74,9 +74,10 @@ class BioGwLabEnvDynamics:
                 continue
 
             state.food_scents[:, :, :, k] = 0
-            state.food_scent = state.get_food_scent(state.food_scents)
+            state.n_foods -= 1
             state.food_mask[i, j] = False
             state.food_map[i, j] = -1
+            state.food_scent = state.get_food_scent(state.food_scents)
             return self.food_rewards[food_type]
 
     def generate_scent_map(self, state: BioGwLabEnvState, rnd: Generator):
@@ -205,12 +206,12 @@ class BioGwLabStateScentRepresenter:
         channels = state.n_scent_channels
         size = state.size
         scent_map = np.zeros((size, size, channels), dtype=np.int8)
-        for channel in range(channels):
-            scent = state.food_scent[:, :, channel].ravel()
-            n_cells = scent.size
-            n_active = int(.2 * n_cells)
-            activations = self.rnd.choice(n_cells, p=scent, size=n_active)
-            scent_map[np.divmod(activations, size), channel] = 1
+        # for channel in range(channels):
+        #     scent = state.food_scent[:, :, channel].ravel()
+        #     n_cells = scent.size
+        #     n_active = int(.2 * n_cells)
+        #     activations = self.rnd.choice(n_cells, p=scent, size=n_active)
+        #     scent_map[np.divmod(activations, size), channel] = 1
         return scent_map
 
     def init_outer_cells(self, repr):
@@ -250,7 +251,7 @@ class BioGwLabEnvRepresentationWrapper(BioGwLabEnv):
 
         visual_representation = self.visual_representation
         scent_representation = self.scent_representer.generate_scent_map(self.state)
-        return visual_representation, scent_representation
+        return (visual_representation, scent_representation), reward, is_done, {}
 
 
 class BioGwLabEnvObservationWrapper(BioGwLabEnvRepresentationWrapper):
@@ -272,16 +273,27 @@ class BioGwLabEnvObservationWrapper(BioGwLabEnvRepresentationWrapper):
 
     def reset(self):
         vis_repr, scent_repr = super().reset()
[…]

### 2021-02-01 htm_rl/htm_rl/envs/biogwlab/env_state.py

@@ -26,14 +26,15 @@ class BioGwLabEnvState:
     agent_initial_direction: Optional[int]
     agent_position: Optional[Tuple[int, int]]
     agent_direction: Optional[int]
+    n_foods: int
 
     def __init__(
             self, size: int, seed: int, n_area_types: int,
             n_wall_colors: int, n_food_types: int,
             n_scent_channels: int, obstacle_mask: np.ndarray,
             areas_map: np.ndarray, wall_colors: np.ndarray,
-            food_items: List[Tuple[int, int, int]], food_mask: np.ndarray,
-            food_map: np.ndarray, food_scents: np.ndarray
+            food_items: List[Tuple[int, int, int]], food_map: np.ndarray,
+            food_mask: np.ndarray, food_scents: np.ndarray
     ):
         self.size = size
         self.seed = seed
@@ -47,6 +48,7 @@ class BioGwLabEnvState:
         self.food_items = food_items
         self.food_map = food_map
         self.food_mask = food_mask
+        self.n_foods = food_mask.sum()
         self.food_scents = food_scents
         self.food_scent = self.get_food_scent(food_scents)
 
@@ -57,12 +59,13 @@ class BioGwLabEnvState:
 
     def get_food_scent(self, food_scents):
         food_scent = food_scents.sum(axis=-1)
-        normalize_factors = (
-            food_scent
-                .reshape((-1, food_scent.shape[-1]))
-                .sum(axis=0)
-                .reshape((1, 1, -1))
-        )
-        food_scent /= normalize_factors
+        if self.n_foods > 0:
+            normalize_factors = (
+                food_scent
+                    .reshape((-1, food_scent.shape[-1]))
+                    .sum(axis=0)
+                    .reshape((1, 1, -1))
+            )
+            food_scent /= normalize_factors + 1e-5
         return food_scent
 

### 2021-02-01 htm_rl/htm_rl/envs/biogwlab/map_generator.py

@@ -14,12 +14,12 @@ class BioGwLabEnvGenerator:
     seed: int
     verbosity: int
 
-    def __init__(self, size, density, n_areas, seed):
+    def __init__(self, size, density, n_areas, seed, verbosity):
         self.size = size
         self.density = density
         self.n_areas = n_areas
         self.seed = seed
-        self.verbosity = 2
+        self.verbosity = verbosity
 
     def generate(self):
         size = self.size
@@ -58,7 +58,7 @@ class BioGwLabEnvGenerator:
         final_map[food_map == 1] = -5
         final_map[food_map == 3] = -10
         final_map[obstacle_mask] = wall_colors[obstacle_mask]
-        trace_image(self.verbosity, 3, final_map)
+        trace_image(self.verbosity, 1, final_map)
 
     def _plot_food_scent_map(self, food_scents, channel):
         food_scent = food_scents.sum(axis=-1)

### 2021-04-14 htm_rl/htm_rl/envs/biogwlab/environment.py
WIP: add real rgb rendering

@@ -83,12 +83,13 @@ class Environment(Env):
 
         self._run_handlers('reset')
         self.generate()
-        self.observe()
         # noinspection PyTypeChecker
         self.agent = self.entities['agent']
 
+        self.observe()
+
         # from htm_rl.common.plot_utils import plot_grid_images
-        # plot_grid_images([self.render_rgb()])
+        # plot_grid_images(self.render_rgb())
 
     def add_module(self, module: Module):
         self.remove_module(module)
@@ -132,9 +133,6 @@ class Environment(Env):
         reward = self.step_reward
         obs = self.render()
         is_first = self.episode_step == 0
-
-        # from htm_rl.common.plot_utils import plot_grid_images
-        # plot_grid_images([self.render_rgb()])
         return reward, obs, is_first
 
     def act(self, action: int):
@@ -192,8 +190,7 @@ class Environment(Env):
         return False
 
     def render(self):
-        # noinspection PyTypeChecker
-        agent: Agent = self.entities['agent']
+        agent = self.agent
 
         return self.renderer.render(
             position=agent.position, view_direction=agent.view_direction,
@@ -209,22 +206,11 @@ class Environment(Env):
         return self.renderer.output_sdr_size
 
     def render_rgb(self):
-        img = np.zeros(self.shape, dtype=np.int8)
-
-        # areas = self.get_module('areas')
-        # img[:] = areas.map
-        #
-        # obstacles = self.get_module('obstacles')
-        # img[obstacles.mask] = 8
-        #
-        # food = self.get_module('food')
-        # norm_rewards = 5 * food._rewards[food.map[food.mask]] / np.abs(food._rewards).max()
-        # norm_rewards[norm_rewards > 0] = norm_rewards[norm_rewards > 0].astype(np.int8) + 12
-        # norm_rewards[norm_rewards < 0] = norm_rewards[norm_rewards < 0].astype(np.int8) - 4
-        # img[food.mask] = norm_rewards
-        #
-        # img[self.agent.position] = 24
-        return img
+        agent = self.agent
+        return self.renderer.render_rgb(
+            position=agent.position, view_direction=agent.view_direction,
+            entities=self.entity_slices
+        )
 
 
 def ensure_all_actions_supported(actions, supported_actions):

### 2021-04-14 htm_rl/htm_rl/envs/biogwlab/food.py

@@ -81,13 +81,12 @@ class Food(Entity):
         if view_clip is None:
             return np.array(list(self.positions_fl)), self.env.shape[0] * self.env.shape[1]
 
-        view_shape, abs_indices, view_indices = view_clip
         indices = []
-        for abs_ind, view_ind in zip(abs_indices, view_indices):
+        for abs_ind, view_ind in zip(view_clip.abs_indices, view_clip.view_indices):
             if abs_ind in self.positions_fl:
                 indices.append(view_ind)
 
-        return np.array(indices), view_shape[0] * view_shape[1]
+        return np.array(indices), view_clip.shape[0] * view_clip.shape[1]
 
     def append_mask(self, mask: np.ndarray):
         for position_fl in self.positions_fl:

### 2021-04-14 htm_rl/htm_rl/envs/biogwlab/obstacle.py

@@ -60,10 +60,8 @@ class BorderObstacle(Entity):
             # won't be included because of zero size
             return None, 0
 
-        view_shape, abs_indices, view_indices = view_clip
-
-        clipped_mask = np.ones(view_shape, dtype=np.bool).flatten()
-        clipped_mask[view_indices] = 0
+        clipped_mask = np.ones(view_clip.shape, dtype=np.bool).flatten()
+        clipped_mask[view_clip.view_indices] = 0
         return np.flatnonzero(clipped_mask), clipped_mask.size
 
     def append_mask(self, mask: np.ndarray):

### 2021-04-14 htm_rl/htm_rl/envs/biogwlab/renderer.py

@@ -1,9 +1,9 @@
-from typing import Tuple, List, Optional, Iterable
+from typing import Tuple, List, Optional, Iterable, Dict
 
 import numpy as np
 
 from htm_rl.common.sdr_encoders import SdrConcatenator
-from htm_rl.envs.biogwlab.module import Entity
+from htm_rl.envs.biogwlab.module import Entity, EntityType
 from htm_rl.envs.biogwlab.view_clipper import ViewClipper, ViewClip
 
 
@@ -25,12 +25,7 @@ class Renderer:
         self.sdr_concatenator = None
 
     def render(self, position, view_direction, entities: Iterable[Entity]):
-        view_clip = None
-        if self.view_clipper is not None:
-            abs_indices, view_indices = self.view_clipper.clip(position, view_direction)
-            abs_indices = abs_indices.flatten()
-            view_indices = view_indices.flatten()
-            view_clip = self.view_shape, view_indices, abs_indices
+        view_clip = self.make_view_clip(position, view_direction)
 
         layers_with_size = []
         for entity in entities:
@@ -51,17 +46,56 @@ class Renderer:
         observation = self.sdr_concatenator.concatenate(*layers)
         return observation
 
+    def render_rgb(self, position, view_direction, entities: Dict[EntityType, List[Entity]]):
+        img = np.empty(self.shape + (3, ), dtype=np.int)
+        img[:] = np.array([255, 3, 209])
+
+        areas = entities[EntityType.Area]
+        area_color, dc = [117, 198, 230], [-12, -15, -6]
+        self._draw_entities(img, areas, area_color, dc)
+
+        obstacles = entities[EntityType.Obstacle]
+        obstacle_color, dc = [70, 70, 110], [-8, -8, 0]
+        self._draw_entities(img, obstacles, obstacle_color, dc)
+
+        # TODO: reward based coloring
+        food = entities[EntityType.Consumable]
+        food_color, dc = [112, 212, 17], [-4, -10, 4]
+        self._draw_entities(img, food, food_color, dc)
+
+        agent = entities[EntityType.Agent]
+        agent_color, dc = [255, 255, 0], [0, 0, 0]
+        self._draw_entities(img, agent, agent_color, dc)
+
+        # view_clip = self.make_view_clip(position, view_direction)
+        return [img]
+
     @property
     def output_sdr_size(self):
         return self.sdr_concatenator.output_sdr_size
+
+    def make_view_clip(self, position, view_direction):
[…]

### 2021-04-14 htm_rl/htm_rl/envs/biogwlab/view_clipper.py

@@ -1,3 +1,5 @@
+import dataclasses
+from dataclasses import dataclass
 from typing import Tuple
 
 import numpy as np
@@ -8,10 +10,14 @@ from htm_rl.common.utils import clip
 from htm_rl.envs.biogwlab.move_dynamics import DIRECTIONS_ORDER
 
 
+@dataclass
 class ViewClip:
     shape: Tuple[int, int]
     abs_indices: np.ndarray
     view_indices: np.ndarray
+
+    def __iter__(self):
+        yield from dataclasses.astuple(self)
 
 
 class ViewClipper:

### 2021-04-14 htm_rl/htm_rl/common/plot_utils.py
WIP: add agent observation to rgb rendering

@@ -1,11 +1,11 @@
 import os
-from typing import List
+from typing import List, Optional, Union
 
 import matplotlib.pyplot as plt
 import numpy as np
 
 
-def plot_grid_images(images: List[np.ndarray], uncenter=False, grid=False):
+def plot_grid_images(images: List[np.ndarray]):
     if not isinstance(images, List):
         images = [images]
 
@@ -14,57 +14,58 @@ def plot_grid_images(images: List[np.ndarray], uncenter=False, grid=False):
         axes = [axes]
 
     for ax, img in zip(axes, images):
-        _plot_grid_image(ax, img, uncenter, grid)
+        _plot_grid_image(ax, img)
     plt.show()
 
 
-def _plot_grid_image(ax, img: np.ndarray, uncenter=False, grid=False):
-    ax.imshow(img)
-    if uncenter:
-        shape = img.shape
-        x_ticks = np.arange(shape[1] + 1)
-        y_ticks = np.arange(shape[0], -1, -1)
-        ax.xticks(x_ticks - .5, x_ticks)
-        ax.yticks(y_ticks - .5, y_ticks)
-
-    ax.xaxis.tick_top()
-    if grid:
-        ax.grid()
-
-
-def store_environment_map(ind, env_map, env_name, seed, test_dir):
-    h, w = env_map.shape
-
+def store_environment_map(
+        ind: int, env_map: Union[np.ndarray, List[np.ndarray]],
+        env_name: str, seed: int, test_dir: str
+):
     fig: plt.Figure
     ax: plt.Axes
-    fig, ax = plt.subplots(1, 1, figsize=(12, 6))
-    ax.set_xticks(np.arange(-.5, w, 1))
-    ax.set_yticks(np.arange(-.5, h, 1))
-    ax.set_xticklabels(np.arange(w))
-    ax.set_yticklabels(np.arange(h))
-    ax.grid(color='grey', linestyle='-', linewidth=1)
+
+    if isinstance(env_map, list):
+        env_map, observation = env_map
+        fig, axes = plt.subplots(1, 2, figsize=(12, 6))
+    else:
+        observation = None
+        fig, axes = plt.subplots(1, 1, figsize=(6, 6))
+        axes = [axes]
+
+    # plot env map
+    ax = axes[0]
+    _plot_grid_image(ax, env_map)
[…]

### 2021-04-14 htm_rl/htm_rl/envs/biogwlab/environment.py

@@ -88,9 +88,6 @@ class Environment(Env):
 
         self.observe()
 
-        # from htm_rl.common.plot_utils import plot_grid_images
-        # plot_grid_images(self.render_rgb())
-
     def add_module(self, module: Module):
         self.remove_module(module)
 

### 2021-04-14 htm_rl/htm_rl/envs/biogwlab/food.py

@@ -51,7 +51,7 @@ class Food(Entity):
 
     def generate(self, seeds):
         seed = seeds['food']
-        empty_mask = ~self.env.aggregated_mask[EntityType.NonEmpty]
+        empty_mask = ~self.env.aggregated_mask[EntityType.Obstacle]
         areas = self.env.entity_slices[EntityType.Area]
 
         area_masks = []

### 2021-04-14 htm_rl/htm_rl/envs/biogwlab/renderer.py

@@ -17,8 +17,6 @@ class Renderer:
 
     def __init__(self, env, view_rectangle=None):
         self.shape = env.shape
-        self.view_shape = env.shape
-
         self.view_clipper = None
         if view_rectangle is not None:
             self.view_clipper = ViewClipper(env.shape, view_rectangle)
@@ -47,28 +45,41 @@ class Renderer:
         return observation
 
     def render_rgb(self, position, view_direction, entities: Dict[EntityType, List[Entity]]):
-        img = np.empty(self.shape + (3, ), dtype=np.int)
-        img[:] = np.array([255, 3, 209])
+        default_filler = np.array([255, 3, 209])
+
+        img_map = np.empty(self.shape + (3, ), dtype=np.int)
+        img_map[:] = default_filler
 
         areas = entities[EntityType.Area]
-        area_color, dc = [117, 198, 230], [-12, -15, -6]
-        self._draw_entities(img, areas, area_color, dc)
+        area_color, area_dc = [117, 198, 230], [-12, -15, -6]
+        self._draw_entities(img_map, areas, area_color, area_dc)
 
         obstacles = entities[EntityType.Obstacle]
-        obstacle_color, dc = [70, 70, 110], [-8, -8, 0]
-        self._draw_entities(img, obstacles, obstacle_color, dc)
+        obstacle_color, obstacle_dc = [70, 40, 100], [-7, -4, -10]
+        self._draw_entities(img_map, obstacles, obstacle_color, obstacle_dc)
 
         # TODO: reward based coloring
         food = entities[EntityType.Consumable]
-        food_color, dc = [112, 212, 17], [-4, -10, 4]
-        self._draw_entities(img, food, food_color, dc)
+        food_color, food_dc = [112, 212, 17], [-4, -10, 4]
+        self._draw_entities(img_map, food, food_color, food_dc)
 
         agent = entities[EntityType.Agent]
-        agent_color, dc = [255, 255, 0], [0, 0, 0]
-        self._draw_entities(img, agent, agent_color, dc)
+        agent_color, agent_dc = [255, 255, 0], [0, 0, 0]
+        self._draw_entities(img_map, agent, agent_color, agent_dc)
 
-        # view_clip = self.make_view_clip(position, view_direction)
-        return [img]
+        view_clip = self.make_view_clip(position, view_direction)
+        if view_clip is None:
+            return img_map
+
[…]

### 2021-04-14 htm_rl/htm_rl/envs/biogwlab/renderer.py
rewrite view clipper to fix its work

@@ -45,25 +45,30 @@ class Renderer:
         return observation
 
     def render_rgb(self, position, view_direction, entities: Dict[EntityType, List[Entity]]):
+        # fill with magenta to catch non-colored cells
         default_filler = np.array([255, 3, 209])
 
         img_map = np.empty(self.shape + (3, ), dtype=np.int)
         img_map[:] = default_filler
 
         areas = entities[EntityType.Area]
+        # areas: light blue
         area_color, area_dc = [117, 198, 230], [-12, -15, -6]
         self._draw_entities(img_map, areas, area_color, area_dc)
 
         obstacles = entities[EntityType.Obstacle]
+        # obstacles: dark blue
         obstacle_color, obstacle_dc = [70, 40, 100], [-7, -4, -10]
         self._draw_entities(img_map, obstacles, obstacle_color, obstacle_dc)
 
         # TODO: reward based coloring
         food = entities[EntityType.Consumable]
+        # consumables: salad green
         food_color, food_dc = [112, 212, 17], [-4, -10, 4]
         self._draw_entities(img_map, food, food_color, food_dc)
 
         agent = entities[EntityType.Agent]
+        # agent: yellow
         agent_color, agent_dc = [255, 255, 0], [0, 0, 0]
         self._draw_entities(img_map, agent, agent_color, agent_dc)
 
@@ -75,8 +80,12 @@ class Renderer:
         abs_indices = np.divmod(view_clip.abs_indices, img_map.shape[1])
         view_indices = np.divmod(view_clip.view_indices, img_obs.shape[1])
 
+        # fill with `out-of-map` obstacles: black
         img_obs[:] = np.array([0, 0, 0])
         img_obs[view_indices] = img_map[abs_indices].copy()
+        img_obs = np.flip(img_obs, axis=[0, 1])     # from ij to xy
+
+        # `grey`-out view area
         img_map[abs_indices] += (.5 * (255 - img_map[abs_indices])).astype(np.int)
 
         return [img_map, img_obs]
@@ -88,14 +97,7 @@ class Renderer:
     def make_view_clip(self, position, view_direction):
         if self.view_clipper is None:
             return None
-        abs_indices, view_indices = self.view_clipper.clip(position, view_direction)
-        abs_indices = abs_indices.flatten()
-        view_indices = view_indices.flatten()
-        return ViewClip(
-            shape=self.view_clipper.view_shape,
-            abs_indices=abs_indices,
-            view_indices=view_indices
[…]

### 2021-04-14 htm_rl/htm_rl/envs/biogwlab/view_clipper.py

@@ -4,10 +4,7 @@ from typing import Tuple
 
 import numpy as np
 
-from htm_rl.common.plot_utils import plot_grid_images
-from htm_rl.common.sdr import dense_to_sparse
-from htm_rl.common.utils import clip
-from htm_rl.envs.biogwlab.move_dynamics import DIRECTIONS_ORDER
+from htm_rl.envs.biogwlab.move_dynamics import DIRECTIONS_ORDER, MoveDynamics
 
 
 @dataclass
@@ -21,11 +18,13 @@ class ViewClip:
 
 
 class ViewClipper:
-    i_transformation = [1, -1, -1, 1, ]
-    j_transformation = [1, 1, -1, -1, ]
-
     base_shape: Tuple[int, int]
     view_rectangle: Tuple[Tuple[int, int], Tuple[int, int]]
+
+    natural_direction: int
+
+    _map_abs_indices_cache: np.ndarray
+    _view_indices_cache: np.ndarray
 
     def __init__(
             self,
@@ -33,7 +32,14 @@ class ViewClipper:
             view_rectangle_xy: Tuple[Tuple[int, int], Tuple[int, int]],
     ):
         self.base_shape = base_shape
-        self.view_rectangle = self.xy_to_ij(view_rectangle_xy)
+        # view direction with natural indices traversal
+        self.natural_direction = DIRECTIONS_ORDER.index('down')
+        self.view_rectangle = _xy_to_ij(view_rectangle_xy)
+
+        h, w = self.base_shape
+        self._map_abs_indices_cache = np.arange(h*w).reshape(self.base_shape)
+        h, w = self.view_shape
+        self._view_indices_cache = np.arange(h*w).reshape(self.view_shape)
 
     # noinspection PyRedundantParentheses
     @property
@@ -42,97 +48,121 @@ class ViewClipper:
         return (ui-bi+1, rj-lj+1)
 
     def clip(self, position, view_direction):
-        # position = 1, 5
-        # view_direction = 3
-        i, j = position
-        i_high, j_high = self.base_shape
+        # get relative direction
+        view_direction = self.natural_direction - view_direction
 
-        # print(i, j, forward_dir, i_high, j_high)
-        # print(obs_rect)
+        # abs rectangle from view rectangle
+        abs_rect = self._get_abs_rectangle(self.view_rectangle, position, view_direction)
+        abs_indices = self._get_abs_indices(abs_rect, view_direction)
 
[…]

### 2021-04-15 htm_rl/htm_rl/agents/htm/htm_agent.py
update

@@ -13,12 +13,14 @@ import random
 import yaml
 import matplotlib.pyplot as plt
 import wandb
+#from memory_profiler import profile
 
 
 class BioGwLabAction:
     """
     Muscles adapter to BioGwLabEnvironment.
     """
+
     def __init__(self, muscles_size, n_actions, patterns, noise_tolerance=0.1, do_nothing_action='random'):
         self.muscles_size = muscles_size
         self.n_actions = n_actions
@@ -52,7 +54,7 @@ class HTMAgent:
 
         self.muscles = Muscles(**config['muscles'])
 
-        self.total_patterns = 2**self.action.muscles_size
+        self.total_patterns = 2 ** self.action.muscles_size
         # there is no first action
         self.action_pattern = np.empty(0)
         self.state_pattern = SDR(config['state_size'])
@@ -114,7 +116,7 @@ class HTMAgent:
         n_patterns_to_gen = np.clip(self.alpha * (self.total_patterns - len(self.hierarchy.output_block.sm)),
                                     a_min=0, a_max=self.total_patterns)
         if n_patterns_to_gen > 0:
-            return np.random.choice([0, 1], size=(int(n_patterns_to_gen), self.muscles.muscles_size//2))
+            return np.random.choice([0, 1], size=(int(n_patterns_to_gen), self.muscles.muscles_size // 2))
         else:
             return np.empty(0)
 
@@ -191,10 +193,9 @@ class HTMAgentRunner:
         self.agent = HTMAgent(config['agent'], hierarchy)
         self.environment = BioGwLabEnvironment(**config['environment'])
 
-    def run_episodes(self, n_episodes, train_patterns=True, logger=None, log_q_table=False, log_every_episode=50, log_patterns=False, log_options=False,
+    def run_episodes(self, n_episodes, train_patterns=True, logger=None, log_q_table=False, log_every_episode=50,
+                     log_patterns=False, log_options=False,
                      log_segments=True):
-        history = {'steps': list(), 'reward': list()}
-
         total_reward = 0
         steps = 0
         episode = 0
@@ -216,19 +217,23 @@ class HTMAgentRunner:
                         for i in range(steps):
                             image = imageio.imread(f'/tmp/{logger.run.id}_episode_{episode}_step_{i}.png')
                             writer.append_data(image)
-                    logger.log({f'animation': logger.Video(f'/tmp/{logger.run.id}_episode_{episode}.gif', fps=4, format='gif')}, step=episode)
-
[…]

### 2021-04-15 htm_rl/htm_rl/envs/biogwlab/agent.py
update

@@ -1,43 +1,97 @@
-from typing import Tuple, Optional
+from __future__ import annotations
+
+from typing import Tuple, Optional, List, TYPE_CHECKING
 
 import numpy as np
 
-from htm_rl.envs.biogwlab.entity import Entity
-from htm_rl.envs.biogwlab.move_dynamics import MoveDynamics
+from htm_rl.common.sdr_encoders import IntBucketEncoder
+from htm_rl.envs.biogwlab.module import Entity, EntityType
+from htm_rl.envs.biogwlab.move_dynamics import MoveDynamics, DIRECTIONS_ORDER, MOVE_DIRECTIONS
+from htm_rl.envs.biogwlab.view_clipper import ViewClip
+
+if TYPE_CHECKING:
+    from htm_rl.envs.biogwlab.environment import Environment
+
+
+class AgentRenderer:
+    position_encoder: Optional[IntBucketEncoder]
+    view_direction_encoder: Optional[IntBucketEncoder]
+
+    def __init__(self, env_shape: Tuple[int, int], what: List[str], **encoder):
+        n_cells = env_shape[0] * env_shape[1]
+        self.position_encoder = None
+        self.view_direction_encoder = None
+
+        if 'position' in what:
+            self.position_encoder = IntBucketEncoder(n_values=n_cells, **encoder)
+        if 'view direction' in what:
+            self.view_direction_encoder = IntBucketEncoder(
+                n_values=len(MOVE_DIRECTIONS), **encoder
+            )
+
+    def render(self, position, view_direction):
+        result = []
+        if self.position_encoder:
+            result.append(self._render_one(self.position_encoder, position))
+        if self.view_direction_encoder:
+            result.append(self._render_one(self.view_direction_encoder, view_direction))
+        return result
+
+    @staticmethod
+    def _render_one(encoder: IntBucketEncoder, value: int):
+        return encoder.encode(value), encoder.output_sdr_size
 
 
 class Agent(Entity):
-    entity = 'agent'
+    family = 'agent'
+    type = EntityType.Agent
 
     position: Tuple[int, int]
     view_direction: Optional[int]
 
-    _obstacles: Entity
-    _mask: np.ndarray
+    renderer: AgentRenderer
+    env: Environment
 
-    def __init__(self, obstacles: Entity, **agent):
-        super(Agent, self).__init__(**agent)
-        self._obstacles = obstacles
+    def __init__(self, env: Environment, rendering=False, **entity):
[…]

### 2021-04-15 htm_rl/htm_rl/envs/biogwlab/area.py

@@ -1,0 +1,59 @@
+from typing import Optional, Dict
+
+import numpy as np
+
+from htm_rl.envs.biogwlab.environment import Environment
+from htm_rl.envs.biogwlab.generation.areas import MultiAreaMapGenerator
+from htm_rl.envs.biogwlab.module import Entity, EntityType, Module
+from htm_rl.envs.biogwlab.renderer import render_mask
+from htm_rl.envs.biogwlab.view_clipper import ViewClip
+
+
+class Area(Entity):
+    family = 'area'
+    type = EntityType.Area
+
+    mask: np.ndarray
+
+    def __init__(self, mask: np.ndarray, **entity):
+        super(Area, self).__init__(**entity)
+        self.mask = mask
+        self.initialized = True
+
+    def render(self, view_clip: ViewClip = None):
+        return render_mask(self.mask, view_clip)
+
+    def append_mask(self, mask: np.ndarray):
+        mask |= self.mask
+
+    def append_position(self, exist: bool, position):
+        return exist or self.mask[position]
+
+
+class MultiAreaGenerator(Module):
+    env: Environment
+    generator: MultiAreaMapGenerator
+    last_seed: Optional[int]
+
+    area_config: Dict
+
+    def __init__(self, n_types: int, env: Environment, name: str, **area_entity):
+        super(MultiAreaGenerator, self).__init__(name=name)
+
+        self.env = env
+        self.generator = MultiAreaMapGenerator(shape=env.shape, n_types=n_types)
+        self.last_seed = None
+        self.area_config = area_entity
+
+    def generate(self, seeds):
+        seed = seeds['map']
+        if self.last_seed == seed:
+            return
+
+        self.last_seed = seed
+        areas_map = self.generator.generate(seed)
+        for i in range(self.generator.n_types):
+            mask = areas_map == i
+            name = f'{self.name}%{i}'
+            area = Area(mask=mask, name=name, **self.area_config)
+            self.env.add_module(area)

### 2021-04-15 htm_rl/htm_rl/envs/biogwlab/areas.py

@@ -1,59 +1,0 @@
-from typing import Tuple, Optional
-
-import numpy as np
-
-from htm_rl.common.sdr_encoders import IntArrayEncoder
-from htm_rl.envs.biogwlab.entity import Entity
-from htm_rl.envs.biogwlab.generation.areas import AreasGenerator
-
-
-class Areas(Entity):
-    entity = 'areas'
-
-    _generator: AreasGenerator
-    _last_seed: Optional[int]
-
-    def __init__(self, **areas):
-        super(Areas, self).__init__(**areas)
-        self._generator = AreasGenerator(shape=self.shape, n_types=self.n_types)
-        self._last_seed = None
-
-    def generate(self, seeds):
-        seed = seeds['map']
-        if self._last_seed is not None and self._last_seed == seed:
-            return
-
-        self.set(
-            mask=None,
-            map=self._generator.generate(seed)
-        )
-
-
-class AreasRenderer:
-    """TODO use or remove it"""
-    areas: Entity
-    view_shape: Tuple[int, int]
-
-    _encoder: IntArrayEncoder
-
-    def __init__(self, view_shape):
-        self.view_shape = view_shape
-
-    def set_renderer(self, view_shape):
-        self.view_shape = view_shape
-        self._encoder = IntArrayEncoder(shape=view_shape, n_types=self.areas.n_types)
-        return self._encoder
-
-    def render(self, view_clip=None):
-        if view_clip is not None:
-            view_indices, abs_indices = view_clip
-
-            area_map = np.zeros(self.view_shape, dtype=np.int).flatten()
-            area_map[view_indices] = self.areas.map.flatten()[abs_indices]
-        else:
-            area_map = self.areas.map
-
-        return self._encoder.encode(area_map)
-
-    def render_rgb(self, img: np.ndarray):
-        img[:] = self.areas.map

### 2021-04-15 htm_rl/htm_rl/envs/biogwlab/entity.py

@@ -1,23 +1,0 @@
-from typing import Tuple, Optional
-
-import numpy as np
-
-from htm_rl.common.utils import isnone
-
-
-class Entity:
-    shape: Tuple[int, int]
-    mask: Optional[np.ndarray]
-    map: Optional[np.ndarray]
-    n_types: int
-
-    def __init__(self, shape=None, n_types=1, env=None):
-        shape = isnone(shape, env.shape)
-        self.shape = shape
-        self.n_types = n_types
-        self.mask = None
-        self.map = None
-
-    def set(self, mask: np.ndarray = None, map: np.ndarray = None):
-        self.mask = mask
-        self.map = map

### 2021-04-15 htm_rl/htm_rl/envs/biogwlab/env.py

@@ -1,21 +1,27 @@
-from typing import Tuple, Dict
+from typing import Tuple, Dict, Any
 
-from htm_rl.envs.biogwlab.areas import Areas
+from htm_rl.envs.biogwlab.agent import Agent
+from htm_rl.envs.biogwlab.area import MultiAreaGenerator
 from htm_rl.envs.biogwlab.environment import Environment
 from htm_rl.envs.biogwlab.episode_terminator import EpisodeTerminator
-from htm_rl.envs.biogwlab.food import add_food
-from htm_rl.envs.biogwlab.obstacles import Obstacles
+from htm_rl.envs.biogwlab.food import Food
+from htm_rl.envs.biogwlab.modules.actions_cost import ActionsCost
+from htm_rl.envs.biogwlab.obstacle import Obstacle, BorderObstacle
 from htm_rl.envs.biogwlab.regenerator import Regenerator
 from htm_rl.envs.biogwlab.renderer import Renderer
 from htm_rl.envs.wrapper import Wrapper
 
 registry = {
-    'areas': Areas,
-    'obstacles': Obstacles,
-    'food': add_food,
+    'areas': MultiAreaGenerator,
+    'obstacle': Obstacle,
+    'food': Food,
     'rendering': Renderer,
     'terminate': EpisodeTerminator,
     'regenerate': Regenerator,
+    'actions_cost': ActionsCost,
+
+    Agent.family: Agent,
+    BorderObstacle.family: BorderObstacle,
 }
 
 
@@ -23,37 +29,39 @@ class BioGwLabEnvironment(Wrapper):
     output_sdr_size: int
 
     def __init__(
-            self, shape_xy: Tuple[int, int], seed: int,
-            action_costs, actions=None,
+            self, shape_xy: Tuple[int, int], seed: int, rendering: Dict = None, actions=None,
             **modules
     ):
         env = Environment(
-            shape_xy=shape_xy, seed=seed
+            shape_xy=shape_xy, seed=seed, actions=actions, rendering=rendering
         )
-        env.set_actions(actions)
-        env.set_action_costs(**action_costs)
 
-        add_module(env, modules, 'areas')
-        add_module(env, modules, 'obstacles')
-        add_module(env, modules, 'food')
-        env.add_agent()
+        default_modules = [
+            Agent.family, BorderObstacle.family, 'regenerate', 'terminate'
+        ]
+        for module_name in default_modules:
+            append_module(module_name, modules)
 
[…]

### 2021-04-15 htm_rl/htm_rl/envs/biogwlab/environment.py

@@ -4,9 +4,12 @@ import numpy as np
 
 from htm_rl.common.utils import isnone
 from htm_rl.envs.biogwlab.agent import Agent
+from htm_rl.envs.biogwlab.flags import CachedFlagDict, CachedEntityAggregation
+from htm_rl.envs.biogwlab.module import Module, Entity, EntityType
 from htm_rl.envs.biogwlab.move_dynamics import (
     MOVE_DIRECTIONS, DIRECTIONS_ORDER, TURN_DIRECTIONS,
 )
+from htm_rl.envs.biogwlab.renderer import Renderer
 from htm_rl.envs.env import Env
 
 
@@ -21,39 +24,53 @@ class Environment(Env):
         'move left', 'move up', 'move right', 'move down',
         'move forward', 'turn left', 'turn right'
     ]
-    supported_entities = {'obstacles', 'areas', 'food', 'agent'}
+    supported_entities = {'obstacle', 'area', 'food', 'agent'}
     supported_events = [
         'reset', 'generate', 'generate_seeds',
         'move', 'turn', 'collect',
         'collected',
         'is_terminal',
-        'render'
     ]
 
     seed: int
     shape: Tuple[int, int]
 
-    modules: Dict[str, Any]
+    agent: Agent
+    modules: Dict[str, Module]
     handlers: Dict[str, List[Any]]
 
-    agent: Agent
+    entities: Dict[str, Entity]
+    entity_slices: Dict[EntityType, List[Entity]]
+    aggregated_mask: Dict[EntityType, np.ndarray]
+
     episode_step: int
     step_reward: float
 
-    action_cost: float
-    action_weight: Dict[str, float]
+    renderer: Renderer
 
     actions: List[str]
 
-    def __init__(self, shape_xy: Tuple[int, int], seed: int):
+    def __init__(
+            self, shape_xy: Tuple[int, int], seed: int, actions: List[str],
+            rendering: Dict = None
+    ):
         # convert from x,y to i,j
         width, height = shape_xy
         self.shape = (height, width)
+        self.seed = seed
+
+        self.actions = isnone(actions, self.supported_actions.copy())
+        ensure_all_actions_supported(self.actions, self.supported_actions)
 
         self.modules = dict()
         self.handlers = dict()
+        self.entities = dict()
+        self.entity_slices = CachedFlagDict(self.entities)
+        self.aggregated_mask = CachedEntityAggregation(self.entity_slices, self.shape)
 
[…]

### 2021-04-15 htm_rl/htm_rl/envs/biogwlab/episode_terminator.py

@@ -3,39 +3,42 @@ from typing import Union
 import numpy as np
 
 from htm_rl.common.utils import isnone
-from htm_rl.envs.biogwlab.entity import Entity
 from htm_rl.envs.biogwlab.environment import Environment
+from htm_rl.envs.biogwlab.module import Module, EntityType
 
 
-class EpisodeTerminator:
+class EpisodeTerminator(Module):
     max_steps: int
     early_stop: bool
 
     terminated: bool
 
-    _food: Entity
+    env: Environment
 
     def __init__(
-            self, env: Environment,
-            episode_max_steps: Union[int, str] = None,
+            self, env: Environment, episode_max_steps: Union[int, str] = None,
             early_stop: bool = False,
+            **module
     ):
+        super().__init__(**module)
+
         if isnone(episode_max_steps, 'auto') == 'auto':
             h, w = env.shape
             episode_max_steps = 2 * h * w
 
+        self.env = env
         self.max_steps = episode_max_steps
         self.early_stop = early_stop
-        self._food = env.get_module('food')
         self.terminated = False
 
     def reset(self):
         self.terminated = False
 
     def collected(self):
-        food = self._food
-        # WARN: ugly workaround
-        self.terminated = not np.any(food._rewards[food.map[food.mask]])
+        if self.early_stop:
+            # WARN: ugly workaround
+            food = self.env.aggregated_mask[EntityType.Consumable]
+            self.terminated = not np.any(food)
 
     def is_terminal(self, episode_step):
         if episode_step >= self.max_steps:

### 2021-04-15 htm_rl/htm_rl/envs/biogwlab/flags.py

@@ -1,0 +1,59 @@
+from typing import Dict, List, Tuple
+
+import numpy as np
+
+from htm_rl.envs.biogwlab.module import Entity, EntityType
+
+
+class CachedFlagDict(dict):
+    _base_dict: Dict[str, Entity]
+
+    def __init__(self, base_dict: Dict[str, Entity]):
+        super().__init__()
+        self._base_dict = base_dict
+
+    def flush(self):
+        self.clear()
+
+    def __getitem__(self, flag: EntityType):
+        if True or flag not in self:
+            self[flag] = self._find_items(flag)
+        return super(CachedFlagDict, self).__getitem__(flag)
+
+    def _find_items(self, flag: EntityType):
+        return [
+            entity
+            for entity in self._base_dict.values()
+            if entity.type & flag
+        ]
+
+
+class CachedEntityAggregation(dict):
+    _shape: Tuple[int, int]
+    _flag_dict: Dict[EntityType, List[Entity]]
+
+    def __init__(
+            self, flag_dict: Dict[EntityType, List[Entity]],
+            shape: Tuple[int, int]
+    ):
+        super().__init__()
+        self._flag_dict = flag_dict
+        self._shape = shape
+
+    def flush(self):
+        self.clear()
+        self._flag_dict.clear()
+
+    def __getitem__(self, flag: EntityType):
+        if True or flag not in self:
+            self[flag] = self._aggregate_entities(flag)
+        return super(CachedEntityAggregation, self).__getitem__(flag)
+
+    def _aggregate_entities(self, flag: EntityType):
+        mask = np.zeros(self._shape, dtype=np.bool)
+        for entity in self._flag_dict[flag]:
+            if entity.initialized:
+                entity.append_mask(mask)
+        return mask
+
+

### 2021-04-15 htm_rl/htm_rl/envs/biogwlab/food.py

@@ -1,9 +1,12 @@
-from typing import Optional
+from typing import List, Set
 
 import numpy as np
 
-from htm_rl.envs.biogwlab.entity import Entity
-from htm_rl.envs.biogwlab.generation.food import FoodGenerator
+from htm_rl.common.plot_utils import plot_grid_images
+from htm_rl.envs.biogwlab.environment import Environment
+from htm_rl.envs.biogwlab.generation.food import FoodPositionsGenerator
+from htm_rl.envs.biogwlab.module import Entity, EntityType
+from htm_rl.envs.biogwlab.view_clipper import ViewClip
 
 
 def add_food(env, types=None, **food):
@@ -24,58 +27,77 @@ def add_food(env, types=None, **food):
 
 
 class Food(Entity):
-    entity = 'food'
+    family = 'food'
+    type = EntityType.Consumable
 
-    _areas: Entity
-    _obstacles: Entity
+    reward: float
+    positions_fl: Set[int]
 
-    _rewards: np.ndarray
+    generator: FoodPositionsGenerator
+    env: Environment
 
-    _generator: FoodGenerator
-    _initial_mask: np.ndarray
-    _last_seed: Optional[int]
+    def __init__(
+            self, env: Environment, reward: float, n_items: int = 1,
+            area_weights: List[float] = None,
+            **entity
+    ):
+        super(Food, self).__init__(**entity)
 
-    def __init__(self, types, areas, obstacles, n_items=None, **food):
-        n_types = len(types)
-        super(Food, self).__init__(n_types=n_types, **food)
-
-        self._areas = areas
-        self._obstacles = obstacles
-
-        if n_items is None:
-            n_items = sum(food_type['n_items'] for food_type in types.values())
-
-        self._rewards = np.array([food_type['reward'] for food_type in types.values()], dtype=np.float)
-
-        self._generator = FoodGenerator(
-            shape=self.shape, n_types=self.n_types, n_items=n_items
+        self.reward = reward
+        self.generator = FoodPositionsGenerator(
+            shape=env.shape, n_items=n_items, area_weights=area_weights
         )
-        self._last_seed = None
+        self.env = env
 
     def generate(self, seeds):
         seed = seeds['food']
-        if self._last_seed == seed:
-            self.mask = self._initial_mask.copy()
-            return
+        empty_mask = ~self.env.aggregated_mask[EntityType.Obstacle]
+        areas = self.env.entity_slices[EntityType.Area]
 
[…]

### 2021-04-15 htm_rl/htm_rl/envs/biogwlab/generation/areas.py

@@ -2,10 +2,9 @@ from itertools import product
 from typing import Tuple
 
 import numpy as np
-from matplotlib import pyplot as plt
 
 
-class AreasGenerator:
+class MultiAreaMapGenerator:
     shape: Tuple[int, int]
     n_types: int
 
@@ -52,7 +51,9 @@ class AreasGenerator:
         area_types = rng.integers(self.n_types, size=n_areas)
         return area_centers, area_types
 
-    def _plot_area_centers(self, area_centers, area_types):
+    def _debug_plot_area_centers(self, area_centers, area_types):
+        from matplotlib import pyplot as plt
+
         size = self.shape
         area = np.zeros((size, size), dtype=np.int8)
         for i, t in enumerate(area_types):

### 2021-04-15 htm_rl/htm_rl/envs/biogwlab/generation/food.py

@@ -1,12 +1,44 @@
 from itertools import product
-from typing import Tuple
+from typing import Tuple, List, Optional
 
 import numpy as np
 
+from htm_rl.common.plot_utils import plot_grid_images
 from htm_rl.common.utils import isnone
 
 
-class FoodGenerator:
+class FoodPositionsGenerator:
+    shape: Tuple[int, int]
+    n_items: int
+    area_weights: Optional[np.ndarray]
+
+    def __init__(self, shape: Tuple[int, int], n_items, area_weights=None):
+        self.shape = shape
+        self.n_items = n_items
+
+        self.area_weights = None
+        if area_weights is not None:
+            self.area_weights = np.array(area_weights).reshape((-1, 1))
+
+    def generate(self, empty_mask: np.ndarray, area_masks: List[np.ndarray], seed):
+        rng = np.random.default_rng(seed)
+        n_cells = self.shape[0] * self.shape[1]
+
+        if self.area_weights is not None:
+            area_weights_masks = self.area_weights * area_masks
+            p = np.mean(area_weights_masks, axis=0)
+            p[~empty_mask] = 0
+            p /= p.sum()
+            indices_fl = rng.choice(n_cells, size=self.n_items, p=p, replace=False)
+        else:
+            empty_indices_fl = np.flatnonzero(empty_mask)
+            indices_fl = rng.choice(empty_indices_fl, size=self.n_items, replace=False)
+
+        # indices = list(zip(*np.divmod(indices_fl, self.shape[1])))
+        return indices_fl
+
+
+class LegacyFoodGenerator:
     BEANS_DISTRIBUTION = [.3, .1, .25, .15, .1]
     RANCID_BEANS_DISTRIBUTION = [.25, .15, .1, .35, .15]
     FRUIT_DISTRIBUTION = [.15, .35, .2, .1, .2]
@@ -18,8 +50,8 @@ class FoodGenerator:
     n_items: int
 
     def __init__(self, shape, n_types, n_items):
+        assert n_types <= 4, 'Up to 4 types of food are supported rn'
         self.shape = shape
-
         self.n_types = n_types
 
         n_cells = shape[0] * shape[1]
@@ -41,6 +73,7 @@ class FoodGenerator:
         rng = np.random.default_rng(seed=seed)
         shape = areas_map.shape
         n_cells = shape[0] * shape[1]
+        areas_modulo = self.food_distribution.shape[1]
 
[…]

### 2021-04-15 htm_rl/htm_rl/envs/biogwlab/module.py

@@ -1,0 +1,45 @@
+from enum import Flag, auto
+from typing import Tuple
+
+import numpy as np
+
+from htm_rl.envs.biogwlab.view_clipper import ViewClip
+
+
+class Module:
+    """Basic module that extends Environment functionality."""
+    name: str
+
+    def __init__(self, name: str = '???'):
+        self.name = name
+
+
+class EntityType(Flag):
+    Unknown = 0
+    Area = auto()
+    Obstacle = auto()
+    Consumable = auto()
+    Agent = auto()
+    NonEmpty = Obstacle | Consumable | Agent
+
+
+class Entity(Module):
+    family: str = '???'
+    type: EntityType = EntityType.Unknown
+
+    rendering: bool
+    initialized: bool
+
+    def __init__(self, name: str, rendering: bool = True):
+        super(Entity, self).__init__(name=name)
+        self.rendering = rendering
+        self.initialized = False
+
+    def render(self, view_clip: ViewClip = None):
+        raise NotImplementedError()
+
+    def append_mask(self, mask: np.ndarray):
+        raise NotImplementedError()
+
+    def append_position(self, exist: bool, position):
+        raise NotImplementedError()

### 2021-04-15 htm_rl/htm_rl/envs/biogwlab/modules/actions_cost.py

@@ -1,0 +1,30 @@
+from typing import List, Dict
+
+from htm_rl.envs.biogwlab.environment import Environment
+from htm_rl.envs.biogwlab.module import Module
+
+
+class ActionsCost(Module):
+    base_coast: float
+    weights: Dict[str, float]
+
+    env: Environment
+
+    def __init__(
+            self, env: Environment, base_cost: float, weights: Dict[str, float],
+            **module
+    ):
+        super().__init__(**module)
+
+        self.base_coast = base_cost
+        self.weights = weights
+        self.env = env
+
+    def stay(self):
+        self.env.step_reward += self.weights['stay'] * self.base_coast
+
+    def move(self, direction):
+        self.env.step_reward += self.weights['move'] * self.base_coast
+
+    def turn(self, turn_direction: int):
+        self.env.step_reward += self.weights['turn'] * self.base_coast

### 2021-04-15 htm_rl/htm_rl/envs/biogwlab/obstacle.py

@@ -1,0 +1,75 @@
+from typing import Optional, Tuple
+
+import numpy as np
+
+from htm_rl.envs.biogwlab.environment import Environment
+from htm_rl.envs.biogwlab.module import Entity, EntityType
+from htm_rl.envs.biogwlab.generation.obstacles import ObstacleMaskGenerator
+from htm_rl.envs.biogwlab.renderer import render_mask
+from htm_rl.envs.biogwlab.view_clipper import ViewClip
+
+
+class Obstacle(Entity):
+    family = 'obstacle'
+    type = EntityType.Obstacle
+
+    generator: ObstacleMaskGenerator
+    mask: np.ndarray
+    last_seed: Optional[int]
+
+    def __init__(self, env: Environment, density, **entity):
+        super(Obstacle, self).__init__(**entity)
+
+        self.generator = ObstacleMaskGenerator(shape=env.shape, density=density)
+        self.mask = np.zeros(env.shape, dtype=np.bool)
+        self.last_seed = None
+
+    def generate(self, seeds):
+        seed = seeds['map']
+        if self.last_seed == seed:
+            return
+
+        self.mask = self.generator.generate(seed)
+        self.initialized = True
+
+    def render(self, view_clip: ViewClip = None):
+        return render_mask(self.mask, view_clip)
+
+    def append_mask(self, mask: np.ndarray):
+        mask |= self.mask
+
+    def append_position(self, exist: bool, position):
+        return exist or self.mask[position]
+
+
+class BorderObstacle(Entity):
+    family = 'obstacle.border'
+    type = EntityType.Obstacle
+
+    shape: Tuple[int, int]
+
+    def __init__(self, env: Environment, rendering=True, **entity):
+        super(BorderObstacle, self).__init__(rendering=rendering, **entity)
+
+        self.shape = env.shape
+        self.initialized = True
+
+    def render(self, view_clip: ViewClip = None):
+        if view_clip is None:
+            # it's not needed without observation window
+            # won't be included because of zero size
+            return None, 0
+
+        clipped_mask = np.ones(view_clip.shape, dtype=np.bool).flatten()
+        clipped_mask[view_clip.view_indices] = 0
+        return np.flatnonzero(clipped_mask), clipped_mask.size
+
+    def append_mask(self, mask: np.ndarray):
+        ...
+
+    def append_position(self, exist: bool, position):
+        return exist or not (
+            0 <= position[0] < self.shape[0]
+            and 0 <= position[1] < self.shape[1]
+        )
+

### 2021-04-15 htm_rl/htm_rl/envs/biogwlab/obstacles.py

@@ -1,27 +1,0 @@
-from typing import Optional
-
-from htm_rl.envs.biogwlab.entity import Entity
-from htm_rl.envs.biogwlab.generation.obstacles import ObstaclesGenerator
-
-
-class Obstacles(Entity):
-    entity = 'obstacles'
-
-    _generator: ObstaclesGenerator
-    _last_seed: Optional[int]
-
-    def __init__(self, density, **obstacles):
-        super(Obstacles, self).__init__(**obstacles)
-        self._generator = ObstaclesGenerator(shape=self.shape, density=density)
-        self._last_seed = None
-
-    def generate(self, seeds):
-        seed = seeds['map']
-        if self._last_seed is not None and self._last_seed == seed:
-            return
-
-        mask = self._generator.generate(seed)
-        self.set(
-            mask=mask,
-            map=None
-        )

### 2021-04-15 htm_rl/htm_rl/envs/biogwlab/regenerator.py

@@ -3,24 +3,29 @@ from typing import Dict
 import numpy as np
 
 from htm_rl.envs.biogwlab.environment import Environment
+from htm_rl.envs.biogwlab.module import Module
 
 
-class Regenerator:
+class Regenerator(Module):
     base_rates = {
         'map': None, 'food': None, 'agent': None
     }
+    base_modulo: int = 1_000_000
 
     seed: int
     seeds: Dict[str, int]
     rates: Dict[str, int]
     episode: int
 
-    def __init__(self, env: Environment, **rates):
+    def __init__(self, env: Environment, name: str, **rates):
+        super().__init__(name=name)
+
         self.seed = env.seed
         self.rates = {**self.base_rates, **rates}
         self.episode = -1
+
         rng = np.random.default_rng(self.seed)
-        self.seeds = {key: rng.integers(1_000_000) for key in self.rates}
+        self.seeds = {key: rng.integers(self.base_modulo) for key in self.rates}
 
     def generate_seeds(self):
         self.episode += 1
@@ -30,7 +35,7 @@ class Regenerator:
             if regenerate or (rate is not None and (self.episode + 1) % rate == 0):
                 seed = self.seeds[key]
                 rng = np.random.default_rng(seed)
-                self.seeds[key] = rng.integers(1_000_000)
+                self.seeds[key] = rng.integers(self.base_modulo)
                 regenerate = True
 
         return self.seeds

### 2021-04-15 htm_rl/htm_rl/envs/biogwlab/renderer.py

@@ -1,109 +1,119 @@
-from typing import Tuple, List, Optional
+from typing import Tuple, List, Optional, Iterable, Dict
 
 import numpy as np
 
-from htm_rl.common.sdr_encoders import SdrConcatenator, IntBucketEncoder, IntArrayEncoder
-from htm_rl.envs.biogwlab.move_dynamics import MOVE_DIRECTIONS
-from htm_rl.envs.biogwlab.view_clipper import ViewClipper
+from htm_rl.common.sdr_encoders import SdrConcatenator
+from htm_rl.envs.biogwlab.module import Entity, EntityType
+from htm_rl.envs.biogwlab.view_clipper import ViewClipper, ViewClip
 
 
 class Renderer:
     output_sdr_size: int
 
     shape: Tuple[int, int]
-    view_shape: Tuple[int, int]
     render: List[str]
-    sdr_concatenator: SdrConcatenator
+    sdr_concatenator: Optional[SdrConcatenator]
     view_clipper: Optional[ViewClipper]
 
-    def __init__(self, env, render: List[str], view_rectangle=None, **renderer):
-        self.encoders = dict()
-        self._cached_renderings = dict()
-
+    def __init__(self, env, view_rectangle=None):
         self.shape = env.shape
-        self.view_shape = env.shape
         self.view_clipper = None
         if view_rectangle is not None:
             self.view_clipper = ViewClipper(env.shape, view_rectangle)
-            self.view_shape = self.view_clipper.view_shape
+        self.sdr_concatenator = None
 
-        for data_name in render:
-            if data_name == 'position':
-                n_cells = env.shape[0] * env.shape[1]
-                encoder = self.get_agent_position_renderer(env.agent, n_cells, **renderer)
-            elif data_name == 'direction':
-                encoder = self.get_agent_direction_renderer(env.agent, **renderer)
-            else:
-                encoder = self.get_renderer(env.get_module(data_name))
+    def render(self, position, view_direction, entities: Iterable[Entity]):
+        view_clip = self.make_view_clip(position, view_direction)
 
-            if encoder is not None:
-                self.encoders[data_name] = encoder
+        layers_with_size = []
+        for entity in entities:
+            if not entity.rendering:
+                continue
+            layer = entity.render(view_clip)
+            if isinstance(layer, list):
+                layers_with_size.extend(layer)
+            elif layer[1]:
+                layers_with_size.append(layer)
 
-        self.output_sdr_size = 0
[…]

### 2021-04-15 htm_rl/htm_rl/envs/biogwlab/view_clipper.py

@@ -1,19 +1,30 @@
+import dataclasses
+from dataclasses import dataclass
 from typing import Tuple
 
 import numpy as np
 
-from htm_rl.common.plot_utils import plot_grid_images
-from htm_rl.common.sdr import dense_to_sparse
-from htm_rl.common.utils import clip
-from htm_rl.envs.biogwlab.move_dynamics import DIRECTIONS_ORDER
+from htm_rl.envs.biogwlab.move_dynamics import DIRECTIONS_ORDER, MoveDynamics
+
+
+@dataclass
+class ViewClip:
+    shape: Tuple[int, int]
+    abs_indices: np.ndarray
+    view_indices: np.ndarray
+
+    def __iter__(self):
+        yield from dataclasses.astuple(self)
 
 
 class ViewClipper:
-    i_transformation = [1, -1, -1, 1, ]
-    j_transformation = [1, 1, -1, -1, ]
-
     base_shape: Tuple[int, int]
     view_rectangle: Tuple[Tuple[int, int], Tuple[int, int]]
+
+    natural_direction: int
+
+    _map_abs_indices_cache: np.ndarray
+    _view_indices_cache: np.ndarray
 
     def __init__(
             self,
@@ -21,7 +32,14 @@ class ViewClipper:
             view_rectangle_xy: Tuple[Tuple[int, int], Tuple[int, int]],
     ):
         self.base_shape = base_shape
-        self.view_rectangle = self.xy_to_ij(view_rectangle_xy)
+        # view direction with natural indices traversal
+        self.natural_direction = DIRECTIONS_ORDER.index('down')
+        self.view_rectangle = _xy_to_ij(view_rectangle_xy)
+
+        h, w = self.base_shape
+        self._map_abs_indices_cache = np.arange(h*w).reshape(self.base_shape)
+        h, w = self.view_shape
+        self._view_indices_cache = np.arange(h*w).reshape(self.view_shape)
 
     # noinspection PyRedundantParentheses
     @property
@@ -30,97 +48,121 @@ class ViewClipper:
         return (ui-bi+1, rj-lj+1)
 
     def clip(self, position, view_direction):
-        # position = 1, 5
-        # view_direction = 3
-        i, j = position
-        i_high, j_high = self.base_shape
+        # get relative direction
+        view_direction = self.natural_direction - view_direction
 
-        # print(i, j, forward_dir, i_high, j_high)
-        # print(obs_rect)
[…]

### 2021-04-15 htm_rl/htm_rl/envs/wrappers/recorders.py

@@ -4,8 +4,9 @@ import numpy as np
 
 from htm_rl.common.plot_utils import store_heatmap
 from htm_rl.common.sdr import SparseSdr
+from htm_rl.envs.biogwlab.environment import Environment
 from htm_rl.envs.env import Env
-from htm_rl.envs.wrapper import Wrapper
+from htm_rl.envs.wrapper import Wrapper, unwrap
 
 
 class EpisodeLengthRecorder(Wrapper):
@@ -37,9 +38,11 @@ class HeatmapRecorder(Wrapper):
     name_str: str
     acted: bool
 
-    def __init__(self, env: Wrapper, n_episodes, test_dir, name_str):
-        super(HeatmapRecorder, self).__init__(env.env)
-        self.heatmap = np.zeros(env.env.shape, dtype=np.float)
+    def __init__(self, env: Env, n_episodes, test_dir, name_str):
+        super(HeatmapRecorder, self).__init__(unwrap(env))
+
+        env: Environment = self.env
+        self.heatmap = np.zeros(env.shape, dtype=np.float)
         self.heatmap_ind = 0
         self.current_episode = 0
         self.n_episodes = n_episodes

### 2021-04-15 htm_rl/htm_rl/agents/htm/htm_agent.py
update agent and configs

@@ -201,6 +201,7 @@ class HTMAgentRunner:
         episode = 0
         animation = False
         prev_reward = 0
+        agent_pos = list()
         while episode < n_episodes:
             if train_patterns:
                 self.agent.train_patterns()
@@ -235,7 +236,7 @@ class HTMAgentRunner:
                              },
                             step=episode)
 
-                if ((episode % log_every_episode) == 0) and (logger is not None) and (episode > 0):
+                if (((episode + 1) % log_every_episode) == 0) and (logger is not None) and (episode > 0):
                     if log_patterns:
                         if log_q_table:
                             q = self.agent.hierarchy.output_block.bg.input_weights_d1 - self.agent.hierarchy.output_block.bg.input_weights_d2
@@ -342,7 +343,6 @@ class HTMAgentRunner:
                         self.agent.hierarchy.unfreeze()
                         logger.log({'options': logger.Table(data=np.array(options_actions).reshape((-1, 1)),
                                                             columns=['actions'])}, step=episode)
-                    pic = self.environment.callmethod('render_rgb')
 
                 if ((((episode + 1) % log_every_episode) == 0) or (episode == 0)) and (logger is not None):
                     animation = True
@@ -357,12 +357,18 @@ class HTMAgentRunner:
                 total_reward += reward
 
             if animation:
-                if self.agent.hierarchy.blocks[5].made_decision:
-                    pic[pic == 24] = 30
-                    pic[self.environment.callmethod('render_rgb') == 24] = 24
+                pic = self.environment.callmethod('render_rgb')[0]
+                if self.agent.hierarchy.blocks[5].made_decision and log_options:
+                    agent_pos.append(self.environment.env.agent.position)
+                    if len(agent_pos) > 1:
+                        pic[tuple(zip(*agent_pos))] = [[255, 255, 150]]*len(agent_pos)
+                    else:
+                        pic[agent_pos[0]] = [255, 255, 255]
                 else:
-                    pic = self.environment.callmethod('render_rgb')
-                plt.imsave(f'/tmp/{logger.run.id}_episode_{episode}_step_{steps}.png', pic, vmax=30)
+                    if len(agent_pos) > 0:
+                        agent_pos.clear()
+
[…]

### 2021-04-15 htm_rl/htm_rl/experiment.py
add new 8x8 obs final seeds and adapted ucb agent config

@@ -13,6 +13,7 @@ from htm_rl.common.utils import timed
 from htm_rl.config import Config
 from htm_rl.envs.biogwlab.env import BioGwLabEnvironment
 from htm_rl.envs.env import Env
+from htm_rl.envs.wrapper import unwrap
 from htm_rl.envs.wrappers.recorders import HeatmapRecorder
 
 
@@ -39,14 +40,14 @@ class Experiment:
 
         print(f'AGENT: {agent.name}     SEED: {env_seed} {agent_seed}')
         if run_results_processor is not None:
-            # store_environment_map(
-            #     seed_ind, env.callmethod('render_rgb'),
-            #     env_config.name, seed, run_results_processor.test_dir
-            # )
-            env = HeatmapRecorder(
-                env, 25, run_results_processor.test_dir,
-                f'{agent_config.name}_{env_seed}'
+            store_environment_map(
+                seed_ind, env.callmethod('render_rgb'),
+                env_config.name, env_seed, run_results_processor.test_dir
             )
+            # env = HeatmapRecorder(
+            #     env, 25, run_results_processor.test_dir,
+            #     f'{agent_config.name}_{env_seed}'
+            # )
 
         # temporal dirty flag-based solution
         run = None
@@ -80,6 +81,10 @@ class Experiment:
 
         while True:
             reward, obs, first = env.observe()
+
+            # from htm_rl.common.plot_utils import plot_grid_images
+            # plot_grid_images(unwrap(env).render_rgb())
+
             if first and step > 0:
                 break
 

### 2021-04-16 reports/v1_0/talk_eng/slides.tex
initial commit

@@
 
-\begin{frame}{Introduction and motivation}
+\begin{frame}{Model-based RL}
 	\begin{itemize}
-		\item why we need a model
-		\item why we draw inspirations from human memory models
+		\item learning the environment allows using planning methods
+		\item can remedy sample inefficiency of model-free RL
+		\item popular choice - Artificial Neural Networks
+		\item we used more human-like memory model - Hierarchical Temporal Memory (HTM)
 	\end{itemize}
@@
 \begin{frame}{Agent's architecture}
-	- figure
+    \includegraphics[width=.65\linewidth]{agent_scheme}
 	- explain how it works on a high level	
@@
 
-\begin{frame}{Results. Experimental setup 1}
-	- explain setup
-	- figures
-	- results	
+\begin{frame}{Experiments in fixed environments}
+\begin{columns}
+	\begin{column}{.55\textwidth}
+	Setup:
+	\begin{itemize}
+		\item fixed handcrafted mazes
+		\item reward position changes every 100 episodes
+	\end{itemize}
+
+	\vspace*{12pt}
+	Conclusions:
+	\begin{itemize}
+		\item better than random policy in any situations
+		\item learns and adapt faster than DQN
+		\item optimal solution requires planning horizon to be large enough to reach the reward from the initial state
+	\end{itemize}
+
+	\end{column}
+
+	\begin{column}{.45\textwidth}
+	\begin{center}
+		\includesvg[width=\linewidth]{multi_way_v0__steps}
+		\includesvg[width=\linewidth]{multi_way_v2__steps}
+	\end{center}
+	\end{column}
+\end{columns}	
 \end{frame}
 
-\begin{frame}{Results. Experimental setup 2}
-	- explain setup
-	- figures
-	- results	
+\begin{frame}{Experiments in changing environments}
+\begin{columns}
+	\begin{column}{.55\textwidth}
+	Setup:
+	\begin{itemize}
+		\item randomly generated mazes
+		\item maze periodically changes too
+		\begin{itemize}
+			\item up: every 200 episodes
+			\item bottom: every 20 episodes
+		\end{itemize}
+	\end{itemize}
+
+	\vspace*{12pt}
+	Conclusions:
+	\begin{itemize}
+		\item learns and adapt faster than DQN
+		\item more stable performance - less degradation spikes
+		\item suboptimal solutions with faster adaptation win over potentially optimal solutions unable to adapt in time
+	\end{itemize}
+
+	\end{column}
+
+	\begin{column}{.45\textwidth}
+	\begin{center}
+		\includesvg[width=\linewidth]{gridworld_5x5_100_2_8_1337__steps}
+		\includesvg[width=\linewidth]{gridworld_5x5_20_1_20_1337__steps}
+	\end{center}
+	\end{column}
+\end{columns}
 \end{frame}

### 2021-04-16 htm_rl/htm_rl/agents/htm/hierarchy.py

@@ -77,7 +77,7 @@ class SpatialMemory:
             overlaps = np.dot(dense_pattern, self.patterns.T)
             pattern_sizes = self.patterns.sum(axis=1)
             overlaps = overlaps / pattern_sizes
-            indices = np.flatnonzero(overlaps > self.overlap_threshold)
+            indices = np.flatnonzero(overlaps >= self.overlap_threshold)
             options = self.patterns[indices]
             # convert to list of sparse representations
             if return_indices:

### 2021-04-16 htm_rl/htm_rl/agents/htm/htm_agent.py

@@ -218,7 +218,7 @@ class HTMAgentRunner:
                         for i in range(steps):
                             image = imageio.imread(f'/tmp/{logger.run.id}_episode_{episode}_step_{i}.png')
                             writer.append_data(image)
-                    logger.log({f'animation': logger.Video(f'/tmp/{logger.run.id}_episode_{episode}.gif', fps=4,
+                    logger.log({f'animation': logger.Video(f'/tmp/{logger.run.id}_episode_{episode}.gif', fps=2,
                                                            format='gif')}, step=episode)
 
                 if (logger is not None) and (episode > 0):
@@ -236,7 +236,7 @@ class HTMAgentRunner:
                              },
                             step=episode)
 
-                if (((episode + 1) % log_every_episode) == 0) and (logger is not None) and (episode > 0):
+                if ((episode % log_every_episode) == 0) and (logger is not None) and (episode > 0):
                     if log_patterns:
                         if log_q_table:
                             q = self.agent.hierarchy.output_block.bg.input_weights_d1 - self.agent.hierarchy.output_block.bg.input_weights_d2

### 2021-04-16 htm_rl/htm_rl/agents/htm/htm_apical_basal_feeedback.py

@@ -424,7 +424,12 @@ class ApicalBasalFeedbackTM:
         self.active_basal_cells.sparse = np.unique(new_active_cells.astype('uint32'))
         self.winner_basal_cells.sparse = np.unique(new_winner_cells)
 
-        anomaly = len(bursting_columns) / self.basal_columns
+        n_active_columns = self.active_columns.sparse.size
+        if n_active_columns != 0:
+            anomaly = len(bursting_columns) / n_active_columns
+        else:
+            anomaly = 1.0
+
         self.anomaly_threshold = self.anomaly_threshold + (anomaly - self.anomaly[0]) / self.anomaly_window
         self.anomaly.append(anomaly)
         self.anomaly.pop(0)

### 2021-04-17 htm_rl/htm_rl/agents/htm/htm_agent.py
add simple intrinsic motivation

@@ -13,7 +13,6 @@ import random
 import yaml
 import matplotlib.pyplot as plt
 import wandb
-#from memory_profiler import profile
 
 
 class BioGwLabAction:
@@ -48,6 +47,8 @@ class BioGwLabAction:
 
 class HTMAgent:
     def __init__(self, config, hierarchy: Hierarchy):
+        self.use_intrinsic_reward = config['intrinsic_reward']
+        self.theta = config['theta']
         self.punish_reward = config['punish_reward']
         self.hierarchy = hierarchy
         self.action = BioGwLabAction(**config['action'])
@@ -85,6 +86,15 @@ class HTMAgent:
         action = self.action.get_action(action_pattern)
         return action
 
+    def get_intrinsic_reward(self):
+        confidence = self.hierarchy.blocks[2].tm.confidence[-1]
+        conf_threshold = self.hierarchy.blocks[2].confidence_threshold
+        if confidence < conf_threshold:
+            reward = conf_threshold - confidence
+        else:
+            reward = 0
+        return reward
+
     def reinforce(self, reward, pseudo_rewards=None):
         """
         Reinforce BasalGanglia.
@@ -95,6 +105,8 @@ class HTMAgent:
         List should be length of number of blocks in hierarchy.
         :return:
         """
+        if self.use_intrinsic_reward:
+            reward += (self.theta * self.get_intrinsic_reward())
         self.hierarchy.output_block.add_reward(reward)
         self.hierarchy.output_block.reinforce()
 
